### 2022-03-15 manuscript.tex
Switches from Pandoc to LaTeX. closes #91

\documentclass{article}
\usepackage{caption}
\usepackage{algorithm2e}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{array}
\usepackage{courier}

\usepackage[backend=bibtex,sorting=none]{biblatex}
\addbibresource{manuscript}
\graphicspath{{assets/images/}}
\emergencystretch=1em

\def\ttt{\texttt}

\begin{document}

\hyphenation{organo-metallics}
\newcolumntype{L}{>{\arraybackslash}m{3cm}}
\lstset{basicstyle=\ttfamily,breaklines=true}

\title{Dialect: A Compact Line Notation Based on SMILES}
\author{Richard L. Apodaca\thanks{rapodaca@metamolecular.com}}

\maketitle

\section{Introduction}

Simplified Molecular Input Line Entry System (SMILES) was first described by Weininger in 1988.\cite{weininger:1988} As a line notation,\cite{wiswesser:1968} SMILES represents molecules as single line character sequences, or strings. SMILES strings can be read and written algorithmically, and can be canonicalized\cite{weininger:1989,oBoyle:2012} to yield unique molecular identifiers.

SMILES has since been widely adopted. Read and write functionalities are routinely supported by popular cheminformatics toolkits, including: Open Babel;\cite{oBoyle:2011} RDKit;\cite{rdkit} Chemistry Development Kit;\cite{steinbeck2003} JChem Base;\cite{jchemBase} the Daylight Toolkit;\cite{daylightToolkit} and OEChem TK.\cite{oeChemTk} SMILES encodings can be found in many public-facing databases, including: PubChem;\cite{kim2016} ChEBI;\cite{degtyarenko2007} ZINC;\cite{irwin2005} ChEMBL;\cite{gaulton2012} and Wikipedia.\cite{wikipedia} Raw SMILES strings have been used extensively in machine learning applications.\cite{sousa2021} SMILES has also been extended to carry various forms of metadata, as exemplified by Jmol SMILES,\cite{hanson2016} CurleySMILES,\cite{drefahl2011} BigSMILES,\cite{lin2019} and CXSMILES.\cite{cxsmiles}

Despite its many applications, SMILES remains a language with an incomplete specification. Weininger's 1988 paper (referred to herein as "the Paper") leaves the following either fully or incompletely specified: (1) stereochemical configuration; (2) double bond conformation; (3) the meaning handling of "aromatic" features; (4) the computation of implicit hydrogen count, especially as related to aromatic features; (5) formal syntax; (6) constraints on quantities such as mass number and charge; and (7) enumeration of error states.

Earlier descriptions of SMILES were published in various media, although these are not cited by Weininger. An accessible example is a 1987 report to the US Environmental Protection Agency.\cite{anderson1987} Although the language described in that report shares many features with later descriptions, there are important differences as well.

Since 1988, a few sources have attempted to clarify or revise Weininger's 1988 description of SMILES. A 2003 book chapter by Weininger (referred to herein as "the Book Chapter") addressed some of the limitations of the paper. The Daylight Theory Manual ("the Manual"), a website maintained by SMILES's corporate sponsor, Daylight Chemical Information Systems, Inc. ("Daylight"), recapitulates much of the material in the Paper and the Book Chapter.\cite{daylightTheory}.

In 2007 a documentation effort that would become known as OpenSMILES began.\cite{openSMILES} OpenSMILES was conceived as "a non-proprietary specification for the SMILES language," and it addressed many of the points left open by previous SMILES documentation efforts. Noteworthy contributions included: the first formal grammar; many refinements around stereochemistry; and introduction of the idea of "standard form." Absent were detailed procedures for assigning and interpreting aromatic features, and a detailed procedure for computing implicit hydrogen count. OpenSMILES also left several points of syntactic and semantic ambiguity unaddressed.

In 2019 IUPAC announced the SMILES+ initiative.\cite{smilesPlus} Noting the limitations of existing SMILES documentation, the SMILES+ effort seeks to "establish a formalized recommended up-to-date specification of the SMILES format." SMILES+ took as its starting point the documentation produced by the OpenSMILES project. Efforts to extend this starting point are in progress online through a public repository, but no formal recommendation has to date resulted.\cite{smilesPlusRepo}

The ongoing lack of a comprehensive SMILES documentation suite has caused several problems. First, authors of new SMILES implementations have limited guidance when resolving ambiguities and so must invent them on an ad hoc basis. Second, maintainers of existing SMILES implementations lack a blueprint for working toward a common feature set. Third, standards bodies draw from a limited set of source material when preparing recommendations. Fourth, a lack of detailed documentation hinders the development of compliance suites, thereby interfering with efforts to validate cross-implementation compatibility. Finally, extensions to SMILES only make sense in the context of a rigorously defined base language.

The lack of a detailed, comprehensive, and universally-applied SMILES specification also hampers broader data integrity efforts. The FAIR Guiding Principles\cite{wilkinson2016} identify four qualities essential for extracting maximum utility from published scholarly data: Findability; Accessibility; Interoperability; and Reusability. Interoperability in particular requires that "(meta)data" use a formal, accessible, shared, and broadly applicable language for knowledge representation." Although SMILES may be a shared, accessible, and broadly applicable language, its formal underpinnings leave room for improvement.

This situation has persisted long enough that the notion of a SMILES "standard" may not currently be feasible. A dozen or more slightly different implementations are now in use. None of them can be considered a reference standard. Some have introduced extensions not recognized by the others, contributing to language "drift" while compounding the already formidable difficulties faced by implementors. Others claim SMILES compatibility while ignoring certain aspects of written documentation. The Daylight implementation, like some of the others, is distributed under a non-Open Source license that restricts use.\cite{daylightToolkit} For several years Daylight operated a publicly-accessible web service based on its toolkit that depicted SMILES strings, but it has since been decommissioned.\cite{depict} Although SMILES may underpin modern chemical information exchange, the foundations of this system are weathered and showing signs of stress.

This paper attempts to resolve these these issues with Dialect, a fully-specified subset of the SMILES language. The following sections present Dialect from three perspectives: as a compact system for molecular representation; as a text-based language; and as a specification for software intended to convey chemical structure information across organizational and temporal boundaries.

\section{Goals, Design, and Tradeoffs}

Dialect's purpose is to solve the \textit{molecular serialization} problem. Serialization refers to the process of translating a data structure into a format that can be transmitted over a communication channel for later reconstruction. Serializations utility lies in enabling general-purpose devices and networks to convey complex pieces of domain-specific information. Receiver and recipient may be separated by time, space, or both. Molecular serialization extends the concept of serialization to molecular structure.

Two concerns factor prominently in any serialization scheme: \textit{syntax} and \textit{semantics}. Syntax sets the vocabulary of symbols to be used and how they may be arranged. Semantics set the rules determining what various arrangements of symbols mean. Failure to fully specify syntax can lead to needless loss of data. Failure to fully specify semantics can result in needless data corruption. Semantics in particular play an important role in Dialect due the the inherently complex nature of molecular representations.

Beyond syntax and semantics, a serialization scheme should be useful. Utility in this context follows directly from the scope of supported molecular representations. Constrain this scope too much and the system will cover too few molecules to be interesting. Relax the constraints too much and the system could become verbose, difficult to implement, or both. Dialect aims squarely for the middle ground: a system of molecular representation that is specific enough to yield a compact language, but general enough to cover a large amount of interesting chemical space.

Dialect's molecular representation system is based on four mutually-exclusive components, each of which is constrained as follows:

\begin{itemize}
\item \textit{Constitution}. The atoms present in a molecule and how they are connected with each other through bonds. Dialect supports elements having two-letter IUPAC designations and all bonding arrangements available through the valence bond model (as described in detail later).
\item \textit{Conformation}. The kind of stereoisomerism resulting from restricted rotation about a bond. Dialect can encode those conformations arising from restricted rotation about double bonds, which includes both alkene and cumulene stereoisomerism.
\item \textit{Configuration}. The kind of stereoisomerism resulting from the arrangement of neighboring atoms in three-dimensional space. Dialect supports configurations for tetracoordinate atoms (those attached to four neighbors) having tetrahedral symmetry.
\item \textit{Delocalization}. A type of bonding relationship in which electrons are distributed over two or more bonds, among three or more atoms, or across some combination of atoms and bonds. Dialect supports a limited form of delocalization operating over paths of alternating single and double bonds.
\end{itemize}

\begin{figure}
\centering
\includegraphics{filler}
\caption{Representative Dialect examples.}
\label{fig:myfig}
\end{figure}

Dialect's system of representation is based on the \textit{molecular graph} concept.\cite{balaban1985} A graph is a data structure comprised of a set of nodes (which map to "atoms") and a set of pairwise relationships between them called edges (which map to "bonds"). A molecular graph is a specialized graph onto which atomic and bonding metadata have been overlaid. Molecular graphs are ubiquitous in cheminformatics, appearing in contexts such as data formats like Chemical Markup Language\cite{murrayrust2011}, CDXML\cite{cdxml}, and Molfile\cite{ctfileFormats}, as well as in-memory data structures found in cheminformatics toolkits. Like these systems, Dialect augments its underlying graph with node and edge labels.

A Dialect molecule consists of a graph with zero or more nodes and zero or more edges. The "empty molecule", devoid of atoms and bonds, is allowed. Dialect imposes no upper bound on the number of atoms or bonds. However, practical limitations related to memory, storage, and CPU time will likely limit the size of molecules in practice.

The edges of a Dialect molecular graph are defined as a set of unordered pairs of non-identical nodes. The practical consequences of this are threefold. First, \textit{loops} are not allowed. A loop is an edge joining a node to itself. Second, edges are \textit{undirected}. An undirected edge is equivalent to another edge whose terminals are swapped. An undirected edge between nodes A and B is therefore identical to an undirected edge between nodes B and A. Third, \textit{parallel edges} are disallowed. Edges are said to be parallel if they connect the same two nodes.

A molecular graph contains one or more \textit{connected components}. A connected component is a graph in which any two vertices are connected to each other by at least one path. Dialect places no restriction on the number of connected components, although practical limitations are likely to impose one.

\section{Data Model}

Dialect uses several abstractions collectively known as the \textit{data model}. The data model defines data types and relationships between them for the purpose of lossless molecular representation.

The data model is expressed in terms of a small set of \textit{primitives}. A primitive is an irreducible data type. Dialects primitives are summarized in Table \ref{table:primitives}.

\begin{table}
\caption{Data Model Primitives.}
\centering
\begin{tabular}{l l l}
    \hline
    Name & Description & Values \\
    \hline
    \ttt{Option<Type>} & Optional Value & \ttt{None}, \ttt{Type} \\
    \ttt{Index} & Unsigned integer & 0{\ldots}2\textsuperscript{32} \\
    \ttt{Ten} & Integer count & 0{\ldots}10 \\
    \ttt{SignedTen} & Integer count & -9{\ldots}10 \\
    \ttt{Thousand} & Integer count & 0{\ldots}1000 \\
    \ttt{Element} & IUPAC-\-approved element symbol & \ttt{Ac}, \ttt{Ag}, \ttt{Al}, {\ldots}\ttt{Zn}\\
    \ttt{HydrogenCount} & Hydrogen count & \ttt{Implicit}, \ttt{Ten} \\
    \ttt{Configuration} & Configurational descriptor & \ttt{Right}, \ttt{Left} \\
    \ttt{Direction} & Conformational descriptor & \ttt{Up}, \ttt{Down} \\
    \ttt{boolean} & Boolean & \ttt{true}, \ttt{false} \\
    \hline
\end{tabular}
\label{table:primitives}
\end{table}

Table \ref{table:atomic-state} summarizes atomic attributes. Not all combinations of values ("atomic state") are valid. Implementations must ensure either the impossibility of an invalid state, or an error condition in the event that one is created.

\begin{table}
\caption{Atomic State.}
\centering
\begin{tabular}{l L L l}
    \hline
    Attribute & Description & Type & Default \\
    \hline
    \ttt{index} & Unique identifier &\ttt{Index} & - \\
    \ttt{element} & Elemental symbol & \ttt{Option} \ttt{<Element>} & \ttt{None} \\
    \ttt{hydrogens} & Hydrogen count & \ttt{HydrogenCount} & 0 \\
    \ttt{configuration} & Configuration & \ttt{Configuration} & \ttt{None} \\
    \ttt{charge} & Formal charge & \ttt{SignedTen} & 0 \\
    \ttt{selected} & Selection status & \ttt{boolean} & \ttt{false} \\
    \hline
\end{tabular}
\label{table:atomic-state}
\end{table}

Three restrictions apply to atomic state:

\begin{enumerate}
    \item{The value of \ttt{index} must be unique over the molecular graph.}
    \item{If \ttt{hydrogens} equals \ttt{Implicit}, then \ttt{isotope}, \ttt{configuration}, and \ttt{charge} must equal their default values. Furthermore, \ttt{element} must equal one of \ttt{B}; \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; \ttt{S}; \ttt{F}; \ttt{Cl}; \ttt{Br}; \ttt{I}.}
    \item{If \ttt{selected} equals \ttt{true}, then \ttt{element} must equal one of: \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; or \ttt{S}.}
\end{enumerate}

The following restrictions apply to bond state:

\begin{enumerate}
    \item{The value of \ttt{index} must be unique over the molecular graph.}
    \item{The \ttt{source} and \ttt{target} attributes must reference valid atomic \ttt{index} attributes.}
    \item{If \ttt{direction} is non-default (any value other than \ttt{None}), then \ttt{order} must equal 1.}
    \item{If the value of \ttt{elided} is \ttt{true}, the values of \ttt{order} and \ttt{direction} must equal 1 and \ttt{None}, respectively.}
\end{enumerate}

The data model provides a link between syntax and semantics. Dialect-compatible software may, of course, use any suitable internal data model. Dialect implementations must, however, ensure that their internal data model is consistent with the one provided here.

\section{Constitution}

A Dialect molecule is described at the lowest level by its constitution. Constitution consists of a set of atomic nuclei (nodes), a set of pairwise bonding relationships between them (edges), and those attributes needed to associate every valence electron with a specific node or edge. Constitution excludes attributes related to stereochemistry, or any other feature whose attributes extend over more than one atom or bond.

Two attributes characterize an atom's nucleus: \ttt{element} and \ttt{isotope}. The `element` attribute is an optional one- or two-letter character sequence selected from the set designated by the IUPAC/IUPAP Working Party.\cite{iupac2016} When the Working Party authorizes additional symbols, they will become valid values for the \ttt{element} attribute. If the elemental identity of an atom is unknown, its \ttt{element} attribute can be omitted. The \ttt{isotope} attribute is an optional integer value representing an atom's nuclear mass number, where mass number is the sum of proton and neutron count. Omitting the \ttt{isotope} property means that the isotopic composition is unspecified. When present, the lower bound on \ttt{isotope} equals one. This constraint leaves room for physically meaningless negative implied neutron count as in, for example, carbon-5. The lower bound for isotope applies even if the \ttt{element} attribute is not defined.

The atomic \ttt{hydrogens} attribute records the number of associated \textit{virtual hydrogens}. A virtual hydrogen is an atom whose presence is recorded, not as a node and edge, but rather as a unit contribution to an integer tally. For example, methane can be represented by a molecular graph having five nodes and four edges. But methane can also be represented as a graph of one node having a \ttt{hydrogens} attribute set to four. The \ttt{hydrogens} attribute, if present, may assume integer values ranging from zero to nine. Only monovalent hydrogens with unspecified isotopic composition are eligible for virtualization.

The two remaining constitutional attributes support \textit{electron counting}. The purpose of electron counting is to ensure that every valence electron, whether bonding or nonbonding, can be associated with either an atom or a bond. Accurate electron counting is crucial for any molecular representation system given their influence on charge, bond order, and mass.

Electron counting in Dialect is based on the well-known \textit{valence bond model} (VB Model).\cite{lewis1916} Although not commonly considered as such, the VB model can be thought of as an algorithm for molecular assembly. The perspective is especially useful given that Dialect itself can be viewed as a programming language for molecular assembly.

A simplified algorithm for electron counting can be summarized as follows. A molecular graph starts as a set of atoms, where each atom consists of a nucleus (as described above) and an integer electron count equal to the atomic number of the atom's element. Two nodes (nuclei) are selected as the terminals of a bond. From each atom, the same positive electron count is deducted. The total electron count deducted from each atom is simultaneously credited to the bond's electron count. For example, if each atom contributes one electron to a bond, then the bond's electron count will increment by two.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Generalized Valence Bond Electron Counting. Two unconnected nodes are chosen for bonding (left). One electron is subtracted from each atom, and both electrons are added to the bond's count (right).}
    \label{fig:electron-counting}
\end{figure}

For convenience, the actual model used by Dialect differs slightly from the above description. Neither nodes nor edges carry an explicit electron count attribute. Instead, edges carry an \ttt{order} attribute, and nodes carry a \ttt{charge} attribute. Bond order equals electron count divided by two. Formal charge equals the number of electrons gained or lost due to ionizations events outside of bond formation. Gaining an electron decrements the \ttt{charge} attribute, whereas losing an electron increments it. If no electrons are gained or lost to ionization outside of bond formation, the formal charge equals zero.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Electron counting shorthand. Rather than explicit electron counts, atoms and bonds possess charge and order attributes. The atomic charge attribute captures deviations from generalized counting.}
    \label{fig:electron-counting-shorthand}
\end{figure}

From these rules follow some important consequences. For example, bond order must be an integer greater than zero. This follows from the definition of bond formation. The electron count deducted from each atom is $n$, an integer greater than zero. Given bond formation deducts an equal number of electrons from each atom, the total number of electrons associated with the newly-formed bond is $2n$. By definition, bond order is the bond's electron count divided by two, therefore $n$ is the bond order. Fractional, negative, and zero bond orders are all disallowed by Dialect for this reason.

Dialect imposes an upper limit of three on formal bond order. The relative scarcity of bond orders greater than three in molecules conforming the the VB Model makes this restriction unlikely to meaningfully restrict applicability.

Whereas negative bond order is disallowed by definition, Dialect places no restrictions on \textit{hypervalence}. Hypervalence occurs when an atom undergoes enough bonding operations to leave it with a negative implied valence electron count. Consider lithium, which possesses one valence electron. Formation of one single bond leaves lithium with zero implied valence electrons. Application of a second bond formation leaves lithium with a zero charge and an implied valence electron count of -1. Such an arrangement may be physically meaningless, but Dialect explicitly supports it. Software using Dialect may or may not reject such species for semantic reasons.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Hypervalence. Depletion of atomic electron count through excessive bonding. Hypervalent atoms are valid by default, although chemically nonsensical.}
    \label{fig:hypervalence}
\end{figure}

\section{Bond Elision}

To promote compact string representations, Dialect supports \textit{bond elision}. An single bond may be elided by marking it as such. An elided bond nevertheless has a bond order of one. A bond may be elided for one of two reasons. First, elided bonds are not written during serialization, allowing for increased information density. The second reason relates to a feature described in the next section.

\section{Delocalization Subgraph}

A molecular representation based solely on the valence bond model can yield artifacts resulting from \textit{delocalization induced molecular equality} (DIME). DIME occurs in a molecular graph when one or more equivalent representations exist, each one differing from the original only in the distribution of single and double bonds. DIME may be recognized as "resonance" or "aromaticity," but those terms are avoided here due to their ambiguity.\cite{randic2018,kerber2006}

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Delocalization-induced molecular equality (DIME). Two chemically equivalent molecular graphs differ on the basis of delocalization.}
    \label{fig:dime}
\end{figure}

DIME can interfere with \textit{canonicalization}, or the selection of a single representation for a molecular graph. The presence of multiple equivalent molecular graphs differing only in the placement of single and double bonds complicates selection rules and invariants, which must be adapted to account for the artificial asymmetry.

To eliminate DIME and thereby streamline canonicalization, Dialect representations support a \textit{delocalization subgraph} (DS). A DS is a possibly empty node-induced subgraph of a molecular graph. The membership of a DS is drawn from the set of atoms and bonds that participate in DIME within a given molecular graph.

A non-empty DS must possess a \textit{perfect matching}. A matching is a subgraph in which no two edges share a common node. Equivalently, a matching is a subgraph in which all nodes have degree one. A perfect matching includes all the nodes of its parent graph. Every atom added to a DS must therefore be part of a perfect matching over it.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Perfect Matching. A subgraph with all of the nodes of its parent, but in which all nodes have degree one.}
    \label{fig:perfect-matching}
\end{figure}

Only some atoms are eligible for inclusion (or "selection") in a DS. Atoms whose \ttt{element} values are one of \ttt{C}, \ttt{N}, \ttt{O}, \ttt{P}, or \ttt{S}, or \ttt{None}, may be selected --- regardless of cycle membership. Atoms having other \ttt{element} values must not be selected.

To support the construction of a DS, eligible atoms carry a \ttt{selected} boolean attribute. Setting this attribute to \ttt{true}, adds the atom to the DS. Otherwise, the atom is excluded from the DS. All ineligible atoms are excluded from the DS.

A bond will be added to the DS only if both of the following conditions are met: (1) both terminals are selected; and (2) the bond itself is elided. No other bond will be added to the DS. This behavior makes it possible to exclude a specific bond from the DS by making its bond order explicit.

A filled DS can be emptied through a two-step process of \textit{deselection}. First, a perfect matching over the DS is found. Next, each matched edge is replaced by a double bond. Because the presence of a filled DS implies a perfect matching over it, kekulization always succeeds. A popular algorithm for matching, the Edmunds "blossom algorithm,"\cite{edmunds65} has a time complexity of $O(|E||V|^2)$, where $|E|$ is the number of edges and $|V|$ is the number of nodes. Although more efficient algorithms are known, they are either much more difficult to implement or lack generality.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Deselection. Unsetting the atomic selected flag with simultaneous promotion of matching bonds.}
    \label{fig:deselection}
\end{figure}

The opposite operation can be accomplished with a \textit{selection algorithm}. A selection algorithm selects two or more eligible atoms, thereby adding them to the DS. The only requirement for a selection algorithm is that the resulting DS must have a perfect matching. Depending on the application, other criteria may be applied. For example, a selection algorithm can restrict candidate atoms to those found in cycles. Electron-counting techniques can also be introduced to approximate the chemical concept of "aromaticity." A double bond between two selected atoms may or may not be elided, depending on the application.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Selection Algorithm. Those atoms with bonds contributing to DIME are added to the delocalization subgraph.}
    \label{fig:selection}
\end{figure}

\section{Implicit Hydrogens and Subvalence}

In addition to virtual hydrogen count, Dialect supports a second form of hydrogen suppression called \textit{implicit hydrogens}. Like a virtual hydrogen, an implicit hydrogen is monovalent, has only default attributes, and is present as an integer tally associated with a particular atom. But unlike a virtual hydrogen, the presence of an implicit hydrogen can only be deduced through computation. Implicit hydrogens are an integral yet invisible component of many molecular graphs.

To support implicit hydrogen counting, Dialect uses the concept of \textit{valence}. Valence is a non-negative integer computed as the sum of bond orders at a given atom. Single bonds contribute one to the tally, double bonds two, and triple bonds three. Every other bond contributes one, as does each virtual hydrogen. For example, the valence of a methyl carbon having a \ttt{hydrogens} attribute of three is four. In contrast, the valence of a methyl carbon with a \ttt{hydrogens} attribute of \ttt{Implicit} is one.

Only some atoms are eligible for implicit hydrogen counting. These are called \textit{eligible atoms}. An atom becomes eligible by fulfilling two requirements: (1) its \ttt{hydrogens} attribute equals \ttt{Implicit}; and (2) its \ttt{element} attribute is associated with at least one \textit{default valence}.

A default valence is a value associated with an element that represents the number of hydrogens that can be attached to an isolated atom. Table \ref{table:target-valences} lists those elements possessing at least one default valence. All other elements have no default valences. The target valence of 4 for carbon, for example, means that a fully-saturated carbon atom will be bound to four hydrogen atoms. Likewise, a fully-saturated oxygen atom will be bound to two hydrogen atoms. However, iron has no default valences. Some elements such as nitrogen have multiple target valences. In these cases, multiple saturated forms are possible. For example, nitrogen has the target valences three and five. Both ammonia and nitrogen pentahydride (NH\textsubscript{5}) are fully saturated forms of nitrogen according to Table \ref{table:target-valences}.

\begin{table}
    \caption{Default Valences.}
    \centering
    \begin{tabular}{l l}
        \hline
        Element & Defaults \\
        \hline
        B & 3 \\
        C & 4 \\
        N & 3,5 \\
        O & 2 \\
        F & 1 \\
        P & 3,5 \\
        S & 2,4,6 \\
        Cl & 1 \\
        Br & 1 \\
        I & 1 \\
        \hline
    \end{tabular}
    \label{table:target-valences}
\end{table}

Given one or more default valences, \textit{subvalence} can be computed. Subvalence is the number of hydrogens that can be added to an eligible atom without exceeding the lowest possible default valence. If no default valence less than a default valence exists, then the subvalence is zero. The subvalence for an eligible atom (\ttt{a}) can be computed with Algorithm \ref{algorithm:implicit-hydrogen}.

\begin{algorithm}
    \SetKwInOut{Input}{input}
    \SetKwInOut{Output}{output}
    \SetKwFunction{BondOrderSum}{BondOrderSum}
    \SetKwFunction{TargetValences}{TargetValences}
    \caption{Compute subvalence}
    \label{algorithm:implicit-hydrogen}

    \Input{An eligible atom $a$}
    \Output{The subvalence of $a$}
    \Begin{
      $v \leftarrow$ Valence($a$)\;
      $T \leftarrow$ TargetValences($a$)\;
      \For{$t \in T$}{
        $d \leftarrow t-v$\;
        \If{$d >= 0$}{
          \Return{$d$}\;
        }
      }
      \Return{$0$}\;
    }
\end{algorithm}

The algorithm accepts an eligible atom \ttt{a} as input and returns its subvalence. First, the valence (sum of bond orders) for \ttt{a} is computed. Next, the ordered list of target valences (\ttt{T}) is obtained from Table {table:target-valences}. For each target valence (\ttt{t}), the difference (\ttt{d}) between \ttt{t} and \ttt{v} is computed. If this difference is non-negative, \ttt{d} is returned as the implicit hydrogen count. If no suitable target valence can be found, zero is returned.

Consider an isolated atom whose values of \ttt{element} and \ttt{hydrgogens} are \ttt{N} and \ttt{Implicit}, respectively. The atom's bond order sum is zero. Its default valences are 3 and 5. The difference \ttt{d} is found to be three (3 - 0). Therefore, subvalence of this atom is three.

The carbonyl carbon atom of acetaldehyde illustrates the effects of substitution and multiple bonding. This atom uses implicit hydrogen counting (\ttt{hydrogens} equals \ttt{Implicit}). Valence is therefore three (2 + 1). The first and only default valence for carbon is four. Subtracting three from four yields one, which is returned as the atom's subvalence.

The phosphorous atom in phosphorous acid (H\textsubscript{3}PO\textsubscript{3}) illustrates the use of Algorithm \ref{algorithm:implicit-hydrogen} for atoms with multiple target valences. Given a \ttt{hydrogens} attribute of \ttt{Implicit}, the atom's bond order sum is four (2 + 1 + 1). The first target valence is 3, but subtracting that value from four yields a negative number (-1). Continuing to the next default valence, 5, a difference of 1 is obtained. Therefore, the subvalence of the phosphorous-bearing atom is reported as 1.

The valence of some atoms exceeds the largest default valence. In these cases, subvalence is reported as zero. Consider sodium perchlorate (NaClO\textsubscript{4}). The chlorine atom has a bond order sum of seven (2 + 2 + 2 + 1). From Table 1, the only target valence for chlorine is one. Subtracting seven yields a negative number (-6). Therefore, the subvalence is reported as zero.

The subvalence for ineligible atoms is zero.

\section{Computing Implicit Hydrogen Count}

For unselected atoms, implicit hydrogen count equals subvalence. Consider an oxygen atom with a \ttt{hydrogens} attribute of \ttt{Implicit}. Subvalence equals two and so does implicit hydrogen count. However, an oxygen atom with a \ttt{hydrogens} attribute equal to one will have an implicit hydrogen count of zero.

For selected atoms, a modified procedure is used. If subvalence is greater than zero, then the implicit hydrogen count equals subvalence minus one. This subtraction models the unpaired electron that would would be required for delocalized bonding, or localized bonding if the atom were deselected. Consider a selected carbon atom in benzene whose \ttt{hydrogens} attribute equals \ttt{Implicit}. Subvalence equals two, so implicit hydrogen count equals one (2 - 1).

A selected atom with a \ttt{hydrogens} attribute of \ttt{Implicit} and subvalence equal to zero has an implicit hydrogen count of zero. The semantics of such a state may seem suspect. If an atom has no unpaired electrons, how can it participate in delocalized bonding? As will be explained in more detail later (Pruning), this situation can arise through gratuitous atom selection. Returning zero avoids miscalculation of the implicit hydrogen count.

An implicit hydrogen count may or may not correlate with chemical intuition or experimental data. Consider the phosphorous-bearing atom of hypophosphorous acid (HOP(O)H\textsubscript{2}). We might expect the implicit hydrogen count to equal the experimentally-determined hydrogen count (2). However, the subvalence for the phosphorous atom is found to be three (2 + 1 - 3). The implicit hydrogen count is therefore zero (3 - 3) rather than the expected two. Similar atoms must be encoded using virtual hydrogens.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Implicit hydrogen count. The number of hydrogens is set algorithmically.}
    \label{fig:implicit-hydrogen-count}
\end{figure}

\section{Atom Index}

To support the assignment and interpretation stereochemical features, each node is assigned a sequential, unique, non-negative integer \ttt{index} attribute ("index"). The value of an atom's index equals zero for a molecular's graph's first atom. The index for each subsequently-added atom equals the order (node count) of the graph prior to the addition. For example, the index for the second atom is one, the index for the third atom is two, and so on. The maximum value of an atomic index is 2\textsuperscript{32} - 1.

Indexes impose an ordering over the atoms in a molecular graph. An atom with an index less than another atom is said to \textit{precede} it. An atom with an index grater than another atom \textit{succeeds} it. No special significance is ascribed to the atom whose index is zero, except that every other atom succeeds it.

\section{Conformation}

The second major component of molecular representation in Dialect is \textit{conformation}. Conformation is a rotational restriction about one or more bonds. Dialect limits conformation to just one of many possible types: rotational restrictions occurring at individual double bonds.

Organic chemistry uses the parity descriptors \textit{entgegen} (\textit{E}) and \textit{zusammen} (\textit{Z}) to label conformational parity. This system is based on prioritization of the atoms neighboring the double bond. One pattern of priority yields the \textit{E} parity, and the other yields \textit{Z}. Conveniently, the parity descriptor can be localized at the double bond.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Traditional conformational stereodescriptors.}
    \label{fig:traditional-conformational-stereodescriptors}
\end{figure}

Dialect takes a different approach by introducing \textit{partial parity bonds} (PPB). As the name implies, a PPB expresses a portion of the parity characterizing a conformationally-restricted double bond. Constructing the full parity requires the double bond itself, and at least two PPBs. At least two PPBs must flank the double bond.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Partial parity bond. Bond conformational parity is distributed over two or more bonds.}
    \label{fig:partial-parity-bonds}
\end{figure}

To support this system, each bond carries a nullable \ttt{direction} attribute ("direction"). When present, this attribute may assume one of the two values \ttt{Up} or \ttt{Down}. These values refer to a geometrical model in which the two terminals of a double bond and their neighbors are assigned local relative coordinates in a plane. The double bond terminals are placed on the x-axis such that the terminal with the lowest \ttt{index} attribute appears to the left of the terminal with the higher atomic \ttt{index} attribute. These terminals are designated "left terminal" and "right terminal," respectively. Each neighbor of a terminal is then assigned a relative coordinate based on the \ttt{index} attribute and the direction of its PPB bond.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Interpreting partial parity bonds.}
    \label{fig:interpret-ppb}
\end{figure}

The following procedure assigns a relative coordinate to a neighbor of the left terminal. First, determine the relative order of the indexes for the neighbor and the left terminal. If the neighbor succeeds the left terminal, place the neighbor to the upper left of the terminal if the bond direction is \ttt{Up}, or to the lower left if the bond direction is \ttt{Down}. If the terminal succeeds its neighbor, reverse these assignments.

An analogous procedure assigns relative coordinates to a neighbor of the right terminal. If a neighbor succeeds the right terminal, place the neighbor to the upper right given an \ttt{Up} direction, or to the lower right given a \ttt{Down} direction. Reverse these assignments if the right terminal succeeds its neighbor.

In some cases the placement of a terminal neighbor can be deduced from the placement of a sibling, without the presence of an explicit PPB. For example, a left terminal has two neighbors, but only one of them uses a PPB. The neighbor with the PPB can then be placed explicitly. Doing so allows the remaining neighbor without a PPB to be placed as well. For example, if the neighbor of a left terminal is placed in the upper left quadrant, then the sibling without a PPB can only occupy the lower left quadrant.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Implied neighbor placement. Bond directions can be omitted when inferrable.}
    \label{fig:implied-neighbor-placement}
\end{figure}

The assignment of relative coordinates to both terminals and all neighbors yields a double bond conformation.

Consider the encoding of PPBs for (\textit{E})-2-butene. Assume that indexes are assigned sequentially from left to right and the goal is to arrive at a trans (or anti) substituent orientation. Begin by placing the two double bond terminals on the x-axis. Left and right terminals are placed on the basis of succession. The left terminal (Atom 1) succeeds its neighbor (Atom 0), which should be placed in the lower-left quadrant. To achieve this, assign a PPB direction of \ttt{Up}. The right terminal (Atom 2) precedes its neighbor (Atom 3), which should be placed in the upper-right quadrant. To active this, assign a PPB direction of \ttt{Up}.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Assigning partial parity bonds to *trans*-2-butene.}
    \label{fig:assigning-ppb}
\end{figure}

In a similar manner assigned PPBs can be decoded. The process starts by placing Atom 1 to the left of Atom 2 along the x-axis. The PPB between the left terminal and its neighbor (Atom 0) uses the \ttt{Up} direction. The neighbor precedes the left terminal, so Atom 0 should be placed in the lower-left quadrant. The PPB between the right terminal and its neighbor (Atom 3) uses the \ttt{Up} direction. The neighbor (Atom 3) succeeds the right terminal (Atom 2). Therefore, Atom 3 is placed in the upper right quadrant. This results in the expected conformation (\textit{E}).

The above procedures assume that all PPBs are expressed in \textit{normal form}. In normal form, a PPB's source precedes it target. In other words, \ttt{source} is less than \ttt{target}. If a PPB is not in normal form, it must be \textit{inverted} before quadrants can be assigned. Inversion swaps the values of `source` and target attributes while simultaneously toggling the `direction` attribute. Inversion occurs in two contexts: conjugated polyenes and cycles. In the case of polyenes, the same PPB carries partial conformation for at least two different double bond systems. In the case of cycles, non-normal PPBs can be generated at bonds forming cycles. However, readers and writers must be capable of inverting PPBs wherever they appear.

Consider (\textit{E})-2-butene. The PPB spanning atoms 0 and 1 will typically be encoded as the tuple \ttt{(0, 1, Up)} because this orientation tracks the order of atom index. However, the PPB can be equivalently encoded as the tuple \ttt{(1, 0, Down)}. Before this representation can be used, it must first be inverted into the equivalent form represented by the tuple \ttt{(0, 1, Up)}.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Partial parity bond inversion. Direction must be inverted for bonds not expressed in normal form.}
    \label{fig:ppb-inversion}
\end{figure}

The distributed nature of conformational encoding means that several additive error states are possible:

\begin{itemize}
    \item{Overspecification. This state occurs when two or more PPBs share a common ancestor and simultaneously use the same parity, thus placing multiple neighbors into the same quadrant.}
    \item{Underspecification. A required and non-deducible PPB is absent. This state occurs when a double bond terminal connects to at least one neighbor through a PPB, but the terminal's mate does not.}
    \item{ Misspecification. A PPB is used between atoms lacking at least one double bond.}
\end{itemize}

A reader that detects one of these error conditions must abort the molecular graph being read and report the error state. It is possible for a conformation to be both overspecified and underspecified.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Error bond states. Because conformational descriptions are defined over three or more bonds, a variety of error states are possible.}
    \label{fig:bond-error-states}
\end{figure}

An exception to the underspecification rule applies in the case of conjugated polyenes. In this case a double bond with a terminal bearing a PPB does not require the opposing terminal to also bear a neighbor with a PPB. If the opposing terminal lacks neighbors, no error is generated and reading continues normally. If the opposing terminal has at least one neighbor, the bond to it need not be a PPB. In either case the conformation of the double bond in question remains undefined.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Error state exceptions. Conjugated dienes sharing a common PPB can lead to exceptional cases.}
    \label{fig:bond-error-state-exceptions}
\end{figure}

Given that conformational specification is distributed over three or more bonds, some conformations will be difficult or possible to represent. Cyclooctatetraene offers an example. Although the (\textit{E}, \textit{E}, \textit{E}, \textit{E}) conformation is accessible, the (\textit{Z}, \textit{E}, \textit{E}, \textit{E}) conformation is not. This problem arises because the same PPB encodes a partial conformation of two different double bonds.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Cyclooctatetraene. Some double bond conformations are not expressable.}
    \label{fig:cyclooctatetraene}
\end{figure}

\section{Configuration}

The third major element of molecular representation is \textit{configuration}. Configuration is the three-dimensional arrangement of neighbors about an atom. Dialect limits configuration to the special case of an atom with exactly four substituents placed at the vertexes of a tetrahedron. Here, "substituent" means an atomic neighbor; lone electron pairs are not considered substituents. One of the four substituents may be a virtual hydrogen.

A configuration is comprised of two components: an ordering of bonds to a central atom; and a \textit{configurational descriptor}. A configurational descriptor is a template for the relative three-dimensional positioning of neighbors about a central atom. Dialect supports two configurational descriptors: \ttt{Left} and \ttt{Right}. The context for the terms "left" and "right" will be explained shortly.

The descriptors \ttt{Left} and \ttt{Right} restrict neighboring atoms to the vertices of a tetrahedron whose center is the central atom. First, the mate of the first bond to the central atom is placed at an arbitrary vertex. Then shift the frame of reference by sighting along the first bond from the neighbor toward the central atom. Finally, arrange the remaining neighbors using the ordering of bonds to the central atom. For the \ttt{Left} variant, use a counterclockwise distribution. For the \ttt{Right} variant, use a clockwise distribution.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Configurational descriptor. Configuration is either encoded or reconstructed using the same model.}
    \label{fig:configurational-descriptor}
\end{figure}

If an atom bears one virtual hydrogen and three substituents, a modified procedure is used. The virtual hydrogen is first placed at an arbitrary vertex of the tetrahedron. The tetrahedron is then oriented along the axis pointing from the virtual hydrogen to the center. Neighbor atoms are then distributed in a counterclockwise (\ttt{Counterclockwise}) or clockwise (\ttt{Clockwise}) pattern to the remaining vertices.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Configurational descriptor with one virtual hydrogen. In this case, hydrogen assumes the role of the first substituent.}
    \label{fig:configurational-descriptor-vh}
\end{figure}

The association of a configurational descriptor with an atom implies a neighbor count of either four if no virtual hydrogens are present, or three if exactly one virtual hydrogen is present. In any other context, the use of a configurational descriptor is considered an error.

This restriction places some constraints around the use of configuration. For example, the assignment of stereodescriptors to central carbon atom of allenes and other cumulenes might seem like a reasonable extension. However, to do so is an error because the central atom lacks four substituents. Similarly, lone pairs must not take the place of atomic substituents, as might be the case with sulfoxides for example.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Invalid uses of configurational descriptors. Only configuration about four-coordinate tetrahedral atoms is supported.}
    \label{fig:invalid-configurational-descriptors}
\end{figure}

A configurational descriptor may be applied without regard to bond type. In other words, single, double, triple, and partial parity bond types can all be present. The only requirement that must be satisfied is four substituents or, equivalently, three substituents and a virtual hydrogen.

It is sometimes useful to manipulate a configuration in a way that preserves the relative three-dimensional positioning of neighbor atoms. This process is called \textit{transformation}. Five rules suffice to transform any configuration into any other:

\begin{itemize}
    \item{Virtualize. Replaces an atomic hydrogen neighbor with a virtual hydrogen. If the deleted bond was first in the bond ordering, toggle the configurational descriptor.}
    \item{Reify. Replaces a virtual hydrogen with an atomic hydrogen neighbor. The bond ordering must place the new neighbor in the second position.}
    \item{Swap. Exchanges any two bonds ordered second or higher while simultaneously toggling the configurational descriptor.}
    \item{Slide Left. Re-orders a bond from the first position to the second position. Disabled if the central atom carries a virtual hydrogen.}
    \item{Slide Right. Re-orders a bond from the second position to the first position. Disabled if the central atom carries a virtual hydrogen.}
\end{itemize}

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Configurational Transformations.}
    \label{fig:iconfigurational-transformations}
\end{figure}

It is an error to assign a stereodescriptor to any atom that is not tetracoordinate, or which has more than one virtual hydrogen.

Not all tetracoordinate atoms will be \textit{stereocenters}. A stereocenter is an atom whose "ligand permutation produces stereoisomers," as defined by Mislow and Siegel.\cite{mislow1984} In the context of Dialect, a stereocenter must be tetracoordinate. Readers should avoid the assignment of stereodescriptors to atoms that are not stereocenters.

Special handling is required for \textit{undefined stereocenters}. A stereocenter is undefined if it lacks a stereodescriptor. Omitting a stereodescriptor conveys to readers that no information about the stereocenter's configuration is known. This could imply that either a single configuration or a configurational mixture are present. This interpretation is consistent with the one used by molfile format.\cite{ctfileFormats}

\section{Syntax}

A Dialect string ("string") is a UTF-encoded sequence of zero or more characters chosen from the set: \ttt{A}; \ttt{B}; \ttt{C}; \ttt{D}; \ttt{E}; \ttt{F}; \ttt{G}; \ttt{H}; \ttt{I}; \ttt{K}; \ttt{L}; \ttt{M}; \ttt{N}; \ttt{O}; \ttt{P}; \ttt{R}; \ttt{S}; \ttt{T}; \ttt{U}; \ttt{V}; \ttt{W}; \ttt{X}; \ttt{Y}; \ttt{Z}; \ttt{a}; \ttt{b}; \ttt{c}; \ttt{d}; \ttt{e}; \ttt{f}; \ttt{g}; \ttt{h}; \ttt{i}; \ttt{k}; \ttt{l}; \ttt{m}; \ttt{n}; \ttt{o}; \ttt{p}; \ttt{r}; \ttt{s}; \ttt{t}; \ttt{u}; \ttt{v}; \ttt{0}; \ttt{1}; \ttt{2}; \ttt{3}; \ttt{4}; \ttt{5}; \ttt{6}; \ttt{7}; \ttt{8}; \ttt{9}; \ttt{+}; \ttt{-}; \ttt{.}; \ttt{\%}; \ttt{@}; \ttt{*}; \ttt{[}; \ttt{]}; \ttt{/}; \ttt{\char`\\}; \ttt{=}; and \ttt{\#}. The internal structure of a string reflects a depth-first traversal of the corresponding molecular graphs. As such, the syntax supports branches, cycles, and disconnected components.

Strings conform to an \textit{LL(1) grammar}. An LL(1) grammar is a context-free grammar whose strings can be parsed one character at a time from left to right with at most one character of lookahead. Additionally, LL(1) grammars expand the leftmost non-terminal first. This set of features makes LL(1) grammars such as the one used by Dialect a good fit for manually-written recursive descent parsers. LL(1) grammars can also be used as a basis for auto-generated parsers through packages such as ANTLR.\cite{parr2014} The full grammar for Dialect strings is available as a text file in this paper's Supporting Material.

Dialect's formal grammar is presented as a series of \textit{production rules} (aka "productions"). A production rule defines a transformation allowed under the grammar. These transformations collectively define the set of valid Dialect strings. Production rules can be used in the forward direction, when writing a string, or in the reverse direction, when reading a string.

A production rule is composed of two kinds of elements: \textit{terminals} and \textit{non-terminals}. A terminal is a character literal (e.g., "A"). A non-terminal is a reference to another production rule. This reference occurs through a name, which appears to the left of a separator (\ttt{::=}) in a production rule. To the right of a separator appear the allowed terminals and non-terminals for the rule.

Consider a hypothetical language composed of variable-length sequences of the lowercase letter "a". Such a language could be cast as the following two production rules:

\begin{lstlisting}
<text> ::= <a>*
<a>    ::= "a"
\end{lstlisting}

The quantifier (\ttt{*}) indicates that a text in this language consists of a sequence of zero or more instances of the production \ttt{<a>}. This production in turn is defined as the lowercase letter "a". Therefore, valid texts in this language include the empty string, \ttt{a}, \ttt{aa}, and \ttt{aaaaaa} to name a few.

\subsection{Atom}

Atoms carry most of the information in a Dialect string. The non-terminal \ttt{<atom>} is a selection among four alternatives.

\begin{lstlisting}
<atom> ::= <star> | <shortcut> | <selection> | <bracket>
\end{lstlisting}

The first atomic non-terminal, \ttt{<star>}, can assume one terminal value, the asterisk character (\ttt{*}). This "star atom" represents an \ttt{Atom} in which every attribute assumes its default value. The implicit hydrogen count of a star atom is zero.

\begin{lstlisting}
<star> ::= "*"
\end{lstlisting}

The next atomic production rule, \ttt{<shortcut>} is a non-terminal selected from the list: "B"; "C"; "N"; "O"; "P"; "S"; "F"; "Cl"; "Br"; and "I". An \ttt{Atom} encoded in this way ("shortcut atom") assigns the corresponding symbol to the \ttt{elemen} attribute. All other attributes retain their default values. The implicit hydrogen count of a shortcut atom is determined by the algorithm previously given.

\begin{lstlisting}
<shortcut> ::= "B" "r"? | "C" "l"? | "N" | "O" | "P"
             | "S" | "F" | "I"
\end{lstlisting}

The third atomic production rule, \ttt{<selection>} is a non-terminal selected from the list: "c"; "n"; "o"; "p"; and "s". An \ttt{Atom} encoded in this way ("selected shortcut atom") assigns the corresponding atom symbol to the \ttt{element} attribute and sets the \ttt{selected} attribute to `true`. All other atomic attributes retain their default values. The implicit hydrogen count of a selected shortcut atom is determined as described previously.

\begin{lstlisting}
<selection> ::= "c" | "n" | "o" | "p" | "s"
\end{lstlisting}

The fourth and most complex atomic production rule is \ttt{<bracket>} ("bracket atom"). A bracket atom can be used to set any atomic attribute. A bracket atom must be used when the `element` attribute of an \ttt{Atom} makes it ineligible for implicit hydrogen counting. Attributes not set within the \ttt{<bracket>} production rule will leave the corresponding atomic values in their default states.

\begin{lstlisting}
<bracket> ::= "[" <isotope>? <symbol> <stereodescriptor>?
              <virtual_hydrogen>? <charge>? "]"
\end{lstlisting}

The value of a bracket atom's \ttt{isotope} attribute is determined by the optional \ttt{<isotope>} non-terminal. It consists of between one and three digits encoding the integers 0-999. Leading zeros are disregarded when assigning the value of the \ttt{isotope} attribute. In other words, "007" and "7" are considered equivalent expressions of the value 7.

\begin{lstlisting}
<isotope> ::= <digit> <digit>? <digit>?
\end{lstlisting}

The values of a bracket atom's \ttt{element} and \ttt{selected} attributes are determined by the \ttt{<symbol>} production rule ("symbol"). Three non-terminal options are available: \ttt{<star>}; \ttt{<element>}; and \ttt{<selection>}.

\begin{lstlisting}
<symbol> ::= <star> | <element> | <selection>
\end{lstlisting}

The \ttt{<element>} non-terminal option assigns the atomic \ttt{element} attribute to the corresponding value while leaving the \ttt{selected} attribute as its default value. Given the large number of choices within the \ttt{<element>} non-terminal, the following production rule only defines the first several. For a complete list of alternative, see the full grammar in the Supporting Information.

\begin{lstlisting}
<element> ::= "A" ( "c" | "g" | "l" | "m" | "r" | "s"
            | "t" | "u" )
            | "B" ( "a" | "e" | "h" | "i" | "k" | "r" )?
            ...
\end{lstlisting}

A \ttt{<selection>} non-terminal found within a symbol sets the atomic \ttt{selected} attribute to \ttt{true}. The corresponding \ttt{element} attribute is assigned by capitalization. For example, the terminal \ttt{p} would assign the atomic \ttt{element} and \ttt{selected} attributes to the values \ttt{P} and \ttt{true}, respectively.

The \ttt{stereodescriptor} attribute of a bracket atom is determined by the \ttt{<stereodescriptor>} non-terminal. Allowed values are "@" and "@@", corresponding to the \ttt{stereodescriptor} values \ttt{Counterclockwise} and \ttt{Clockwise}, respectively.

\begin{lstlisting}
<stereodescriptor> ::= "@" "@"?
\end{lstlisting}

The \ttt{hydrogens} attribute of a bracket atom is controlled by the \\ \ttt{<virtual{\_}hydrogen>} non-terminal. This non-terminal is comprised of the terminal "H" followed by an optional \ttt{<digit>} non-terminal. A digit appearing after the "H" terminal sets \ttt{hydrogens} to the corresponding virtual hydrogen count. A digit of 0 (i.e., "H0") sets the \ttt{hydrogens} attribute to zero. If no digit is present, \ttt{hydrogens} is set to one.

\begin{lstlisting}
<virtual_hydrogen> ::= "H" <digit>?
\end{lstlisting}

The \ttt{<charge>} production rule sets the \ttt{charge} attribute of a bracket atom. This non-terminal begins with either the plus or minus terminals ("+", "-", respectively) and ends with an optional \ttt{<digit>} non-terminal. A missing digit causes the atomic \ttt{charge} attribute to be set to one. In other words, "+" sets the \ttt{charge} attribute to one and "-" sets it to minus one. A digit of zero (e.g., "+0" or "-0") sets the \ttt{charge} attribute to zero.

\begin{lstlisting}
<charge> ::= ( "+" | "-" ) <digit>?
\end{lstlisting}

Should an optional bracket non-terminal not be present, the atom must retain its corresponding default value. For example, the bracket "[C@H+]" lacks the \ttt{<isotope>} non-terminal so the value of the \ttt{isotope} attribute will remain as \ttt{None}. Similarly, the bracket "[13CH+]" lacks the \ttt{<stereodescriptor>} non-terminal, so the corresponding \ttt{configuration} attribute remains \ttt{None}.

\subsection{Bond}

An \ttt{Atom} may be connected to zero or more neighbors through a \ttt{Bond}, encoded with the non-terminal \ttt{<bond>}.

\begin{lstlisting}
<bond> ::= "-" | "=" | "#" | | "/" | "\"
\end{lstlisting}

Five terminals are available ("-", "=", "\#", "/", and "\textbackslash"). The first three ("-", "=", and "\#") set the \ttt{order} attribute of a \ttt{Bond} to 1; 2; or 3, respectively. The last two terminals, "/" and "\textbackslash", set the \ttt{order} attribute to 1 while also setting the \ttt{direction} attribute to \ttt{Up} and \ttt{Down}, respectively.

\subsection{Sequence}

An \ttt{Atom} may be assigned zero or more children through the \ttt{<sequence>} non-terminal ("sequence"). A sequence starts with a required \ttt{<atom>} non-terminal. If no allowed non-terminals follow, the corresponding \ttt{Atom} will have no children. Allowed non-terminals are chosen from the list: \ttt{<union>}; \ttt{<branch>}; and \ttt{<sequence>}.

\begin{lstlisting}
<sequence> ::= <atom> ( <union> | <branch> | <split> )*
\end{lstlisting}

The \ttt{<union>} non-terminal consists of an optional \ttt{<bond>} non-terminal followed by a mandatory non-terminal selected from the list: \ttt{<cut>} or \ttt{<sequence>}. If either of these latter non-terminals are detected but \ttt{<bond>} is not, bond elision is used.

\begin{lstlisting}
<union> ::= <bond>? ( <cut> | <sequence> )
\end{lstlisting}

A sequence can contain a union, which in turn can contain a sequence. This is an example of recursion, transitive though it may be. Although left-recursion is disallowed in LL(1) grammars, right recursion of the kind in \ttt{<sequence>} is allowed. Right recursion also occurs within the \ttt{<union>} and \ttt{<branch>} non-terminals.

The \ttt{<cut>} non-terminal ("cut") can takes two forms. A digit non-terminal can be used, enabling single-digit cut indexes. Double- or single-digit cut indexes, supporting the values zero through 99 inclusive, are available by prepending the percent character (\ttt{\%}).

\begin{lstlisting}
<cut> ::= <digit> | "%" <digit> <digit>
\end{lstlisting}

A cut signifies the disconnection of an \ttt{Atom} from its neighbor. The most common use for cut is to encode a cycle. The cut index serves as a placeholder for the neighboring \ttt{Atom}, which is located by scanning either forward or backward along a string. If the number of previous appearances of the cut index is even, then the string is scanned right. Otherwise, the string is scanned left. Although often used for cycles, cuts can be used for any bond, regardless of cycle membership. The only requirement is that a cut must be paired with another cut having the same index.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Cut. Each cycle yields a bond that must be cut.}
    \label{fig:cut}
\end{figure}

To prevent overflow, a cut index may be reused provided that it appears to the right of its last paired appearance.

An alternative to \ttt{<union>} within a sequence is the \ttt{<branch>} non-terminal ("branch"). Like union, branch joins a parent and child node through a bond. Wrapped by opening and closing parenthesis terminals ("(" and ")", respectively), branch encodes a sequence that may or may not be attached to its parent. Attachment occurs if the \ttt{<bond>} non-terminal is included. Alternatively, the subgraph will be detached if the \ttt{<detachment>} non-terminal appears. If neither \ttt{<bond>} nor \ttt{<detachment>} non-terminals are detected between the parentheses but a sequence is, then bonding of parent and child occurs through an elided bond.

\begin{lstlisting}
<branch> ::= "(" ( <detachment> | <bond> )?
             <sequence> ")"
\end{lstlisting}

The third option for associating a parent Atom with a child within a sequence is the \ttt{<split>} non-terminal ("split"). A split is a sequence of \ttt{<detachment>} and \ttt{<sequence>} non-terminals. The `<detachment>` non-terminal serves the same purpose as it does within a branch: the prevent the bonding of the parent \ttt{Atom} with a child.

\begin{lstlisting}
<split> ::= <detachment> <sequence>
\end{lstlisting}

Having defined sequence, it's now possible to define a string as an optional sequence. In other words, a Dialect string is either empty or contains a sequence. A string without a sequence encodes a molecular graph of zero nodes and zero edges.

\begin{lstlisting}
<string> ::= <sequence>?
\end{lstlisting}

\section{Reading Strings}

The goal of a Dialect reader is to transform a string input into a data structure output consistent with the string's content. The output data structure can take many forms. For example, a reader can merely validate a string by returning a boolean type. A more sophisticated reader can return a molecular graph capturing all \ttt{Atom} and \ttt{Bond} attributes and connectivity relationships.

Dialect strings are read one character at a time starting at the leftmost character and finishing at the rightmost character. The first character sets an initial reader state, and each subsequent character causes a state transition. The cumulative application of these state transitions yields the data structure to be returned.

Readers that capture \ttt{Atom}-\ttt{Atom} connectivity will typically maintain a reference to a \textit{root atom}. The root atom is initially undefined. On reading the first complete \ttt{<atom>} non-terminal, the corresponding \ttt{Atom} is constructed and set as the root atom. Subsequently processing a complete \ttt{<union>}, \ttt{<branch>}, or \ttt{<split>} non-terminal triggers three changes: (1) a child \ttt{Atom} is constructed; (2) the child is connected to the current root atom, unless a \ttt{<detachment>} non-terminal intervenes; and (3) the root atom is replaced with the child atom.

The presence of a branch adds some nuances over a union. The leading open parenthesis terminal ("(") signifies that the current root atom will later be re-exposed. This operation can be supported by a stack. At the start of a branch, the current root is pushed to the stack. At the end of the branch, the stack is popped and its top value is assigned as the new root atom.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Stack for branches.}
    \label{fig:branches-stack}
\end{figure}

Further workflow adjustments can handle cuts. The necessary state can be maintained by a 100-element array. The first occurrence of a cut creates an entry in the array at the cut index consisting of the bond and the root atom. The second occurrence of a cut takes the entry at the cut index, attaching the atom in the entry to the current root atom. Either the leftmost or rightmost bond can be used to make the connection, provided they are \textit{compatible}.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Array for cuts. The array's fixed length reflects the syntactic constraint on cut index.}
    \label{fig:cuts-array}
\end{figure}

Two cut bonds are compatible if they possess the same \ttt{order} attributes, and their \ttt{direction} attributes are complementary. Bonds with undefined \ttt{direction} are always complementary. A bond with a defined \ttt{direction} is complementary to any \ttt{Bond} with an undefined state, or a \ttt{Bond} with an opposing \ttt{direction}. For example, an elided \ttt{Bond} is compatible with a single \ttt{Bond}. Likewise, two elided \ttt{Bond}s are compatible. A \ttt{Bond} with a \ttt{direction} of \ttt{Up} is incompatible with another \ttt{Bond} with a \ttt{direction} of \ttt{Up}, but compatible with one having a \ttt{direction} of \ttt{Down}.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Bond compatibility. Two non-elided bonds across a cut must have compatible `order` and \ttt{direction} attributes.}
    \label{fig:bond-compatibility}
\end{figure}

Decoding atomic configuration from a string requires the order of attachment from each \ttt{Atom} to its neighbors to be recorded. Cuts are treated as if the neighboring atom itself were detected. In this regard, the first member of a cut pair requires special attention. The neighboring \ttt{Atom} won't be available until the paired cut index is found. Nevertheless, it must ultimately appear before the neighbors in subsequent unions, branches, and splits. This can be accomplished though the use of a placeholder or some other mechanism that preserves the overall relative order of attachment. No such special treatment is needed for the second member of a cut pair.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Order of attachment. A placeholder target index preserves the order of attachment.}
    \label{fig:order-of-attachment}
\end{figure}

Readers must not assume that detachment (the period terminal, ".") implies the presence of disconnected components. This assumption is most likely to arise in the context of ad-hoc parsers using regular expressions, string matching, and the like. For example, the single molecule propane can be encoded using the string "C1C.C1".

The partial parity bond model is unusual in that it spreads conformational information over three or more bonds. Regardless, readers must report as an error any string containing an isolated PPB (e.g., "C/C"). Beyond this check, generating (\textit{E}) and (\textit{Z}) stereodescriptors from PPBs is not a straightforward process. Performing such a conversion requires a method capable of translating PPB into a localized bond descriptor. One approach is to first consider the parity of each double bond terminal individually. A separate step would then relate the individual terminal parities to a double bond conformation parity. This intermediate parity can then be used to perform the final conversion.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Localized configurational descriptor. Such a representation can be useful when bridging Dialect to other representations.}
    \label{fig:localized-configurational-descriptor}
\end{figure}

A reader must assume that any input string can contain errors, and take appropriate steps to report them. The most useful errors will report a specific cause. Some will also report one or more cursor indexes. The most common mandatory errors are:

\begin{enumerate}
    \item{Invalid character (position). An unexpected character was encountered. A list of acceptable characters is helpful, but not required.}
    \item{Unexpected end-of-line. Input ended unexpectedly.}
    \item{Unbalanced cut (position). A cut with a given index appears an odd number of times.}
    \item{Incompatible cut bonds (position, position). The bonds to a pair of cuts are incompatible.}
    \item{Delocalization subgraph lacks perfect matching. Before reporting this error, steps to remove ineligible atoms should be taken as described in the next section.}
    \item{Partial parity bond not allowed (position). Neither terminal of a PPB possesses a double bond. Strings such as "C/C" and "C{\textbackslash}C" contain isolated PPBs, which are invalid. A reader encountering such strings must report an error.}
\end{enumerate}

A reader may also report optional errors, including:

\begin{enumerate}
    \item{Impossible isotope. A negative implied mass number results from the atom. For example, "[2C]".}
    \item{Impossible valence. The valence at an `Atom` is impossibly high. For example, "C(C)(C)(C)(C)C".}
    \item{Impossible charge. An \ttt{Atom}'s charge gives it an apparent negative electron count. For example, "[C+7]".}
\end{enumerate}

\section{Pruning}

As noted previously, a delocalization subgraph is invalid if it lacks a perfect matching. The one exception is when a selected atom can be deleted from the delocalization subgraph through \textit{pruning}. Pruning unsets the \ttt{selected} attribute of a selected atom without corresponding promotion of any attached bonds.

An atom must be pruned if its subvalence equals zero. None of the bonds to such an atom can be promoted without altering the atom's \ttt{charge} attribute. Pruning the atom ensures the stability of its \ttt{charge} attribute, without interfering with bond promotion elsewhere. Viewed from another perspective, an atom with zero subvalence lacks unpaired electrons - at least within the narrow boundaries of the Dialect valence model. Such atoms can only form double bonds through changes to atomic charge.

If a selected atom bears a non-zero \ttt{charge} attribute, subvalence is computed using the isoelectronic element's default valences. For example, a selected nitrogen atom with a charge of +1 would use the default valences for carbon. A selected phosphorous atom with a charge of -1 would use the default valences of sulfur. And so on. If no default valences are found in this way (e.g., "[c+2]"), a reader must generate an error. Writers must not encode such atoms.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Gratuitous selection.}
    \label{fig:gratuitous-selection}
\end{figure}

Pruning becomes necessary in cases of gratuitous atomic selection. This occurs whenever style, tradition, or convenience overrides necessity. Consider furan represented as the string "c1ccco1". Selecting any atom is unnecessary because furan does not exhibit DIME. But selecting the oxygen atom is particularly unnecessary because it lacks an unpaired electron and so will never lead to DIME. It is nevertheless convenient to select the carbon atoms because all bonds can then be elided. The resulting representation, "c1cccO1" leads to a delocalization subgraph with a perfect matching and hydrogen counts consistent with the original encoding.

Writers are encouraged, but not required, to avoid gratuitous atom selection. Readers, however, must always be prepared to prune.

\section{Writing Strings}

Whereas a reader transforms a string into a data structure, the goal of a writer is the opposite: to transform a data structure into a valid string. This data structure will most commonly take the form of a molecular graph, but other forms are possible. Although a graph itself could be used as the basis for an input representation, a more primitive representation such as an adjacency list could also be used. For practical reasons, the input data structure is likely to resemble the output from a reader. Such an arrangement makes it possible to encode and decode strings with minimal intermediate translation.

Regardless of the form taken by the input data structure, it must be traversable in \textit{depth-first} order. A depth-first traversal operates over a set of nodes bound by a set of connectivity relationships, which are typically edges. Traversal proceeds by successively replacing each node as the center of focus, or root. Each new root is selected from the untraversed neighbors of the current root. Iteration eventually selects all nodes, at which point the traversal ends.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Depth-first traversal.}
    \label{fig:depth-first-traversal}
\end{figure}

A writer intercepts the depth-first traversal of an input data structure to write node and edge representations. The \ttt{<atom>} and \ttt{<bond>} non-terminals are used for this purpose. There are no requirements around style. For example, it's equally valid to represent the carbon atom of methane as either "C" or "[CH4]". Single bonds may be elided or not. Similarly, selection can be used or not for eligible atoms. Although an organization may seek to standardize certain styles of string output, any syntactically-valid string must be considered valid by a reader.

The presence of branches within an input data structure is encoded via the \ttt{<branch>} data structure. A useful tool for this purpose is a \ttt{stack}. A stack is a data structure that allows items to be added individually and removed ("popped") in the reverse order of addition. A writer begins by pushing the current branch onto the stack and extending it. When a new branch is encountered, it is pushed to the stack and extended. When the branch terminates, the current branch is popped and its contents are appended to the stack's new top item.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Stack for branch assembly.}
    \label{fig:branch-stack}
\end{figure}

Cycles are encoded using the \ttt{<cut>} non-terminal. This one- or two-digit integer replaces one terminal of a cycle closure bond. The challenge is to unambiguously supply these integers while allowing re-use to avoid overflow. This can be accomplished with a \textit{pool}. A pool is a data structure that yields a numerical index given an ordered pairing of atomic indexes. The numerical index will not be re-issued until the pool receives the corresponding reversed pairing.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Pool for assigning cut indexes.}
    \label{fig:pool}
\end{figure}

A pool can be used by a writer in the following way. The presence of a cycle during depth-first traversal is indicated by an atom that has already been traversed. On encountering a cycle, a writer requests an index from the pool, submitting the corresponding atomic indexes as an ordered pair. Later, the same bond must be reversed in the reverse direction. When it is, the writer once again requests an index, but this time using a reversed pairing. Doing so yields the same index, while simultaneously freeing it for later use.

The encoding of cuts is likely to pose special challenges for writers. For a cut across a PPB, care must be taken to report the correct parity. Consider \textit{trans}-cyclooctene. The correct encoding accounts for the reverse in parity at the left-hand side (e.g., "C{\textbackslash}1C=C/CCCCC/1"). Moreover, the double-encoding of the PPB bond type can lead to inconsistencies that must be reported by a reader (e.g., the invalid string "C/1C=C/CCCCC/1"). Such errors can be avoided by encoding the bond type of one side or the other side of a cut, but not both. (e.g., "C{\textbackslash}1C=C/CCCCC/1" or "C1C=C/CCCCC/1"). A cut across a bond attached to an atom with a stereodescriptor must take bond ordering into consideration. For example, the string "O[C@H]1NC1" encodes the same absolute configuration as the string "O[C@H](C1)N1", but the \textit{opposite} configuration as "O[C@H](N1)C1".

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Difficult cuts.}
    \label{fig:difficult-cuts}
\end{figure}

The presence of conformational restriction about a double bond presents a special challenge to writers. Few other molecular graph systems distribute a conformation descriptor over two or more bonds the way that PPBs does. Instead, descriptors are more likely to be localized at the double bond. Producing the necessary PPBs from a descriptor localized at the double bond is likely to be non-trivial. An approach to the reverse problem was presented previously (Reading Strings section). That approach can also be adapted to the reverse problem.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Writing partial parity bonds through a localized conformation descriptor.}
    \label{fig:writing-ppb}
\end{figure}

Writers should carefully weigh the non-negligible costs of atom selection. Algorithms for doing so sometimes involve the perceptions of cycles, and often exhibit superlinear time complexity. Very often a reader must perform a global deselection to arrive at a localized valence bond representation, which at the very least requires pruning and a maximal matching procedure. In this sense atom selection imposes two sets of costs: one on the writer and the other on every subsequent reader forever into the future.

Writers must ensure that all selected atoms can be deselected. Consider pyrrole, erroneously encoded as the string "n1cccc1". The nitrogen atom can not be pruned because its subvalence is 1 (3 - 2). The DS therefore contains all five atoms and all five bonds. A perfect matching does not exist. A reader receiving such a string must report an error. To avoid this outcome, a writer can consume the subvalence by adding a virtual hydrogen, as in the string "[nH]1cccc1". The subvalence of nitrogen in this case is zero, so the atom can be pruned. Doing so leaves a DS with four nodes, four edges, and a perfect matching. Readers will consider the string valid. Nevertheless, the larger question of gratuitous selection remains.

\section{SMILES Compatibility}

To maximize compatibility with contemporary software, Dialect is based on the SMILES language. Given the widespread adoption of "SMILES," arriving at a specific definition of the language is surprisingly challenging. No single document completely specifies SMILES syntax and semantics. The first description of the SMILES language in the primary literature is contained within a 1988 article (referenced here as "the article").\cite{weininger:1988} The article, by its own admission, is incomplete: "... isomeric SMILES [defined in the book chapter as covering isotopism, conformation, and configuration (p. 91)] is not otherwise covered in this paper." (p. 34). Weininger later refined and expanded SMILES in a 2003 book chapter,\cite{weininger2008} referred to here as "the book chapter." A third major source of information is the Daylight Theory Manual (referred to here as "the manual"), a website operated by Daylight Chemical Information Systems, Inc.\cite{daylightTheory} Information about SMILES is also contained in various software packages and associated documentation, slide decks, websites, mailing lists, and private communications.

For the purpose of providing a foundation for Dialect, "SMILES" was defined as the language described in the article and the book chapter. All other sources --- irrespective of quality or apparent claim to authority --- were considered out of scope. The manual was disregarded because it merely recapitulated material in the article and chapter, and due to the possibility of future alteration. The remaining sources were disregarded due to lack of authority, poor accessibility, or both. This narrow focus simplifies the task of evaluating compatibility between Dialect and SMILES. However, it does not provide a method for testing compatibility due to the lack of an authoritative SMILES compliance suite.

It is nevertheless possible to compare Dialect and SMILES on a feature-by-feature basis. Dialect is intended to be a subset of SMILES, meaning that every feature present in Dialect is ideally also present in SMILES. As will be seen, certain factors make this undesirable in the strictest sense. With a few notable exceptions, however, Dialect can be viewed as a subset of SMILES.

The book chapter defines a fixed set of element symbols which even at the time of publication was obsolete (p. 83). The symbol for element 105 is given as Ha (Hahnium), despite IUPAC's resolution of this naming controversy in favor of the symbol Db (Dubnium) years prior.\cite{ohrstrom1997} Accordingly, Dialect disallows the symbol Ha. To support the full set of currently-named transuranium elements, Dialect adds the symbols: Db; Sg; Bh; Hs; Mt; Rg; Cn; Nh; Fl; Mc; Lv; Ts; and Og. Moreover, Dialect explicitly accepts all future element symbols approved by IUPAC. These additions are one of the reasons Dialect can't be considered a strict subset of SMILES.

The article implies that comma (\ttt{,}) is either an element symbol or a valid SMILES character: "This subset uses only the symbols H, C, N, O, P, S, F, Cl, Br, I, and (,) and digits, with the following four rules..." (p. 33). The comma character is explicitly disallowed in Dialect.

The book chapter defines a recursive grammar for branching which, perhaps without the knowledge of the author, results in multiple branching (p. 86). Dialect disallows the multiple branching constructs, including for example "*((*))*" and "*(((*)))*".

"Reaction SMILES" is an extension described in the book chapter (p. 89), but disallowed in Dialect. Specifically, the greater than symbol (\ttt{>}) is not a valid Dialect character. This precludes strings such as "*>>*" in Dialect. Furthermore, Dialect lacks the "map" attribute, used for atom-atom mapping in reactions, and its accompanying syntax.

Dialect only supports two stereodescriptors, encoded with the terminals "@" and "@@". The book chapter provides a recursive grammar for stereodescriptors that allows multi-symbol descriptors such as "@@@" and "@@@@" (p. 94). The book chapter also discusses non-tetrahedral descriptors including "@AL1", "@AL2", "@1", and "@SP1". None of these are supported by Dialect. Nor does Dialect support the application of tetracoordinate stereodescriptors to odd cumulenes as described in the book chapter. Although some of these forms of configuration are supported today, it's unusual to find them all supported. The reason is simple: in the context in which SMILES is most often used, these forms of stereochemistry are rare. To support them places a burden on implementors with little payoff.

Along these lines, Dialect does not recognize atoms bearing either arsenic (As) or selenium (Se) as selectable. Doing so would increase the complexity of Dialect unnecessarily. Although the paper implies selectability ("Elements other than C, N, O, P, S, As, and Se are not yet dealt with in an aromatic context." p35), default valences for the latter two atoms are not provided. Even if they were, the elements As and Se must be bracketed. As such, special rules for handling them in the context of pruning would be required. Moreover, molecules bearing the elements As or Se rarely exhibit DIME. Consider the arsenic analog of pyrrole ("arsole"). Selecting the arsenic-bearing atom is gratuitous because it does not contribute to DIME. Finally, allowing the selection of atoms bearing As or Se does nothing to address molecules with atoms bearing germanium (Ge), tin (Sn), or other heavy main group elements that may be discovered in the future to participate in DIME.

It's not clear whether SMILES allows a virtual hydrogen count on hydrogen itself. The article implies that the element symbol H (hydrogen) must not have an implicit hydrogen count when used within a bracket sequence: "There are few exceptions to the hydrogen-suppression convention, the most obvious being specification of a proton, [H+], and molecular hydrogen, [H][H]." (p. 34) If so, the author may have erroneously believed that parsers would be incapable of processing constructs such as `[HH]`. The book chapter appears to reiterate this view, noting that "explicit hydrogen specification is required" in the case of "hydrogens connected to other hydrogens, e.g. [H][H], molecular hydrogen" (p. 97). To be clear, Dialect supports any valid virtual hydrogen count on hydrogen itself. For example, the following are all valid, if chemically nonsensical, strings: "[HH]"; "[HH2]"; and "[HH3]". Depending on interpretation, this support may or may not render Dialect a superset of SMILES.

The book chapter ascribes ambiguous meaning to detachments: "... In terms of the valence model being represented, the dot literally represents a bond of formal order zero: the atoms on either side of the dot are explicitly not bonded to each other." (p. 88). The state of not being bonded and the state of having a zero-order bond are not necessarily equivalent.\cite{clark2011} To be clear, Dialect explicitly disallows the zero bond order interpretation of detachments and the dot character.

SMILES assigns no explicit upper or lower bounds to numerical atomic attributes. These boundaries are nevertheless crucial for implementors, who often seek a data type just large enough to prevent underflow or overflow. For this reason, Dialect sets both upper and lower bounds on the following atomic attributes: isotope (0 $\leq$ value \textless 1000); charge (-10, \textless value \textless 10); and virtual hydrogen (\ttt{hydrogens}, 0 $\leq$ value \textless 10). Although it might be argued that lower bounds on physical quantities such as isotope and hydrogen count should be obvious in SMILES, the lack of specificity forces implementors to complete work that should have been done by the language's creator.

Dialect provides a few semantic clarifications not fully addressed in SMILES. A detailed algorithm for determining implicit hydrogen count is provided, together with the required valence table. Unlike SMILES, Dialect explicitly considers the case of computing implicit hydrogens on selected atoms. Dialect also provides detailed algorithms, absent in SMILES, for selection and deselection. These are based in graph theory rather than the ambiguous and overloaded chemical concept of "aromaticity" used in SMILES. Uniquely, Dialect introduces the concepts of "pruning" and "gratuitous selection." Neither the paper nor the book chapter indicate those bonds that can be promoted during deselection, but Dialect does. Finally, Dialect syntax is based on a formal grammar and tooling rooted in decades of computer science research, whereas SMILES syntax is based for the most part on imprecise natural language descriptions.

Dialect also resolves some internal contradictions within SMILES. The article reports 3 as the only default valence for nitrogen (p. 34), whereas the book chapter reports two default valences: 3 and 5 (p. 84). Dialect uses the latter default valences. Eligible atoms ("the organic subset") are reported in the article as: B; C; N; O; P; S; F; Cl; Br; and I (p. 32). However, an example later in the article indicates that "Hn1cccc1" with its unbracketed hydrogen is a valid string (p. 35). Dialect explicitly rejects strings containing unbracketed hydrogen atoms. On the topic of atom selection ("aromaticity"), the article implies that selected atoms must be part of a cycle: "To qualify as aromatic, all atoms in the ring must be sp2 hybridized and the number of available 'excess' $\pi$ electrons must satisfy Hückel's 4{\textit{N}} + 2 criterion." (p. 34). The book chapter, however, presents two examples of valid strings that do not encode cycles ("cc" and "cccc", p. 85). Consistent with the latter view, Dialect explicitly allows the selection of atoms regardless of cycle membership.

\begin{table}
    \caption{Some differences between SMILES and Dialect.}
    \centering
    \begin{tabular}{p{0.55\linewidth} p{0.15\linewidth} p{0.15\linewidth}}
        \hline
        Feature & SMILES & Dialect \\
        \hline
        element symbol Ha & yes & no \\
        element symbols Db; Sg; Bh; Hs; Mt; Rg; Cn; Nh; Fl; Mc; Lv; Ts; and Og & no & yes \\
        future element symbols approved by IUPAC & no & yes \\
        comma symbol (\ttt{,}) & maybe & no \\
        multiple branching such as "*((*))*" & yes & no \\
        reactions using greater than symbol (\ttt{>}) & yes & no \\
        atomic "map" attribute & yes & no \\
        extended stereodescriptors & yes & no \\
        use of stereodescriptors on odd cumulene centers & yes & no \\
        virtual hydrogen count on hydrogen & no & yes \\
        detachments are bonds of "formal order zero" & maybe & no \\
        upper and lower bounds on atomic properties & no & yes \\
        nitrogen default valence includes 5 & maybe & yes \\
        unbracketed hydrogen atom & maybe & no \\
        acyclic atom selection & maybe & yes \\
        atoms with elements As and Se are selectable & yes & no \\
        non-elided bonds may be promoted during selection & maybe & no \\
        \hline
    \end{tabular}
    \label{table:feature-comparison}
\end{table}

\section{Discussion}

Dialect's main advantage as a molecular serialization format is high information density. The most common atom types can be represented with just one character. The worst case atom, using every available atomic attribute, requires 14 characters. The most common bond types can be encoded implicitly. In the worst case one character per bond is required.

These metrics compare favorably with alternatives. Consider the molfile (V2000) format. 32 characters per atom are required in the best case, and 51 characters in the worst. All bonds, regardless of type, require between nine and 12 characters. Atomic charges and isotopes typically require additional characters.

The information density of Dialect strings makes them attractive for several applications. In memory-constrained environments such as those found on handheld devices and network serialization, many more Dialect strings can be present than alternative encodings. In-memory structure search over large collections becomes feasible. Because a Dialect string often fits within one line on a terminal, it can be used for data entry in manual interactive shells such as a real-eval-print loops (REPLs) or notebooks.

Another advantage is lossless serialization and deserialization. The line notation InChI might appear to compete with Dialect in this sense. Although InChI can be read and written, the InChI Technical manual notes that InChI's value lies in molecular identification. The authors of InChI have repeatedly noted that InChI is not a solution to the molecular serialization problem. The reason presumably lies with the fact that both serialization and deserialization require intimate knowledge of the InChI canonicalization algorithm. To date, no third-party reader or writer of InChI has ever been released.

Despite its information density, Dialect can faithfully encode and decode most of what chemists would consider "organic molecules." This is evidenced by the widespread presence of Dialect strings within large, public-facing databases such as PubChem, ChEMBL, ChEBI, and eMolecules.

Dialect's compact representation requires some tradeoffs regarding expressiveness. Extended bonding beyond the delocalization subgraph is not supported. This limitation excludes many types of molecule, including organometallics and delocalization-\-stabilized ions. Zero-\-order bonds are not supported, making it difficult to accurately encode coordination complexes. Only four-\-coordinate, tetrahedral stereocenters can be encoded, excluding many forms of chirality, including helical chirality, all forms of non-\-tetrahedral stereochemistry, and lone-\-pair tetrahedral centers. Conformational restrictions beyond the isolated double bond are not supported, which excludes important molecules such as those exhibiting atropisomerism. Other formats such as CDXML and molfile support enhanced stereochemical features enabling the differentiation of various kinds of partial stereochemical information, but Dialect does not.

Broader expansion of Dialect's capabilities could be possible through \textit{metaformats}. A metaformat embeds one or more Dialect strings within a surrounding serialization format.For example, two- or three-dimensional coordinates could be associated with each atom through a metaformat that includes a dictionary mapping atomic index to coordinate. Collections of atoms or bonds could likewise be encoded to replicate the enhanced stereochemistry features of other formats. And so on. However, the utility of such extensions should be weighed against Dialect's main value proposition: high information density. An application attempting to use a verbose metaformat may benefit from adopting better-suited format instead.

Setting aside the many technical and usability issues a metaformat would raise, versioning is likely to play an important role. Dialect itself lacks any mechanism to convey the concept of version. This stands in contrast to InChI, which not only encodes a version identifier, but has done so from its first release. Adding a version identifier would, unfortunately, break compatibility with the large number of existing SMILES software. Metaformats offer an opportunity to address this limitation.

The most noteworthy feature of Dialect is its compatibility with SMILES. Contemporary SMILES implementations will be able to read and write Dialect strings --- at least to a point. Incompatibilities will arise from two main sources: (1) those features of SMILES that Dialect has deliberately omitted; and (2) those aspects of the SMILES language that are incompletely-specified, ambiguous, or self-contradictory.

The availability of a minimal yet highly functional, fully-specified core language offers many opportunities to improve data quality. One of the most important will be a reference implementation, the design and source code of which will be discussed later. A freely-available reference implementation in turn makes automated validation suites possible. These suites can improve data quality by reporting syntax and semantic differences among implementations, preferably before release. A reference implementation taken together with the guidelines for readers and writers in this paper should make it possible to write software that conforms to a very high level of precision, regardless of programming language or paradigm. Given verified implementations, performance optimizations can be considered. The existence of a core language specification should also aid standardization efforts, either for Dialect itself, or SMILES. Finally, the development of better line notations is only possible given a thorough understanding of the scope and limitations of existing options. Here, metaformats could offer a bridge from the present to the future.

\section{Conclusion}

This paper describes Dialect, a subset of the widely-\-used SMILES line notation. Dialect's features are detailed at four levels: constitution; delocalization; conformation; and configuration. Constitutionally, Dialect can encode any molecule conforming to the valence bond model. In the event of undesired symmetry artifacts due to delocalization, Dialect offers a mitigation based in graph theory. Conformational isomerism of alkenes is supported by partial parity bonds. The configurations of tetracoordinate, stereogenic atoms are encoded through the use of a parity enumeration and conventions around its use.

Dialect's syntax is described in detail through both graphical diagrams and a formal grammar. The latter method concisely summarizes the complete set of strings that could be considered valid Dialect representations. The formal grammar was deliberately structured to be used either directly with an automated parser generator or with a hand-written recursive-descent parser.

The complete set of operations needed to interpret Dialect's syntax semantically are described in detail. This is a crucial component of the language's definition because Dialect achieves its information density largely by favoring convention over encoding. As an aid to difficult cases, guidelines for readers and writers are included.

As a subset of SMILES, Dialect can be used with a wide range of SMILES software today. In this sense, Dialect may seem to offer nothing new or even of value. However, Dialect has been defined at a level of detail that SMILES never was. This difference makes it possible to use Dialect in unique ways. Open reference implementations and validation suites can now be developed and deployed. Families of extensions can be built, each one based on the same unambiguous foundation. Formal standardization becomes more feasible given detailed reference material on which to draw. Finally, it is only through the clear demarcation of boundaries that the frontier becomes visible.

\clearpage
\printbibliography

\end{document}

### 2022-03-15 bin/build.sh

@@ -1,4 +1,11 @@
-#!/usr/bin/env bash
+mkdir -p build
 
-mkdir -p ./build
-pandoc --citeproc --pdf-engine=xelatex --resource-path=src -s src/manuscript.md -o build/manuscript.pdf
+cp manuscript.bib build
+cp manuscript.tex build
+cp -r assets build
+
+cd build
+
+pdflatex manuscript
+bibtex manuscript
+pdflatex manuscript

### 2022-03-15 src/railroad/atom.js

@@ -1,11 +1,0 @@
-const { Diagram, Choice, NonTerminal } = require('../../lib/railroad-diagrams');
-
-module.exports = Diagram(
-  Choice(
-    0,
-    NonTerminal('star'),
-    NonTerminal('shortcut'),
-    NonTerminal('selection'),
-    NonTerminal('bracket'),
-  )
-)

### 2022-03-15 src/railroad/bond.js

@@ -1,12 +1,0 @@
-const { Diagram, Choice, Terminal } = require('../../lib/railroad-diagrams');
-
-module.exports = Diagram(
-    Choice(
-        0,
-        Terminal('-'),
-        Terminal('='),
-        Terminal('#'),
-        Terminal('/'),
-        Terminal('\\')
-    )
-);

### 2022-03-15 src/railroad/bracket.js

@@ -1,22 +1,0 @@
-const { Diagram, NonTerminal, Optional, Stack } = require('../../lib/railroad-diagrams');
-
-module.exports = Diagram(
-  NonTerminal('['),
-  Stack(
-    Optional(NonTerminal('isotope')),
-    NonTerminal('symbol'),
-    Optional(
-      NonTerminal('stereodescriptor')
-    ),
-    Optional(
-      NonTerminal('virtual-hydrogen')
-    ),
-    Optional(
-      NonTerminal('charge')
-    ),
-    Optional(
-      NonTerminal('extension')
-    )
-  ),
-  NonTerminal(']')
-);

### 2022-03-15 src/railroad/branch.js

@@ -1,14 +1,0 @@
-const {Diagram, NonTerminal, Terminal, Optional, Choice } = require('../../lib/railroad-diagrams');
-
-module.exports = Diagram(
-  Terminal('('),
-  Optional(
-    Choice(
-      0,
-      Terminal('.'),
-      NonTerminal('bond')
-    )
-  ),
-  NonTerminal('sequence'),
-  Terminal(')')
-);

### 2022-03-15 src/railroad/charge.js

@@ -1,10 +1,0 @@
-const { Diagram, Choice, Terminal, NonTerminal, Optional } = require('../../lib/railroad-diagrams');
-
-module.exports = Diagram(
-  Choice(
-    0,
-    Terminal('+'),
-    Terminal('-')
-  ),
-  Optional(NonTerminal('non_zero'))
-);

### 2022-03-15 src/railroad/cut.js

@@ -1,13 +1,0 @@
-const {Diagram, Choice, Terminal, Sequence, NonTerminal } = require('../../lib/railroad-diagrams');
-
-module.exports = Diagram(
-  Choice(
-    0,
-    NonTerminal('digit'),
-    Sequence(
-      Terminal('%'),
-      NonTerminal('non_zero'),
-      NonTerminal('digit')
-    )
-  )
-);

### 2022-03-15 src/railroad/digit.js

@@ -1,17 +1,0 @@
-const {Diagram, Terminal, Choice } = require('../../lib/railroad-diagrams');
-
-module.exports = Diagram(
-  Choice(
-    0,
-    Terminal('0'),
-    Terminal('1'),
-    Terminal('2'),
-    Terminal('3'),
-    Terminal('4'),
-    Terminal('5'),
-    Terminal('6'),
-    Terminal('7'),
-    Terminal('8'),
-    Terminal('9')
-  )
-);

### 2022-03-15 src/railroad/element.js

@@ -1,51 +1,0 @@
-const {Diagram, HorizontalChoice, Sequence, Terminal, Choice , Optional, NonTerminal } = require('../../lib/railroad-diagrams.js');
-
-module.exports = Diagram(
-  HorizontalChoice(
-    Sequence(
-      Terminal('A'),
-      Choice(
-        0,
-        Terminal('c'),
-        Terminal('g'),
-        Terminal('l'),
-        Terminal('m'),
-        Terminal('r'),
-        Terminal('s'),
-        Terminal('t'),
-        Terminal('u')
-      )
-    ),
-    Sequence(
-       Terminal('B'),
-        Optional(
-          Choice(
-            0,
-            Terminal('a'),
-            Terminal('e'),
-            Terminal('h'),
-            Terminal('i'),
-            Terminal('k'),
-            Terminal('r')
-          )
-        )
-    ),
-    Sequence(
-      Terminal('C'),
-      Optional(
-        Choice(
-          0,
-          Terminal('a'),
-          Terminal('d'),
-          Terminal('e'),
-          Terminal('f'),
-          Terminal('l'),
-          Terminal('m'),
-          Terminal('n'),
-          Terminal('o')
-        )
-      )
-    ),
-    NonTerminal('...')
-  )
-)

### 2022-03-15 src/railroad/extension.js

@@ -1,9 +1,0 @@
-const { Diagram, Terminal, NonTerminal, Optional } = require('../../lib/railroad-diagrams.js');
-
-module.exports = Diagram(
- Terminal(':'),
-  NonTerminal('digit'),
-  Optional(NonTerminal('digit')),
-  Optional(NonTerminal('digit')),
-  Optional(NonTerminal('digit'))
-)

### 2022-03-15 src/railroad/isotope.js

@@ -1,13 +1,0 @@
-const { Diagram, Sequence, Optional, NonTerminal } = require('../../lib/railroad-diagrams.js');
-
-module.exports = Diagram(
-  Sequence(
-    NonTerminal('non_zero'),
-    Optional(
-      NonTerminal('digit')
-    ),
-    Optional(
-      NonTerminal('digit')
-    )
-  )
-)

### 2022-03-15 src/railroad/selection.js

@@ -1,13 +1,0 @@
-const { Diagram, Choice, Terminal } = require('../../lib/railroad-diagrams.js');
-
-module.exports = Diagram(
-  Choice(
-    0,
-    Terminal('b'),
-    Terminal('c'),
-    Terminal('n'),
-    Terminal('o'),
-    Terminal('p'),
-    Terminal('s')
-  )
-)

### 2022-03-15 src/railroad/sequence.js

@@ -1,13 +1,0 @@
-const {Diagram, ZeroOrMore, Choice, NonTerminal } = require('../../lib/railroad-diagrams.js');
-
-module.exports = Diagram(
-  NonTerminal('atom'),
-  ZeroOrMore(
-    Choice(
-      0,
-      NonTerminal('union'),
-      NonTerminal('branch'),
-      NonTerminal('split')
-    )
-  )
-)

### 2022-03-15 src/railroad/shortcut.js

@@ -1,15 +1,0 @@
-const { Diagram, Choice, Sequence, Terminal, Optional } = require('../../lib/railroad-diagrams.js');
-
-module.exports = Diagram(
-  Choice(
-    0,
-    Sequence(Terminal('B'), Optional(Terminal('r'))),
-    Sequence(Terminal('C'), Optional(Terminal('l'))),
-    Terminal('N'),
-    Terminal('O'),
-    Terminal('P'),
-    Terminal('S'),
-    Terminal('F'),
-    Terminal('I'),
-  )
-)

### 2022-03-15 src/railroad/split.js

@@ -1,6 +1,0 @@
-const { Diagram, Terminal, NonTerminal } = require('../../lib/railroad-diagrams.js');
-
-module.exports = Diagram(
-  Terminal('.'),
-  NonTerminal('string')
-)

### 2022-03-15 src/railroad/star.js

@@ -1,5 +1,0 @@
-const { Diagram, Terminal } = require('../../lib/railroad-diagrams');
-
-module.exports = Diagram(
-  Terminal('*')
-);

### 2022-03-15 src/railroad/stereodescriptor.js

@@ -1,6 +1,0 @@
-const { Diagram, NonTerminal, Optional } = require('../../lib/railroad-diagrams.js');
-
-module.exports = Diagram(
-    NonTerminal('@'),
-    Optional(NonTerminal('@'))
-)

### 2022-03-15 src/railroad/string.js

@@ -1,7 +1,0 @@
-const {Diagram, Optional, NonTerminal } = require('../../lib/railroad-diagrams');
-
-module.exports = Diagram(
-  Optional(
-    NonTerminal('sequence')
-  )
-);

### 2022-03-15 src/railroad/symbol.js

@@ -1,10 +1,0 @@
-const { Diagram, Choice, NonTerminal } = require('../../lib/railroad-diagrams');
-
-module.exports = Diagram(
-  Choice(
-    0,
-    NonTerminal('star'),
-    NonTerminal('element'),
-    NonTerminal('selection')
-  )
-);

### 2022-03-15 src/railroad/union.js

@@ -1,12 +1,0 @@
-const { Diagram, Optional, NonTerminal, Choice } = require('../../lib/railroad-diagrams.js');
-
-module.exports = Diagram(
-  Optional(
-    NonTerminal('bond')
-  ),
-  Choice(
-    0,
-    NonTerminal('cut'),
-    NonTerminal('sequence'),
-  )
-)

### 2022-03-15 src/railroad/virtual-hydrogen.js

@@ -1,8 +1,0 @@
-const { Diagram, Terminal, NonTerminal, Optional } = require('../../lib/railroad-diagrams.js');
-
-module.exports = Diagram(
-  Terminal('H'),
-  Optional(
-    NonTerminal('digit')
-  )
-)

### 2022-03-16 manuscript.tex
Removes section numbers. closes #92
Unifies use of hydrogens attribute. closes #80

@@
 
-\section{Introduction}
+\section*{Introduction}
 
@@
 
-\section{Goals, Design, and Tradeoffs}
+\section*{Goals, Design, and Tradeoffs}
 
@@
 
-\section{Data Model}
+\section*{Data Model}
 
@@
     \ttt{Option<Type>} & Optional Value & \ttt{None}, \ttt{Type} \\
-    \ttt{Index} & Unsigned integer & 0{\ldots}2\textsuperscript{32} \\
-    \ttt{Ten} & Integer count & 0{\ldots}10 \\
-    \ttt{SignedTen} & Integer count & -9{\ldots}10 \\
-    \ttt{Thousand} & Integer count & 0{\ldots}1000 \\
+    \ttt{Index} & Unsigned integer & 0..2\textsuperscript{32} \\
+    \ttt{Ten} & Integer count & 0..10 \\
+    \ttt{SignedTen} & Integer count & -9..10 \\
+    \ttt{Thousand} & Integer count & 0..1000 \\
     \ttt{Element} & IUPAC-\-approved element symbol & \ttt{Ac}, \ttt{Ag}, \ttt{Al}, {\ldots}\ttt{Zn}\\
@@
 
-\section{Constitution}
+\section*{Constitution}
 
@@
 
-The atomic \ttt{hydrogens} attribute records the number of associated \textit{virtual hydrogens}.
-A virtual hydrogen is an atom whose presence is recorded, not as a node and edge, but rather as a unit contribution to an integer tally.
+The \ttt{hydrogens} attribute encodes an atom's mode of \textit{hydrogen suppression}.
+Hydrogen suppression replaces hydrogen atoms in a molecular graph with a more compact representation.
+Despite the removal of its node, a suppressed hydrogen and its bonds to the parent atom are understood to be present.
+The abundance and regular bonding of hydrogen atoms in molecular graphs makes hydrogen suppression a popular strategy for increasing information density.
+
+Dialect supports two mutually-exclusive modes of hydrogen suppression.
+The first is \textit{virtual hydrogen}.
+A virtual hydrogen is one whose node and connecting bond have been replaced by an integer tally attribute on the parent atom.
 For example, methane can be represented by a molecular graph having five nodes and four edges.
-But methane can also be represented as a graph of one node having a \ttt{hydrogens} attribute set to four.
-The \ttt{hydrogens} attribute, if present, may assume integer values ranging from zero to nine.
-Only monovalent hydrogens with unspecified isotopic composition are eligible for virtualization.
-
-The two remaining constitutional attributes support \textit{electron counting}.
-The purpose of electron counting is to ensure that every valence electron, whether bonding or nonbonding, can be associated with either an atom or a bond.
-Accurate electron counting is crucial for any molecular representation system given their influence on charge, bond order, and mass.
+But methane can also be represented as a graph of one node whose \ttt{hydrogens} attribute equals four.
+When using this variant, the \ttt{hydrogens} attribute may assume integer values from zero to nine, inclusive.
+The process of transforming one or more hydrogen nodes into a virtual hydrogen is called "virtualization." The opposite process, transforming one or more virtual hydrogens into a node, is called "reification."
+
+The second mode of suppressed hydrogen is \textit{implicit hydrogen}.
+An implicit hydrogen is similar to a virtual hydrogen in that both the hydrogen node and bond have been replaced with a tally.
+Unlike a virtual hydrogen, however, this tally is not present as attribute on the parent atom but rather as a computation.
+The computation will be described in detail in a subsequent section.
+Subject to certain limitations, an an atom identifies itself as having an implicit hydrogen count by setting its \ttt{hydrogens} attribute to \ttt{Implicit}.
+
+The remaining constitutional atomic attribute, \ttt{charge}, enables electron counting.
+Electron counting associates each electron in a molecular graph with either an atom or a bond.
+This is a fundamental requirement for the principle of mass conversion.
+In a more practical sense, electron counting lays the foundation for calculations of formal charge and bond order.\cite{dietz:1995}
 
@@
 Bond order equals electron count divided by two.
-Formal charge equals the number of electrons gained or lost due to ionizations events outside of bond formation.
+Formal charge equals the number of electrons gained or lost due to ionization events outside of bond formation.
 Gaining an electron decrements the \ttt{charge} attribute, whereas losing an electron increments it.
@@
 
-From these rules follow some important consequences.
-For example, bond order must be an integer greater than zero.
+From these rules follow some practical consequences.
+For one, bond order must be an integer greater than zero.
 This follows from the definition of bond formation.
@@
 
-\section{Bond Elision}
+\section*{Bond Elision}
 
 To promote compact string representations, Dialect supports \textit{bond elision}.
-An single bond may be elided by marking it as such.
-An elided bond nevertheless has a bond order of one.
+A bond of order one may be setting its \ttt{elided} attribute to \ttt{true}.
+As such, an elided bond always has a bond order of one.
 A bond may be elided for one of two reasons.
@@
 
-\section{Delocalization Subgraph}
+\section*{Delocalization Subgraph}
 
@@
 
-\section{Implicit Hydrogens and Subvalence}
-
-In addition to virtual hydrogen count, Dialect supports a second form of hydrogen suppression called \textit{implicit hydrogens}.
-Like a virtual hydrogen, an implicit hydrogen is monovalent, has only default attributes, and is present as an integer tally associated with a particular atom.
-But unlike a virtual hydrogen, the presence of an implicit hydrogen can only be deduced through computation.
-Implicit hydrogens are an integral yet invisible component of many molecular graphs.
+\section*{Valence and Subvalence}
 
@@
 
-\section{Computing Implicit Hydrogen Count}
+\section*{Computing Implicit Hydrogen Count}
 
@@
 
-\section{Atom Index}
+\section*{Atom Index}
 
@@
 
-\section{Conformation}
+\section*{Conformation}
 
@@
 
-\section{Configuration}
+\section*{Configuration}
 
@@
 
-\section{Syntax}
+\section*{Syntax}
 
@@
 
-\subsection{Atom}
+\subsection*{Atom}
 
@@
 
-\subsection{Bond}
+\subsection*{Bond}
 
@@
 
-\subsection{Sequence}
+\subsection*{Sequence}
 
@@
 
-\section{Reading Strings}
+\section*{Reading Strings}
 
@@
 
-\section{Pruning}
+\section*{Pruning}
 
@@
 
-\section{Writing Strings}
+\section*{Writing Strings}
 
@@
 
-\section{SMILES Compatibility}
+\section*{SMILES Compatibility}
 
@@
 
-\section{Discussion}
+\section*{Discussion}
 
@@
 
-\section{Conclusion}
+\section*{Conclusion}
 

### 2022-03-16 manuscript.tex
Restores bond state table. closes #93

@@
     \ttt{Index} & Unsigned integer & 0..2\textsuperscript{32} \\
-    \ttt{Ten} & Integer count & 0..10 \\
+    \ttt{Three} & Integer count & 1..4 \\
+    \ttt{Ten} & Unsigned integer & 0..10 \\
     \ttt{SignedTen} & Integer count & -9..10 \\
@@
 
-Three restrictions apply to atomic state:
+Restrictions on atomic state are:
 
@@
 
-The following restrictions apply to bond state:
+Bond attributes are summarized in Table \ref{table:bond-state}. As with atomic attributes, not all bond states are valid. Three restrictions apply:
+
+\begin{table}
+\caption{Bond State.}
+\centering
+\begin{tabular}{l L L l}
+    \hline
+    Attribute & Description & Type & Default \\
+    \hline
+    \ttt{index} & Unique identifier & \ttt{Index} & - \\
+    \ttt{source} & Source atom index & \ttt{Index} & - \\
+    \ttt{target} & Target atom index & \ttt{Index} & - \\
+    \ttt{order} & Bond order & \ttt{Three} & 1 \\
+    \ttt{elided} & Elision status & \ttt{boolean} & \ttt{false} \\
+    \ttt{direction} & Partial conformational parity & \ttt{Option} \ttt{<Direction>} & \ttt{None} \\
+    \hline
+\end{tabular}
+\label{table:bond-state}
+\end{table}
 

### 2022-03-16 manuscript.tex
Explains Table 1 range notation. closes #94
Unifies atom/bond capitalization and style. closes #82

@@
 \begin{table}
+\begin{minipage}{\textwidth}
+\renewcommand*\footnoterule{}
 \caption{Data Model Primitives.}
+\label{table:primitives}
 \centering
@@
     \hline
-    Name & Description & Values \\
+    Name & Description & Values\footnote{Ranges defined as \ttt{a..b} include \ttt{a} and exclude \ttt{b}.} \\
     \hline
@@
 \end{tabular}
-\label{table:primitives}
+\end{minipage}
 \end{table}
@@
 The first atomic non-terminal, \ttt{<star>}, can assume one terminal value, the asterisk character (\ttt{*}).
-This "star atom" represents an \ttt{Atom} in which every attribute assumes its default value.
+This "star atom" represents an atom in which every attribute assumes its default value.
 The implicit hydrogen count of a star atom is zero.
@@
 The next atomic production rule, \ttt{<shortcut>} is a non-terminal selected from the list: "B"; "C"; "N"; "O"; "P"; "S"; "F"; "Cl"; "Br"; and "I".
-An \ttt{Atom} encoded in this way ("shortcut atom") assigns the corresponding symbol to the \ttt{elemen} attribute.
+An atom encoded in this way ("shortcut atom") assigns the corresponding symbol to the \ttt{element} attribute.
 All other attributes retain their default values.
@@
 The third atomic production rule, \ttt{<selection>} is a non-terminal selected from the list: "c"; "n"; "o"; "p"; and "s".
-An \ttt{Atom} encoded in this way ("selected shortcut atom") assigns the corresponding atom symbol to the \ttt{element} attribute and sets the \ttt{selected} attribute to `true`.
+An atom encoded in this way ("selected shortcut atom") assigns the corresponding atom symbol to the \ttt{element} attribute and sets the \ttt{selected} attribute to `true`.
 All other atomic attributes retain their default values.
@@
 A bracket atom can be used to set any atomic attribute.
-A bracket atom must be used when the `element` attribute of an \ttt{Atom} makes it ineligible for implicit hydrogen counting.
+A bracket atom must be used when the `element` attribute of an atom makes it ineligible for implicit hydrogen counting.
 Attributes not set within the \ttt{<bracket>} production rule will leave the corresponding atomic values in their default states.
@@
 
-An \ttt{Atom} may be connected to zero or more neighbors through a \ttt{Bond}, encoded with the non-terminal \ttt{<bond>}.
+An atom may be connected to zero or more neighbors through a bond, encoded with the non-terminal \ttt{<bond>}.
 
@@
 Five terminals are available ("-", "=", "\#", "/", and "\textbackslash").
-The first three ("-", "=", and "\#") set the \ttt{order} attribute of a \ttt{Bond} to 1; 2; or 3, respectively.
+The first three ("-", "=", and "\#") set the \ttt{order} attribute of a bond to 1; 2; or 3, respectively.
 The last two terminals, "/" and "\textbackslash", set the \ttt{order} attribute to 1 while also setting the \ttt{direction} attribute to \ttt{Up} and \ttt{Down}, respectively.
@@
 
-An \ttt{Atom} may be assigned zero or more children through the \ttt{<sequence>} non-terminal ("sequence").
+An atom may be assigned zero or more children through the \ttt{<sequence>} non-terminal ("sequence").
 A sequence starts with a required \ttt{<atom>} non-terminal.
-If no allowed non-terminals follow, the corresponding \ttt{Atom} will have no children.
+If no allowed non-terminals follow, the corresponding atom will have no children.
 Allowed non-terminals are chosen from the list: \ttt{<union>}; \ttt{<branch>}; and \ttt{<sequence>}.
@@
 
-A cut signifies the disconnection of an \ttt{Atom} from its neighbor.
+A cut signifies the disconnection of an atom from its neighbor.
 The most common use for cut is to encode a cycle.
-The cut index serves as a placeholder for the neighboring \ttt{Atom}, which is located by scanning either forward or backward along a string.
+The cut index serves as a placeholder for the neighboring atom, which is located by scanning either forward or backward along a string.
 If the number of previous appearances of the cut index is even, then the string is scanned right.
@@
 A split is a sequence of \ttt{<detachment>} and \ttt{<sequence>} non-terminals.
-The `<detachment>` non-terminal serves the same purpose as it does within a branch: the prevent the bonding of the parent \ttt{Atom} with a child.
+The `<detachment>` non-terminal serves the same purpose as it does within a branch: the prevent the bonding of the parent atom with a child.
 
@@
 For example, a reader can merely validate a string by returning a boolean type.
-A more sophisticated reader can return a molecular graph capturing all \ttt{Atom} and \ttt{Bond} attributes and connectivity relationships.
+A more sophisticated reader can return a molecular graph capturing all atom and bond attributes and connectivity relationships.
 
@@
 
-Readers that capture \ttt{Atom}-\ttt{Atom} connectivity will typically maintain a reference to a \textit{root atom}.
+Readers that capture atom-atom connectivity will typically maintain a reference to a \textit{root atom}.
 The root atom is initially undefined.
-On reading the first complete \ttt{<atom>} non-terminal, the corresponding \ttt{Atom} is constructed and set as the root atom.
-Subsequently processing a complete \ttt{<union>}, \ttt{<branch>}, or \ttt{<split>} non-terminal triggers three changes: (1) a child \ttt{Atom} is constructed; (2) the child is connected to the current root atom, unless a \ttt{<detachment>} non-terminal intervenes; and (3) the root atom is replaced with the child atom.
+On reading the first complete \ttt{<atom>} non-terminal, the corresponding atom is constructed and set as the root atom.
+Subsequently processing a complete \ttt{<union>}, \ttt{<branch>}, or \ttt{<split>} non-terminal triggers three changes: (1) a child atom is constructed; (2) the child is connected to the current root atom, unless a \ttt{<detachment>} non-terminal intervenes; and (3) the root atom is replaced with the child atom.
 
@@
 Bonds with undefined \ttt{direction} are always complementary.
-A bond with a defined \ttt{direction} is complementary to any \ttt{Bond} with an undefined state, or a \ttt{Bond} with an opposing \ttt{direction}.
-For example, an elided \ttt{Bond} is compatible with a single \ttt{Bond}.
-Likewise, two elided \ttt{Bond}s are compatible.
-A \ttt{Bond} with a \ttt{direction} of \ttt{Up} is incompatible with another \ttt{Bond} with a \ttt{direction} of \ttt{Up}, but compatible with one having a \ttt{direction} of \ttt{Down}.
+A bond with a defined \ttt{direction} is complementary to any bond with an undefined state, or a bond with an opposing \ttt{direction}.
+For example, an elided bond is compatible with a single bond.
+Likewise, two elided bonds are compatible.
+A bond with a \ttt{direction} of \ttt{Up} is incompatible with another bond with a \ttt{direction} of \ttt{Up}, but compatible with one having a \ttt{direction} of \ttt{Down}.
 
@@
 
-Decoding atomic configuration from a string requires the order of attachment from each \ttt{Atom} to its neighbors to be recorded.
+Decoding atomic configuration from a string requires the order of attachment from each atom to its neighbors to be recorded.
 Cuts are treated as if the neighboring atom itself were detected.
 In this regard, the first member of a cut pair requires special attention.
-The neighboring \ttt{Atom} won't be available until the paired cut index is found.
+The neighboring atom won't be available until the paired cut index is found.
 Nevertheless, it must ultimately appear before the neighbors in subsequent unions, branches, and splits.
@@
     \item{Impossible valence. The valence at an `Atom` is impossibly high. For example, "C(C)(C)(C)(C)C".}
-    \item{Impossible charge. An \ttt{Atom}'s charge gives it an apparent negative electron count. For example, "[C+7]".}
+    \item{Impossible charge. An atom's charge gives it an apparent negative electron count. For example, "[C+7]".}
 \end{enumerate}

### 2022-03-16 manuscript.tex
Replaces "eligible" uses meaning "selectable" with "selectable." refs #69

@@
 
+An atom is added to a DS by setting its \ttt{selected} attribute to \ttt{true} ("selection").
+For an atom to be selectable, its \ttt{element} attribute must be contained in the list: \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; and \ttt{S}.
+Atoms having other elements (or \ttt{None}) must not be selected.
+Only selected atoms belong to the DS.
+
+\begin{figure}
+    \centering
+    \includegraphics{filler}
+    \caption{Perfect Matching. A subgraph with all of the nodes of its parent, but in which all nodes have degree one.}
+    \label{fig:perfect-matching}
+\end{figure}
+
+A bond will be added to the DS only if both of the following conditions are met: (1) both terminals are selected; and (2) the bond itself is elided.
+No other bond will be added to the DS.
+This behavior makes it possible to exclude a specific bond from the DS by making its bond order explicit.
+
 A non-empty DS must possess a \textit{perfect matching}.
@@
 Every atom added to a DS must therefore be part of a perfect matching over it.
-
-\begin{figure}
-    \centering
-    \includegraphics{filler}
-    \caption{Perfect Matching. A subgraph with all of the nodes of its parent, but in which all nodes have degree one.}
-    \label{fig:perfect-matching}
-\end{figure}
-
-Only some atoms are eligible for inclusion (or "selection") in a DS.
-Atoms whose \ttt{element} values are one of \ttt{C}, \ttt{N}, \ttt{O}, \ttt{P}, or \ttt{S}, or \ttt{None}, may be selected --- regardless of cycle membership.
-Atoms having other \ttt{element} values must not be selected.
-
-To support the construction of a DS, eligible atoms carry a \ttt{selected} boolean attribute.
-Setting this attribute to \ttt{true}, adds the atom to the DS.
-Otherwise, the atom is excluded from the DS.
-All ineligible atoms are excluded from the DS.
-
-A bond will be added to the DS only if both of the following conditions are met: (1) both terminals are selected; and (2) the bond itself is elided.
-No other bond will be added to the DS.
-This behavior makes it possible to exclude a specific bond from the DS by making its bond order explicit.
 
@@
 The opposite operation can be accomplished with a \textit{selection algorithm}.
-A selection algorithm selects two or more eligible atoms, thereby adding them to the DS.
+A selection algorithm selects two or more atoms, thereby adding them to the DS.
 The only requirement for a selection algorithm is that the resulting DS must have a perfect matching.
@@
     \item{Incompatible cut bonds (position, position). The bonds to a pair of cuts are incompatible.}
-    \item{Delocalization subgraph lacks perfect matching. Before reporting this error, steps to remove ineligible atoms should be taken as described in the next section.}
+    \item{Delocalization subgraph lacks perfect matching. Before reporting this error, steps to remove unnecessarily selected atoms should be taken as described in the next section.}
     \item{Partial parity bond not allowed (position).

### 2022-03-16 manuscript.tex
Removes most uses of "eligible." closes #69

@@
     \item{If \ttt{hydrogens} equals \ttt{Implicit}, then \ttt{isotope}, \ttt{configuration}, and \ttt{charge} must equal their default values.
-Furthermore, \ttt{element} must equal one of \ttt{B}; \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; \ttt{S}; \ttt{F}; \ttt{Cl}; \ttt{Br}; \ttt{I}.}
+Furthermore, \ttt{element} must equal one of \ttt{B}; \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; \ttt{S}; \ttt{F}; \ttt{Cl}; \ttt{Br}; \ttt{I}; or \ttt{None}.}
     \item{If \ttt{selected} equals \ttt{true}, then \ttt{element} must equal one of: \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; or \ttt{S}.}
@@
 Single bonds contribute one to the tally, double bonds two, and triple bonds three.
-Every other bond contributes one, as does each virtual hydrogen.
-For example, the valence of a methyl carbon having a \ttt{hydrogens} attribute of three is four.
-In contrast, the valence of a methyl carbon with a \ttt{hydrogens} attribute of \ttt{Implicit} is one.
-
-Only some atoms are eligible for implicit hydrogen counting.
-These are called \textit{eligible atoms}.
-An atom becomes eligible by fulfilling two requirements: (1) its \ttt{hydrogens} attribute equals \ttt{Implicit}; and (2) its \ttt{element} attribute is associated with at least one \textit{default valence}.
-
-A default valence is a value associated with an element that represents the number of hydrogens that can be attached to an isolated atom.
-Table \ref{table:target-valences} lists those elements possessing at least one default valence.
-All other elements have no default valences.
-The target valence of 4 for carbon, for example, means that a fully-saturated carbon atom will be bound to four hydrogen atoms.
-Likewise, a fully-saturated oxygen atom will be bound to two hydrogen atoms.
-However, iron has no default valences.
-Some elements such as nitrogen have multiple target valences.
+All other bonds contribute one, as does each virtual hydrogen.
+For example, the valence of a methyl carbon having three virtual hydrogens and a hydrogen neighbor is four.
+The valence of an acetaldehyde oxygen atom is two.
+And so on.
+
+Some elements are associated with one or more \textit{default valences} (Table~\ref{table:default-valences}).
+A default valence is the number of hydrogens that can be attached to an isolated, fully-saturated atom using the corresponding element.
+For example, the default valence for a carbon-bearing atom is 4.
+This means that a fully saturated atom of carbon will have four hydrogens.
+Likewise, a fully-saturated atom of oxygen will have two hydrogens.
+Some elements such as nitrogen are associated with multiple default valences.
 In these cases, multiple saturated forms are possible.
-For example, nitrogen has the target valences three and five.
-Both ammonia and nitrogen pentahydride (NH\textsubscript{5}) are fully saturated forms of nitrogen according to Table \ref{table:target-valences}.
+For example, nitrogen has the default valences three and five.
+Both ammonia (NH\textsubscript{3}) and nitrogen pentahydride (NH\textsubscript{5}) are therefore fully saturated forms of nitrogen.
 
@@
         \hline
-        Element & Defaults \\
+        Element & Valences \\
         \hline
@@
     \end{tabular}
-    \label{table:target-valences}
+    \label{table:default-valences}
 \end{table}
 
-Given one or more default valences, \textit{subvalence} can be computed.
-Subvalence is the number of hydrogens that can be added to an eligible atom without exceeding the lowest possible default valence.
-If no default valence less than a default valence exists, then the subvalence is zero.
-The subvalence for an eligible atom (\ttt{a}) can be computed with Algorithm \ref{algorithm:implicit-hydrogen}.
+Given an atom whose element has one or more default valences \textit{subvalence} can be computed.
+Subvalence is the number of hydrogens that can be added to an atom without exceeding the lowest possible default valence.
+If no suitable default valence exists, then the subvalence is zero.
+This computation can be expressed algorithmically (Algorithm~\ref{algorithm:subvalence}).
+First, the atom's valence is computed.
+Next, the ordered list of default valences of the atom's element is found.
+For each default valence, the difference between it and the valence is compute.
+If this differences is positive, it is returned as the subvalence.
+If no suitable default valence is found, zero is returned.
 
@@
     \SetKwFunction{TargetValences}{TargetValences}
-    \caption{Compute subvalence}
-    \label{algorithm:implicit-hydrogen}
-
-    \Input{An eligible atom $a$}
+    \caption{Computing subvalence.}
+    \label{algorithm:subvalence}
+
+    \Input{An atom $a$ with one or more default valences}
     \Output{The subvalence of $a$}
@@
       $v \leftarrow$ Valence($a$)\;
-      $T \leftarrow$ TargetValences($a$)\;
+      $T \leftarrow$ DefaultValences($a$)\;
       \For{$t \in T$}{
@@
 
-The algorithm accepts an eligible atom \ttt{a} as input and returns its subvalence.
-First, the valence (sum of bond orders) for \ttt{a} is computed.
-Next, the ordered list of target valences (\ttt{T}) is obtained from Table {table:target-valences}.
-For each target valence (\ttt{t}), the difference (\ttt{d}) between \ttt{t} and \ttt{v} is computed.
-If this difference is non-negative, \ttt{d} is returned as the implicit hydrogen count.
-If no suitable target valence can be found, zero is returned.
-
-Consider an isolated atom whose values of \ttt{element} and \ttt{hydrgogens} are \ttt{N} and \ttt{Implicit}, respectively.
+The subvalence of an atom whose element has no default valences (i.e.
+no listing in Table~\ref{table:default-valences}) is zero.
+For example, an iron atom always has a subvalence of zero, regardless of how it is bonded.
+
+For a representative subvalence computation, consider an isolated nitrogen atom without virtual hydrogens.
 The atom's bond order sum is zero.
 Its default valences are 3 and 5.
-The difference \ttt{d} is found to be three (3 - 0).
-Therefore, subvalence of this atom is three.
+The difference is found to be three ($3 - 0$).
+Therefore, the subvalence of this atom is three.
 
 The carbonyl carbon atom of acetaldehyde illustrates the effects of substitution and multiple bonding.
-This atom uses implicit hydrogen counting (\ttt{hydrogens} equals \ttt{Implicit}).
-Valence is therefore three (2 + 1).
-The first and only default valence for carbon is four.
-Subtracting three from four yields one, which is returned as the atom's subvalence.
-
-The phosphorous atom in phosphorous acid (H\textsubscript{3}PO\textsubscript{3}) illustrates the use of Algorithm \ref{algorithm:implicit-hydrogen} for atoms with multiple target valences.
-Given a \ttt{hydrogens} attribute of \ttt{Implicit}, the atom's bond order sum is four (2 + 1 + 1).
-The first target valence is 3, but subtracting that value from four yields a negative number (-1).
-Continuing to the next default valence, 5, a difference of 1 is obtained.
-Therefore, the subvalence of the phosphorous-bearing atom is reported as 1.
+Bond order sum is three (2 + 1).
+From Table~\ref{table:default-valences}, default valence is four.
+Subtracting bond order sum from default valence yields one ($4 - 3$), which is returned as the atom's subvalence.
+
+The phosphorous atom in phosphorous acid (H\textsubscript{3}PO\textsubscript{3}) illustrates the use of Algorithm \ref{algorithm:subvalence} for atoms with multiple target valences.
+Bond order sum is four ($2 + 1 + 1$).
+The first default valence is 3, but subtracting valence from that value yields a negative number ($3 - 4$).
+Continuing to the next default valence, five, a difference of 1 is obtained.
+Therefore, the subvalence of the phosphorous-bearing atom is reported as one.
 
@@
 Consider sodium perchlorate (NaClO\textsubscript{4}).
-The chlorine atom has a bond order sum of seven (2 + 2 + 2 + 1).
-From Table 1, the only target valence for chlorine is one.
+The chlorine atom has a bond order sum of seven ($2 + 2 + 2 + 1$).
+From Table~\ref{table:default-valences}, the only default valence for chlorine is one.
 Subtracting seven yields a negative number (-6).
@@
 
-The subvalence for ineligible atoms is zero.
-
 \section*{Computing Implicit Hydrogen Count}
 
-For unselected atoms, implicit hydrogen count equals subvalence.
-Consider an oxygen atom with a \ttt{hydrogens} attribute of \ttt{Implicit}.
-Subvalence equals two and so does implicit hydrogen count.
-However, an oxygen atom with a \ttt{hydrogens} attribute equal to one will have an implicit hydrogen count of zero.
+To qualify for implicit hydrogen counting, an atom must have a \ttt{hydrogens} attribute set to \ttt{Implicit}.
+As noted previously, this setting requires: (1) \ttt{isotope}, \ttt{charge}, and \ttt{configuration} must be set to their default values; and (2) the value of \ttt{element} must equal one of: \ttt{B}; \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; \ttt{S}; \ttt{F}; \ttt{Cl}; \ttt{Br}; \ttt{I}; or \ttt{None}.
+
+For unselected atoms, implicit hydrogen count equals subvalence, with one exception.
+An oxygen atom with one single bond has an implicit hydrogen count of one because subvalence is one ($2 - 1$).
+However, an oxygen atom with two singly-bonded neighbors has an implicit hydrogen count of zero ($2 - 2$).
+For atoms whose \ttt{element} attribute is \ttt{None}, implicit hydrogen count equals zero.
 
@@
 This subtraction models the unpaired electron that would would be required for delocalized bonding, or localized bonding if the atom were deselected.
-Consider a selected carbon atom in benzene whose \ttt{hydrogens} attribute equals \ttt{Implicit}.
-Subvalence equals two, so implicit hydrogen count equals one (2 - 1).
-
-A selected atom with a \ttt{hydrogens} attribute of \ttt{Implicit} and subvalence equal to zero has an implicit hydrogen count of zero.
-The semantics of such a state may seem suspect.
+Consider a selected carbon atom in benzene with a \ttt{hydrogens} attribute of \ttt{Implicit}.
+Subvalence equals two, so implicit hydrogen count equals one ($2 - 1$).
+
+If the subvalence of a selected atom equals zero, then an implicit hydrogen count of zero is reported.
+The meaning of such an atomic state may seem suspect.
 If an atom has no unpaired electrons, how can it participate in delocalized bonding?
-As will be explained in more detail later (Pruning), this situation can arise through gratuitous atom selection.
+As will be explained in detail later (Pruning), this situation can arise for reasons of convenience or tradition.
 Returning zero avoids miscalculation of the implicit hydrogen count.
@@
 We might expect the implicit hydrogen count to equal the experimentally-determined hydrogen count (2).
-However, the subvalence for the phosphorous atom is found to be three (2 + 1 - 3).
-The implicit hydrogen count is therefore zero (3 - 3) rather than the expected two.
-Similar atoms must be encoded using virtual hydrogens.
+However, the subvalence for the phosphorous atom is found to be three ($2 + 1 - 3$).
+The implicit hydrogen count is therefore zero ($3 - 3$) rather than the expected two.
+The hydrogen count of this atom, and others like it, must be encoded as a virtual hydrogen count.
 
@@
 Single bonds may be elided or not.
-Similarly, selection can be used or not for eligible atoms.
+Similarly, the DS may or may not be empty.
 Although an organization may seek to standardize certain styles of string output, any syntactically-valid string must be considered valid by a reader.
@@
 Dialect uses the latter default valences.
-Eligible atoms ("the organic subset") are reported in the article as: B; C; N; O; P; S; F; Cl; Br; and I (p.
+Atoms whose \ttt{hydrogens} attribute may be set to \ttt{Implicit} ("the organic subset") are reported in the article as: B; C; N; O; P; S; F; Cl; Br; and I (p.
 32).

### 2022-03-18 manuscript.tex
Imports auto-generated bibliograph from Zotero. Fixes IUPAC element references. closes #97

@@
 \emergencystretch=1em
+\DeclareUnicodeCharacter{2212}{-}
 
@@
 SMILES has since been widely adopted.
-Read and write functionalities are routinely supported by popular cheminformatics toolkits, including: Open Babel;\cite{oBoyle:2011} RDKit;\cite{rdkit} Chemistry Development Kit;\cite{steinbeck2003} JChem Base;\cite{jchemBase} the Daylight Toolkit;\cite{daylightToolkit} and OEChem TK.\cite{oeChemTk} SMILES encodings can be found in many public-facing databases, including: PubChem;\cite{kim2016} ChEBI;\cite{degtyarenko2007} ZINC;\cite{irwin2005} ChEMBL;\cite{gaulton2012} and Wikipedia.\cite{wikipedia} Raw SMILES strings have been used extensively in machine learning applications.\cite{sousa2021} SMILES has also been extended to carry various forms of metadata, as exemplified by Jmol SMILES,\cite{hanson2016} CurleySMILES,\cite{drefahl2011} BigSMILES,\cite{lin2019} and CXSMILES.\cite{cxsmiles}
+Read and write functionalities are routinely supported by popular cheminformatics toolkits, including: Open Babel;\cite{oBoyle:2011} RDKit;\cite{rdkit} Chemistry Development Kit;\cite{steinbeck:2003} JChem Base;\cite{jchemBase} the Daylight Toolkit;\cite{daylightToolkit} and OEChem TK.\cite{oetk} SMILES encodings can be found in many public-facing databases, including: PubChem;\cite{kim:2016} ChEBI;\cite{degtyarenko:2007} ZINC;\cite{irwin:2005} ChEMBL;\cite{gaulton:2012} and Wikipedia.\cite{wikipediaCAS} Raw SMILES strings have been used extensively in machine learning applications.\cite{sousa:2021} SMILES has also been extended to carry various forms of metadata, as exemplified by Jmol SMILES,\cite{hanson:2016} CurleySMILES,\cite{drefahl:2011} BigSMILES,\cite{lin:2019} and CXSMILES.\cite{cxsmiles}
 
@@
 Earlier descriptions of SMILES were published in various media, although these are not cited by Weininger.
-An accessible example is a 1987 report to the US Environmental Protection Agency.\cite{anderson1987} Although the language described in that report shares many features with later descriptions, there are important differences as well.
+An accessible example is a 1987 report to the US Environmental Protection Agency.\cite{anderson:1987} Although the language described in that report shares many features with later descriptions, there are important differences as well.
 
@@
 
-In 2019 IUPAC announced the SMILES+ initiative.\cite{smilesPlus} Noting the limitations of existing SMILES documentation, the SMILES+ effort seeks to "establish a formalized recommended up-to-date specification of the SMILES format." SMILES+ took as its starting point the documentation produced by the OpenSMILES project.
-Efforts to extend this starting point are in progress online through a public repository, but no formal recommendation has to date resulted.\cite{smilesPlusRepo}
+In 2019 IUPAC announced the SMILES+ initiative.\cite{smilesPlusSpecification} Noting the limitations of existing SMILES documentation, the SMILES+ effort seeks to "establish a formalized recommended up-to-date specification of the SMILES format." SMILES+ took as its starting point the documentation produced by the OpenSMILES project.
+Efforts to extend this starting point are in progress online through a public repository, but no formal recommendation has to date resulted.\cite{smilesPlusGitHub}
 
@@
 The lack of a detailed, comprehensive, and universally-applied SMILES specification also hampers broader data integrity efforts.
-The FAIR Guiding Principles\cite{wilkinson2016} identify four qualities essential for extracting maximum utility from published scholarly data: Findability; Accessibility; Interoperability; and Reusability.
+The FAIR Guiding Principles\cite{wilkinson:2016} identify four qualities essential for extracting maximum utility from published scholarly data: Findability; Accessibility; Interoperability; and Reusability.
 Interoperability in particular requires that "(meta)data" use a formal, accessible, shared, and broadly applicable language for knowledge representation." Although SMILES may be a shared, accessible, and broadly applicable language, its formal underpinnings leave room for improvement.
@@
 Others claim SMILES compatibility while ignoring certain aspects of written documentation.
-The Daylight implementation, like some of the others, is distributed under a non-Open Source license that restricts use.\cite{daylightToolkit} For several years Daylight operated a publicly-accessible web service based on its toolkit that depicted SMILES strings, but it has since been decommissioned.\cite{depict} Although SMILES may underpin modern chemical information exchange, the foundations of this system are weathered and showing signs of stress.
+The Daylight implementation, like some of the others, is distributed under a non-Open Source license that restricts use.\cite{daylightToolkit} For several years Daylight operated a publicly-accessible web service based on its toolkit that depicted SMILES strings, but it has since been decommissioned.\cite{depictArchive} Although SMILES may underpin modern chemical information exchange, the foundations of this system are weathered and showing signs of stress.
 
@@
 
-Dialect's system of representation is based on the \textit{molecular graph} concept.\cite{balaban1985} A graph is a data structure comprised of a set of nodes (which map to "atoms") and a set of pairwise relationships between them called edges (which map to "bonds").
+Dialect's system of representation is based on the \textit{molecular graph} concept.\cite{balaban:1985} A graph is a data structure comprised of a set of nodes (which map to "atoms") and a set of pairwise relationships between them called edges (which map to "bonds").
 A molecular graph is a specialized graph onto which atomic and bonding metadata have been overlaid.
-Molecular graphs are ubiquitous in cheminformatics, appearing in contexts such as data formats like Chemical Markup Language\cite{murrayrust2011}, CDXML\cite{cdxml}, and Molfile\cite{ctfileFormats}, as well as in-memory data structures found in cheminformatics toolkits.
+Molecular graphs are ubiquitous in cheminformatics, appearing in contexts such as data formats like Chemical Markup Language\cite{murray-rust:2011}, CDXML\cite{cdxml}, and Molfile\cite{ctfileFormats}, as well as in-memory data structures found in cheminformatics toolkits.
 Like these systems, Dialect augments its underlying graph with node and edge labels.
@@
 Two attributes characterize an atom's nucleus: \ttt{element} and \ttt{isotope}.
-The `element` attribute is an optional one- or two-letter character sequence selected from the set designated by the IUPAC/IUPAP Working Party.\cite{iupac2016} When the Working Party authorizes additional symbols, they will become valid values for the \ttt{element} attribute.
+The `element` attribute is an optional one- or two-letter character sequence selected from the set designated by the IUPAC.\cite{periodicTable} When IUPAC authorizes additional symbols, they will become valid values for the \ttt{element} attribute.
 If the elemental identity of an atom is unknown, its \ttt{element} attribute can be omitted.
@@
 
-Electron counting in Dialect is based on the well-known \textit{valence bond model} (VB Model).\cite{lewis1916} Although not commonly considered as such, the VB model can be thought of as an algorithm for molecular assembly.
+Electron counting in Dialect is based on the well-known \textit{valence bond model} (VB Model).\cite{lewis:1916} Although not commonly considered as such, the VB model can be thought of as an algorithm for molecular assembly.
 The perspective is especially useful given that Dialect itself can be viewed as a programming language for molecular assembly.
@@
 DIME occurs in a molecular graph when one or more equivalent representations exist, each one differing from the original only in the distribution of single and double bonds.
-DIME may be recognized as "resonance" or "aromaticity," but those terms are avoided here due to their ambiguity.\cite{randic2018,kerber2006}
+DIME may be recognized as "resonance" or "aromaticity," but those terms are avoided here due to their ambiguity.\cite{randic:2018,kerber:2006}
 
@@
 Because the presence of a filled DS implies a perfect matching over it, kekulization always succeeds.
-A popular algorithm for matching, the Edmunds "blossom algorithm,"\cite{edmunds65} has a time complexity of $O(|E||V|^2)$, where $|E|$ is the number of edges and $|V|$ is the number of nodes.
+A popular algorithm for matching, the Edmunds "blossom algorithm,"\cite{edmonds:1965} has a time complexity of $O(|E||V|^2)$, where $|E|$ is the number of edges and $|V|$ is the number of nodes.
 Although more efficient algorithms are known, they are either much more difficult to implement or lack generality.
@@
 Not all tetracoordinate atoms will be \textit{stereocenters}.
-A stereocenter is an atom whose "ligand permutation produces stereoisomers," as defined by Mislow and Siegel.\cite{mislow1984} In the context of Dialect, a stereocenter must be tetracoordinate.
+A stereocenter is an atom whose "ligand permutation produces stereoisomers," as defined by Mislow and Siegel.\cite{mislow:1984} In the context of Dialect, a stereocenter must be tetracoordinate.
 Readers should avoid the assignment of stereodescriptors to atoms that are not stereocenters.
@@
 This set of features makes LL(1) grammars such as the one used by Dialect a good fit for manually-written recursive descent parsers.
-LL(1) grammars can also be used as a basis for auto-generated parsers through packages such as ANTLR.\cite{parr2014} The full grammar for Dialect strings is available as a text file in this paper's Supporting Material.
+LL(1) grammars can also be used as a basis for auto-generated parsers through packages such as ANTLR.\cite{parr:2014} The full grammar for Dialect strings is available as a text file in this paper's Supporting Material.
 
@@
 34).
-Weininger later refined and expanded SMILES in a 2003 book chapter,\cite{weininger2008} referred to here as "the book chapter." A third major source of information is the Daylight Theory Manual (referred to here as "the manual"), a website operated by Daylight Chemical Information Systems, Inc.\cite{daylightTheory} Information about SMILES is also contained in various software packages and associated documentation, slide decks, websites, mailing lists, and private communications.
+Weininger later refined and expanded SMILES in a 2003 book chapter,\cite{weininger:2008} referred to here as "the book chapter." A third major source of information is the Daylight Theory Manual (referred to here as "the manual"), a website operated by Daylight Chemical Information Systems, Inc.\cite{daylightTheory} Information about SMILES is also contained in various software packages and associated documentation, slide decks, websites, mailing lists, and private communications.
 
@@
 83).
-The symbol for element 105 is given as Ha (Hahnium), despite IUPAC's resolution of this naming controversy in favor of the symbol Db (Dubnium) years prior.\cite{ohrstrom1997} Accordingly, Dialect disallows the symbol Ha.
+The symbol for element 105 is given as Ha (Hahnium), despite IUPAC's resolution of this naming controversy in favor of the symbol Db (Dubnium) years prior.\cite{sageson:1997} Accordingly, Dialect disallows the symbol Ha.
 To support the full set of currently-named transuranium elements, Dialect adds the symbols: Db; Sg; Bh; Hs; Mt; Rg; Cn; Nh; Fl; Mc; Lv; Ts; and Og.
@@
 88).
-The state of not being bonded and the state of having a zero-order bond are not necessarily equivalent.\cite{clark2011} To be clear, Dialect explicitly disallows the zero bond order interpretation of detachments and the dot character.
+The state of not being bonded and the state of having a zero-order bond are not necessarily equivalent.\cite{clark:2011} To be clear, Dialect explicitly disallows the zero bond order interpretation of detachments and the dot character.
 

### 2022-03-18 manuscript.tex
Cites paper reporting reluctance to publish. closes #96
Cites den Joed. closes #95

@@
 ("Daylight"), recapitulates much of the material in the Paper and the Book Chapter.\cite{daylightTheory}.
+The lack of other primary literature sources is consistent with Weininger's reported disinterest in publication.\cite{gasteiger:2018}
 
 In 2007 a documentation effort that would become known as OpenSMILES began.\cite{openSMILES} OpenSMILES was conceived as "a non-proprietary specification for the SMILES language," and it addressed many of the points left open by previous SMILES documentation efforts.
-Noteworthy contributions included: the first formal grammar; many refinements around stereochemistry; and introduction of the idea of "standard form." Absent were detailed procedures for assigning and interpreting aromatic features, and a detailed procedure for computing implicit hydrogen count.
+Noteworthy contributions included: the first formal grammar, later later adapted;\cite{denjoed:2010} many refinements around stereochemistry; and introduction of the idea of "standard form." Absent were detailed procedures for assigning and interpreting aromatic features, and a detailed procedure for computing implicit hydrogen count.
 OpenSMILES also left several points of syntactic and semantic ambiguity unaddressed.

### 2022-03-18 manuscript.tex
Clarifies introductory discussion of SMILES. refs #85

--- 2022-03-18 manuscript.tex
@@
 Despite its many applications, SMILES remains a language with an incomplete specification.
-Weininger's 1988 paper (referred to herein as "the Paper") leaves the following either fully or incompletely specified: (1) stereochemical configuration; (2) double bond conformation; (3) the meaning handling of "aromatic" features; (4) the computation of implicit hydrogen count, especially as related to aromatic features; (5) formal syntax; (6) constraints on quantities such as mass number and charge; and (7) enumeration of error states.
+Weininger's 1988 paper leaves the following either fully or incompletely specified: (1) stereochemical configuration; (2) double bond conformation; (3) the meaning handling of "aromatic" features; (4) the computation of implicit hydrogen count, especially as related to aromatic features; (5) formal syntax; (6) constraints on quantities such as mass number and charge; and (7) enumeration of error states.
 
 Earlier descriptions of SMILES were published in various media, although these are not cited by Weininger.
-An accessible example is a 1987 report to the US Environmental Protection Agency.\cite{anderson:1987} Although the language described in that report shares many features with later descriptions, there are important differences as well.
-
-Since 1988, a few sources have attempted to clarify or revise Weininger's 1988 description of SMILES.
-A 2003 book chapter by Weininger (referred to herein as "the Book Chapter") addressed some of the limitations of the paper.
-The Daylight Theory Manual ("the Manual"), a website maintained by SMILES's corporate sponsor, Daylight Chemical Information Systems, Inc.
-("Daylight"), recapitulates much of the material in the Paper and the Book Chapter.\cite{daylightTheory}.
-
-In 2007 a documentation effort that would become known as OpenSMILES began.\cite{openSMILES} OpenSMILES was conceived as "a non-proprietary specification for the SMILES language," and it addressed many of the points left open by previous SMILES documentation efforts.
-Noteworthy contributions included: the first formal grammar; many refinements around stereochemistry; and introduction of the idea of "standard form." Absent were detailed procedures for assigning and interpreting aromatic features, and a detailed procedure for computing implicit hydrogen count.
-OpenSMILES also left several points of syntactic and semantic ambiguity unaddressed.
+An currently-accessible example is a 1987 report to the US Environmental Protection Agency.\cite{anderson:1987} Although the language described there shares many features with later descriptions, there are important differences as well.
+
+Since 1988, two authoritative sources have attempted to clarify or revise Weininger's 1988 description of SMILES.
+A 2003 book chapter by Weininger addressed some of the limitations of the paper.\cite{weininger:2008} The Daylight Theory Manual, a website maintained by SMILES's corporate sponsor, Daylight Chemical Information Systems, Inc.
+("Daylight"), recapitulates previously-published material.\cite{daylightTheory}.
+The lack of other written sources describing SMILES is consistent with Weininger's reported disinterest in traditional scientific publication.\cite{gasteiger:2018}
+
+In 2007 a documentation effort that would become known as OpenSMILES began.\cite{openSMILES} OpenSMILES was conceived as "a non-proprietary specification for the SMILES language." Many of the points left open by previous SMILES documentation efforts were addressed.
+Noteworthy contributions included: the first formal grammar, which was later adapted;\cite{denjoed:2010} many refinements around stereochemistry; and the introduction of "standard form." Absent were detailed procedures for assigning and interpreting aromatic features, and a detailed procedure for computing implicit hydrogen count.
+OpenSMILES also left several points of syntactic and semantic ambiguity unsettled, while introducing extensions of its own.
 

### 2022-03-20 manuscript.tex
Avoids normative SMILES. closes #85

@@
 
-\hyphenation{organo-metallics}
+\hyphenation{organo-metallics Proto-SMILES}
 \newcolumntype{L}{>{\arraybackslash}m{3cm}}
@@
 
-\section*{SMILES Compatibility}
-
-To maximize compatibility with contemporary software, Dialect is based on the SMILES language.
-Given the widespread adoption of "SMILES," arriving at a specific definition of the language is surprisingly challenging.
-No single document completely specifies SMILES syntax and semantics.
-The first description of the SMILES language in the primary literature is contained within a 1988 article (referenced here as "the article").\cite{weininger:1988} The article, by its own admission, is incomplete: "...
-isomeric SMILES [defined in the book chapter as covering isotopism, conformation, and configuration (p.
-91)] is not otherwise covered in this paper." (p.
-34).
-Weininger later refined and expanded SMILES in a 2003 book chapter,\cite{weininger:2008} referred to here as "the book chapter." A third major source of information is the Daylight Theory Manual (referred to here as "the manual"), a website operated by Daylight Chemical Information Systems, Inc.\cite{daylightTheory} Information about SMILES is also contained in various software packages and associated documentation, slide decks, websites, mailing lists, and private communications.
-
-For the purpose of providing a foundation for Dialect, "SMILES" was defined as the language described in the article and the book chapter.
-All other sources --- irrespective of quality or apparent claim to authority --- were considered out of scope.
-The manual was disregarded because it merely recapitulated material in the article and chapter, and due to the possibility of future alteration.
-The remaining sources were disregarded due to lack of authority, poor accessibility, or both.
-This narrow focus simplifies the task of evaluating compatibility between Dialect and SMILES.
-However, it does not provide a method for testing compatibility due to the lack of an authoritative SMILES compliance suite.
-
-It is nevertheless possible to compare Dialect and SMILES on a feature-by-feature basis.
-Dialect is intended to be a subset of SMILES, meaning that every feature present in Dialect is ideally also present in SMILES.
-As will be seen, certain factors make this undesirable in the strictest sense.
-With a few notable exceptions, however, Dialect can be viewed as a subset of SMILES.
-
-The book chapter defines a fixed set of element symbols which even at the time of publication was obsolete (p.
-83).
-The symbol for element 105 is given as Ha (Hahnium), despite IUPAC's resolution of this naming controversy in favor of the symbol Db (Dubnium) years prior.\cite{sageson:1997} Accordingly, Dialect disallows the symbol Ha.
+\section*{Compatibility}
+
+For maximum compatibility with existing software, Dialect was designed as a \textit{language subset} of SMILES.
+A language subset contains all of the syntax and semantics of its parent language, but adds none of its own.
+In principle this means that a feature present in Dialect should also be present in SMILES.
+Conversely, a feature present in SMILES may or may not also be present in Dialect.
+
+Language subsets have ample precedent in computer science.
+A well-known example is MISRA C,\cite{misraC} a subset of the C programming language that aims to eliminate "known undefined or otherwise dangerous behavior."\cite{hatton:2007} Another example is the subset of JavaScript described by Crockford whose purpose is to "chip away at the features that are not beautiful until the language's true nature reveals itself."\cite{crockford:2008} Language subsets arise for two main reasons.
+First, designing a language presents many opportunities for errors small and large.
+Second, popular languages inevitably bring with them many users with diverse problems that demand new features.
+Some of these features cause future problems of their own.
+A language subset can improve both situations by eliminating the most problematic features.
+
+Although similar considerations apply to SMILES, its case is different in one important way: SMILES was never publicly described with high precision.
+Overviews of the language have been published by both Weininger and Daylight.
+But the kind of technical documentation suitable for software development was never made public.
+Both the OpenSMILES and SMILES+ initiatives cite this shortfall as motivation.
+
+Nevertheless, several SMILES software implementations have been released as noted previously.
+Each one represents its own unique set of decisions about how to reduce the published broad SMILES descriptions to a specification detailed enough for software development.
+Some implementations have themselves been reverse-engineered for clues.
+The Daylight implementation in particular is sometimes invoked as a de facto SMILES specification.
+Unfortunately, access to this implementation is restricted by commercial licensing, as is the case with several SMILES implementations.
+Moreover, reverse engineering is unlikely to address every question because of the vast search space embodied by SMILES.
+Although some implementations are released under Open Source licenses, they are neither authoritative nor unanimous in their interpretations.
+
+A publicly-available and precise subset of SMILES would offer a new way to address the problem.
+If sufficiently constrained, the subset should be interpretable by all existing software without modification.
+Moreover, most strings written by existing software would be interpretable by software based on the language subset.
+Because its description would be both public and precise, the language subset could serve as the basis for many implementations with strong consensus around both syntax and semantics.
+
+Unfortunately, the choice of the Dialect base language is fraught with difficulty on technical, historical, and cultural grounds.
+Simply put, "SMILES" has come to mean different things to different people.
+To some, SMILES is the language defined in public documentation.
+To others, SMILES is the Daylight implementation source code.
+Some may view SMILES as the collective definitions contained in both published documents and all software.
+To others, SMILES isn't even definable.
+
+These problems can be avoided by considering as a base language not SMILES itself, but a \textit{protolanguage}.
+A protolanguage is a hypothetical language from which a family of languages can be derived.
+A suitable SMILES protolanguage would ideally contain all of the features consistent with the many ways in which the term "SMILES" is used today.
+
+Accordingly, Dialect's base language is "ProtoSMILES," a SMILES protolanguage.
+ProtoSMILES was developed through a process of elimination that started with every SMILES document ever published and every SMILES software implementation, Open Source or not.
+The field was narrowed through two limitations: (1) the source must be authoritative; and (2) the source must be publicly accessible.
+In this context, authoritative means that the source was created by either Weininger himself or Daylight.
+They alone can speak most authoritatively about what SMILES is.
+The requirement for public accessibility ensures that, even if inconvenient, claims of compatibility can be tested.
+The following documents were identified as candidate ProtoSMILES sources:
+
+\begin{itemize}
+    \item Weininger's report to the US Environmental Protection Agency (herein "The Article").\cite{anderson:1987}
+    \item Weininger's original publication (herein "The Article").\cite{weininger:1988}
+    \item Weininger's 2003 book chapter, revised in 2008 (herein "The Book Chapter").\cite{weininger:2008}
+    \item An online manual maintained by Daylight (herein "The Manual").
+    \item The Daylight's toolkit, which implements a SMILES reader and writer (herein "The Toolkit").\cite{daylightToolkit}
+\end{itemize}
+
+The Toolkit was eliminated for three reasons.
+First, its source code is not available, forcing an indirect reverse-engineering approach to language definition.
+Second, as a product of the company Weininger started, the Toolkit is likely to recapitulate the other sources.
+Third, as a commercial product without source code, the Toolkit may not always be available in the future.
+
+The Report was eliminated mainly because the contents of this older document contradict the three remaining sources.
+For example, The Report contains no mention of bracket syntax, using markedly different notation instead.
+The report also supports unusual characters including the exclamation mark (\ttt{!}).
+Semantically, The Report supports "special states" for nitrogen-containing atoms.
+
+The Manual was disregarded as a ProtoSMILES source because it merely recapitulates content already present in The Paper and The Book Chapter.
+Removing this source simplifies the construction of ProtoSMILES without changing its definition.
+A secondary consideration was the possibility that updates to this web page could change the meaning of ProtoSMILES.
+However web archives indicate that neither styling nor content of the Manual have changed in at least ten years.
+
+The only remaining sources are The Paper and the Book Chapter.
+These are the two works that collectively define ProtoSMILES.
+The remainder of this section compares and contrasts ProtoSMILES with Dialect.
+The main difficulty with this approach is that the works defining ProtoSMILES not only contradict each other in some ways, but they also contradict themselves internally.
+These contradictions will be noted when meaningful.
+
+The Book Chapter defines a fixed set of element symbols which even at the time of publication was obsolete.\cite[p.
+83]{weininger:2008} The symbol for element 105 is given as Ha (Hahnium), despite IUPAC's resolution of this naming controversy in favor of the symbol Db (Dubnium) years prior.\cite{sageson:1997} Accordingly, Dialect disallows the symbol Ha.
 To support the full set of currently-named transuranium elements, Dialect adds the symbols: Db; Sg; Bh; Hs; Mt; Rg; Cn; Nh; Fl; Mc; Lv; Ts; and Og.
 Moreover, Dialect explicitly accepts all future element symbols approved by IUPAC.
-These additions are one of the reasons Dialect can't be considered a strict subset of SMILES.
-
-The article implies that comma (\ttt{,}) is either an element symbol or a valid SMILES character: "This subset uses only the symbols H, C, N, O, P, S, F, Cl, Br, I, and (,) and digits, with the following four rules..." (p.
-33).
-The comma character is explicitly disallowed in Dialect.
-
-The book chapter defines a recursive grammar for branching which, perhaps without the knowledge of the author, results in multiple branching (p.
-86).
-Dialect disallows the multiple branching constructs, including for example "*((*))*" and "*(((*)))*".
-
-"Reaction SMILES" is an extension described in the book chapter (p.
-89), but disallowed in Dialect.
+The addition of these element symbols means that Dialect can't be considered a strict subset of SMILES.
+The alternative would, however, leave dialect without the ability to express every kind of IUPAC-recognized element.
+
+The Article implies that comma (\ttt{,}) is either an element symbol or a valid SMILES character: "This subset uses only the symbols H, C, N, O, P, S, F, Cl, Br, I, and (,) and digits, with the following four rules...".\cite[p.
+33]{weininger:1988} The comma character is explicitly disallowed in Dialect.
+
+The Book Chapter defines a recursive grammar for branching which, perhaps without the knowledge of the author, defines sequential parentheses.\cite[p.
+86]{weininger:2008} Dialect disallows such constructs, including for example "*((*))*" and "*(((*)))*".
+
+"Reaction SMILES" is an extension described in the Book Chapter,\cite[p.
+89]{weininger:2008} but disallowed in Dialect.
 Specifically, the greater than symbol (\ttt{>}) is not a valid Dialect character.
@@
 Dialect only supports two stereodescriptors, encoded with the terminals "@" and "@@".
-The book chapter provides a recursive grammar for stereodescriptors that allows multi-symbol descriptors such as "@@@" and "@@@@" (p.
-94).
-The book chapter also discusses non-tetrahedral descriptors including "@AL1", "@AL2", "@1", and "@SP1".
+The Book Chapter provides a recursive grammar for stereodescriptors that allows multi-symbol descriptors such as "@@@" and "@@@@".\cite[p.
+94]{weininger:2008} The book chapter also discusses non-tetrahedral descriptors including "@AL1", "@AL2", "@1", and "@SP1".
 None of these are supported by Dialect.
@@
 Although some of these forms of configuration are supported today, it's unusual to find them all supported.
-The reason is simple: in the context in which SMILES is most often used, these forms of stereochemistry are rare.
-To support them places a burden on implementors with little payoff.
+The reason is simple: in the context in which Dialect will most often used, these forms of stereochemistry are rare.
+To support them burdens implementors with little payoff.
+
+ProtoSMILES supports the concept of "aromatic bond" (":"), but Dialect does not.
+Despite clear inclusion within the syntax of ProtoSMILES, no precise guidance on using this feature is provided.
+The Book Chapter notes "Adjacent atoms without an intervening bond symbol are connected by a valence-dictated bond (typically a single or aromatic bond).
+"-" (single) and ":" (aromatic) bond symbols may always be omitted on input".\cite[p 85]{weininger:2008} This passage implies that aromatic bonds serve no purpose.
+Accordingly, Dialect does not allow them and the syntax does not include the colon symbol (\ttt{:}).
 
 Along these lines, Dialect does not recognize atoms bearing either arsenic (As) or selenium (Se) as selectable.
-Doing so would increase the complexity of Dialect unnecessarily.
-Although the paper implies selectability ("Elements other than C, N, O, P, S, As, and Se are not yet dealt with in an aromatic context." p35), default valences for the latter two atoms are not provided.
+Doing so would increase the complexity of the language unnecessarily.
+Although The Paper implies selectability ("Elements other than C, N, O, P, S, As, and Se are not yet dealt with in an aromatic context."),\cite[p.
+35]{weininger:1988} default valences for the latter two atoms are not provided.
 Even if they were, the elements As and Se must be bracketed.
 As such, special rules for handling them in the context of pruning would be required.
-Moreover, molecules bearing the elements As or Se rarely exhibit DIME.
-Consider the arsenic analog of pyrrole ("arsole").
-Selecting the arsenic-bearing atom is gratuitous because it does not contribute to DIME.
-Finally, allowing the selection of atoms bearing As or Se does nothing to address molecules with atoms bearing germanium (Ge), tin (Sn), or other heavy main group elements that may be discovered in the future to participate in DIME.
-
-It's not clear whether SMILES allows a virtual hydrogen count on hydrogen itself.
-The article implies that the element symbol H (hydrogen) must not have an implicit hydrogen count when used within a bracket sequence: "There are few exceptions to the hydrogen-suppression convention, the most obvious being specification of a proton, [H+], and molecular hydrogen, [H][H]." (p.
-34) If so, the author may have erroneously believed that parsers would be incapable of processing constructs such as `[HH]`.
-The book chapter appears to reiterate this view, noting that "explicit hydrogen specification is required" in the case of "hydrogens connected to other hydrogens, e.g.
-[H][H], molecular hydrogen" (p.
-97).
-To be clear, Dialect supports any valid virtual hydrogen count on hydrogen itself.
+Moreover, molecules bearing the elements As or Se are often made "aromatic" unnecessarily in SMILES.
+Finally, allowing the selection of atoms bearing As or Se does nothing to address molecules with atoms bearing silicon, germanium, tin, or other heavy main group elements that may be discovered in the future to participate in DIME.
+
+It's unclear whether ProtoSMILES allows a virtual hydrogen count on hydrogen itself.
+The Article implies that the element symbol H (hydrogen) must not have an implicit hydrogen count when used within a bracket sequence ("There are few exceptions to the hydrogen-suppression convention, the most obvious being specification of a proton, [H+], and molecular hydrogen, [H][H].").\cite[p.
+34]{weininger:1988} If so, the author may have erroneously believed that parsers would be incapable of processing constructs such as "[HH]".
+The Book Chapter appears to reiterate this view, noting that "explicit hydrogen specification is required" in the case of "hydrogens connected to other hydrogens, e.g.
+[H][H], molecular hydrogen".\cite[p 97]{weininger:2008} To be clear, Dialect supports any valid virtual hydrogen count on hydrogen itself.
 For example, the following are all valid, if chemically nonsensical, strings: "[HH]"; "[HH2]"; and "[HH3]".
-Depending on interpretation, this support may or may not render Dialect a superset of SMILES.
-
-The book chapter ascribes ambiguous meaning to detachments: "...
-In terms of the valence model being represented, the dot literally represents a bond of formal order zero: the atoms on either side of the dot are explicitly not bonded to each other." (p.
-88).
-The state of not being bonded and the state of having a zero-order bond are not necessarily equivalent.\cite{clark:2011} To be clear, Dialect explicitly disallows the zero bond order interpretation of detachments and the dot character.
-
-SMILES assigns no explicit upper or lower bounds to numerical atomic attributes.
-These boundaries are nevertheless crucial for implementors, who often seek a data type just large enough to prevent underflow or overflow.
-For this reason, Dialect sets both upper and lower bounds on the following atomic attributes: isotope (0 $\leq$ value \textless 1000); charge (-10, \textless value \textless 10); and virtual hydrogen (\ttt{hydrogens}, 0 $\leq$ value \textless 10).
-Although it might be argued that lower bounds on physical quantities such as isotope and hydrogen count should be obvious in SMILES, the lack of specificity forces implementors to complete work that should have been done by the language's creator.
-
-Dialect provides a few semantic clarifications not fully addressed in SMILES.
+Depending on interpretation, this feature may or may not render Dialect a superset of SMILES.
+
+ProtoSMILES ascribes ambiguous meaning to detachments.
+As noted in the Book Chapter, "...
+In terms of the valence model being represented, the dot literally represents a bond of formal order zero: the atoms on either side of the dot are explicitly not bonded to each other.".\cite[p 88]{weininger:2008} The state of not being bonded and the state of having a zero-order bond are not necessarily equivalent.\cite{clark:2011} To be clear, Dialect explicitly disallows the zero bond order interpretation of detachments and the dot character.
+
+ProtoSMILES assigns no explicit upper or lower bounds to numerical atomic attributes.
+These boundaries are nevertheless crucial for implementors, who seek the data types necessary to prevent underflow and overflow.
+For this reason, Dialect sets both upper and lower bounds on the following atomic attributes: isotope (0 $\leq$ value \textless 1000); charge (-10, \textless value \textless 10); and hydrogens, 0 $\leq$ value \textless 10).
+Although it might be argued that lower bounds on physical quantities such as isotope and hydrogen count should be implicit, the lack of precision forces implementors to improvise, which can lead to divergent behavior.
+ProtoSMILES places no upper bound on the charge notation "+", "++", "+++", and their corresponding negative forms ("The SMILES program also recognizes constructions of the form [Fe+++] as being synonymous with the form [Fe+3].").\cite[p 32]{weininger:1988} However, Dialect disallows duplicated plus and minus characters (\ttt{+} and \ttt{-}, respectively) from bracket atoms.
+
+Dialect provides a few semantic clarifications not fully addressed in ProtoSMILES.
 A detailed algorithm for determining implicit hydrogen count is provided, together with the required valence table.
-Unlike SMILES, Dialect explicitly considers the case of computing implicit hydrogens on selected atoms.
+Unlike ProtoSMILES, Dialect explicitly considers the case of computing implicit hydrogens on selected atoms.
 Dialect also provides detailed algorithms, absent in SMILES, for selection and deselection.
-These are based in graph theory rather than the ambiguous and overloaded chemical concept of "aromaticity" used in SMILES.
-Uniquely, Dialect introduces the concepts of "pruning" and "gratuitous selection." Neither the paper nor the book chapter indicate those bonds that can be promoted during deselection, but Dialect does.
-Finally, Dialect syntax is based on a formal grammar and tooling rooted in decades of computer science research, whereas SMILES syntax is based for the most part on imprecise natural language descriptions.
-
-Dialect also resolves some internal contradictions within SMILES.
-The article reports 3 as the only default valence for nitrogen (p.
-34), whereas the book chapter reports two default valences: 3 and 5 (p.
-84).
-Dialect uses the latter default valences.
-Atoms whose \ttt{hydrogens} attribute may be set to \ttt{Implicit} ("the organic subset") are reported in the article as: B; C; N; O; P; S; F; Cl; Br; and I (p.
-32).
-However, an example later in the article indicates that "Hn1cccc1" with its unbracketed hydrogen is a valid string (p.
-35).
-Dialect explicitly rejects strings containing unbracketed hydrogen atoms.
-On the topic of atom selection ("aromaticity"), the article implies that selected atoms must be part of a cycle: "To qualify as aromatic, all atoms in the ring must be sp2 hybridized and the number of available 'excess' $\pi$ electrons must satisfy Hückel's 4{\textit{N}} + 2 criterion." (p.
-34).
-The book chapter, however, presents two examples of valid strings that do not encode cycles ("cc" and "cccc", p.
-85).
-Consistent with the latter view, Dialect explicitly allows the selection of atoms regardless of cycle membership.
+These are based in graph theory rather than the ambiguous and overloaded chemical concept of "aromaticity" used in ProtoSMILES.
+Uniquely, Dialect introduces the concepts of "pruning" and "gratuitous selection." ProtoSMILES does not specify those bonds that can be promoted during deselection, but Dialect does.
+ProtoSMILES does not restrict partial parity bonds to those atoms also possessing at least one double bond, whereas Dialect does.
+Finally, Dialect syntax is based on a formal grammar and tooling rooted in decades of computer science research, whereas ProtoSMILES syntax is based for the most part on imprecise natural language descriptions.
+
+Dialect also resolves some internal contradictions within ProtoSMILES.
+The Article reports the only default valence for nitrogen as three (p.
+34),\cite[p 34]{weininger:1988} whereas the Book Chapter reports two default valences: 3 and 5.\cite[p 84]{weininger:2008} Dialect uses the latter default valences.
+Atoms whose \ttt{hydrogens} attribute may be set to \ttt{Implicit} ("the organic subset") are reported in the Article as: B; C; N; O; P; S; F; Cl; Br; and I.\cite[p 32]{weininger:2008} However, an example later in the Article indicates that "Hn1cccc1" with its unbracketed hydrogen is a valid string.\cite[p 35]{weininger:1988} Dialect explicitly rejects strings containing unbracketed hydrogen atoms.
+On the topic of atom selection ("aromaticity"), the Article implies that selected atoms must be part of a cycle: "To qualify as aromatic, all atoms in the ring must be sp2 hybridized and the number of available 'excess' $\pi$ electrons must satisfy Hückel's 4{\textit{N}} + 2 criterion."\cite[p 34]{weininger:1988} The Book Chapter, however, presents two examples of valid strings that do not encode cycles ("cc" and "cccc").\cite[p 85]{weininger:2008} Consistent with the latter view, Dialect explicitly allows the selection of atoms regardless of cycle membership.
 
 \begin{table}
-    \caption{Some differences between SMILES and Dialect.}
-    \centering
-    \begin{tabular}{p{0.55\linewidth} p{0.15\linewidth} p{0.15\linewidth}}
+\begin{minipage}{\textwidth}
+\renewcommand*\footnoterule{}
+    \caption{Some differences between ProtoSMILES and Dialect.}
+    \centering
+    \begin{tabular}{p{0.50\linewidth} p{0.20\linewidth} p{0.15\linewidth}}
         \hline
-        Feature & SMILES & Dialect \\
+        Feature & ProtoSMILES\footnote{Values of "maybe" indicate ambiguous source descriptions.} & Dialect \\
         \hline
@@
         atomic "map" attribute & yes & no \\
-        extended stereodescriptors & yes & no \\
+        extended stereodescriptors (e.g., "@@@", "@OH1", etc.) & yes & no \\
         use of stereodescriptors on odd cumulene centers & yes & no \\
-        virtual hydrogen count on hydrogen & no & yes \\
+        virtual hydrogen count on hydrogen (e.g., "[HH]") & maybe & yes \\
         detachments are bonds of "formal order zero" & maybe & no \\
-        upper and lower bounds on atomic properties & no & yes \\
+        upper and lower bounds on numerical atomic properties & no & yes \\
         nitrogen default valence includes 5 & maybe & yes \\
@@
         non-elided bonds may be promoted during selection & maybe & no \\
+        isolated partial parity bond & yes & no \\
+        repeated charge notation (e.g., "++", "+++", etc.) & yes & no \\
+        aromatic bond (":") & yes & no \\
         \hline
@@
     \label{table:feature-comparison}
+\end{minipage}
 \end{table}
+
+With only one clear-cut exception, Table~\ref{table:feature-comparison} shows Dialect to be a subset of ProtoSMILES.
+The exceptional case deals with the validity certain transfermium element symbols already recognized by IUPAC and to be recognized in the future.
+Ambiguous exceptions include those for which the ProtoSMILES column feature support is uncertain ("maybe") but the Dialect column is affirmative ("yes").
+If ProtoSMILES support tends toward "no," then support by Dialect would break out of the ProtoSMILES subset.
+There are three such cases.
+The first is dihydrogen represented using the string "[HH]".
+The second ambiguous case is whether or not the default valence of nitrogen includes the value five.
+The third case is whether or not ProtoSMILES recognizes acyclic "aromatic" atoms as valid.
 

### 2022-03-24 manuscript.tex
Clarifies definition of DIME. closes #65

@@
 A molecular representation based solely on the valence bond model can yield artifacts resulting from \textit{delocalization induced molecular equality} (DIME).
-DIME occurs in a molecular graph when one or more equivalent representations exist, each one differing from the original only in the distribution of single and double bonds.
-DIME may be recognized as "resonance" or "aromaticity," but those terms are avoided here due to their ambiguity.\cite{randic:2018,kerber:2006}
+DIME occurs when a molecular graph exists in two or more equal forms, each differing from the others only through electron delocalization.
+Two molecular graphs are equal if they are isomorphic and a 1:1 mapping exists between the atoms and bonds in each graph.
+Electron delocalization results when the electrons of a valence bond representation can be distributed over more than one atom pair, between two or more atoms, or across some combination of atoms and bonds.
+DIME is similar to "resonance" or "aromaticity," but those terms are avoided here due to their ambiguity.\cite{randic:2018,kerber:2006}
 

### 2022-03-24 manuscript.tex
Removes redundant statement about configuration error. closes #78

@@
 A molecular representation based solely on the valence bond model can yield artifacts resulting from \textit{delocalization induced molecular equality} (DIME).
-DIME occurs when a molecular graph exists in two or more equal forms, each differing from the others only through electron delocalization.
+DIME occurs when a molecular graph exists in two or more unequal forms, each differing from the others only through electron delocalization.
 Two molecular graphs are equal if they are isomorphic and a 1:1 mapping exists between the atoms and bonds in each graph.
-Electron delocalization results when the electrons of a valence bond representation can be distributed over more than one atom pair, between two or more atoms, or across some combination of atoms and bonds.
+Electron delocalization results when the electrons of a valence bond representation can be distributed over more than one pair of atoms, as is the case with benzene and related molecules.
 DIME is similar to "resonance" or "aromaticity," but those terms are avoided here due to their ambiguity.\cite{randic:2018,kerber:2006}
@@
 \end{figure}
-
-It is an error to assign a stereodescriptor to any atom that is not tetracoordinate, or which has more than one virtual hydrogen.
 

### 2022-03-24 manuscript.tex
Unifies page citation. closes #98

@@
 \begin{itemize}
-    \item Weininger's report to the US Environmental Protection Agency (herein "The Article").\cite{anderson:1987}
+    \item Weininger's report to the US Environmental Protection Agency ("The Article").\cite{anderson:1987}
     \item Weininger's original publication (herein "The Article").\cite{weininger:1988}
-    \item Weininger's 2003 book chapter, revised in 2008 (herein "The Book Chapter").\cite{weininger:2008}
-    \item An online manual maintained by Daylight (herein "The Manual").
-    \item The Daylight's toolkit, which implements a SMILES reader and writer (herein "The Toolkit").\cite{daylightToolkit}
+    \item Weininger's 2003 book chapter, revised in 2008 ("The Book Chapter").\cite{weininger:2008}
+    \item An online manual maintained by Daylight ("The Manual").
+    \item The Daylight's toolkit, which implements a SMILES reader and writer ("The Toolkit").\cite{daylightToolkit}
 \end{itemize}
@@
 A secondary consideration was the possibility that updates to this web page could change the meaning of ProtoSMILES.
-However web archives indicate that neither styling nor content of the Manual have changed in at least ten years.
+However, web archives indicate that neither styling nor content of the Manual have changed in at least ten years.
 
@@
 
-The Book Chapter defines a fixed set of element symbols which even at the time of publication was obsolete.\cite[p.
-83]{weininger:2008} The symbol for element 105 is given as Ha (Hahnium), despite IUPAC's resolution of this naming controversy in favor of the symbol Db (Dubnium) years prior.\cite{sageson:1997} Accordingly, Dialect disallows the symbol Ha.
+The Book Chapter defines a fixed set of element symbols which even at the time of publication was obsolete.\cite[p.~83]{weininger:2008} The symbol for element 105 is given as Ha (Hahnium), despite IUPAC's resolution of this naming controversy in favor of the symbol Db (Dubnium) years prior.\cite{sageson:1997} Accordingly, Dialect disallows the symbol Ha.
 To support the full set of currently-named transuranium elements, Dialect adds the symbols: Db; Sg; Bh; Hs; Mt; Rg; Cn; Nh; Fl; Mc; Lv; Ts; and Og.
@@
 
-The Article implies that comma (\ttt{,}) is either an element symbol or a valid SMILES character: "This subset uses only the symbols H, C, N, O, P, S, F, Cl, Br, I, and (,) and digits, with the following four rules...".\cite[p.
-33]{weininger:1988} The comma character is explicitly disallowed in Dialect.
-
-The Book Chapter defines a recursive grammar for branching which, perhaps without the knowledge of the author, defines sequential parentheses.\cite[p.
-86]{weininger:2008} Dialect disallows such constructs, including for example "*((*))*" and "*(((*)))*".
-
-"Reaction SMILES" is an extension described in the Book Chapter,\cite[p.
-89]{weininger:2008} but disallowed in Dialect.
+The Article implies that comma (\ttt{,}) is either an element symbol or a valid SMILES character: "This subset uses only the symbols H, C, N, O, P, S, F, Cl, Br, I, and (,) and digits, with the following four rules...".\cite[p.~33]{weininger:1988} The comma character is explicitly disallowed in Dialect.
+
+The Book Chapter defines a recursive grammar for branching which, perhaps without the knowledge of the author, defines sequential parentheses.\cite[p.~86]{weininger:2008} Dialect disallows such constructs, including for example "*((*))*" and "*(((*)))*".
+
+"Reaction SMILES" is an extension described in the Book Chapter,\cite[p.~89]{weininger:2008} but disallowed in Dialect.
 Specifically, the greater than symbol (\ttt{>}) is not a valid Dialect character.
@@
 Dialect only supports two stereodescriptors, encoded with the terminals "@" and "@@".
-The Book Chapter provides a recursive grammar for stereodescriptors that allows multi-symbol descriptors such as "@@@" and "@@@@".\cite[p.
-94]{weininger:2008} The book chapter also discusses non-tetrahedral descriptors including "@AL1", "@AL2", "@1", and "@SP1".
+The Book Chapter provides a recursive grammar for stereodescriptors that allows multi-symbol descriptors such as "@@@" and "@@@@".\cite[p.~94]{weininger:2008} The book chapter also discusses non-tetrahedral descriptors including "@AL1", "@AL2", "@1", and "@SP1".
 None of these are supported by Dialect.
@@
 The Book Chapter notes "Adjacent atoms without an intervening bond symbol are connected by a valence-dictated bond (typically a single or aromatic bond).
-"-" (single) and ":" (aromatic) bond symbols may always be omitted on input".\cite[p 85]{weininger:2008} This passage implies that aromatic bonds serve no purpose.
+"-" (single) and ":" (aromatic) bond symbols may always be omitted on input".\cite[p.~85]{weininger:2008} This passage implies that aromatic bonds serve no purpose.
 Accordingly, Dialect does not allow them and the syntax does not include the colon symbol (\ttt{:}).
@@
 Doing so would increase the complexity of the language unnecessarily.
-Although The Paper implies selectability ("Elements other than C, N, O, P, S, As, and Se are not yet dealt with in an aromatic context."),\cite[p.
-35]{weininger:1988} default valences for the latter two atoms are not provided.
+Although The Paper implies selectability ("Elements other than C, N, O, P, S, As, and Se are not yet dealt with in an aromatic context."),\cite[p.~35]{weininger:1988} default valences for the latter two atoms are not provided.
 Even if they were, the elements As and Se must be bracketed.
@@
 It's unclear whether ProtoSMILES allows a virtual hydrogen count on hydrogen itself.
-The Article implies that the element symbol H (hydrogen) must not have an implicit hydrogen count when used within a bracket sequence ("There are few exceptions to the hydrogen-suppression convention, the most obvious being specification of a proton, [H+], and molecular hydrogen, [H][H].").\cite[p.
-34]{weininger:1988} If so, the author may have erroneously believed that parsers would be incapable of processing constructs such as "[HH]".
+The Article implies that the element symbol H (hydrogen) must not have an implicit hydrogen count when used within a bracket sequence ("There are few exceptions to the hydrogen-suppression convention, the most obvious being specification of a proton, [H+], and molecular hydrogen, [H][H].").\cite[p.~34]{weininger:1988} If so, the author may have erroneously believed that parsers would be incapable of processing constructs such as "[HH]".
 The Book Chapter appears to reiterate this view, noting that "explicit hydrogen specification is required" in the case of "hydrogens connected to other hydrogens, e.g.
-[H][H], molecular hydrogen".\cite[p 97]{weininger:2008} To be clear, Dialect supports any valid virtual hydrogen count on hydrogen itself.
+[H][H], molecular hydrogen".\cite[p.~97]{weininger:2008} To be clear, Dialect supports any valid virtual hydrogen count on hydrogen itself.
 For example, the following are all valid, if chemically nonsensical, strings: "[HH]"; "[HH2]"; and "[HH3]".
@@
 As noted in the Book Chapter, "...
-In terms of the valence model being represented, the dot literally represents a bond of formal order zero: the atoms on either side of the dot are explicitly not bonded to each other.".\cite[p 88]{weininger:2008} The state of not being bonded and the state of having a zero-order bond are not necessarily equivalent.\cite{clark:2011} To be clear, Dialect explicitly disallows the zero bond order interpretation of detachments and the dot character.
+In terms of the valence model being represented, the dot literally represents a bond of formal order zero: the atoms on either side of the dot are explicitly not bonded to each other.".\cite[p.~88]{weininger:2008} The state of not being bonded and the state of having a zero-order bond are not necessarily equivalent.\cite{clark:2011} To be clear, Dialect explicitly disallows the zero bond order interpretation of detachments and the dot character.
 
@@
 Although it might be argued that lower bounds on physical quantities such as isotope and hydrogen count should be implicit, the lack of precision forces implementors to improvise, which can lead to divergent behavior.
-ProtoSMILES places no upper bound on the charge notation "+", "++", "+++", and their corresponding negative forms ("The SMILES program also recognizes constructions of the form [Fe+++] as being synonymous with the form [Fe+3].").\cite[p 32]{weininger:1988} However, Dialect disallows duplicated plus and minus characters (\ttt{+} and \ttt{-}, respectively) from bracket atoms.
+ProtoSMILES places no upper bound on the charge notation "+", "++", "+++", and their corresponding negative forms ("The SMILES program also recognizes constructions of the form [Fe+++] as being synonymous with the form [Fe+3].").\cite[p.~32]{weininger:1988} However, Dialect disallows duplicated plus and minus characters (\ttt{+} and \ttt{-}, respectively) from bracket atoms.
 
@@
 The Article reports the only default valence for nitrogen as three (p.
-34),\cite[p 34]{weininger:1988} whereas the Book Chapter reports two default valences: 3 and 5.\cite[p 84]{weininger:2008} Dialect uses the latter default valences.
-Atoms whose \ttt{hydrogens} attribute may be set to \ttt{Implicit} ("the organic subset") are reported in the Article as: B; C; N; O; P; S; F; Cl; Br; and I.\cite[p 32]{weininger:2008} However, an example later in the Article indicates that "Hn1cccc1" with its unbracketed hydrogen is a valid string.\cite[p 35]{weininger:1988} Dialect explicitly rejects strings containing unbracketed hydrogen atoms.
-On the topic of atom selection ("aromaticity"), the Article implies that selected atoms must be part of a cycle: "To qualify as aromatic, all atoms in the ring must be sp2 hybridized and the number of available 'excess' $\pi$ electrons must satisfy Hückel's 4{\textit{N}} + 2 criterion."\cite[p 34]{weininger:1988} The Book Chapter, however, presents two examples of valid strings that do not encode cycles ("cc" and "cccc").\cite[p 85]{weininger:2008} Consistent with the latter view, Dialect explicitly allows the selection of atoms regardless of cycle membership.
+34),\cite[p.~34]{weininger:1988} whereas the Book Chapter reports two default valences: 3 and 5.\cite[p.~84]{weininger:2008} Dialect uses the latter default valences.
+Atoms whose \ttt{hydrogens} attribute may be set to \ttt{Implicit} ("the organic subset") are reported in the Article as: B; C; N; O; P; S; F; Cl; Br; and I.\cite[p.~32]{weininger:2008} However, an example later in the Article indicates that "Hn1cccc1" with its unbracketed hydrogen is a valid string.\cite[p.~35]{weininger:1988} Dialect explicitly rejects strings containing unbracketed hydrogen atoms.
+On the topic of atom selection ("aromaticity"), the Article implies that selected atoms must be part of a cycle: "To qualify as aromatic, all atoms in the ring must be sp2 hybridized and the number of available 'excess' $\pi$ electrons must satisfy Hückel's 4{\textit{N}} + 2 criterion."\cite[p.~34]{weininger:1988} The Book Chapter, however, presents two examples of valid strings that do not encode cycles ("cc" and "cccc").\cite[p 85]{weininger:2008} Consistent with the latter view, Dialect explicitly allows the selection of atoms regardless of cycle membership.
 

### 2022-03-24 manuscript.tex
Standardizes placement of references. closes #99

@@
 
-Simplified Molecular Input Line Entry System (SMILES) was first described by Weininger in 1988.\cite{weininger:1988} As a line notation,\cite{wiswesser:1968} SMILES represents molecules as single line character sequences, or strings.
-SMILES strings can be read and written algorithmically, and can be canonicalized\cite{weininger:1989,oBoyle:2012} to yield unique molecular identifiers.
+Simplified Molecular Input Line Entry System (SMILES) was first described by Weininger in 1988~\cite{weininger:1988}.
+As a line notation~\cite{wiswesser:1968}, SMILES represents molecules as single line character sequences, or strings.
+SMILES strings can be read and written algorithmically~\cite{weininger:1989,oBoyle:2012}, and can be canonicalized to yield unique molecular identifiers.
 
 SMILES has since been widely adopted.
-Read and write functionalities are routinely supported by popular cheminformatics toolkits, including: Open Babel;\cite{oBoyle:2011} RDKit;\cite{rdkit} Chemistry Development Kit;\cite{steinbeck:2003} JChem Base;\cite{jchemBase} the Daylight Toolkit;\cite{daylightToolkit} and OEChem TK.\cite{oetk} SMILES encodings can be found in many public-facing databases, including: PubChem;\cite{kim:2016} ChEBI;\cite{degtyarenko:2007} ZINC;\cite{irwin:2005} ChEMBL;\cite{gaulton:2012} and Wikipedia.\cite{wikipediaCAS} Raw SMILES strings have been used extensively in machine learning applications.\cite{sousa:2021} SMILES has also been extended to carry various forms of metadata, as exemplified by Jmol SMILES,\cite{hanson:2016} CurleySMILES,\cite{drefahl:2011} BigSMILES,\cite{lin:2019} and CXSMILES.\cite{cxsmiles}
+Read and write functionalities are routinely supported by popular cheminformatics toolkits, including: Open Babel~\cite{oBoyle:2011}; RDKit~\cite{rdkit}; Chemistry Development Kit~\cite{steinbeck:2003}; JChem Base~\cite{jchemBase}; the Daylight Toolkit~\cite{daylightToolkit}; and OEChem TK~\cite{oetk}.
+SMILES encodings can be found in many public-facing databases, including: PubChem~\cite{kim:2016}; ChEBI~\cite{degtyarenko:2007}; ZINC~\cite{irwin:2005}; ChEMBL~\cite{gaulton:2012}; and Wikipedia~\cite{wikipediaCAS}.
+Raw SMILES strings have been used extensively in machine learning applications~\cite{sousa:2021}.
+SMILES has also been extended to carry various forms of metadata, as exemplified by Jmol SMILES~\cite{hanson:2016}, CurleySMILES~\cite{drefahl:2011}, BigSMILES~\cite{lin:2019}, and CXSMILES~\cite{cxsmiles}.
 
@@
 Earlier descriptions of SMILES were published in various media, although these are not cited by Weininger.
-An currently-accessible example is a 1987 report to the US Environmental Protection Agency.\cite{anderson:1987} Although the language described there shares many features with later descriptions, there are important differences as well.
+A currently-accessible example is a 1987 report to the US Environmental Protection Agency~\cite{anderson:1987}.
+Although the language described there shares many features with later descriptions, there are important differences as well.
 
 Since 1988, two authoritative sources have attempted to clarify or revise Weininger's 1988 description of SMILES.
-A 2003 book chapter by Weininger addressed some of the limitations of the paper.\cite{weininger:2008} The Daylight Theory Manual, a website maintained by SMILES's corporate sponsor, Daylight Chemical Information Systems, Inc.
-("Daylight"), recapitulates previously-published material.\cite{daylightTheory}.
-The lack of other written sources describing SMILES is consistent with Weininger's reported disinterest in traditional scientific publication.\cite{gasteiger:2018}
-
-In 2007 a documentation effort that would become known as OpenSMILES began.\cite{openSMILES} OpenSMILES was conceived as "a non-proprietary specification for the SMILES language." Many of the points left open by previous SMILES documentation efforts were addressed.
-Noteworthy contributions included: the first formal grammar, which was later adapted;\cite{denjoed:2010} many refinements around stereochemistry; and the introduction of "standard form." Absent were detailed procedures for assigning and interpreting aromatic features, and a detailed procedure for computing implicit hydrogen count.
+A 2003 book chapter by Weininger addressed some of the limitations of the paper~\cite{weininger:2008}.
+The Daylight Theory Manual, a website maintained by SMILES's corporate sponsor, Daylight Chemical Information Systems, Inc.
+("Daylight"), recapitulates previously-published material~\cite{daylightTheory}.
+The lack of other written sources describing SMILES is consistent with Weininger's reported disinterest in traditional scientific publication~\cite{gasteiger:2018}.
+
+In 2007 a documentation effort that would become known as OpenSMILES began~\cite{openSMILES}.
+OpenSMILES was conceived as "a non-proprietary specification for the SMILES language." Many of the points left open by previous SMILES documentation efforts were addressed.
+Noteworthy contributions included: the first formal grammar, which was later adapted~\cite{denjoed:2010}; many refinements around stereochemistry; and the introduction of "standard form." Absent were detailed procedures for assigning and interpreting aromatic features, and a detailed procedure for computing implicit hydrogen count.
 OpenSMILES also left several points of syntactic and semantic ambiguity unsettled, while introducing extensions of its own.
 
-In 2019 IUPAC announced the SMILES+ initiative.\cite{smilesPlusSpecification} Noting the limitations of existing SMILES documentation, the SMILES+ effort seeks to "establish a formalized recommended up-to-date specification of the SMILES format." SMILES+ took as its starting point the documentation produced by the OpenSMILES project.
-Efforts to extend this starting point are in progress online through a public repository, but no formal recommendation has to date resulted.\cite{smilesPlusGitHub}
+In 2019 IUPAC announced the SMILES+ initiative~\cite{smilesPlusSpecification}.
+Noting the limitations of existing SMILES documentation, the SMILES+ effort seeks to "establish a formalized recommended up-to-date specification of the SMILES format." SMILES+ took as its starting point the documentation produced by the OpenSMILES project.
+Efforts to extend this starting point are in progress online through a public repository, but no formal recommendation has to date resulted~\cite{smilesPlusGitHub}.
 
@@
 The lack of a detailed, comprehensive, and universally-applied SMILES specification also hampers broader data integrity efforts.
-The FAIR Guiding Principles\cite{wilkinson:2016} identify four qualities essential for extracting maximum utility from published scholarly data: Findability; Accessibility; Interoperability; and Reusability.
+The FAIR Guiding Principles identify four qualities essential for extracting maximum utility from published scholarly data: Findability; Accessibility; Interoperability; and Reusability~\cite{wilkinson:2016}.
 Interoperability in particular requires that "(meta)data" use a formal, accessible, shared, and broadly applicable language for knowledge representation." Although SMILES may be a shared, accessible, and broadly applicable language, its formal underpinnings leave room for improvement.
@@
 Others claim SMILES compatibility while ignoring certain aspects of written documentation.
-The Daylight implementation, like some of the others, is distributed under a non-Open Source license that restricts use.\cite{daylightToolkit} For several years Daylight operated a publicly-accessible web service based on its toolkit that depicted SMILES strings, but it has since been decommissioned.\cite{depictArchive} Although SMILES may underpin modern chemical information exchange, the foundations of this system are weathered and showing signs of stress.
+The Daylight implementation, like some of the others, is distributed under a non-Open Source license that restricts use~\cite{daylightToolkit}.
+For several years Daylight operated a publicly-accessible web service based on its toolkit that depicted SMILES strings, but it has since been decommissioned~\cite{depictArchive}.
+Although SMILES may underpin modern chemical information exchange, the foundations of this system are weathered and showing signs of stress.
 
@@
 
-Dialect's system of representation is based on the \textit{molecular graph} concept.\cite{balaban:1985} A graph is a data structure comprised of a set of nodes (which map to "atoms") and a set of pairwise relationships between them called edges (which map to "bonds").
+Dialect's system of representation is based on the \textit{molecular graph} concept~\cite{balaban:1985}.
+A graph is a data structure comprised of a set of nodes (which map to "atoms") and a set of pairwise relationships between them called edges (which map to "bonds").
 A molecular graph is a specialized graph onto which atomic and bonding metadata have been overlaid.
-Molecular graphs are ubiquitous in cheminformatics, appearing in contexts such as data formats like Chemical Markup Language\cite{murray-rust:2011}, CDXML\cite{cdxml}, and Molfile\cite{ctfileFormats}, as well as in-memory data structures found in cheminformatics toolkits.
+Molecular graphs are ubiquitous in cheminformatics, appearing in contexts such as data formats like Chemical Markup Language~\cite{murray-rust:2011}, CDXML~\cite{cdxml}, and Molfile~\cite{ctfileFormats}, as well as in-memory data structures found in cheminformatics toolkits.
 Like these systems, Dialect augments its underlying graph with node and edge labels.
@@
 Two attributes characterize an atom's nucleus: \ttt{element} and \ttt{isotope}.
-The `element` attribute is an optional one- or two-letter character sequence selected from the set designated by the IUPAC.\cite{periodicTable} When IUPAC authorizes additional symbols, they will become valid values for the \ttt{element} attribute.
+The `element` attribute is an optional one- or two-letter character sequence selected from the set designated by the IUPAC~\cite{periodicTable}.
+When IUPAC authorizes additional symbols, they will become valid values for the \ttt{element} attribute.
 If the elemental identity of an atom is unknown, its \ttt{element} attribute can be omitted.
@@
 This is a fundamental requirement for the principle of mass conversion.
-In a more practical sense, electron counting lays the foundation for calculations of formal charge and bond order.\cite{dietz:1995}
-
-Electron counting in Dialect is based on the well-known \textit{valence bond model} (VB Model).\cite{lewis:1916} Although not commonly considered as such, the VB model can be thought of as an algorithm for molecular assembly.
+In a more practical sense, electron counting lays the foundation for calculations of formal charge and bond order~\cite{dietz:1995}.
+
+Electron counting in Dialect is based on the well-known \textit{valence bond model} (VB Model)~\cite{lewis:1916}.
+Although not commonly considered as such, the VB model can be thought of as an algorithm for molecular assembly.
 The perspective is especially useful given that Dialect itself can be viewed as a programming language for molecular assembly.
@@
 Electron delocalization results when the electrons of a valence bond representation can be distributed over more than one pair of atoms, as is the case with benzene and related molecules.
-DIME is similar to "resonance" or "aromaticity," but those terms are avoided here due to their ambiguity.\cite{randic:2018,kerber:2006}
+DIME is similar to "resonance" or "aromaticity," but those terms are avoided here due to their ambiguity~\cite{randic:2018,kerber:2006}.
 
@@
 Because the presence of a filled DS implies a perfect matching over it, kekulization always succeeds.
-A popular algorithm for matching, the Edmunds "blossom algorithm,"\cite{edmonds:1965} has a time complexity of $O(|E||V|^2)$, where $|E|$ is the number of edges and $|V|$ is the number of nodes.
+A popular algorithm for matching, the Edmunds "blossom algorithm," has a time complexity of $O(|E||V|^2)$, where $|E|$ is the number of edges and $|V|$ is the number of nodes~\cite{edmonds:1965}.
 Although more efficient algorithms are known, they are either much more difficult to implement or lack generality.
@@
 Not all tetracoordinate atoms will be \textit{stereocenters}.
-A stereocenter is an atom whose "ligand permutation produces stereoisomers," as defined by Mislow and Siegel.\cite{mislow:1984} In the context of Dialect, a stereocenter must be tetracoordinate.
+A stereocenter is an atom whose "ligand permutation produces stereoisomers," as defined by Mislow and Siegel~\cite{mislow:1984}.
+In the context of Dialect, a stereocenter must be tetracoordinate.
 Readers should avoid the assignment of stereodescriptors to atoms that are not stereocenters.
@@
 This could imply that either a single configuration or a configurational mixture are present.
-This interpretation is consistent with the one used by molfile format.\cite{ctfileFormats}
+This interpretation is consistent with the one used by molfile format~\cite{ctfileFormats}.
 
@@
 This set of features makes LL(1) grammars such as the one used by Dialect a good fit for manually-written recursive descent parsers.
-LL(1) grammars can also be used as a basis for auto-generated parsers through packages such as ANTLR.\cite{parr:2014} The full grammar for Dialect strings is available as a text file in this paper's Supporting Material.
+LL(1) grammars can also be used as a basis for auto-generated parsers through packages such as ANTLR~\cite{parr:2014}.
+The full grammar for Dialect strings is available as a text file in this paper's Supporting Material.
 
@@
 Language subsets have ample precedent in computer science.
-A well-known example is MISRA C,\cite{misraC} a subset of the C programming language that aims to eliminate "known undefined or otherwise dangerous behavior."\cite{hatton:2007} Another example is the subset of JavaScript described by Crockford whose purpose is to "chip away at the features that are not beautiful until the language's true nature reveals itself."\cite{crockford:2008} Language subsets arise for two main reasons.
+A well-known example is MISRA C~\cite{misraC}, a subset of the C programming language that aims to eliminate "known undefined or otherwise dangerous behavior"~\cite{hatton:2007}.
+Another example is the subset of JavaScript described by Crockford whose purpose is to "chip away at the features that are not beautiful until the language's true nature reveals itself."~\cite{crockford:2008}.
+Language subsets arise for two main reasons.
 First, designing a language presents many opportunities for errors small and large.
@@
 \begin{itemize}
-    \item Weininger's report to the US Environmental Protection Agency ("The Article").\cite{anderson:1987}
-    \item Weininger's original publication (herein "The Article").\cite{weininger:1988}
-    \item Weininger's 2003 book chapter, revised in 2008 ("The Book Chapter").\cite{weininger:2008}
+    \item Weininger's report to the US Environmental Protection Agency ("The Article")~\cite{anderson:1987}.
+    \item Weininger's original publication (herein "The Article")~\cite{weininger:1988}.
+    \item Weininger's 2003 book chapter, revised in 2008 ("The Book Chapter")~\cite{weininger:2008}.
     \item An online manual maintained by Daylight ("The Manual").
-    \item The Daylight's toolkit, which implements a SMILES reader and writer ("The Toolkit").\cite{daylightToolkit}
+    \item The Daylight's toolkit, which implements a SMILES reader and writer ("The Toolkit")~\cite{daylightToolkit}.
 \end{itemize}
@@
 
-The Book Chapter defines a fixed set of element symbols which even at the time of publication was obsolete.\cite[p.~83]{weininger:2008} The symbol for element 105 is given as Ha (Hahnium), despite IUPAC's resolution of this naming controversy in favor of the symbol Db (Dubnium) years prior.\cite{sageson:1997} Accordingly, Dialect disallows the symbol Ha.
+The Book Chapter defines a fixed set of element symbols which even at the time of publication was obsolete~\cite[p.~83]{weininger:2008}.
+The symbol for element 105 is given as Ha (Hahnium), despite IUPAC's resolution of this naming controversy in favor of the symbol Db (Dubnium) years prior~\cite{sageson:1997}.
+Accordingly, Dialect disallows the symbol Ha.
 To support the full set of currently-named transuranium elements, Dialect adds the symbols: Db; Sg; Bh; Hs; Mt; Rg; Cn; Nh; Fl; Mc; Lv; Ts; and Og.
@@
 
-The Article implies that comma (\ttt{,}) is either an element symbol or a valid SMILES character: "This subset uses only the symbols H, C, N, O, P, S, F, Cl, Br, I, and (,) and digits, with the following four rules...".\cite[p.~33]{weininger:1988} The comma character is explicitly disallowed in Dialect.
-
-The Book Chapter defines a recursive grammar for branching which, perhaps without the knowledge of the author, defines sequential parentheses.\cite[p.~86]{weininger:2008} Dialect disallows such constructs, including for example "*((*))*" and "*(((*)))*".
-
-"Reaction SMILES" is an extension described in the Book Chapter,\cite[p.~89]{weininger:2008} but disallowed in Dialect.
+The Article implies that comma (\ttt{,}) is either an element symbol or a valid SMILES character: "This subset uses only the symbols H, C, N, O, P, S, F, Cl, Br, I, and (,) and digits, with the following four rules..."~\cite[p.~33]{weininger:1988}.
+The comma character is explicitly disallowed in Dialect.
+
+The Book Chapter defines a recursive grammar for branching which, perhaps without the knowledge of the author, defines sequential parentheses~\cite[p.~86]{weininger:2008}.
+Dialect disallows such constructs, including for example "*((*))*" and "*(((*)))*".
+
+"Reaction SMILES" is an extension described in the Book Chapter~\cite[p.~89]{weininger:2008}, but disallowed in Dialect.
 Specifically, the greater than symbol (\ttt{>}) is not a valid Dialect character.
@@
 Dialect only supports two stereodescriptors, encoded with the terminals "@" and "@@".
-The Book Chapter provides a recursive grammar for stereodescriptors that allows multi-symbol descriptors such as "@@@" and "@@@@".\cite[p.~94]{weininger:2008} The book chapter also discusses non-tetrahedral descriptors including "@AL1", "@AL2", "@1", and "@SP1".
+The Book Chapter provides a recursive grammar for stereodescriptors that allows multi-symbol descriptors such as "@@@" and "@@@@"~\cite[p.~94]{weininger:2008}.
+The book chapter also discusses non-tetrahedral descriptors including "@AL1", "@AL2", "@1", and "@SP1".
 None of these are supported by Dialect.
@@
 The Book Chapter notes "Adjacent atoms without an intervening bond symbol are connected by a valence-dictated bond (typically a single or aromatic bond).
-"-" (single) and ":" (aromatic) bond symbols may always be omitted on input".\cite[p.~85]{weininger:2008} This passage implies that aromatic bonds serve no purpose.
+"-" (single) and ":" (aromatic) bond symbols may always be omitted on input"~\cite[p.~85]{weininger:2008}.
+This passage implies that aromatic bonds serve no purpose.
 Accordingly, Dialect does not allow them and the syntax does not include the colon symbol (\ttt{:}).
@@
 Doing so would increase the complexity of the language unnecessarily.
-Although The Paper implies selectability ("Elements other than C, N, O, P, S, As, and Se are not yet dealt with in an aromatic context."),\cite[p.~35]{weininger:1988} default valences for the latter two atoms are not provided.
+Although The Paper implies selectability ("Elements other than C, N, O, P, S, As, and Se are not yet dealt with in an aromatic context.")~\cite[p.~35]{weininger:1988}, default valences for the latter two atoms are not provided.
 Even if they were, the elements As and Se must be bracketed.
@@
 It's unclear whether ProtoSMILES allows a virtual hydrogen count on hydrogen itself.
-The Article implies that the element symbol H (hydrogen) must not have an implicit hydrogen count when used within a bracket sequence ("There are few exceptions to the hydrogen-suppression convention, the most obvious being specification of a proton, [H+], and molecular hydrogen, [H][H].").\cite[p.~34]{weininger:1988} If so, the author may have erroneously believed that parsers would be incapable of processing constructs such as "[HH]".
+The Article implies that the element symbol H (hydrogen) must not have an implicit hydrogen count when used within a bracket sequence ("There are few exceptions to the hydrogen-suppression convention, the most obvious being specification of a proton, [H+], and molecular hydrogen, [H][H].")~\cite[p.~34]{weininger:1988}.
+If so, the author may have erroneously believed that parsers would be incapable of processing constructs such as "[HH]".
 The Book Chapter appears to reiterate this view, noting that "explicit hydrogen specification is required" in the case of "hydrogens connected to other hydrogens, e.g.
-[H][H], molecular hydrogen".\cite[p.~97]{weininger:2008} To be clear, Dialect supports any valid virtual hydrogen count on hydrogen itself.
+[H][H], molecular hydrogen"~\cite[p.~97]{weininger:2008}.
+To be clear, Dialect supports any valid virtual hydrogen count on hydrogen itself.
 For example, the following are all valid, if chemically nonsensical, strings: "[HH]"; "[HH2]"; and "[HH3]".
@@
 As noted in the Book Chapter, "...
-In terms of the valence model being represented, the dot literally represents a bond of formal order zero: the atoms on either side of the dot are explicitly not bonded to each other.".\cite[p.~88]{weininger:2008} The state of not being bonded and the state of having a zero-order bond are not necessarily equivalent.\cite{clark:2011} To be clear, Dialect explicitly disallows the zero bond order interpretation of detachments and the dot character.
+In terms of the valence model being represented, the dot literally represents a bond of formal order zero: the atoms on either side of the dot are explicitly not bonded to each other."~\cite[p.~88]{weininger:2008}.
+The state of not being bonded and the state of having a zero-order bond are not necessarily equivalent~\cite{clark:2011}.
+To be clear, Dialect explicitly disallows the zero bond order interpretation of detachments and the dot character.
 
@@
 Although it might be argued that lower bounds on physical quantities such as isotope and hydrogen count should be implicit, the lack of precision forces implementors to improvise, which can lead to divergent behavior.
-ProtoSMILES places no upper bound on the charge notation "+", "++", "+++", and their corresponding negative forms ("The SMILES program also recognizes constructions of the form [Fe+++] as being synonymous with the form [Fe+3].").\cite[p.~32]{weininger:1988} However, Dialect disallows duplicated plus and minus characters (\ttt{+} and \ttt{-}, respectively) from bracket atoms.
+ProtoSMILES places no upper bound on the charge notation "+", "++", "+++", and their corresponding negative forms ("The SMILES program also recognizes constructions of the form [Fe+++] as being synonymous with the form [Fe+3].")~\cite[p.~32]{weininger:1988}.
+However, Dialect disallows duplicated plus and minus characters (\ttt{+} and \ttt{-}, respectively) from bracket atoms.
 
@@
 The Article reports the only default valence for nitrogen as three (p.
-34),\cite[p.~34]{weininger:1988} whereas the Book Chapter reports two default valences: 3 and 5.\cite[p.~84]{weininger:2008} Dialect uses the latter default valences.
-Atoms whose \ttt{hydrogens} attribute may be set to \ttt{Implicit} ("the organic subset") are reported in the Article as: B; C; N; O; P; S; F; Cl; Br; and I.\cite[p.~32]{weininger:2008} However, an example later in the Article indicates that "Hn1cccc1" with its unbracketed hydrogen is a valid string.\cite[p.~35]{weininger:1988} Dialect explicitly rejects strings containing unbracketed hydrogen atoms.
-On the topic of atom selection ("aromaticity"), the Article implies that selected atoms must be part of a cycle: "To qualify as aromatic, all atoms in the ring must be sp2 hybridized and the number of available 'excess' $\pi$ electrons must satisfy Hückel's 4{\textit{N}} + 2 criterion."\cite[p.~34]{weininger:1988} The Book Chapter, however, presents two examples of valid strings that do not encode cycles ("cc" and "cccc").\cite[p 85]{weininger:2008} Consistent with the latter view, Dialect explicitly allows the selection of atoms regardless of cycle membership.
+34)~\cite[p.~34]{weininger:1988}, whereas the Book Chapter reports two default valences: 3 and 5~\cite[p.~84]{weininger:2008}.
+Dialect uses the latter default valences.
+Atoms whose \ttt{hydrogens} attribute may be set to \ttt{Implicit} ("the organic subset") are reported in the Article as: B; C; N; O; P; S; F; Cl; Br; and I~\cite[p.~32]{weininger:2008}.
+However, an example later in the Article indicates that "Hn1cccc1" with its unbracketed hydrogen is a valid string~\cite[p.~35]{weininger:1988}.
+Dialect explicitly rejects strings containing unbracketed hydrogen atoms.
+On the topic of atom selection ("aromaticity"), the Article implies that selected atoms must be part of a cycle: "To qualify as aromatic, all atoms in the ring must be sp2 hybridized and the number of available 'excess' $\pi$ electrons must satisfy Hückel's 4{\textit{N}} + 2 criterion."~\cite[p.~34]{weininger:1988}.
+The Book Chapter, however, presents two examples of valid strings that do not encode cycles ("cc" and "cccc")~\cite[p 85]{weininger:2008}.
+Consistent with the latter view, Dialect explicitly allows the selection of atoms regardless of cycle membership.
 

### 2022-03-24 manuscript.tex
Uses smart quotes. closes #100

@@
 Despite its many applications, SMILES remains a language with an incomplete specification.
-Weininger's 1988 paper leaves the following either fully or incompletely specified: (1) stereochemical configuration; (2) double bond conformation; (3) the meaning handling of "aromatic" features; (4) the computation of implicit hydrogen count, especially as related to aromatic features; (5) formal syntax; (6) constraints on quantities such as mass number and charge; and (7) enumeration of error states.
+Weininger's 1988 paper leaves the following either fully or incompletely specified: (1) stereochemical configuration; (2) double bond conformation; (3) the meaning handling of ``aromatic'' features; (4) the computation of implicit hydrogen count, especially as related to aromatic features; (5) formal syntax; (6) constraints on quantities such as mass number and charge; and (7) enumeration of error states.
 
@@
 The Daylight Theory Manual, a website maintained by SMILES's corporate sponsor, Daylight Chemical Information Systems, Inc.
-("Daylight"), recapitulates previously-published material~\cite{daylightTheory}.
+(``Daylight''), recapitulates previously-published material~\cite{daylightTheory}.
 The lack of other written sources describing SMILES is consistent with Weininger's reported disinterest in traditional scientific publication~\cite{gasteiger:2018}.
@@
 In 2007 a documentation effort that would become known as OpenSMILES began~\cite{openSMILES}.
-OpenSMILES was conceived as "a non-proprietary specification for the SMILES language." Many of the points left open by previous SMILES documentation efforts were addressed.
-Noteworthy contributions included: the first formal grammar, which was later adapted~\cite{denjoed:2010}; many refinements around stereochemistry; and the introduction of "standard form." Absent were detailed procedures for assigning and interpreting aromatic features, and a detailed procedure for computing implicit hydrogen count.
+OpenSMILES was conceived as ``a non-proprietary specification for the SMILES language.'' Many of the points left open by previous SMILES documentation efforts were addressed.
+Noteworthy contributions included: the first formal grammar, which was later adapted~\cite{denjoed:2010}; many refinements around stereochemistry; and the introduction of ``standard form.'' Absent were detailed procedures for assigning and interpreting aromatic features, and a detailed procedure for computing implicit hydrogen count.
 OpenSMILES also left several points of syntactic and semantic ambiguity unsettled, while introducing extensions of its own.
@@
 In 2019 IUPAC announced the SMILES+ initiative~\cite{smilesPlusSpecification}.
-Noting the limitations of existing SMILES documentation, the SMILES+ effort seeks to "establish a formalized recommended up-to-date specification of the SMILES format." SMILES+ took as its starting point the documentation produced by the OpenSMILES project.
+Noting the limitations of existing SMILES documentation, the SMILES+ effort seeks to ``establish a formalized recommended up-to-date specification of the SMILES format.'' SMILES+ took as its starting point the documentation produced by the OpenSMILES project.
 Efforts to extend this starting point are in progress online through a public repository, but no formal recommendation has to date resulted~\cite{smilesPlusGitHub}.
@@
 The FAIR Guiding Principles identify four qualities essential for extracting maximum utility from published scholarly data: Findability; Accessibility; Interoperability; and Reusability~\cite{wilkinson:2016}.
-Interoperability in particular requires that "(meta)data" use a formal, accessible, shared, and broadly applicable language for knowledge representation." Although SMILES may be a shared, accessible, and broadly applicable language, its formal underpinnings leave room for improvement.
-
-This situation has persisted long enough that the notion of a SMILES "standard" may not currently be feasible.
+Interoperability in particular requires that ``(meta)data'' use a formal, accessible, shared, and broadly applicable language for knowledge representation.'' Although SMILES may be a shared, accessible, and broadly applicable language, its formal underpinnings leave room for improvement.
+
+This situation has persisted long enough that the notion of a SMILES standard may not currently be feasible.
 A dozen or more slightly different implementations are now in use.
 None of them can be considered a reference standard.
-Some have introduced extensions not recognized by the others, contributing to language "drift" while compounding the already formidable difficulties faced by implementors.
+Some have introduced extensions not recognized by the others, contributing to language drift while compounding the already formidable difficulties faced by implementors.
 Others claim SMILES compatibility while ignoring certain aspects of written documentation.
@@
 Dialect's system of representation is based on the \textit{molecular graph} concept~\cite{balaban:1985}.
-A graph is a data structure comprised of a set of nodes (which map to "atoms") and a set of pairwise relationships between them called edges (which map to "bonds").
+A graph is a data structure comprised of a set of nodes (which map to atoms) and a set of pairwise relationships between them called edges (which map to bonds).
 A molecular graph is a specialized graph onto which atomic and bonding metadata have been overlaid.
@@
 A Dialect molecule consists of a graph with zero or more nodes and zero or more edges.
-The "empty molecule", devoid of atoms and bonds, is allowed.
+The empty molecule, devoid of atoms and bonds, is allowed.
 Dialect imposes no upper bound on the number of atoms or bonds.
@@
 Table \ref{table:atomic-state} summarizes atomic attributes.
-Not all combinations of values ("atomic state") are valid.
+Not all combinations of values (``atomic state'') are valid.
 Implementations must ensure either the impossibility of an invalid state, or an error condition in the event that one is created.
@@
 When using this variant, the \ttt{hydrogens} attribute may assume integer values from zero to nine, inclusive.
-The process of transforming one or more hydrogen nodes into a virtual hydrogen is called "virtualization." The opposite process, transforming one or more virtual hydrogens into a node, is called "reification."
+The process of transforming one or more hydrogen nodes into a virtual hydrogen is called ``virtualization.'' The opposite process, transforming one or more virtual hydrogens into a node, is called ``reification.''
 
@@
 Electron delocalization results when the electrons of a valence bond representation can be distributed over more than one pair of atoms, as is the case with benzene and related molecules.
-DIME is similar to "resonance" or "aromaticity," but those terms are avoided here due to their ambiguity~\cite{randic:2018,kerber:2006}.
+DIME is similar to ``resonance'' or ``aromaticity,'' but those terms are avoided here due to their ambiguity~\cite{randic:2018,kerber:2006}.
 
@@
 
-An atom is added to a DS by setting its \ttt{selected} attribute to \ttt{true} ("selection").
+An atom is added to a DS by setting its \ttt{selected} attribute to \ttt{true} (``selection'').
 For an atom to be selectable, its \ttt{element} attribute must be contained in the list: \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; and \ttt{S}.
@@
 Because the presence of a filled DS implies a perfect matching over it, kekulization always succeeds.
-A popular algorithm for matching, the Edmunds "blossom algorithm," has a time complexity of $O(|E||V|^2)$, where $|E|$ is the number of edges and $|V|$ is the number of nodes~\cite{edmonds:1965}.
+A popular algorithm for matching, the Edmunds ``blossom algorithm,'' has a time complexity of $O(|E||V|^2)$, where $|E|$ is the number of edges and $|V|$ is the number of nodes~\cite{edmonds:1965}.
 Although more efficient algorithms are known, they are either much more difficult to implement or lack generality.
@@
 For example, a selection algorithm can restrict candidate atoms to those found in cycles.
-Electron-counting techniques can also be introduced to approximate the chemical concept of "aromaticity." A double bond between two selected atoms may or may not be elided, depending on the application.
+Electron-counting techniques can also be introduced to approximate the chemical concept of ``aromaticity.'' A double bond between two selected atoms may or may not be elided, depending on the application.
 
@@
 
-To support the assignment and interpretation stereochemical features, each node is assigned a sequential, unique, non-negative integer \ttt{index} attribute ("index").
+To support the assignment and interpretation stereochemical features, each node is assigned a sequential, unique, non-negative integer \ttt{index} attribute (``index'').
 The value of an atom's index equals zero for a molecular's graph's first atom.
@@
 
-To support this system, each bond carries a nullable \ttt{direction} attribute ("direction").
+To support this system, each bond carries a nullable \ttt{direction} attribute (``direction'').
 When present, this attribute may assume one of the two values \ttt{Up} or \ttt{Down}.
@@
 The double bond terminals are placed on the x-axis such that the terminal with the lowest \ttt{index} attribute appears to the left of the terminal with the higher atomic \ttt{index} attribute.
-These terminals are designated "left terminal" and "right terminal," respectively.
+These terminals are designated ``left terminal'' and ``right terminal,'' respectively.
 Each neighbor of a terminal is then assigned a relative coordinate based on the \ttt{index} attribute and the direction of its PPB bond.
@@
 Dialect limits configuration to the special case of an atom with exactly four substituents placed at the vertexes of a tetrahedron.
-Here, "substituent" means an atomic neighbor; lone electron pairs are not considered substituents.
+Here, ``substituent'' means an atomic neighbor; lone electron pairs are not considered substituents.
 One of the four substituents may be a virtual hydrogen.
@@
 Dialect supports two configurational descriptors: \ttt{Left} and \ttt{Right}.
-The context for the terms "left" and "right" will be explained shortly.
+The context for the terms ``left'' and ``right'' will be explained shortly.
 
@@
 Not all tetracoordinate atoms will be \textit{stereocenters}.
-A stereocenter is an atom whose "ligand permutation produces stereoisomers," as defined by Mislow and Siegel~\cite{mislow:1984}.
+A stereocenter is an atom whose ``ligand permutation produces stereoisomers,'' as defined by Mislow and Siegel~\cite{mislow:1984}.
 In the context of Dialect, a stereocenter must be tetracoordinate.
@@
 
-A Dialect string ("string") is a UTF-encoded sequence of zero or more characters chosen from the set: \ttt{A}; \ttt{B}; \ttt{C}; \ttt{D}; \ttt{E}; \ttt{F}; \ttt{G}; \ttt{H}; \ttt{I}; \ttt{K}; \ttt{L}; \ttt{M}; \ttt{N}; \ttt{O}; \ttt{P}; \ttt{R}; \ttt{S}; \ttt{T}; \ttt{U}; \ttt{V}; \ttt{W}; \ttt{X}; \ttt{Y}; \ttt{Z}; \ttt{a}; \ttt{b}; \ttt{c}; \ttt{d}; \ttt{e}; \ttt{f}; \ttt{g}; \ttt{h}; \ttt{i}; \ttt{k}; \ttt{l}; \ttt{m}; \ttt{n}; \ttt{o}; \ttt{p}; \ttt{r}; \ttt{s}; \ttt{t}; \ttt{u}; \ttt{v}; \ttt{0}; \ttt{1}; \ttt{2}; \ttt{3}; \ttt{4}; \ttt{5}; \ttt{6}; \ttt{7}; \ttt{8}; \ttt{9}; \ttt{+}; \ttt{-}; \ttt{.}; \ttt{\%}; \ttt{@}; \ttt{*}; \ttt{[}; \ttt{]}; \ttt{/}; \ttt{\char`\\}; \ttt{=}; and \ttt{\#}.
+A Dialect string (``string'') is a UTF-encoded sequence of zero or more characters chosen from the set: \ttt{A}; \ttt{B}; \ttt{C}; \ttt{D}; \ttt{E}; \ttt{F}; \ttt{G}; \ttt{H}; \ttt{I}; \ttt{K}; \ttt{L}; \ttt{M}; \ttt{N}; \ttt{O}; \ttt{P}; \ttt{R}; \ttt{S}; \ttt{T}; \ttt{U}; \ttt{V}; \ttt{W}; \ttt{X}; \ttt{Y}; \ttt{Z}; \ttt{a}; \ttt{b}; \ttt{c}; \ttt{d}; \ttt{e}; \ttt{f}; \ttt{g}; \ttt{h}; \ttt{i}; \ttt{k}; \ttt{l}; \ttt{m}; \ttt{n}; \ttt{o}; \ttt{p}; \ttt{r}; \ttt{s}; \ttt{t}; \ttt{u}; \ttt{v}; \ttt{0}; \ttt{1}; \ttt{2}; \ttt{3}; \ttt{4}; \ttt{5}; \ttt{6}; \ttt{7}; \ttt{8}; \ttt{9}; \ttt{+}; \ttt{-}; \ttt{.}; \ttt{\%}; \ttt{@}; \ttt{*}; \ttt{[}; \ttt{]}; \ttt{/}; \ttt{\char`\\}; \ttt{=}; and \ttt{\#}.
 The internal structure of a string reflects a depth-first traversal of the corresponding molecular graphs.
@@
 
-Dialect's formal grammar is presented as a series of \textit{production rules} (aka "productions").
+Dialect's formal grammar is presented as a series of \textit{production rules} (aka ``productions'').
 A production rule defines a transformation allowed under the grammar.
@@
 A production rule is composed of two kinds of elements: \textit{terminals} and \textit{non-terminals}.
-A terminal is a character literal (e.g., "A").
+A terminal is a character literal (e.g., ``A'').
 A non-terminal is a reference to another production rule.
@@
 
-Consider a hypothetical language composed of variable-length sequences of the lowercase letter "a". Such a language could be cast as the following two production rules:
+Consider a hypothetical language composed of variable-length sequences of the lowercase letter ``a''. Such a language could be cast as the following two production rules:
 
@@
 The quantifier (\ttt{*}) indicates that a text in this language consists of a sequence of zero or more instances of the production \ttt{<a>}.
-This production in turn is defined as the lowercase letter "a".
+This production in turn is defined as a lowercase letter (`a`).
 Therefore, valid texts in this language include the empty string, \ttt{a}, \ttt{aa}, and \ttt{aaaaaa} to name a few.
@@
 The first atomic non-terminal, \ttt{<star>}, can assume one terminal value, the asterisk character (\ttt{*}).
-This "star atom" represents an atom in which every attribute assumes its default value.
+This ``star atom'' represents an atom in which every attribute assumes its default value.
 The implicit hydrogen count of a star atom is zero.
@@
 
-The next atomic production rule, \ttt{<shortcut>} is a non-terminal selected from the list: "B"; "C"; "N"; "O"; "P"; "S"; "F"; "Cl"; "Br"; and "I".
-An atom encoded in this way ("shortcut atom") assigns the corresponding symbol to the \ttt{element} attribute.
+The next atomic production rule, \ttt{<shortcut>} is a non-terminal selected from the list: ``B''; ``C''; ``N''; ``O''; ``P''; ``S''; ``F''; ``Cl''; ``Br''; and ``I.'' An atom encoded in this way (``shortcut atom'') assigns the corresponding symbol to the \ttt{element} attribute.
 All other attributes retain their default values.
@@
 
-The third atomic production rule, \ttt{<selection>} is a non-terminal selected from the list: "c"; "n"; "o"; "p"; and "s".
-An atom encoded in this way ("selected shortcut atom") assigns the corresponding atom symbol to the \ttt{element} attribute and sets the \ttt{selected} attribute to `true`.
+The third atomic production rule, \ttt{<selection>} is a non-terminal selected from the list: ``c''; ``n''; ``o''; ``p''; and ``s.'' An atom encoded in this way (``selected shortcut atom'') assigns the corresponding atom symbol to the \ttt{element} attribute and sets the \ttt{selected} attribute to `true`.
 All other atomic attributes retain their default values.
@@
 
-The fourth and most complex atomic production rule is \ttt{<bracket>} ("bracket atom").
+The fourth and most complex atomic production rule is \ttt{<bracket>} (``bracket atom'').
 A bracket atom can be used to set any atomic attribute.
@@
 Leading zeros are disregarded when assigning the value of the \ttt{isotope} attribute.
-In other words, "007" and "7" are considered equivalent expressions of the value 7.
+In other words, ``007'' and ``7'' are considered equivalent expressions of the value 7.
 
@@
 
-The values of a bracket atom's \ttt{element} and \ttt{selected} attributes are determined by the \ttt{<symbol>} production rule ("symbol").
+The values of a bracket atom's \ttt{element} and \ttt{selected} attributes are determined by the \ttt{<symbol>} production rule (``symbol'').
 Three non-terminal options are available: \ttt{<star>}; \ttt{<element>}; and \ttt{<selection>}.
@@
 The \ttt{stereodescriptor} attribute of a bracket atom is determined by the \ttt{<stereodescriptor>} non-terminal.
-Allowed values are "@" and "@@", corresponding to the \ttt{stereodescriptor} values \ttt{Counterclockwise} and \ttt{Clockwise}, respectively.
+Allowed values are ``@'' and ``@@,'' corresponding to the \ttt{stereodescriptor} values \ttt{Counterclockwise} and \ttt{Clockwise}, respectively.
 
@@
 The \ttt{hydrogens} attribute of a bracket atom is controlled by the \\ \ttt{<virtual{\_}hydrogen>} non-terminal.
-This non-terminal is comprised of the terminal "H" followed by an optional \ttt{<digit>} non-terminal.
-A digit appearing after the "H" terminal sets \ttt{hydrogens} to the corresponding virtual hydrogen count.
-A digit of 0 (i.e., "H0") sets the \ttt{hydrogens} attribute to zero.
+This non-terminal is comprised of the terminal ``H'' followed by an optional \ttt{<digit>} non-terminal.
+A digit appearing after the ``H'' terminal sets \ttt{hydrogens} to the corresponding virtual hydrogen count.
+A digit of 0 (i.e., ``H0'') sets the \ttt{hydrogens} attribute to zero.
 If no digit is present, \ttt{hydrogens} is set to one.
@@
 The \ttt{<charge>} production rule sets the \ttt{charge} attribute of a bracket atom.
-This non-terminal begins with either the plus or minus terminals ("+", "-", respectively) and ends with an optional \ttt{<digit>} non-terminal.
+This non-terminal begins with either the plus or minus terminals (``+'' and ``-,'' respectively) and ends with an optional \ttt{<digit>} non-terminal.
 A missing digit causes the atomic \ttt{charge} attribute to be set to one.
-In other words, "+" sets the \ttt{charge} attribute to one and "-" sets it to minus one.
-A digit of zero (e.g., "+0" or "-0") sets the \ttt{charge} attribute to zero.
+In other words, the plus terminal (``+'') sets the \ttt{charge} attribute to one and the minus terminal (``-'') sets it to minus one.
+A charge of zero (e.g., ``+0'' or ``-0'') sets the \ttt{charge} attribute to zero.
 
@@
 Should an optional bracket non-terminal not be present, the atom must retain its corresponding default value.
-For example, the bracket "[C@H+]" lacks the \ttt{<isotope>} non-terminal so the value of the \ttt{isotope} attribute will remain as \ttt{None}.
-Similarly, the bracket "[13CH+]" lacks the \ttt{<stereodescriptor>} non-terminal, so the corresponding \ttt{configuration} attribute remains \ttt{None}.
+For example, the bracket ``[C@H+]'' lacks the \ttt{<isotope>} non-terminal so the value of the \ttt{isotope} attribute will remain as \ttt{None}.
+Similarly, the bracket ``[13CH+]'' lacks the \ttt{<stereodescriptor>} non-terminal, so the corresponding \ttt{configuration} attribute remains \ttt{None}.
 
@@
 
-Five terminals are available ("-", "=", "\#", "/", and "\textbackslash").
-The first three ("-", "=", and "\#") set the \ttt{order} attribute of a bond to 1; 2; or 3, respectively.
-The last two terminals, "/" and "\textbackslash", set the \ttt{order} attribute to 1 while also setting the \ttt{direction} attribute to \ttt{Up} and \ttt{Down}, respectively.
+Five terminals are available (``-,'' ``=,'' ``\#,'' ``/,'' and ``\textbackslash'').
+The first three (``-,'' ``=,'' and ``\#'') set the \ttt{order} attribute of a bond to one, two, or three, respectively.
+The last two terminals, ``/'' and ``\textbackslash,'' set the \ttt{order} attribute to 1 while also setting the \ttt{direction} attribute to \ttt{Up} and \ttt{Down}, respectively.
 
@@
 
-An atom may be assigned zero or more children through the \ttt{<sequence>} non-terminal ("sequence").
+An atom may be assigned zero or more children through the \ttt{<sequence>} non-terminal (``sequence'').
 A sequence starts with a required \ttt{<atom>} non-terminal.
@@
 
-The \ttt{<cut>} non-terminal ("cut") can takes two forms.
+The \ttt{<cut>} non-terminal (``cut'') can takes two forms.
 A digit non-terminal can be used, enabling single-digit cut indexes.
@@
 
-An alternative to \ttt{<union>} within a sequence is the \ttt{<branch>} non-terminal ("branch").
+An alternative to \ttt{<union>} within a sequence is the \ttt{<branch>} non-terminal (``branch'').
 Like union, branch joins a parent and child node through a bond.
-Wrapped by opening and closing parenthesis terminals ("(" and ")", respectively), branch encodes a sequence that may or may not be attached to its parent.
+Wrapped by opening and closing parenthesis terminals (``('' and ``),'' respectively), branch encodes a sequence that may or may not be attached to its parent.
 Attachment occurs if the \ttt{<bond>} non-terminal is included.
@@
 
-The third option for associating a parent Atom with a child within a sequence is the \ttt{<split>} non-terminal ("split").
+The third option for associating a parent Atom with a child within a sequence is the \ttt{<split>} non-terminal (``split'').
 A split is a sequence of \ttt{<detachment>} and \ttt{<sequence>} non-terminals.
@@
 The presence of a branch adds some nuances over a union.
-The leading open parenthesis terminal ("(") signifies that the current root atom will later be re-exposed.
+The leading open parenthesis terminal (``('') signifies that the current root atom will later be re-exposed.
 This operation can be supported by a stack.
@@
 
-Readers must not assume that detachment (the period terminal, ".") implies the presence of disconnected components.
+Readers must not assume that detachment (the period terminal, ``.'') implies the presence of disconnected components.
 This assumption is most likely to arise in the context of ad-hoc parsers using regular expressions, string matching, and the like.
-For example, the single molecule propane can be encoded using the string "C1C.C1".
+For example, the single molecule propane can be encoded using the string ``C1C.C1''.
 
 The partial parity bond model is unusual in that it spreads conformational information over three or more bonds.
-Regardless, readers must report as an error any string containing an isolated PPB (e.g., "C/C").
+Regardless, readers must report as an error any string containing an isolated PPB (e.g., ``C/C'').
 Beyond this check, generating (\textit{E}) and (\textit{Z}) stereodescriptors from PPBs is not a straightforward process.
@@
 Neither terminal of a PPB possesses a double bond.
-Strings such as "C/C" and "C{\textbackslash}C" contain isolated PPBs, which are invalid.
+Strings such as ``C/C'' and ``C{\textbackslash}C'' contain isolated PPBs, which are invalid.
 A reader encountering such strings must report an error.}
@@
 \begin{enumerate}
-    \item{Impossible isotope. A negative implied mass number results from the atom. For example, "[2C]".}
-    \item{Impossible valence. The valence at an `Atom` is impossibly high. For example, "C(C)(C)(C)(C)C".}
-    \item{Impossible charge. An atom's charge gives it an apparent negative electron count. For example, "[C+7]".}
+    \item{Impossible isotope. A negative implied mass number results from the atom (e.g., ``[2C]'').}
+    \item{Impossible valence. The valence at an atom is impossibly high (e.g., ``C(C)(C)(C)(C)C'').}
+    \item{Impossible charge. An atom's charge gives it an apparent negative electron count (e.g., ``[C+7]'').}
 \end{enumerate}
@@
 And so on.
-If no default valences are found in this way (e.g., "[c+2]"), a reader must generate an error.
+If no default valences are found in this way (e.g., ``[c+2]''), a reader must generate an error.
 Writers must not encode such atoms.
@@
 This occurs whenever style, tradition, or convenience overrides necessity.
-Consider furan represented as the string "c1ccco1".
+Consider a string representing furan in which all atoms are selected (e.g., ``c1ccco1'').
 Selecting any atom is unnecessary because furan does not exhibit DIME.
@@
 It is nevertheless convenient to select the carbon atoms because all bonds can then be elided.
-The resulting representation, "c1cccO1" leads to a delocalization subgraph with a perfect matching and hydrogen counts consistent with the original encoding.
+The resulting representation (e.g., ``c1cccO1'') leads to a delocalization subgraph with a perfect matching and hydrogen counts consistent with the original encoding.
 
@@
 There are no requirements around style.
-For example, it's equally valid to represent the carbon atom of methane as either "C" or "[CH4]".
+For example, it's equally valid to represent the carbon atom of methane using either implicit or virtual hydrogens (e.g., ``C'' or ``[CH4]'').
 Single bonds may be elided or not.
@@
 A useful tool for this purpose is a \ttt{stack}.
-A stack is a data structure that allows items to be added individually and removed ("popped") in the reverse order of addition.
+A stack is a data structure that allows items to be added individually and removed (``popped'') in the reverse order of addition.
 A writer begins by pushing the current branch onto the stack and extending it.
@@
 Consider \textit{trans}-cyclooctene.
-The correct encoding accounts for the reverse in parity at the left-hand side (e.g., "C{\textbackslash}1C=C/CCCCC/1").
-Moreover, the double-encoding of the PPB bond type can lead to inconsistencies that must be reported by a reader (e.g., the invalid string "C/1C=C/CCCCC/1").
+The correct encoding accounts for the reverse in parity at the left-hand side (e.g., ``C{\textbackslash}1C=C/CCCCC/1'').
+Moreover, the double-encoding of the PPB bond type can lead to inconsistencies that must be reported by a reader (e.g., the invalid string ``C/1C=C/CCCCC/1'').
 Such errors can be avoided by encoding the bond type of one side or the other side of a cut, but not both.
-(e.g., "C{\textbackslash}1C=C/CCCCC/1" or "C1C=C/CCCCC/1").
+(e.g., ``C{\textbackslash}1C=C/CCCCC/1'' or ``C1C=C/CCCCC/1'').
 A cut across a bond attached to an atom with a stereodescriptor must take bond ordering into consideration.
-For example, the string "O[C@H]1NC1" encodes the same absolute configuration as the string "O[C@H](C1)N1", but the \textit{opposite} configuration as "O[C@H](N1)C1".
+For example, the two strings ``O[C@H]1NC1'' and ``O[C@H](C1)N1'' encode the same configuration, but the string ``O[C@H](N1)C1'' encodes the \textit{opposite} configuration.
 
@@
 Writers must ensure that all selected atoms can be deselected.
-Consider pyrrole, erroneously encoded as the string "n1cccc1".
+Consider pyrrole, erroneously encoded with a selected nitrogen atom (e.g., ``n1cccc1'').
 The nitrogen atom can not be pruned because its subvalence is 1 (3 - 2).
@@
 A reader receiving such a string must report an error.
-To avoid this outcome, a writer can consume the subvalence by adding a virtual hydrogen, as in the string "[nH]1cccc1".
+To avoid this outcome, a writer can consume the subvalence by adding a virtual hydrogen (e.g., ``[nH]1cccc1'').
 The subvalence of nitrogen in this case is zero, so the atom can be pruned.
@@
 Language subsets have ample precedent in computer science.
-A well-known example is MISRA C~\cite{misraC}, a subset of the C programming language that aims to eliminate "known undefined or otherwise dangerous behavior"~\cite{hatton:2007}.
-Another example is the subset of JavaScript described by Crockford whose purpose is to "chip away at the features that are not beautiful until the language's true nature reveals itself."~\cite{crockford:2008}.
+A well-known example is MISRA C~\cite{misraC}, a subset of the C programming language that aims to eliminate ``known undefined or otherwise dangerous behavior''~\cite{hatton:2007}.
+Another example is the subset of JavaScript described by Crockford whose purpose is to ``chip away at the features that are not beautiful until the language's true nature reveals itself.''~\cite{crockford:2008}.
 Language subsets arise for two main reasons.
@@
 Unfortunately, the choice of the Dialect base language is fraught with difficulty on technical, historical, and cultural grounds.
-Simply put, "SMILES" has come to mean different things to different people.
+Simply put, the term ``SMILES'' has come to mean different things to different people.
 To some, SMILES is the language defined in public documentation.
@@
 A protolanguage is a hypothetical language from which a family of languages can be derived.
-A suitable SMILES protolanguage would ideally contain all of the features consistent with the many ways in which the term "SMILES" is used today.
-
-Accordingly, Dialect's base language is "ProtoSMILES," a SMILES protolanguage.
+A suitable SMILES protolanguage would ideally contain all of the features consistent with the many ways in which the term ``SMILES'' is used today.
+
+Accordingly, Dialect's base language is ``ProtoSMILES,'' a SMILES protolanguage.
 ProtoSMILES was developed through a process of elimination that started with every SMILES document ever published and every SMILES software implementation, Open Source or not.
@@
 \begin{itemize}
-    \item Weininger's report to the US Environmental Protection Agency ("The Article")~\cite{anderson:1987}.
-    \item Weininger's original publication (herein "The Article")~\cite{weininger:1988}.
-    \item Weininger's 2003 book chapter, revised in 2008 ("The Book Chapter")~\cite{weininger:2008}.
-    \item An online manual maintained by Daylight ("The Manual").
-    \item The Daylight's toolkit, which implements a SMILES reader and writer ("The Toolkit")~\cite{daylightToolkit}.
+    \item Weininger's report to the US Environmental Protection Agency (``The Article'')~\cite{anderson:1987}.
+    \item Weininger's original publication (herein ``The Article'')~\cite{weininger:1988}.
+    \item Weininger's 2003 book chapter, revised in 2008 (``The Book Chapter'')~\cite{weininger:2008}.
+    \item An online manual maintained by Daylight (``The Manual'').
+    \item The Daylight's toolkit, which implements a SMILES reader and writer (``The Toolkit'')~\cite{daylightToolkit}.
 \end{itemize}
@@
 The report also supports unusual characters including the exclamation mark (\ttt{!}).
-Semantically, The Report supports "special states" for nitrogen-containing atoms.
+Semantically, The Report supports ``special states'' for nitrogen-containing atoms.
 
@@
 
-The Article implies that comma (\ttt{,}) is either an element symbol or a valid SMILES character: "This subset uses only the symbols H, C, N, O, P, S, F, Cl, Br, I, and (,) and digits, with the following four rules..."~\cite[p.~33]{weininger:1988}.
+The Article implies that comma (\ttt{,}) is either an element symbol or a valid SMILES character: ``This subset uses only the symbols H, C, N, O, P, S, F, Cl, Br, I, and (,) and digits, with the following four rules...''~\cite[p.~33]{weininger:1988}.
 The comma character is explicitly disallowed in Dialect.
@@
 The Book Chapter defines a recursive grammar for branching which, perhaps without the knowledge of the author, defines sequential parentheses~\cite[p.~86]{weininger:2008}.
-Dialect disallows such constructs, including for example "*((*))*" and "*(((*)))*".
-
-"Reaction SMILES" is an extension described in the Book Chapter~\cite[p.~89]{weininger:2008}, but disallowed in Dialect.
+Dialect disallows such constructs (e.g., ``*((*))*'' and ``*(((*)))*'').
+
+``Reaction SMILES'' is an extension described in the Book Chapter~\cite[p.~89]{weininger:2008}, but disallowed in Dialect.
 Specifically, the greater than symbol (\ttt{>}) is not a valid Dialect character.
-This precludes strings such as "*>>*" in Dialect.
-Furthermore, Dialect lacks the "map" attribute, used for atom-atom mapping in reactions, and its accompanying syntax.
-
-Dialect only supports two stereodescriptors, encoded with the terminals "@" and "@@".
-The Book Chapter provides a recursive grammar for stereodescriptors that allows multi-symbol descriptors such as "@@@" and "@@@@"~\cite[p.~94]{weininger:2008}.
-The book chapter also discusses non-tetrahedral descriptors including "@AL1", "@AL2", "@1", and "@SP1".
-None of these are supported by Dialect.
+This precludes strings such as ``*>>*'' in Dialect.
+Furthermore, Dialect lacks the ``map'' attribute, used for atom-atom mapping in reactions, and its accompanying syntax.
+
+Dialect only supports two stereodescriptors, encoded with the terminals ``@'' and ``@@.'' The Book Chapter provides a recursive grammar for stereodescriptors that allows multi-symbol descriptors such as ``@@@'' and ``@@@@''~\cite[p.~94]{weininger:2008}.
+The book chapter also discusses non-tetrahedral descriptors including ``@AL1,'' ``@AL2,'' ``@1,'' and ``@SP1.'' None of these are supported by Dialect.
 Nor does Dialect support the application of tetracoordinate stereodescriptors to odd cumulenes as described in the book chapter.
@@
 
-ProtoSMILES supports the concept of "aromatic bond" (":"), but Dialect does not.
+ProtoSMILES supports the concept of ``aromatic bond'' (``:''), but Dialect does not.
 Despite clear inclusion within the syntax of ProtoSMILES, no precise guidance on using this feature is provided.
-The Book Chapter notes "Adjacent atoms without an intervening bond symbol are connected by a valence-dictated bond (typically a single or aromatic bond).
-"-" (single) and ":" (aromatic) bond symbols may always be omitted on input"~\cite[p.~85]{weininger:2008}.
+The Book Chapter notes ``Adjacent atoms without an intervening bond symbol are connected by a valence-dictated bond (typically a single or aromatic bond).
+`-' (single) and `:' (aromatic) bond symbols may always be omitted on input''~\cite[p.~85]{weininger:2008}.
 This passage implies that aromatic bonds serve no purpose.
@@
 Doing so would increase the complexity of the language unnecessarily.
-Although The Paper implies selectability ("Elements other than C, N, O, P, S, As, and Se are not yet dealt with in an aromatic context.")~\cite[p.~35]{weininger:1988}, default valences for the latter two atoms are not provided.
+Although The Paper implies selectability (``Elements other than C, N, O, P, S, As, and Se are not yet dealt with in an aromatic context.'')~\cite[p.~35]{weininger:1988}, default valences for the latter two atoms are not provided.
 Even if they were, the elements As and Se must be bracketed.
 As such, special rules for handling them in the context of pruning would be required.
-Moreover, molecules bearing the elements As or Se are often made "aromatic" unnecessarily in SMILES.
+Moreover, molecules bearing the elements As or Se are often made ``aromatic'' unnecessarily in SMILES.
 Finally, allowing the selection of atoms bearing As or Se does nothing to address molecules with atoms bearing silicon, germanium, tin, or other heavy main group elements that may be discovered in the future to participate in DIME.
@@
 It's unclear whether ProtoSMILES allows a virtual hydrogen count on hydrogen itself.
-The Article implies that the element symbol H (hydrogen) must not have an implicit hydrogen count when used within a bracket sequence ("There are few exceptions to the hydrogen-suppression convention, the most obvious being specification of a proton, [H+], and molecular hydrogen, [H][H].")~\cite[p.~34]{weininger:1988}.
-If so, the author may have erroneously believed that parsers would be incapable of processing constructs such as "[HH]".
-The Book Chapter appears to reiterate this view, noting that "explicit hydrogen specification is required" in the case of "hydrogens connected to other hydrogens, e.g.
-[H][H], molecular hydrogen"~\cite[p.~97]{weininger:2008}.
-To be clear, Dialect supports any valid virtual hydrogen count on hydrogen itself.
-For example, the following are all valid, if chemically nonsensical, strings: "[HH]"; "[HH2]"; and "[HH3]".
-Depending on interpretation, this feature may or may not render Dialect a superset of SMILES.
+The Article implies that the element symbol H (hydrogen) must not have an implicit hydrogen count when used within a bracket sequence (``There are few exceptions to the hydrogen-suppression convention, the most obvious being specification of a proton, [H+], and molecular hydrogen, [H][H].'')~\cite[p.~34]{weininger:1988}.
+If so, the author may have erroneously believed that parsers would be incapable of processing strings encoding bracket hydrogen with a virtual hydrogen count (e.g., ``[HH]'').
+The Book Chapter appears to reiterate this view, noting that ``explicit hydrogen specification is required'' in the case of ``hydrogens connected to other hydrogens, e.g.
+[H][H], molecular hydrogen''~\cite[p.~97]{weininger:2008}.
+To be clear, Dialect supports any valid virtual hydrogen count on hydrogen itself (e.g., ``[HH]'' and ``[HH2]'').
+Subject to interpretation of the original literature, this feature may or may not render Dialect a superset of SMILES.
 
 ProtoSMILES ascribes ambiguous meaning to detachments.
-As noted in the Book Chapter, "...
-In terms of the valence model being represented, the dot literally represents a bond of formal order zero: the atoms on either side of the dot are explicitly not bonded to each other."~\cite[p.~88]{weininger:2008}.
-The state of not being bonded and the state of having a zero-order bond are not necessarily equivalent~\cite{clark:2011}.
-To be clear, Dialect explicitly disallows the zero bond order interpretation of detachments and the dot character.
+As noted in the Book Chapter, ``...
+In terms of the valence model being represented, the dot literally represents a bond of formal order zero: the atoms on either side of the dot are explicitly not bonded to each other.''~\cite[p.~88]{weininger:2008}.
+Other statements appear to contradict this statement.
+To be clear, the state of not being bonded and the state of having a zero-order bond are not necessarily equivalent~\cite{clark:2011}.
+Dialect explicitly disallows the zero bond order interpretation of detachments and the dot character.
 
@@
 Although it might be argued that lower bounds on physical quantities such as isotope and hydrogen count should be implicit, the lack of precision forces implementors to improvise, which can lead to divergent behavior.
-ProtoSMILES places no upper bound on the charge notation "+", "++", "+++", and their corresponding negative forms ("The SMILES program also recognizes constructions of the form [Fe+++] as being synonymous with the form [Fe+3].")~\cite[p.~32]{weininger:1988}.
-However, Dialect disallows duplicated plus and minus characters (\ttt{+} and \ttt{-}, respectively) from bracket atoms.
+ProtoSMILES places no upper bound on the charge notation (e.g., ``++'' and ``+++'') and their corresponding negative forms: ``The SMILES program also recognizes constructions of the form [Fe+++] as being synonymous with the form [Fe+3]''~\cite[p.~32]{weininger:1988}.
+Due to their redundancy, Dialect disallows duplicated plus and minus characters (\ttt{+} and \ttt{-}, respectively) from bracket atoms.
 
@@
 Dialect also provides detailed algorithms, absent in SMILES, for selection and deselection.
-These are based in graph theory rather than the ambiguous and overloaded chemical concept of "aromaticity" used in ProtoSMILES.
-Uniquely, Dialect introduces the concepts of "pruning" and "gratuitous selection." ProtoSMILES does not specify those bonds that can be promoted during deselection, but Dialect does.
+These are based in graph theory rather than the ambiguous and overloaded chemical concept of ``aromaticity'' used in ProtoSMILES.
+Uniquely, Dialect introduces the concepts of ``pruning'' and ``gratuitous selection.'' ProtoSMILES does not specify those bonds that can be promoted during deselection, but Dialect does.
 ProtoSMILES does not restrict partial parity bonds to those atoms also possessing at least one double bond, whereas Dialect does.
@@
 Dialect uses the latter default valences.
-Atoms whose \ttt{hydrogens} attribute may be set to \ttt{Implicit} ("the organic subset") are reported in the Article as: B; C; N; O; P; S; F; Cl; Br; and I~\cite[p.~32]{weininger:2008}.
-However, an example later in the Article indicates that "Hn1cccc1" with its unbracketed hydrogen is a valid string~\cite[p.~35]{weininger:1988}.
+Atoms whose \ttt{hydrogens} attribute may be set to \ttt{Implicit} (``the organic subset'') are reported in the Article as: B; C; N; O; P; S; F; Cl; Br; and I~\cite[p.~32]{weininger:2008}.
+However, an example later in the Article indicates that the string ``Hn1cccc1'' with its unbracketed hydrogen valid~\cite[p.~35]{weininger:1988}.
 Dialect explicitly rejects strings containing unbracketed hydrogen atoms.
-On the topic of atom selection ("aromaticity"), the Article implies that selected atoms must be part of a cycle: "To qualify as aromatic, all atoms in the ring must be sp2 hybridized and the number of available 'excess' $\pi$ electrons must satisfy Hückel's 4{\textit{N}} + 2 criterion."~\cite[p.~34]{weininger:1988}.
-The Book Chapter, however, presents two examples of valid strings that do not encode cycles ("cc" and "cccc")~\cite[p 85]{weininger:2008}.
+On the topic of atom selection (``aromaticity''), The Article implies that selected atoms must be part of a cycle: ``To qualify as aromatic, all atoms in the ring must be sp2 hybridized and the number of available 'excess' $\pi$ electrons must satisfy Hückel's 4{\textit{N}} + 2 criterion''~\cite[p.~34]{weininger:1988}.
+The Book Chapter, however, presents two examples of valid strings that do not encode cycles (``cc'' and ``cccc'')~\cite[p.~85]{weininger:2008}.
 Consistent with the latter view, Dialect explicitly allows the selection of atoms regardless of cycle membership.
@@
         \hline
-        Feature & ProtoSMILES\footnote{Values of "maybe" indicate ambiguous source descriptions.} & Dialect \\
+        Feature & ProtoSMILES\footnote{Values of ``maybe'' indicate ambiguous source descriptions.} & Dialect \\
         \hline
@@
         future element symbols approved by IUPAC & no & yes \\
-        comma symbol (\ttt{,}) & maybe & no \\
-        multiple branching such as "*((*))*" & yes & no \\
+        comma symbol (``,'') & maybe & no \\
+        multiple branching (e.g., ``*((*))*'') & yes & no \\
         reactions using greater than symbol (\ttt{>}) & yes & no \\
-        atomic "map" attribute & yes & no \\
-        extended stereodescriptors (e.g., "@@@", "@OH1", etc.) & yes & no \\
+        atomic ``map'' attribute & yes & no \\
+        extended stereodescriptors (e.g., ``@@@'' and ``@OH1'') & yes & no \\
         use of stereodescriptors on odd cumulene centers & yes & no \\
-        virtual hydrogen count on hydrogen (e.g., "[HH]") & maybe & yes \\
-        detachments are bonds of "formal order zero" & maybe & no \\
+        virtual hydrogen count on hydrogen (e.g., ``[HH]'') & maybe & yes \\
+        detachments are bonds of ``formal order zero'' & maybe & no \\
         upper and lower bounds on numerical atomic properties & no & yes \\
@@
         isolated partial parity bond & yes & no \\
-        repeated charge notation (e.g., "++", "+++", etc.) & yes & no \\
-        aromatic bond (":") & yes & no \\
+        repeated charge notation (e.g., ``++'' and ``+++'') & yes & no \\
+        aromatic bond (``:'') & yes & no \\
         \hline
@@
 With only one clear-cut exception, Table~\ref{table:feature-comparison} shows Dialect to be a subset of ProtoSMILES.
-The exceptional case deals with the validity certain transfermium element symbols already recognized by IUPAC and to be recognized in the future.
-Ambiguous exceptions include those for which the ProtoSMILES column feature support is uncertain ("maybe") but the Dialect column is affirmative ("yes").
-If ProtoSMILES support tends toward "no," then support by Dialect would break out of the ProtoSMILES subset.
+The exceptional case deals with the validity certain transfermium element symbols already recognized by IUPAC and to be recognized in the future by Dialect.
+Ambiguous exceptions include those for which the ProtoSMILES column feature support is uncertain (``maybe'') but the Dialect column is affirmative (``yes'').
+If ProtoSMILES support tends toward ``no,'' then support by Dialect would break out of the ProtoSMILES subset.
 There are three such cases.
-The first is dihydrogen represented using the string "[HH]".
+The first is hydrogen having virtual hydrogens (e.g., ``[HH]'').
 The second ambiguous case is whether or not the default valence of nitrogen includes the value five.
-The third case is whether or not ProtoSMILES recognizes acyclic "aromatic" atoms as valid.
+The third case is whether or not ProtoSMILES recognizes acyclic ``aromatic'' atoms as valid.
 
@@
 
-Despite its information density, Dialect can faithfully encode and decode most of what chemists would consider "organic molecules." This is evidenced by the widespread presence of Dialect strings within large, public-facing databases such as PubChem, ChEMBL, ChEBI, and eMolecules.
+Despite its information density, Dialect can faithfully encode and decode most of what chemists would consider ``organic molecules.'' This is evidenced by the widespread presence of Dialect strings within large, public-facing databases such as PubChem, ChEMBL, ChEBI, and eMolecules.
 

### 2022-03-25 manuscript.tex
Adds csquotes, replacing smart quotes with \enquote. refs #100

@@
 \usepackage{courier}
+\usepackage{csquotes}
+\usepackage[USenglish]{babel}
 
@@
 Despite its many applications, SMILES remains a language with an incomplete specification.
-Weininger's 1988 paper leaves the following either fully or incompletely specified: (1) stereochemical configuration; (2) double bond conformation; (3) the meaning handling of ``aromatic'' features; (4) the computation of implicit hydrogen count, especially as related to aromatic features; (5) formal syntax; (6) constraints on quantities such as mass number and charge; and (7) enumeration of error states.
+Weininger's 1988 paper leaves the following either fully or incompletely specified: (1) stereochemical configuration; (2) double bond conformation; (3) the meaning handling of \enquote{aromatic} features; (4) the computation of implicit hydrogen count, especially as related to aromatic features; (5) formal syntax; (6) constraints on quantities such as mass number and charge; and (7) enumeration of error states.
 
@@
 The Daylight Theory Manual, a website maintained by SMILES's corporate sponsor, Daylight Chemical Information Systems, Inc.
-(``Daylight''), recapitulates previously-published material~\cite{daylightTheory}.
+(\enquote{Daylight}), recapitulates previously-published material~\cite{daylightTheory}.
 The lack of other written sources describing SMILES is consistent with Weininger's reported disinterest in traditional scientific publication~\cite{gasteiger:2018}.
@@
 In 2007 a documentation effort that would become known as OpenSMILES began~\cite{openSMILES}.
-OpenSMILES was conceived as ``a non-proprietary specification for the SMILES language.'' Many of the points left open by previous SMILES documentation efforts were addressed.
-Noteworthy contributions included: the first formal grammar, which was later adapted~\cite{denjoed:2010}; many refinements around stereochemistry; and the introduction of ``standard form.'' Absent were detailed procedures for assigning and interpreting aromatic features, and a detailed procedure for computing implicit hydrogen count.
+OpenSMILES was conceived as \enquote{a non-proprietary specification for the SMILES language.} Many of the points left open by previous SMILES documentation efforts were addressed.
+Noteworthy contributions included: the first formal grammar, which was later adapted~\cite{denjoed:2010}; many refinements around stereochemistry; and the introduction of \enquote{standard form.} Absent were detailed procedures for assigning and interpreting aromatic features, and a detailed procedure for computing implicit hydrogen count.
 OpenSMILES also left several points of syntactic and semantic ambiguity unsettled, while introducing extensions of its own.
@@
 In 2019 IUPAC announced the SMILES+ initiative~\cite{smilesPlusSpecification}.
-Noting the limitations of existing SMILES documentation, the SMILES+ effort seeks to ``establish a formalized recommended up-to-date specification of the SMILES format.'' SMILES+ took as its starting point the documentation produced by the OpenSMILES project.
+Noting the limitations of existing SMILES documentation, the SMILES+ effort seeks to \enquote{establish a formalized recommended up-to-date specification of the SMILES format.} SMILES+ took as its starting point the documentation produced by the OpenSMILES project.
 Efforts to extend this starting point are in progress online through a public repository, but no formal recommendation has to date resulted~\cite{smilesPlusGitHub}.
@@
 The FAIR Guiding Principles identify four qualities essential for extracting maximum utility from published scholarly data: Findability; Accessibility; Interoperability; and Reusability~\cite{wilkinson:2016}.
-Interoperability in particular requires that ``(meta)data'' use a formal, accessible, shared, and broadly applicable language for knowledge representation.'' Although SMILES may be a shared, accessible, and broadly applicable language, its formal underpinnings leave room for improvement.
+Interoperability in particular requires that \enquote{(meta)data} use a formal, accessible, shared, and broadly applicable language for knowledge representation.'' Although SMILES may be a shared, accessible, and broadly applicable language, its formal underpinnings leave room for improvement.
 
@@
 Table \ref{table:atomic-state} summarizes atomic attributes.
-Not all combinations of values (``atomic state'') are valid.
+Not all combinations of values (\enquote{atomic state}) are valid.
 Implementations must ensure either the impossibility of an invalid state, or an error condition in the event that one is created.
@@
 When using this variant, the \ttt{hydrogens} attribute may assume integer values from zero to nine, inclusive.
-The process of transforming one or more hydrogen nodes into a virtual hydrogen is called ``virtualization.'' The opposite process, transforming one or more virtual hydrogens into a node, is called ``reification.''
+The process of transforming one or more hydrogen nodes into a virtual hydrogen is called \enquote{virtualization.} The opposite process, transforming one or more virtual hydrogens into a node, is called \enquote{reification.}
 
@@
 Electron delocalization results when the electrons of a valence bond representation can be distributed over more than one pair of atoms, as is the case with benzene and related molecules.
-DIME is similar to ``resonance'' or ``aromaticity,'' but those terms are avoided here due to their ambiguity~\cite{randic:2018,kerber:2006}.
+DIME is similar to \enquote{resonance} or \enquote{aromaticity,} but those terms are avoided here due to their ambiguity~\cite{randic:2018,kerber:2006}.
 
@@
 
-An atom is added to a DS by setting its \ttt{selected} attribute to \ttt{true} (``selection'').
+An atom is added to a DS by setting its \ttt{selected} attribute to \ttt{true} (\enquote{selection}).
 For an atom to be selectable, its \ttt{element} attribute must be contained in the list: \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; and \ttt{S}.
@@
 Because the presence of a filled DS implies a perfect matching over it, kekulization always succeeds.
-A popular algorithm for matching, the Edmunds ``blossom algorithm,'' has a time complexity of $O(|E||V|^2)$, where $|E|$ is the number of edges and $|V|$ is the number of nodes~\cite{edmonds:1965}.
+A popular algorithm for matching, the Edmunds \enquote{blossom algorithm,} has a time complexity of $O(|E||V|^2)$, where $|E|$ is the number of edges and $|V|$ is the number of nodes~\cite{edmonds:1965}.
 Although more efficient algorithms are known, they are either much more difficult to implement or lack generality.
@@
 For example, a selection algorithm can restrict candidate atoms to those found in cycles.
-Electron-counting techniques can also be introduced to approximate the chemical concept of ``aromaticity.'' A double bond between two selected atoms may or may not be elided, depending on the application.
+Electron-counting techniques can also be introduced to approximate the chemical concept of \enquote{aromaticity.} A double bond between two selected atoms may or may not be elided, depending on the application.
 
@@
 
-To support the assignment and interpretation stereochemical features, each node is assigned a sequential, unique, non-negative integer \ttt{index} attribute (``index'').
+To support the assignment and interpretation stereochemical features, each node is assigned a sequential, unique, non-negative integer \ttt{index} attribute (\enquote{index}).
 The value of an atom's index equals zero for a molecular's graph's first atom.
@@
 
-To support this system, each bond carries a nullable \ttt{direction} attribute (``direction'').
+To support this system, each bond carries a nullable \ttt{direction} attribute (\enquote{direction}).
 When present, this attribute may assume one of the two values \ttt{Up} or \ttt{Down}.
@@
 The double bond terminals are placed on the x-axis such that the terminal with the lowest \ttt{index} attribute appears to the left of the terminal with the higher atomic \ttt{index} attribute.
-These terminals are designated ``left terminal'' and ``right terminal,'' respectively.
+These terminals are designated \enquote{left terminal} and \enquote{right terminal,} respectively.
 Each neighbor of a terminal is then assigned a relative coordinate based on the \ttt{index} attribute and the direction of its PPB bond.
@@
 Dialect limits configuration to the special case of an atom with exactly four substituents placed at the vertexes of a tetrahedron.
-Here, ``substituent'' means an atomic neighbor; lone electron pairs are not considered substituents.
+Here, \enquote{substituent} means an atomic neighbor; lone electron pairs are not considered substituents.
 One of the four substituents may be a virtual hydrogen.
@@
 Dialect supports two configurational descriptors: \ttt{Left} and \ttt{Right}.
-The context for the terms ``left'' and ``right'' will be explained shortly.
+The context for the terms \enquote{left} and \enquote{right} will be explained shortly.
 
@@
 Not all tetracoordinate atoms will be \textit{stereocenters}.
-A stereocenter is an atom whose ``ligand permutation produces stereoisomers,'' as defined by Mislow and Siegel~\cite{mislow:1984}.
+A stereocenter is an atom whose \enquote{ligand permutation produces stereoisomers,} as defined by Mislow and Siegel~\cite{mislow:1984}.
 In the context of Dialect, a stereocenter must be tetracoordinate.
@@
 
-A Dialect string (``string'') is a UTF-encoded sequence of zero or more characters chosen from the set: \ttt{A}; \ttt{B}; \ttt{C}; \ttt{D}; \ttt{E}; \ttt{F}; \ttt{G}; \ttt{H}; \ttt{I}; \ttt{K}; \ttt{L}; \ttt{M}; \ttt{N}; \ttt{O}; \ttt{P}; \ttt{R}; \ttt{S}; \ttt{T}; \ttt{U}; \ttt{V}; \ttt{W}; \ttt{X}; \ttt{Y}; \ttt{Z}; \ttt{a}; \ttt{b}; \ttt{c}; \ttt{d}; \ttt{e}; \ttt{f}; \ttt{g}; \ttt{h}; \ttt{i}; \ttt{k}; \ttt{l}; \ttt{m}; \ttt{n}; \ttt{o}; \ttt{p}; \ttt{r}; \ttt{s}; \ttt{t}; \ttt{u}; \ttt{v}; \ttt{0}; \ttt{1}; \ttt{2}; \ttt{3}; \ttt{4}; \ttt{5}; \ttt{6}; \ttt{7}; \ttt{8}; \ttt{9}; \ttt{+}; \ttt{-}; \ttt{.}; \ttt{\%}; \ttt{@}; \ttt{*}; \ttt{[}; \ttt{]}; \ttt{/}; \ttt{\char`\\}; \ttt{=}; and \ttt{\#}.
+A Dialect string (\enquote{string}) is a UTF-encoded sequence of zero or more characters chosen from the set: \ttt{A}; \ttt{B}; \ttt{C}; \ttt{D}; \ttt{E}; \ttt{F}; \ttt{G}; \ttt{H}; \ttt{I}; \ttt{K}; \ttt{L}; \ttt{M}; \ttt{N}; \ttt{O}; \ttt{P}; \ttt{R}; \ttt{S}; \ttt{T}; \ttt{U}; \ttt{V}; \ttt{W}; \ttt{X}; \ttt{Y}; \ttt{Z}; \ttt{a}; \ttt{b}; \ttt{c}; \ttt{d}; \ttt{e}; \ttt{f}; \ttt{g}; \ttt{h}; \ttt{i}; \ttt{k}; \ttt{l}; \ttt{m}; \ttt{n}; \ttt{o}; \ttt{p}; \ttt{r}; \ttt{s}; \ttt{t}; \ttt{u}; \ttt{v}; \ttt{0}; \ttt{1}; \ttt{2}; \ttt{3}; \ttt{4}; \ttt{5}; \ttt{6}; \ttt{7}; \ttt{8}; \ttt{9}; \ttt{+}; \ttt{-}; \ttt{.}; \ttt{\%}; \ttt{@}; \ttt{*}; \ttt{[}; \ttt{]}; \ttt{/}; \ttt{\char`\\}; \ttt{=}; and \ttt{\#}.
 The internal structure of a string reflects a depth-first traversal of the corresponding molecular graphs.
@@
 
-Dialect's formal grammar is presented as a series of \textit{production rules} (aka ``productions'').
+Dialect's formal grammar is presented as a series of \textit{production rules} (aka \enquote{productions}).
 A production rule defines a transformation allowed under the grammar.
@@
 A production rule is composed of two kinds of elements: \textit{terminals} and \textit{non-terminals}.
-A terminal is a character literal (e.g., ``A'').
+A terminal is a character literal (e.g., \enquote{A}).
 A non-terminal is a reference to another production rule.
@@
 
-Consider a hypothetical language composed of variable-length sequences of the lowercase letter ``a''. Such a language could be cast as the following two production rules:
+Consider a hypothetical language composed of variable-length sequences of the lowercase letter \enquote{a}. Such a language could be cast as the following two production rules:
 
@@
 The first atomic non-terminal, \ttt{<star>}, can assume one terminal value, the asterisk character (\ttt{*}).
-This ``star atom'' represents an atom in which every attribute assumes its default value.
+This \enquote{star atom} represents an atom in which every attribute assumes its default value.
 The implicit hydrogen count of a star atom is zero.
@@
 
-The next atomic production rule, \ttt{<shortcut>} is a non-terminal selected from the list: ``B''; ``C''; ``N''; ``O''; ``P''; ``S''; ``F''; ``Cl''; ``Br''; and ``I.'' An atom encoded in this way (``shortcut atom'') assigns the corresponding symbol to the \ttt{element} attribute.
+The next atomic production rule, \ttt{<shortcut>} is a non-terminal selected from the list: \enquote{B}; \enquote{C}; \enquote{N}; \enquote{O}; \enquote{P}; \enquote{S}; \enquote{F}; \enquote{Cl}; \enquote{Br}; and \enquote{I.} An atom encoded in this way (\enquote{shortcut atom}) assigns the corresponding symbol to the \ttt{element} attribute.
 All other attributes retain their default values.
@@
 
-The third atomic production rule, \ttt{<selection>} is a non-terminal selected from the list: ``c''; ``n''; ``o''; ``p''; and ``s.'' An atom encoded in this way (``selected shortcut atom'') assigns the corresponding atom symbol to the \ttt{element} attribute and sets the \ttt{selected} attribute to `true`.
+The third atomic production rule, \ttt{<selection>} is a non-terminal selected from the list: \enquote{c}; \enquote{n}; \enquote{o}; \enquote{p}; and \enquote{s.} An atom encoded in this way (\enquote{selected shortcut atom}) assigns the corresponding atom symbol to the \ttt{element} attribute and sets the \ttt{selected} attribute to `true`.
 All other atomic attributes retain their default values.
@@
 
-The fourth and most complex atomic production rule is \ttt{<bracket>} (``bracket atom'').
+The fourth and most complex atomic production rule is \ttt{<bracket>} (\enquote{bracket atom}).
 A bracket atom can be used to set any atomic attribute.
@@
 Leading zeros are disregarded when assigning the value of the \ttt{isotope} attribute.
-In other words, ``007'' and ``7'' are considered equivalent expressions of the value 7.
+In other words, \enquote{007} and \enquote{7} are considered equivalent expressions of the value 7.
 
@@
 
-The values of a bracket atom's \ttt{element} and \ttt{selected} attributes are determined by the \ttt{<symbol>} production rule (``symbol'').
+The values of a bracket atom's \ttt{element} and \ttt{selected} attributes are determined by the \ttt{<symbol>} production rule (\enquote{symbol}).
 Three non-terminal options are available: \ttt{<star>}; \ttt{<element>}; and \ttt{<selection>}.
@@
 The \ttt{stereodescriptor} attribute of a bracket atom is determined by the \ttt{<stereodescriptor>} non-terminal.
-Allowed values are ``@'' and ``@@,'' corresponding to the \ttt{stereodescriptor} values \ttt{Counterclockwise} and \ttt{Clockwise}, respectively.
+Allowed values are \enquote{@} and \enquote{@@,} corresponding to the \ttt{stereodescriptor} values \ttt{Counterclockwise} and \ttt{Clockwise}, respectively.
 
@@
 The \ttt{hydrogens} attribute of a bracket atom is controlled by the \\ \ttt{<virtual{\_}hydrogen>} non-terminal.
-This non-terminal is comprised of the terminal ``H'' followed by an optional \ttt{<digit>} non-terminal.
-A digit appearing after the ``H'' terminal sets \ttt{hydrogens} to the corresponding virtual hydrogen count.
-A digit of 0 (i.e., ``H0'') sets the \ttt{hydrogens} attribute to zero.
+This non-terminal is comprised of the terminal \enquote{H} followed by an optional \ttt{<digit>} non-terminal.
+A digit appearing after the \enquote{H} terminal sets \ttt{hydrogens} to the corresponding virtual hydrogen count.
+A digit of 0 (i.e., \enquote{H0}) sets the \ttt{hydrogens} attribute to zero.
 If no digit is present, \ttt{hydrogens} is set to one.
@@
 The \ttt{<charge>} production rule sets the \ttt{charge} attribute of a bracket atom.
-This non-terminal begins with either the plus or minus terminals (``+'' and ``-,'' respectively) and ends with an optional \ttt{<digit>} non-terminal.
+This non-terminal begins with either the plus or minus terminals (\enquote{+} and \enquote{-,} respectively) and ends with an optional \ttt{<digit>} non-terminal.
 A missing digit causes the atomic \ttt{charge} attribute to be set to one.
-In other words, the plus terminal (``+'') sets the \ttt{charge} attribute to one and the minus terminal (``-'') sets it to minus one.
-A charge of zero (e.g., ``+0'' or ``-0'') sets the \ttt{charge} attribute to zero.
+In other words, the plus terminal (\enquote{+}) sets the \ttt{charge} attribute to one and the minus terminal (\enquote{-}) sets it to minus one.
+A charge of zero (e.g., \enquote{+0} or \enquote{-0}) sets the \ttt{charge} attribute to zero.
 
@@
 Should an optional bracket non-terminal not be present, the atom must retain its corresponding default value.
-For example, the bracket ``[C@H+]'' lacks the \ttt{<isotope>} non-terminal so the value of the \ttt{isotope} attribute will remain as \ttt{None}.
-Similarly, the bracket ``[13CH+]'' lacks the \ttt{<stereodescriptor>} non-terminal, so the corresponding \ttt{configuration} attribute remains \ttt{None}.
+For example, the bracket \enquote{[C@H+]} lacks the \ttt{<isotope>} non-terminal so the value of the \ttt{isotope} attribute will remain as \ttt{None}.
+Similarly, the bracket \enquote{[13CH+]} lacks the \ttt{<stereodescriptor>} non-terminal, so the corresponding \ttt{configuration} attribute remains \ttt{None}.
 
@@
 
-Five terminals are available (``-,'' ``=,'' ``\#,'' ``/,'' and ``\textbackslash'').
-The first three (``-,'' ``=,'' and ``\#'') set the \ttt{order} attribute of a bond to one, two, or three, respectively.
-The last two terminals, ``/'' and ``\textbackslash,'' set the \ttt{order} attribute to 1 while also setting the \ttt{direction} attribute to \ttt{Up} and \ttt{Down}, respectively.
+Five terminals are available (\enquote{-,} \enquote{=,} \enquote{\#,} \enquote{/,} and \enquote{\textbackslash}).
+The first three (\enquote{-,} \enquote{=,} and \enquote{\#}) set the \ttt{order} attribute of a bond to one, two, or three, respectively.
+The last two terminals, \enquote{/} and \enquote{\textbackslash,} set the \ttt{order} attribute to 1 while also setting the \ttt{direction} attribute to \ttt{Up} and \ttt{Down}, respectively.
 
@@
 
-An atom may be assigned zero or more children through the \ttt{<sequence>} non-terminal (``sequence'').
+An atom may be assigned zero or more children through the \ttt{<sequence>} non-terminal (\enquote{sequence}).
 A sequence starts with a required \ttt{<atom>} non-terminal.
@@
 
-The \ttt{<cut>} non-terminal (``cut'') can takes two forms.
+The \ttt{<cut>} non-terminal (\enquote{cut}) can takes two forms.
 A digit non-terminal can be used, enabling single-digit cut indexes.
@@
 
-An alternative to \ttt{<union>} within a sequence is the \ttt{<branch>} non-terminal (``branch'').
+An alternative to \ttt{<union>} within a sequence is the \ttt{<branch>} non-terminal (\enquote{branch}).
 Like union, branch joins a parent and child node through a bond.
-Wrapped by opening and closing parenthesis terminals (``('' and ``),'' respectively), branch encodes a sequence that may or may not be attached to its parent.
+Wrapped by opening and closing parenthesis terminals (\enquote{(} and \enquote{),} respectively), branch encodes a sequence that may or may not be attached to its parent.
 Attachment occurs if the \ttt{<bond>} non-terminal is included.
@@
 
-The third option for associating a parent Atom with a child within a sequence is the \ttt{<split>} non-terminal (``split'').
+The third option for associating a parent Atom with a child within a sequence is the \ttt{<split>} non-terminal (\enquote{split}).
 A split is a sequence of \ttt{<detachment>} and \ttt{<sequence>} non-terminals.
@@
 The presence of a branch adds some nuances over a union.
-The leading open parenthesis terminal (``('') signifies that the current root atom will later be re-exposed.
+The leading open parenthesis terminal (\enquote{(}) signifies that the current root atom will later be re-exposed.
 This operation can be supported by a stack.
@@
 
-Readers must not assume that detachment (the period terminal, ``.'') implies the presence of disconnected components.
+Readers must not assume that detachment (the period terminal, \enquote{.}) implies the presence of disconnected components.
 This assumption is most likely to arise in the context of ad-hoc parsers using regular expressions, string matching, and the like.
-For example, the single molecule propane can be encoded using the string ``C1C.C1''.
+For example, the single molecule propane can be encoded using the string \enquote{C1C.C1}.
 
 The partial parity bond model is unusual in that it spreads conformational information over three or more bonds.
-Regardless, readers must report as an error any string containing an isolated PPB (e.g., ``C/C'').
+Regardless, readers must report as an error any string containing an isolated PPB (e.g., \enquote{C/C}).
 Beyond this check, generating (\textit{E}) and (\textit{Z}) stereodescriptors from PPBs is not a straightforward process.
@@
 Neither terminal of a PPB possesses a double bond.
-Strings such as ``C/C'' and ``C{\textbackslash}C'' contain isolated PPBs, which are invalid.
+Strings such as \enquote{C/C} and \enquote{C{\textbackslash}C} contain isolated PPBs, which are invalid.
 A reader encountering such strings must report an error.}
@@
 \begin{enumerate}
-    \item{Impossible isotope. A negative implied mass number results from the atom (e.g., ``[2C]'').}
-    \item{Impossible valence. The valence at an atom is impossibly high (e.g., ``C(C)(C)(C)(C)C'').}
-    \item{Impossible charge. An atom's charge gives it an apparent negative electron count (e.g., ``[C+7]'').}
+    \item{Impossible isotope. A negative implied mass number results from the atom (e.g., \enquote{[2C]}).}
+    \item{Impossible valence. The valence at an atom is impossibly high (e.g., \enquote{C(C)(C)(C)(C)C}).}
+    \item{Impossible charge. An atom's charge gives it an apparent negative electron count (e.g., \enquote{[C+7]}).}
 \end{enumerate}
@@
 And so on.
-If no default valences are found in this way (e.g., ``[c+2]''), a reader must generate an error.
+If no default valences are found in this way (e.g., \enquote{[c+2]}), a reader must generate an error.
 Writers must not encode such atoms.
@@
 This occurs whenever style, tradition, or convenience overrides necessity.
-Consider a string representing furan in which all atoms are selected (e.g., ``c1ccco1'').
+Consider a string representing furan in which all atoms are selected (e.g., \enquote{c1ccco1}).
 Selecting any atom is unnecessary because furan does not exhibit DIME.
@@
 It is nevertheless convenient to select the carbon atoms because all bonds can then be elided.
-The resulting representation (e.g., ``c1cccO1'') leads to a delocalization subgraph with a perfect matching and hydrogen counts consistent with the original encoding.
+The resulting representation (e.g., \enquote{c1cccO1}) leads to a delocalization subgraph with a perfect matching and hydrogen counts consistent with the original encoding.
 
@@
 There are no requirements around style.
-For example, it's equally valid to represent the carbon atom of methane using either implicit or virtual hydrogens (e.g., ``C'' or ``[CH4]'').
+For example, it's equally valid to represent the carbon atom of methane using either implicit or virtual hydrogens (e.g., \enquote{C} or \enquote{[CH4]}).
 Single bonds may be elided or not.
@@
 A useful tool for this purpose is a \ttt{stack}.
-A stack is a data structure that allows items to be added individually and removed (``popped'') in the reverse order of addition.
+A stack is a data structure that allows items to be added individually and removed (\enquote{popped}) in the reverse order of addition.
 A writer begins by pushing the current branch onto the stack and extending it.
@@
 Consider \textit{trans}-cyclooctene.
-The correct encoding accounts for the reverse in parity at the left-hand side (e.g., ``C{\textbackslash}1C=C/CCCCC/1'').
-Moreover, the double-encoding of the PPB bond type can lead to inconsistencies that must be reported by a reader (e.g., the invalid string ``C/1C=C/CCCCC/1'').
+The correct encoding accounts for the reverse in parity at the left-hand side (e.g., \enquote{C{\textbackslash}1C=C/CCCCC/1}).
+Moreover, the double-encoding of the PPB bond type can lead to inconsistencies that must be reported by a reader (e.g., the invalid string \enquote{C/1C=C/CCCCC/1}).
 Such errors can be avoided by encoding the bond type of one side or the other side of a cut, but not both.
-(e.g., ``C{\textbackslash}1C=C/CCCCC/1'' or ``C1C=C/CCCCC/1'').
+(e.g., \enquote{C{\textbackslash}1C=C/CCCCC/1} or \enquote{C1C=C/CCCCC/1}).
 A cut across a bond attached to an atom with a stereodescriptor must take bond ordering into consideration.
-For example, the two strings ``O[C@H]1NC1'' and ``O[C@H](C1)N1'' encode the same configuration, but the string ``O[C@H](N1)C1'' encodes the \textit{opposite} configuration.
+For example, the two strings \enquote{O[C@H]1NC1} and \enquote{O[C@H](C1)N1} encode the same configuration, but the string \enquote{O[C@H](N1)C1} encodes the \textit{opposite} configuration.
 
@@
 Writers must ensure that all selected atoms can be deselected.
-Consider pyrrole, erroneously encoded with a selected nitrogen atom (e.g., ``n1cccc1'').
+Consider pyrrole, erroneously encoded with a selected nitrogen atom (e.g., \enquote{n1cccc1}).
 The nitrogen atom can not be pruned because its subvalence is 1 (3 - 2).
@@
 A reader receiving such a string must report an error.
-To avoid this outcome, a writer can consume the subvalence by adding a virtual hydrogen (e.g., ``[nH]1cccc1'').
+To avoid this outcome, a writer can consume the subvalence by adding a virtual hydrogen (e.g., \enquote{[nH]1cccc1}).
 The subvalence of nitrogen in this case is zero, so the atom can be pruned.
@@
 Language subsets have ample precedent in computer science.
-A well-known example is MISRA C~\cite{misraC}, a subset of the C programming language that aims to eliminate ``known undefined or otherwise dangerous behavior''~\cite{hatton:2007}.
-Another example is the subset of JavaScript described by Crockford whose purpose is to ``chip away at the features that are not beautiful until the language's true nature reveals itself.''~\cite{crockford:2008}.
+A well-known example is MISRA C~\cite{misraC}, a subset of the C programming language that aims to eliminate \enquote{known undefined or otherwise dangerous behavior}~\cite{hatton:2007}.
+Another example is the subset of JavaScript described by Crockford whose purpose is to \enquote{chip away at the features that are not beautiful until the language's true nature reveals itself.}~\cite{crockford:2008}.
 Language subsets arise for two main reasons.
@@
 Unfortunately, the choice of the Dialect base language is fraught with difficulty on technical, historical, and cultural grounds.
-Simply put, the term ``SMILES'' has come to mean different things to different people.
+Simply put, the term \enquote{SMILES} has come to mean different things to different people.
 To some, SMILES is the language defined in public documentation.
@@
 A protolanguage is a hypothetical language from which a family of languages can be derived.
-A suitable SMILES protolanguage would ideally contain all of the features consistent with the many ways in which the term ``SMILES'' is used today.
-
-Accordingly, Dialect's base language is ``ProtoSMILES,'' a SMILES protolanguage.
+A suitable SMILES protolanguage would ideally contain all of the features consistent with the many ways in which the term \enquote{SMILES} is used today.
+
+Accordingly, Dialect's base language is \enquote{ProtoSMILES,} a SMILES protolanguage.
 ProtoSMILES was developed through a process of elimination that started with every SMILES document ever published and every SMILES software implementation, Open Source or not.
@@
 \begin{itemize}
-    \item Weininger's report to the US Environmental Protection Agency (``The Article'')~\cite{anderson:1987}.
-    \item Weininger's original publication (herein ``The Article'')~\cite{weininger:1988}.
-    \item Weininger's 2003 book chapter, revised in 2008 (``The Book Chapter'')~\cite{weininger:2008}.
-    \item An online manual maintained by Daylight (``The Manual'').
-    \item The Daylight's toolkit, which implements a SMILES reader and writer (``The Toolkit'')~\cite{daylightToolkit}.
+    \item Weininger's report to the US Environmental Protection Agency (\enquote{The Article})~\cite{anderson:1987}.
+    \item Weininger's original publication (herein \enquote{The Article})~\cite{weininger:1988}.
+    \item Weininger's 2003 book chapter, revised in 2008 (\enquote{The Book Chapter})~\cite{weininger:2008}.
+    \item An online manual maintained by Daylight (\enquote{The Manual}).
+    \item The Daylight's toolkit, which implements a SMILES reader and writer (\enquote{The Toolkit})~\cite{daylightToolkit}.
 \end{itemize}
@@
 The report also supports unusual characters including the exclamation mark (\ttt{!}).
-Semantically, The Report supports ``special states'' for nitrogen-containing atoms.
+Semantically, The Report supports \enquote{special states} for nitrogen-containing atoms.
 
@@
 
-The Article implies that comma (\ttt{,}) is either an element symbol or a valid SMILES character: ``This subset uses only the symbols H, C, N, O, P, S, F, Cl, Br, I, and (,) and digits, with the following four rules...''~\cite[p.~33]{weininger:1988}.
+The Article implies that comma (\ttt{,}) is either an element symbol or a valid SMILES character: \enquote{This subset uses only the symbols H, C, N, O, P, S, F, Cl, Br, I, and (,) and digits, with the following four rules...}~\cite[p.~33]{weininger:1988}.
 The comma character is explicitly disallowed in Dialect.
@@
 The Book Chapter defines a recursive grammar for branching which, perhaps without the knowledge of the author, defines sequential parentheses~\cite[p.~86]{weininger:2008}.
-Dialect disallows such constructs (e.g., ``*((*))*'' and ``*(((*)))*'').
-
-``Reaction SMILES'' is an extension described in the Book Chapter~\cite[p.~89]{weininger:2008}, but disallowed in Dialect.
-Specifically, the greater than symbol (\ttt{>}) is not a valid Dialect character.
-This precludes strings such as ``*>>*'' in Dialect.
-Furthermore, Dialect lacks the ``map'' attribute, used for atom-atom mapping in reactions, and its accompanying syntax.
-
-Dialect only supports two stereodescriptors, encoded with the terminals ``@'' and ``@@.'' The Book Chapter provides a recursive grammar for stereodescriptors that allows multi-symbol descriptors such as ``@@@'' and ``@@@@''~\cite[p.~94]{weininger:2008}.
-The book chapter also discusses non-tetrahedral descriptors including ``@AL1,'' ``@AL2,'' ``@1,'' and ``@SP1.'' None of these are supported by Dialect.
+Dialect disallows such constructs (e.g., \enquote{*((*))*} and \enquote{*(((*)))*}).
+
+\enquote{Reaction SMILES} is an extension described in the Book Chapter~\cite[p.~89]{weininger:2008}, but disallowed in Dialect.
+Specifically, the greater than symbol (\ttt{$>$}) is not a valid Dialect character.
+This precludes strings such as \enquote{*$>$$>$*} in Dialect.
+Furthermore, Dialect lacks the \enquote{map} attribute, used for atom-atom mapping in reactions, and its accompanying syntax.
+
+Dialect only supports two stereodescriptors, encoded with the terminals \enquote{@} and \enquote{@@.} The Book Chapter provides a recursive grammar for stereodescriptors that allows multi-symbol descriptors such as \enquote{@@@} and \enquote{@@@@}~\cite[p.~94]{weininger:2008}.
+The book chapter also discusses non-tetrahedral descriptors including \enquote{@AL1,} \enquote{@AL2,} \enquote{@1,} and \enquote{@SP1.} None of these are supported by Dialect.
 Nor does Dialect support the application of tetracoordinate stereodescriptors to odd cumulenes as described in the book chapter.
@@
 
-ProtoSMILES supports the concept of ``aromatic bond'' (``:''), but Dialect does not.
+ProtoSMILES supports the concept of \enquote{aromatic bond} (\enquote{:}), but Dialect does not.
 Despite clear inclusion within the syntax of ProtoSMILES, no precise guidance on using this feature is provided.
-The Book Chapter notes ``Adjacent atoms without an intervening bond symbol are connected by a valence-dictated bond (typically a single or aromatic bond).
-`-' (single) and `:' (aromatic) bond symbols may always be omitted on input''~\cite[p.~85]{weininger:2008}.
+The Book Chapter notes \enquote{Adjacent atoms without an intervening bond symbol are connected by a valence-dictated bond (typically a single or aromatic bond) \enquote{-} (single) and \enquote{:} (aromatic) bond symbols may always be omitted on input}~\cite[p.~85]{weininger:2008}.
 This passage implies that aromatic bonds serve no purpose.
@@
 Doing so would increase the complexity of the language unnecessarily.
-Although The Paper implies selectability (``Elements other than C, N, O, P, S, As, and Se are not yet dealt with in an aromatic context.'')~\cite[p.~35]{weininger:1988}, default valences for the latter two atoms are not provided.
+Although The Paper implies selectability (\enquote{Elements other than C, N, O, P, S, As, and Se are not yet dealt with in an aromatic context.})~\cite[p.~35]{weininger:1988}, default valences for the latter two atoms are not provided.
 Even if they were, the elements As and Se must be bracketed.
 As such, special rules for handling them in the context of pruning would be required.
-Moreover, molecules bearing the elements As or Se are often made ``aromatic'' unnecessarily in SMILES.
+Moreover, molecules bearing the elements As or Se are often made \enquote{aromatic} unnecessarily in SMILES.
 Finally, allowing the selection of atoms bearing As or Se does nothing to address molecules with atoms bearing silicon, germanium, tin, or other heavy main group elements that may be discovered in the future to participate in DIME.
@@
 It's unclear whether ProtoSMILES allows a virtual hydrogen count on hydrogen itself.
-The Article implies that the element symbol H (hydrogen) must not have an implicit hydrogen count when used within a bracket sequence (``There are few exceptions to the hydrogen-suppression convention, the most obvious being specification of a proton, [H+], and molecular hydrogen, [H][H].'')~\cite[p.~34]{weininger:1988}.
-If so, the author may have erroneously believed that parsers would be incapable of processing strings encoding bracket hydrogen with a virtual hydrogen count (e.g., ``[HH]'').
-The Book Chapter appears to reiterate this view, noting that ``explicit hydrogen specification is required'' in the case of ``hydrogens connected to other hydrogens, e.g.
-[H][H], molecular hydrogen''~\cite[p.~97]{weininger:2008}.
-To be clear, Dialect supports any valid virtual hydrogen count on hydrogen itself (e.g., ``[HH]'' and ``[HH2]'').
+The Article implies that the element symbol H (hydrogen) must not have an implicit hydrogen count when used within a bracket sequence (\enquote{There are few exceptions to the hydrogen-suppression convention, the most obvious being specification of a proton, [H+], and molecular hydrogen, [H][H].})~\cite[p.~34]{weininger:1988}.
+If so, the author may have erroneously believed that parsers would be incapable of processing strings encoding bracket hydrogen with a virtual hydrogen count (e.g., \enquote{[HH]}).
+The Book Chapter appears to reiterate this view, noting that \enquote{explicit hydrogen specification is required} in the case of \enquote{hydrogens connected to other hydrogens, e.g.
+[H][H], molecular hydrogen}~\cite[p.~97]{weininger:2008}.
+To be clear, Dialect supports any valid virtual hydrogen count on hydrogen itself (e.g., \enquote{[HH]} and \enquote{[HH2]}).
 Subject to interpretation of the original literature, this feature may or may not render Dialect a superset of SMILES.
@@
 ProtoSMILES ascribes ambiguous meaning to detachments.
-As noted in the Book Chapter, ``...
-In terms of the valence model being represented, the dot literally represents a bond of formal order zero: the atoms on either side of the dot are explicitly not bonded to each other.''~\cite[p.~88]{weininger:2008}.
+As noted in the Book Chapter, \enquote{...
+In terms of the valence model being represented, the dot literally represents a bond of formal order zero: the atoms on either side of the dot are explicitly not bonded to each other.}~\cite[p.~88]{weininger:2008}.
 Other statements appear to contradict this statement.
@@
 Although it might be argued that lower bounds on physical quantities such as isotope and hydrogen count should be implicit, the lack of precision forces implementors to improvise, which can lead to divergent behavior.
-ProtoSMILES places no upper bound on the charge notation (e.g., ``++'' and ``+++'') and their corresponding negative forms: ``The SMILES program also recognizes constructions of the form [Fe+++] as being synonymous with the form [Fe+3]''~\cite[p.~32]{weininger:1988}.
+ProtoSMILES places no upper bound on the charge notation (e.g., \enquote{++} and \enquote{+++}) and their corresponding negative forms: \enquote{The SMILES program also recognizes constructions of the form [Fe+++] as being synonymous with the form [Fe+3]}~\cite[p.~32]{weininger:1988}.
 Due to their redundancy, Dialect disallows duplicated plus and minus characters (\ttt{+} and \ttt{-}, respectively) from bracket atoms.
@@
 Dialect also provides detailed algorithms, absent in SMILES, for selection and deselection.
-These are based in graph theory rather than the ambiguous and overloaded chemical concept of ``aromaticity'' used in ProtoSMILES.
-Uniquely, Dialect introduces the concepts of ``pruning'' and ``gratuitous selection.'' ProtoSMILES does not specify those bonds that can be promoted during deselection, but Dialect does.
+These are based in graph theory rather than the ambiguous and overloaded chemical concept of \enquote{aromaticity} used in ProtoSMILES.
+Uniquely, Dialect introduces the concepts of \enquote{pruning} and \enquote{gratuitous selection.} ProtoSMILES does not specify those bonds that can be promoted during deselection, but Dialect does.
 ProtoSMILES does not restrict partial parity bonds to those atoms also possessing at least one double bond, whereas Dialect does.
@@
 Dialect uses the latter default valences.
-Atoms whose \ttt{hydrogens} attribute may be set to \ttt{Implicit} (``the organic subset'') are reported in the Article as: B; C; N; O; P; S; F; Cl; Br; and I~\cite[p.~32]{weininger:2008}.
-However, an example later in the Article indicates that the string ``Hn1cccc1'' with its unbracketed hydrogen valid~\cite[p.~35]{weininger:1988}.
+Atoms whose \ttt{hydrogens} attribute may be set to \ttt{Implicit} (\enquote{the organic subset}) are reported in the Article as: B; C; N; O; P; S; F; Cl; Br; and I~\cite[p.~32]{weininger:2008}.
+However, an example later in the Article indicates that the string \enquote{Hn1cccc1} with its unbracketed hydrogen valid~\cite[p.~35]{weininger:1988}.
 Dialect explicitly rejects strings containing unbracketed hydrogen atoms.
-On the topic of atom selection (``aromaticity''), The Article implies that selected atoms must be part of a cycle: ``To qualify as aromatic, all atoms in the ring must be sp2 hybridized and the number of available 'excess' $\pi$ electrons must satisfy Hückel's 4{\textit{N}} + 2 criterion''~\cite[p.~34]{weininger:1988}.
-The Book Chapter, however, presents two examples of valid strings that do not encode cycles (``cc'' and ``cccc'')~\cite[p.~85]{weininger:2008}.
+On the topic of atom selection (\enquote{aromaticity}), The Article implies that selected atoms must be part of a cycle: \enquote{To qualify as aromatic, all atoms in the ring must be sp2 hybridized and the number of available 'excess' $\pi$ electrons must satisfy Hückel's 4{\textit{N}} + 2 criterion}~\cite[p.~34]{weininger:1988}.
+The Book Chapter, however, presents two examples of valid strings that do not encode cycles (\enquote{cc} and \enquote{cccc})~\cite[p.~85]{weininger:2008}.
 Consistent with the latter view, Dialect explicitly allows the selection of atoms regardless of cycle membership.
@@
         \hline
-        Feature & ProtoSMILES\footnote{Values of ``maybe'' indicate ambiguous source descriptions.} & Dialect \\
+        Feature & ProtoSMILES\footnote{Values of \enquote{maybe} indicate ambiguous source descriptions.} & Dialect \\
         \hline
@@
         future element symbols approved by IUPAC & no & yes \\
-        comma symbol (``,'') & maybe & no \\
-        multiple branching (e.g., ``*((*))*'') & yes & no \\
+        comma symbol (\enquote{,}) & maybe & no \\
+        multiple branching (e.g., \enquote{*((*))*}) & yes & no \\
         reactions using greater than symbol (\ttt{>}) & yes & no \\
-        atomic ``map'' attribute & yes & no \\
-        extended stereodescriptors (e.g., ``@@@'' and ``@OH1'') & yes & no \\
+        atomic \enquote{map} attribute & yes & no \\
+        extended stereodescriptors (e.g., \enquote{@@@} and \enquote{@OH1}) & yes & no \\
         use of stereodescriptors on odd cumulene centers & yes & no \\
-        virtual hydrogen count on hydrogen (e.g., ``[HH]'') & maybe & yes \\
-        detachments are bonds of ``formal order zero'' & maybe & no \\
+        virtual hydrogen count on hydrogen (e.g., \enquote{[HH]}) & maybe & yes \\
+        detachments are bonds of \enquote{formal order zero} & maybe & no \\
         upper and lower bounds on numerical atomic properties & no & yes \\
@@
         isolated partial parity bond & yes & no \\
-        repeated charge notation (e.g., ``++'' and ``+++'') & yes & no \\
-        aromatic bond (``:'') & yes & no \\
+        repeated charge notation (e.g., \enquote{++} and \enquote{+++}) & yes & no \\
+        aromatic bond (\enquote{:}) & yes & no \\
         \hline
@@
 The exceptional case deals with the validity certain transfermium element symbols already recognized by IUPAC and to be recognized in the future by Dialect.
-Ambiguous exceptions include those for which the ProtoSMILES column feature support is uncertain (``maybe'') but the Dialect column is affirmative (``yes'').
-If ProtoSMILES support tends toward ``no,'' then support by Dialect would break out of the ProtoSMILES subset.
+Ambiguous exceptions include those for which the ProtoSMILES column feature support is uncertain (\enquote{maybe}) but the Dialect column is affirmative (\enquote{yes}).
+If ProtoSMILES support tends toward \enquote{no,} then support by Dialect would break out of the ProtoSMILES subset.
 There are three such cases.
-The first is hydrogen having virtual hydrogens (e.g., ``[HH]'').
+The first is hydrogen having virtual hydrogens (e.g., \enquote{[HH]}).
 The second ambiguous case is whether or not the default valence of nitrogen includes the value five.
-The third case is whether or not ProtoSMILES recognizes acyclic ``aromatic'' atoms as valid.
+The third case is whether or not ProtoSMILES recognizes acyclic \enquote{aromatic} atoms as valid.
 
@@
 
-Despite its information density, Dialect can faithfully encode and decode most of what chemists would consider ``organic molecules.'' This is evidenced by the widespread presence of Dialect strings within large, public-facing databases such as PubChem, ChEMBL, ChEBI, and eMolecules.
+Despite its information density, Dialect can faithfully encode and decode most of what chemists would consider \enquote{organic molecules.} This is evidenced by the widespread presence of Dialect strings within large, public-facing databases such as PubChem, ChEMBL, ChEBI, and eMolecules.
 

### 2022-03-25 manuscript.tex
Improves flow of Introduction.

@@
 As a line notation~\cite{wiswesser:1968}, SMILES represents molecules as single line character sequences, or strings.
-SMILES strings can be read and written algorithmically~\cite{weininger:1989,oBoyle:2012}, and can be canonicalized to yield unique molecular identifiers.
-
 SMILES has since been widely adopted.
@@
 Raw SMILES strings have been used extensively in machine learning applications~\cite{sousa:2021}.
+Algorithmic selection of a single SMILES encoding has been used for unique molecular identification~\cite{weininger:1989,oBoyle:2012}.
 SMILES has also been extended to carry various forms of metadata, as exemplified by Jmol SMILES~\cite{hanson:2016}, CurleySMILES~\cite{drefahl:2011}, BigSMILES~\cite{lin:2019}, and CXSMILES~\cite{cxsmiles}.
 
-Despite its many applications, SMILES remains a language with an incomplete specification.
-Weininger's 1988 paper leaves the following either fully or incompletely specified: (1) stereochemical configuration; (2) double bond conformation; (3) the meaning handling of \enquote{aromatic} features; (4) the computation of implicit hydrogen count, especially as related to aromatic features; (5) formal syntax; (6) constraints on quantities such as mass number and charge; and (7) enumeration of error states.
-
-Earlier descriptions of SMILES were published in various media, although these are not cited by Weininger.
-A currently-accessible example is a 1987 report to the US Environmental Protection Agency~\cite{anderson:1987}.
-Although the language described there shares many features with later descriptions, there are important differences as well.
+Despite the central role that SMILES plays in cheminformatics, public descriptions of the language remain imprecise at best.
+A 1987 report to the US Environmental Protection Agency by Weininger described a language similar to SMILES, but with important differences~\cite{anderson:1987}.
+Weininger's 1988 paper offered an overview of the language, but omitted or only partially addressed points crucial to correctly implementing SMILES, including: (1) stereochemical configuration; (2) double bond conformation; (3) the encoding and decoding of \enquote{aromatic} features; (4) the computation of implicit hydrogen count, especially in the context of aromatic features; (5) formal syntax; (6) constraints on quantities such as mass number and charge; and (7) enumeration of error states.
 
 Since 1988, two authoritative sources have attempted to clarify or revise Weininger's 1988 description of SMILES.
-A 2003 book chapter by Weininger addressed some of the limitations of the paper~\cite{weininger:2008}.
+A 2003 book chapter by Weininger addressed some of the limitations of the paper~\cite{weininger:2003}.
 The Daylight Theory Manual, a website maintained by SMILES's corporate sponsor, Daylight Chemical Information Systems, Inc.
 (\enquote{Daylight}), recapitulates previously-published material~\cite{daylightTheory}.
-The lack of other written sources describing SMILES is consistent with Weininger's reported disinterest in traditional scientific publication~\cite{gasteiger:2018}.
+The lack of other authoritative sources describing SMILES is consistent with Weininger's reported disinterest in traditional scientific publication~\cite{gasteiger:2018}.
 
@@
 OpenSMILES was conceived as \enquote{a non-proprietary specification for the SMILES language.} Many of the points left open by previous SMILES documentation efforts were addressed.
-Noteworthy contributions included: the first formal grammar, which was later adapted~\cite{denjoed:2010}; many refinements around stereochemistry; and the introduction of \enquote{standard form.} Absent were detailed procedures for assigning and interpreting aromatic features, and a detailed procedure for computing implicit hydrogen count.
+Noteworthy contributions included: the first formal grammar, later adapted to a standard parser toolchain~\cite{denjoed:2010}; many refinements around stereochemistry; and the introduction of \enquote{standard form.} Absent were detailed procedures for assigning and interpreting aromatic features, and a detailed procedure for computing implicit hydrogen count.
 OpenSMILES also left several points of syntactic and semantic ambiguity unsettled, while introducing extensions of its own.
@@
 
-The ongoing lack of a comprehensive SMILES documentation suite has caused several problems.
-First, authors of new SMILES implementations have limited guidance when resolving ambiguities and so must invent them on an ad hoc basis.
-Second, maintainers of existing SMILES implementations lack a blueprint for working toward a common feature set.
-Third, standards bodies draw from a limited set of source material when preparing recommendations.
-Fourth, a lack of detailed documentation hinders the development of compliance suites, thereby interfering with efforts to validate cross-implementation compatibility.
+The lack of a comprehensive SMILES specification has caused several immediate problems.
+First, authors of new SMILES implementations must address ambiguities on an ad hoc basis.
+At the same time, maintainers of existing SMILES implementations lack a blueprint for working toward a common set of features and behaviors.
+The diversity of implementations combined with language underspecification leaves standards bodies with a limited pool of source material upon which to draw.
+Widespread divergence around the edges of the SMILES language hampers the development of compliance suites, which only amplifies the problems faced by software teams trying to build compliant software.
 Finally, extensions to SMILES only make sense in the context of a rigorously defined base language.
 
-The lack of a detailed, comprehensive, and universally-applied SMILES specification also hampers broader data integrity efforts.
-The FAIR Guiding Principles identify four qualities essential for extracting maximum utility from published scholarly data: Findability; Accessibility; Interoperability; and Reusability~\cite{wilkinson:2016}.
-Interoperability in particular requires that \enquote{(meta)data} use a formal, accessible, shared, and broadly applicable language for knowledge representation.'' Although SMILES may be a shared, accessible, and broadly applicable language, its formal underpinnings leave room for improvement.
-
-This situation has persisted long enough that the notion of a SMILES standard may not currently be feasible.
-A dozen or more slightly different implementations are now in use.
+More broadly, this situation undermines ongoing data integrity efforts underway in other scientific disciplines.
+For example, the FAIR Guiding Principles identify four qualities essential for extracting maximum utility from published scholarly data: Findability; Accessibility; Interoperability; and Reusability~\cite{wilkinson:2016}.
+The lack of a well-developed standard means that chemical data sets based on SMILES will suffer in all four areas.
+
+Even the very notion of a SMILES standard may not be realistic at this point.
+Today a dozen or more slightly different implementations are in use.
 None of them can be considered a reference standard.
 Some have introduced extensions not recognized by the others, contributing to language drift while compounding the already formidable difficulties faced by implementors.
-Others claim SMILES compatibility while ignoring certain aspects of written documentation.
-The Daylight implementation, like some of the others, is distributed under a non-Open Source license that restricts use~\cite{daylightToolkit}.
-For several years Daylight operated a publicly-accessible web service based on its toolkit that depicted SMILES strings, but it has since been decommissioned~\cite{depictArchive}.
-Although SMILES may underpin modern chemical information exchange, the foundations of this system are weathered and showing signs of stress.
-
-This paper attempts to resolve these these issues with Dialect, a fully-specified subset of the SMILES language.
-The following sections present Dialect from three perspectives: as a compact system for molecular representation; as a text-based language; and as a specification for software intended to convey chemical structure information across organizational and temporal boundaries.
+Others claim SMILES compatibility while ignoring certain aspects of the documentation that does exist.
+The Daylight implementation, like some of the others, is distributed under a non-Open Source license that restricts use~\cite{daylightToolkit}, and a publicly-accessible web service operated for many years has been decommissioned~\cite{depictArchive}.
+Although SMILES may underpin a lot of modern chemical information exchange, the foundations of this system are weathered and showing signs of stress.
+
+This paper attempts to solve these problems with Dialect, a language designed as a fully-specified subset of SMILES.
+The following sections present Dialect from three perspectives: as a compact system for molecular representation; as a text-based language; and as a specification for software capable of conveying chemical structure information across organizational and temporal boundaries.
 
@@
     \item Weininger's original publication (herein \enquote{The Article})~\cite{weininger:1988}.
-    \item Weininger's 2003 book chapter, revised in 2008 (\enquote{The Book Chapter})~\cite{weininger:2008}.
+    \item Weininger's 2003 book chapter, revised in 2008 (\enquote{The Book Chapter})~\cite{weininger:2003}.
     \item An online manual maintained by Daylight (\enquote{The Manual}).
@@
 
-The Book Chapter defines a fixed set of element symbols which even at the time of publication was obsolete~\cite[p.~83]{weininger:2008}.
+The Book Chapter defines a fixed set of element symbols which even at the time of publication was obsolete~\cite[p.~83]{weininger:2003}.
 The symbol for element 105 is given as Ha (Hahnium), despite IUPAC's resolution of this naming controversy in favor of the symbol Db (Dubnium) years prior~\cite{sageson:1997}.
@@
 
-The Book Chapter defines a recursive grammar for branching which, perhaps without the knowledge of the author, defines sequential parentheses~\cite[p.~86]{weininger:2008}.
+The Book Chapter defines a recursive grammar for branching which, perhaps without the knowledge of the author, defines sequential parentheses~\cite[p.~86]{weininger:2003}.
 Dialect disallows such constructs (e.g., \enquote{*((*))*} and \enquote{*(((*)))*}).
 
-\enquote{Reaction SMILES} is an extension described in the Book Chapter~\cite[p.~89]{weininger:2008}, but disallowed in Dialect.
+\enquote{Reaction SMILES} is an extension described in the Book Chapter~\cite[p.~89]{weininger:2003}, but disallowed in Dialect.
 Specifically, the greater than symbol (\ttt{$>$}) is not a valid Dialect character.
@@
 
-Dialect only supports two stereodescriptors, encoded with the terminals \enquote{@} and \enquote{@@.} The Book Chapter provides a recursive grammar for stereodescriptors that allows multi-symbol descriptors such as \enquote{@@@} and \enquote{@@@@}~\cite[p.~94]{weininger:2008}.
+Dialect only supports two stereodescriptors, encoded with the terminals \enquote{@} and \enquote{@@.} The Book Chapter provides a recursive grammar for stereodescriptors that allows multi-symbol descriptors such as \enquote{@@@} and \enquote{@@@@}~\cite[p.~94]{weininger:2003}.
 The book chapter also discusses non-tetrahedral descriptors including \enquote{@AL1,} \enquote{@AL2,} \enquote{@1,} and \enquote{@SP1.} None of these are supported by Dialect.
@@
 Despite clear inclusion within the syntax of ProtoSMILES, no precise guidance on using this feature is provided.
-The Book Chapter notes \enquote{Adjacent atoms without an intervening bond symbol are connected by a valence-dictated bond (typically a single or aromatic bond) \enquote{-} (single) and \enquote{:} (aromatic) bond symbols may always be omitted on input}~\cite[p.~85]{weininger:2008}.
+The Book Chapter notes \enquote{Adjacent atoms without an intervening bond symbol are connected by a valence-dictated bond (typically a single or aromatic bond) \enquote{-} (single) and \enquote{:} (aromatic) bond symbols may always be omitted on input}~\cite[p.~85]{weininger:2003}.
 This passage implies that aromatic bonds serve no purpose.
@@
 The Book Chapter appears to reiterate this view, noting that \enquote{explicit hydrogen specification is required} in the case of \enquote{hydrogens connected to other hydrogens, e.g.
-[H][H], molecular hydrogen}~\cite[p.~97]{weininger:2008}.
+[H][H], molecular hydrogen}~\cite[p.~97]{weininger:2003}.
 To be clear, Dialect supports any valid virtual hydrogen count on hydrogen itself (e.g., \enquote{[HH]} and \enquote{[HH2]}).
@@
 As noted in the Book Chapter, \enquote{...
-In terms of the valence model being represented, the dot literally represents a bond of formal order zero: the atoms on either side of the dot are explicitly not bonded to each other.}~\cite[p.~88]{weininger:2008}.
+In terms of the valence model being represented, the dot literally represents a bond of formal order zero: the atoms on either side of the dot are explicitly not bonded to each other.}~\cite[p.~88]{weininger:2003}.
 Other statements appear to contradict this statement.
@@
 The Article reports the only default valence for nitrogen as three (p.
-34)~\cite[p.~34]{weininger:1988}, whereas the Book Chapter reports two default valences: 3 and 5~\cite[p.~84]{weininger:2008}.
+34)~\cite[p.~34]{weininger:1988}, whereas the Book Chapter reports two default valences: 3 and 5~\cite[p.~84]{weininger:2003}.
 Dialect uses the latter default valences.
-Atoms whose \ttt{hydrogens} attribute may be set to \ttt{Implicit} (\enquote{the organic subset}) are reported in the Article as: B; C; N; O; P; S; F; Cl; Br; and I~\cite[p.~32]{weininger:2008}.
+Atoms whose \ttt{hydrogens} attribute may be set to \ttt{Implicit} (\enquote{the organic subset}) are reported in the Article as: B; C; N; O; P; S; F; Cl; Br; and I~\cite[p.~32]{weininger:2003}.
 However, an example later in the Article indicates that the string \enquote{Hn1cccc1} with its unbracketed hydrogen valid~\cite[p.~35]{weininger:1988}.
@@
 On the topic of atom selection (\enquote{aromaticity}), The Article implies that selected atoms must be part of a cycle: \enquote{To qualify as aromatic, all atoms in the ring must be sp2 hybridized and the number of available 'excess' $\pi$ electrons must satisfy Hückel's 4{\textit{N}} + 2 criterion}~\cite[p.~34]{weininger:1988}.
-The Book Chapter, however, presents two examples of valid strings that do not encode cycles (\enquote{cc} and \enquote{cccc})~\cite[p.~85]{weininger:2008}.
+The Book Chapter, however, presents two examples of valid strings that do not encode cycles (\enquote{cc} and \enquote{cccc})~\cite[p.~85]{weininger:2003}.
 Consistent with the latter view, Dialect explicitly allows the selection of atoms regardless of cycle membership.

### 2022-03-25 manuscript.tex
Tweaks introduction.

@@
 The lack of a comprehensive SMILES specification has caused several immediate problems.
-First, authors of new SMILES implementations must address ambiguities on an ad hoc basis.
-At the same time, maintainers of existing SMILES implementations lack a blueprint for working toward a common set of features and behaviors.
+Maintainers of SMILES implementations lack a blueprint for working toward a common set of features and behaviors.
+At the same time, authors of new SMILES implementations must address ambiguities on an ad hoc basis.
 The diversity of implementations combined with language underspecification leaves standards bodies with a limited pool of source material upon which to draw.
@@
 
-Even the very notion of a SMILES standard may not be realistic at this point.
+The very notion of a SMILES standard may not be realistic at this point.
 Today a dozen or more slightly different implementations are in use.
@@
 Others claim SMILES compatibility while ignoring certain aspects of the documentation that does exist.
-The Daylight implementation, like some of the others, is distributed under a non-Open Source license that restricts use~\cite{daylightToolkit}, and a publicly-accessible web service operated for many years has been decommissioned~\cite{depictArchive}.
+The Daylight implementation, like some of the others, is distributed under a commercial license that restricts use~\cite{daylightToolkit}, and a publicly-accessible web service operated for many years has been decommissioned~\cite{depictArchive}.
 Although SMILES may underpin a lot of modern chemical information exchange, the foundations of this system are weathered and showing signs of stress.
 
-This paper attempts to solve these problems with Dialect, a language designed as a fully-specified subset of SMILES.
-The following sections present Dialect from three perspectives: as a compact system for molecular representation; as a text-based language; and as a specification for software capable of conveying chemical structure information across organizational and temporal boundaries.
+This paper attempts to solve these problems with Dialect, a fully-specified language designed as a subset of SMILES.
+The following sections present Dialect from three perspectives: as a compact system for molecular representation; as a text-based language; and as a specification for software capable of the lossless conveyance of chemical structure information across organizational and temporal boundaries.
 

### 2022-03-25 manuscript.tex
Improves the flow of Goals, Design, and Tradeoffs.

@@
 Serialization refers to the process of translating a data structure into a format that can be transmitted over a communication channel for later reconstruction.
-Serializations utility lies in enabling general-purpose devices and networks to convey complex pieces of domain-specific information.
+This is useful because it enables general-purpose devices and networks to convey complex pieces of domain-specific information.
 Receiver and recipient may be separated by time, space, or both.
 Molecular serialization extends the concept of serialization to molecular structure.
-
-Two concerns factor prominently in any serialization scheme: \textit{syntax} and \textit{semantics}.
-Syntax sets the vocabulary of symbols to be used and how they may be arranged.
-Semantics set the rules determining what various arrangements of symbols mean.
-Failure to fully specify syntax can lead to needless loss of data.
-Failure to fully specify semantics can result in needless data corruption.
-Semantics in particular play an important role in Dialect due the the inherently complex nature of molecular representations.
-
-Beyond syntax and semantics, a serialization scheme should be useful.
-Utility in this context follows directly from the scope of supported molecular representations.
-Constrain this scope too much and the system will cover too few molecules to be interesting.
-Relax the constraints too much and the system could become verbose, difficult to implement, or both.
-Dialect aims squarely for the middle ground: a system of molecular representation that is specific enough to yield a compact language, but general enough to cover a large amount of interesting chemical space.
-
-Dialect's molecular representation system is based on four mutually-exclusive components, each of which is constrained as follows:
+Examples of molecular serialization formats other than SMILES include Chemical Markup Language (CML)~\cite{murray-rust:2011}, CDXML~\cite{cdxml}, and CTfile ("molfile")~\cite{ctfileFormats}.
+
+Molecular serialization require careful attention to two complementary rule sets: \textit{syntax} and \textit{semantics}.
+Syntax defines the set of valid strings for a language, whereas semantics determine the meaning of valid strings.
+Incomplete specification of either syntax or semantics can lead to data loss and corruption.
+The semantics for Dialect are complex due to the brevity of its syntax and the inherently complex nature of molecular representation.
+
+As a molecular serialization scheme, Dialect must balance two opposing factors: generality and complexity.
+If the chemical space Dialect spans is too broad, complete implementation could become impractical due to language complexity.
+But if the language is made too simple, the chemical space it spans may be of little interest.
+An important goal for Dialect is therefore to address the most interesting chemical space with the simplest possible syntax and semantics.
+
+Toward this end, Dialect is built on four mutually-exclusive concepts:
 
 \begin{itemize}
-\item \textit{Constitution}.
-The atoms present in a molecule and how they are connected with each other through bonds.
-Dialect supports elements having two-letter IUPAC designations and all bonding arrangements available through the valence bond model (as described in detail later).
-\item \textit{Conformation}.
-The kind of stereoisomerism resulting from restricted rotation about a bond.
-Dialect can encode those conformations arising from restricted rotation about double bonds, which includes both alkene and cumulene stereoisomerism.
-\item \textit{Configuration}.
-The kind of stereoisomerism resulting from the arrangement of neighboring atoms in three-dimensional space.
-Dialect supports configurations for tetracoordinate atoms (those attached to four neighbors) having tetrahedral symmetry.
-\item \textit{Delocalization}.
-A type of bonding relationship in which electrons are distributed over two or more bonds, among three or more atoms, or across some combination of atoms and bonds.
+\item \textit{Constitution}, the atoms present in a molecule and how they are connected with each other through bonds.
+Dialect supports elements having two-letter IUPAC designations and all bonding arrangements available through the valence bond model.
+\item \textit{Conformation}, the kind of stereoisomerism resulting from restricted rotation about a bond. Dialect support restricted rotation about double bonds leading to alkene isomerism.
+\item \textit{Configuration}, the kind of stereoisomerism resulting from the arrangement of neighboring atoms in three-dimensional space.
+Dialect supports configurations for atoms having tetrahedral symmetry attached to four neighboring atoms.
+\item \textit{Delocalization}, a type of bonding in which electrons are shared among three or more atoms, or across a combination of atoms and bonds.
 Dialect supports a limited form of delocalization operating over paths of alternating single and double bonds.
@@
 
-Dialect's system of representation is based on the \textit{molecular graph} concept~\cite{balaban:1985}.
-A graph is a data structure comprised of a set of nodes (which map to atoms) and a set of pairwise relationships between them called edges (which map to bonds).
-A molecular graph is a specialized graph onto which atomic and bonding metadata have been overlaid.
-Molecular graphs are ubiquitous in cheminformatics, appearing in contexts such as data formats like Chemical Markup Language~\cite{murray-rust:2011}, CDXML~\cite{cdxml}, and Molfile~\cite{ctfileFormats}, as well as in-memory data structures found in cheminformatics toolkits.
-Like these systems, Dialect augments its underlying graph with node and edge labels.
-
-A Dialect molecule consists of a graph with zero or more nodes and zero or more edges.
-The empty molecule, devoid of atoms and bonds, is allowed.
-Dialect imposes no upper bound on the number of atoms or bonds.
-However, practical limitations related to memory, storage, and CPU time will likely limit the size of molecules in practice.
+Dialect molecules inherit behavior from an underlying \textit{molecular graph}~\cite{balaban:1985}.
+A graph is a data structure comprised of a set of nodes and a set of pairwise relationships between them called edges.
+A molecular graph is a specialized graph onto which chemically-relevant metadata have been overlaid.
+The presence of this overlay allows nodes and edges to represent atoms and bonds, respectively.
+Molecular graphs are ubiquitous in cheminformatics, appearing in both serialization formats in-memory data structures found in software toolkits.
+
+A Dialect molecule is a molecular graph having zero or more nodes and zero or more edges.
+The empty molecule, devoid of atoms and bonds, is therefore allowed.
+Dialect imposes an upper bound on the number of nodes of $2^{32} - 1$ and the same limit for the number of edges.
+However, practical limitations related to memory, storage, and CPU time will likely further constrain these limits in practice.
 
@@
 Second, edges are \textit{undirected}.
-An undirected edge is equivalent to another edge whose terminals are swapped.
-An undirected edge between nodes A and B is therefore identical to an undirected edge between nodes B and A.
+An undirected edge is equivalent to another edge whose terminals are transposed.
+With one exception that will be discussed in detail, an undirected edge between nodes A and B is therefore identical to an undirected edge between nodes B and A.
 Third, \textit{parallel edges} are disallowed.
@@
 
-A molecular graph contains one or more \textit{connected components}.
+A molecular graph contains zero or more \textit{connected components}.
 A connected component is a graph in which any two vertices are connected to each other by at least one path.
-Dialect places no restriction on the number of connected components, although practical limitations are likely to impose one.
+The maximum number of connected components is equal to the maximum node count, although practical factors may reduce this limit.
 

### 2022-03-26 manuscript.tex
Improves the flow of Data Model.

@@
 
-Dialect uses several abstractions collectively known as the \textit{data model}. The data model defines data types and relationships between them for the purpose of lossless molecular representation.
-
-The data model is expressed in terms of a small set of \textit{primitives}. A primitive is an irreducible data type. Dialects primitives are summarized in Table \ref{table:primitives}.
+Dialect uses several abstractions collectively known as the \textit{data model}.
+The data model defines metadata attributes associated with the components of a molecular graph.
+These attributes are themselves composed of \textit{primitives} (Table~\ref{table:primitives}), or irreducible data types.
 
@@
 
-Table \ref{table:atomic-state} summarizes atomic attributes.
-Not all combinations of values (\enquote{atomic state}) are valid.
+An atom possesses several attributes (Table~\ref{table:atomic-state}), which collectively define an \textit{atomic state}.
+Not all atomic states are valid.
 Implementations must ensure either the impossibility of an invalid state, or an error condition in the event that one is created.
+Atomic state is restricted in the following ways:
 
@@
 
-Restrictions on atomic state are:
-
 \begin{enumerate}
     \item{The value of \ttt{index} must be unique over the molecular graph.}
-    \item{If \ttt{hydrogens} equals \ttt{Implicit}, then \ttt{isotope}, \ttt{configuration}, and \ttt{charge} must equal their default values.
-Furthermore, \ttt{element} must equal one of \ttt{B}; \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; \ttt{S}; \ttt{F}; \ttt{Cl}; \ttt{Br}; \ttt{I}; or \ttt{None}.}
-    \item{If \ttt{selected} equals \ttt{true}, then \ttt{element} must equal one of: \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; or \ttt{S}.}
+    \item{If the value of \ttt{hydrogens} equals \ttt{Implicit}, then the values of \ttt{isotope}, \ttt{configuration}, and \ttt{charge} must equal their respective defaults.
+Furthermore, the value of \ttt{element} must equal one of \ttt{B}; \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; \ttt{S}; \ttt{F}; \ttt{Cl}; \ttt{Br}; \ttt{I}; or \ttt{None}.}
+    \item{If the value of \ttt{selected} equals \ttt{true}, then the value of \ttt{element} must equal one of: \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; or \ttt{S}.}
 \end{enumerate}
 
-Bond attributes are summarized in Table \ref{table:bond-state}. As with atomic attributes, not all bond states are valid. Three restrictions apply:
+Like atoms, bonds carry metadata in the form of attributes (Table \ref{table:bond-state}), which collectively define a \textit{bond state}.
+As with atomic attributes, not all bond states are valid.
+Bond state is restricted in the following ways:
 
@@
     \item{The value of \ttt{index} must be unique over the molecular graph.}
-    \item{The \ttt{source} and \ttt{target} attributes must reference valid atomic \ttt{index} attributes.}
-    \item{If \ttt{direction} is non-default (any value other than \ttt{None}), then \ttt{order} must equal 1.}
-    \item{If the value of \ttt{elided} is \ttt{true}, the values of \ttt{order} and \ttt{direction} must equal 1 and \ttt{None}, respectively.}
+    \item{The \ttt{source} and \ttt{target} attributes must reference atomic \ttt{index} attributes found elsewhere in the molecular graph.}
+    \item{If the value of \ttt{direction} does not equal \ttt{None}, then the value of \ttt{order} must equal one.}
+    \item{If the value of \ttt{elided} equals \ttt{true}, the values of \ttt{order} and \ttt{direction} must equal one and \ttt{None}, respectively.}
 \end{enumerate}
 
-The data model provides a link between syntax and semantics.
+The data model will be referenced throughout this paper as a way to connect syntax and semantics.
 Dialect-compatible software may, of course, use any suitable internal data model.
-Dialect implementations must, however, ensure that their internal data model is consistent with the one provided here.
+Implementations must, however, ensure that any internal data model is consistent with the one provided here.
 

### 2022-03-28 manuscript.tex
Improves the flow of Constitution and Bond Elision.

@@
 
-A Dialect molecule is described at the lowest level by its constitution.
-Constitution consists of a set of atomic nuclei (nodes), a set of pairwise bonding relationships between them (edges), and those attributes needed to associate every valence electron with a specific node or edge.
-Constitution excludes attributes related to stereochemistry, or any other feature whose attributes extend over more than one atom or bond.
-
-Two attributes characterize an atom's nucleus: \ttt{element} and \ttt{isotope}.
-The `element` attribute is an optional one- or two-letter character sequence selected from the set designated by the IUPAC~\cite{periodicTable}.
+A molecular graph is described at the lowest level by its constitution.
+In Dialect, constitution comprises the molecular graph and those attributes locating nucleons and electrons at specific atoms and bonds.
+All other attributes lie outside the scope of constitution.
+Atomic attributes related to constitution are: \ttt{index}; \ttt{element}; \ttt{hydrogens}; and \ttt{charge}.
+Bond attributes related to constitution are \ttt{source}, \ttt{target}, \ttt{order}, and \ttt{elided}.
+
+The attributes \ttt{element} and \ttt{isotope} characterize an atom's nucleus.
+The \ttt{element} attribute is an optional one- or two-letter character sequence selected from the set designated by the IUPAC~\cite{periodicTable}.
 When IUPAC authorizes additional symbols, they will become valid values for the \ttt{element} attribute.
-If the elemental identity of an atom is unknown, its \ttt{element} attribute can be omitted.
+If the elemental identity of an atom is unknown, its \ttt{element} attribute must equal \ttt{None}.
 The \ttt{isotope} attribute is an optional integer value representing an atom's nuclear mass number, where mass number is the sum of proton and neutron count.
-Omitting the \ttt{isotope} property means that the isotopic composition is unspecified.
-When present, the lower bound on \ttt{isotope} equals one.
-This constraint leaves room for physically meaningless negative implied neutron count as in, for example, carbon-5.
-The lower bound for isotope applies even if the \ttt{element} attribute is not defined.
-
-The \ttt{hydrogens} attribute encodes an atom's mode of \textit{hydrogen suppression}.
-Hydrogen suppression replaces hydrogen atoms in a molecular graph with a more compact representation.
-Despite the removal of its node, a suppressed hydrogen and its bonds to the parent atom are understood to be present.
-The abundance and regular bonding of hydrogen atoms in molecular graphs makes hydrogen suppression a popular strategy for increasing information density.
-
-Dialect supports two mutually-exclusive modes of hydrogen suppression.
-The first is \textit{virtual hydrogen}.
-A virtual hydrogen is one whose node and connecting bond have been replaced by an integer tally attribute on the parent atom.
+Setting the \ttt{isotope} property to \ttt{None} means that the element's isotopic composition equals natural abundance.
+Otherwise, the lower bound on the value of \ttt{isotope} is zero.
+This lower limit allows for such physically nonsensical states such as atoms of zero implied mass (\ttt{isotope} equals zero) or negative implied neutron count (e.g., carbon-5).
+Implementations may reject such atomic states as invalid.
+
+The \ttt{hydrogens} attribute sets an atom's mode of \textit{hydrogen suppression}.
+Hydrogen suppression replaces an atomic hydrogen of natural isotopic abundance and its associated bond with an integer tally associated with the parent atom.
+Even so, a suppressed hydrogen and its bond are understood to be present.
+The relative abundance and regular bonding of hydrogen atoms in molecular graphs makes hydrogen suppression an effective technique for increasing information density.
+Two mutually-exclusive types of suppressed hydrogen are supported: \textit{virtual hydrogen} and \textit{implicit hydrogen}.
+
+A virtual hydrogen is one that appears as an integer unit contribution associated with the parent atom.
 For example, methane can be represented by a molecular graph having five nodes and four edges.
 But methane can also be represented as a graph of one node whose \ttt{hydrogens} attribute equals four.
-When using this variant, the \ttt{hydrogens} attribute may assume integer values from zero to nine, inclusive.
-The process of transforming one or more hydrogen nodes into a virtual hydrogen is called \enquote{virtualization.} The opposite process, transforming one or more virtual hydrogens into a node, is called \enquote{reification.}
-
-The second mode of suppressed hydrogen is \textit{implicit hydrogen}.
+When using this mode, the \ttt{hydrogens} attribute may assume integer values from zero to nine, inclusive.
+The process of replacing atomic hydrogens with virtual hydrogen is called \enquote{virtualization.} The opposite process, replacing virtual hydrogen with atomic hydrogen, is called \enquote{reification.}
+
 An implicit hydrogen is similar to a virtual hydrogen in that both the hydrogen node and bond have been replaced with a tally.
 Unlike a virtual hydrogen, however, this tally is not present as attribute on the parent atom but rather as a computation.
-The computation will be described in detail in a subsequent section.
-Subject to certain limitations, an an atom identifies itself as having an implicit hydrogen count by setting its \ttt{hydrogens} attribute to \ttt{Implicit}.
-
-The remaining constitutional atomic attribute, \ttt{charge}, enables electron counting.
-Electron counting associates each electron in a molecular graph with either an atom or a bond.
+Subject to certain limitations, an atom identifies itself as having an implicit hydrogen count by setting its \ttt{hydrogens} attribute to \ttt{Implicit}.
+The computation and its limitations will be described later in this paper.
+
+The atomic attribute, \ttt{charge}, enables electron counting.
+The purpose of electron counting is to associate each electron in a molecular graph with either an atom or a bond.
 This is a fundamental requirement for the principle of mass conversion.
@@
 Although not commonly considered as such, the VB model can be thought of as an algorithm for molecular assembly.
-The perspective is especially useful given that Dialect itself can be viewed as a programming language for molecular assembly.
+This perspective is especially useful given that Dialect itself can be viewed as a programming language for molecular assembly.
 
 A simplified algorithm for electron counting can be summarized as follows.
-A molecular graph starts as a set of atoms, where each atom consists of a nucleus (as described above) and an integer electron count equal to the atomic number of the atom's element.
-Two nodes (nuclei) are selected as the terminals of a bond.
+A molecular graph starts as a set of disconnected atoms and an integer electron count equal to the atomic number of the atom's element.
+Two nodes are selected as the terminals of a bond.
 From each atom, the same positive electron count is deducted.
 The total electron count deducted from each atom is simultaneously credited to the bond's electron count.
-For example, if each atom contributes one electron to a bond, then the bond's electron count will increment by two.
+For example, if each atom contributes one electron to a bond, then the bond's electron count will increase by two.
 
@@
 The electron count deducted from each atom is $n$, an integer greater than zero.
-Given bond formation deducts an equal number of electrons from each atom, the total number of electrons associated with the newly-formed bond is $2n$.
+Given that bond formation deducts an equal number of electrons from each atom, the total number of electrons associated with the newly-formed bond is $2n$.
 By definition, bond order is the bond's electron count divided by two, therefore $n$ is the bond order.
@@
 Dialect imposes an upper limit of three on formal bond order.
-The relative scarcity of bond orders greater than three in molecules conforming the the VB Model makes this restriction unlikely to meaningfully restrict applicability.
+The relative scarcity of bond orders greater than this limit in molecules conforming the the VB Model makes this restriction unlikely to meaningfully restrict applicability.
 
@@
 
+Each atom carries an \ttt{index} attribute, allowing bonds to reference atoms in a molecular graph.
+To do so, the \ttt{source} and \ttt{target} attributes are set with the respective indexes of the source and target atoms.
+The set of all bonds and the nodes they reference determines the connectivity of a molecular graph.
+In the context of determining connectivity, source and target indexes may be reversed.
+
 \section*{Bond Elision}
 
-To promote compact string representations, Dialect supports \textit{bond elision}.
-A bond of order one may be setting its \ttt{elided} attribute to \ttt{true}.
-As such, an elided bond always has a bond order of one.
-A bond may be elided for one of two reasons.
-First, elided bonds are not written during serialization, allowing for increased information density.
-The second reason relates to a feature described in the next section.
+A bond whose attributes are assigned default values can participate in \textit{bond elision}.
+In bond elision, a bond's \ttt{elided} attribute is set to \ttt{true}.
+When written in string form, elided bonds require no characters.
+This makes bond elision useful for data compression.
+The second application for elided bonds will be described in the next section.
 

### 2022-03-29 manuscript.tex
Improves flow of Delocalization Subgraph.

@@
 A molecular representation based solely on the valence bond model can yield artifacts resulting from \textit{delocalization induced molecular equality} (DIME).
-DIME occurs when a molecular graph exists in two or more unequal forms, each differing from the others only through electron delocalization.
-Two molecular graphs are equal if they are isomorphic and a 1:1 mapping exists between the atoms and bonds in each graph.
-Electron delocalization results when the electrons of a valence bond representation can be distributed over more than one pair of atoms, as is the case with benzene and related molecules.
-DIME is similar to \enquote{resonance} or \enquote{aromaticity,} but those terms are avoided here due to their ambiguity~\cite{randic:2018,kerber:2006}.
+DIME arises when a molecular graph exists in two or more unequal forms, each differing from the others only through electron delocalization.
+In this context, two graphs are equal if they are isomorphic and each matched pair of atoms or bonds possess equivalent non-index attributes.
+If shifting electrons through delocalization causes two graphs to become equal, then DIME has occurred.
+DIME is sometimes referred to as \enquote{resonance} or \enquote{aromaticity,} but those terms are avoided here due to their ambiguity~\cite{randic:2018,kerber:2006}.
 
@@
 DIME can interfere with \textit{canonicalization}, or the selection of a single representation for a molecular graph.
-The presence of multiple equivalent molecular graphs differing only in the placement of single and double bonds complicates selection rules and invariants, which must be adapted to account for the artificial asymmetry.
-
-To eliminate DIME and thereby streamline canonicalization, Dialect representations support a \textit{delocalization subgraph} (DS).
+The presence of multiple equivalent molecular graphs differing only in electron delocalization complicates the formulation of selection rules and invariants, which must be adapted to account for the artificial asymmetry.
+
+To eliminate DIME and thereby streamline canonicalization, the molecular graph is augmented with a \textit{delocalization subgraph} (DS).
 A DS is a possibly empty node-induced subgraph of a molecular graph.
-The membership of a DS is drawn from the set of atoms and bonds that participate in DIME within a given molecular graph.
-
-An atom is added to a DS by setting its \ttt{selected} attribute to \ttt{true} (\enquote{selection}).
+The membership of a DS may be drawn from the set of atoms and bonds that participate in DIME within a given molecular graph.
+
+Membership of a DS is determined as follows.
+An atom is added to the DS by setting its \ttt{selected} attribute to \ttt{true} through a process called \textit{selection}.
 For an atom to be selectable, its \ttt{element} attribute must be contained in the list: \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; and \ttt{S}.
 Atoms having other elements (or \ttt{None}) must not be selected.
-Only selected atoms belong to the DS.
+A bond will be added to the DS if both of its terminals are selected and the \ttt{elided} attribute equals \ttt{true}.
+No other bond will be added to the DS.
+This rule allows arbitrary bonds between selected atoms to be excluded from the DS.
 
@@
 \end{figure}
-
-A bond will be added to the DS only if both of the following conditions are met: (1) both terminals are selected; and (2) the bond itself is elided.
-No other bond will be added to the DS.
-This behavior makes it possible to exclude a specific bond from the DS by making its bond order explicit.
 
@@
 A perfect matching includes all the nodes of its parent graph.
-Every atom added to a DS must therefore be part of a perfect matching over it.
+Every atom added to a DS must therefore become part of its perfect matching.
 
@@
 Because the presence of a filled DS implies a perfect matching over it, kekulization always succeeds.
-A popular algorithm for matching, the Edmunds \enquote{blossom algorithm,} has a time complexity of $O(|E||V|^2)$, where $|E|$ is the number of edges and $|V|$ is the number of nodes~\cite{edmonds:1965}.
+A widely-used algorithm for matching, the Edmunds \enquote{blossom algorithm,} has a time complexity of $O(|E||V|^2)$, where $|E|$ is the number of edges and $|V|$ is the number of nodes~\cite{edmonds:1965}.
 Although more efficient algorithms are known, they are either much more difficult to implement or lack generality.
@@
 Depending on the application, other criteria may be applied.
-For example, a selection algorithm can restrict candidate atoms to those found in cycles.
-Electron-counting techniques can also be introduced to approximate the chemical concept of \enquote{aromaticity.} A double bond between two selected atoms may or may not be elided, depending on the application.
+For example, a selection algorithm may restrict candidate atoms to those found in cycles.
+Electron-counting techniques may also be introduced to approximate the chemical concept of \enquote{aromaticity.} Elision of the single bonds to be added to the DS completes the process.
 

### 2022-03-29 manuscript.tex
Streamlines Valence and Subvalence.

@@
 Valence is a non-negative integer computed as the sum of bond orders at a given atom.
-Single bonds contribute one to the tally, double bonds two, and triple bonds three.
-All other bonds contribute one, as does each virtual hydrogen.
+Single and elided bonds contribute one to the tally, double bonds two, and triple bonds three.
+Each virtual hydrogen contributes one.
 For example, the valence of a methyl carbon having three virtual hydrogens and a hydrogen neighbor is four.
@@
 For example, nitrogen has the default valences three and five.
-Both ammonia (NH\textsubscript{3}) and nitrogen pentahydride (NH\textsubscript{5}) are therefore fully saturated forms of nitrogen.
+Both ammonia (NH\textsubscript{3}) and nitrogen pentahydride (NH\textsubscript{5}) are therefore fully saturated forms of nitrogen according to Table~\ref{table:default-valences}.
 
@@
 
-Given an atom whose element has one or more default valences \textit{subvalence} can be computed.
+Given an atom whose element has one or more default valences, \textit{subvalence} can be computed.
 Subvalence is the number of hydrogens that can be added to an atom without exceeding the lowest possible default valence.
-If no suitable default valence exists, then the subvalence is zero.
+If no suitable default valence exists, then subvalence equals zero.
 This computation can be expressed algorithmically (Algorithm~\ref{algorithm:subvalence}).
@@
 If no suitable default valence is found, zero is returned.
+The subvalence of an atom whose element has no default valences (i.e., no listing in Table~\ref{table:default-valences}) is zero.
+For example, an iron atom always has a subvalence of zero, regardless of how it is bonded.
 
@@
 \end{algorithm}
-
-The subvalence of an atom whose element has no default valences (i.e.
-no listing in Table~\ref{table:default-valences}) is zero.
-For example, an iron atom always has a subvalence of zero, regardless of how it is bonded.
 

### 2022-03-29 manuscript.tex
Improves the flow of Implict Hydrogen Count.

@@
 
-To qualify for implicit hydrogen counting, an atom must have a \ttt{hydrogens} attribute set to \ttt{Implicit}.
-As noted previously, this setting requires: (1) \ttt{isotope}, \ttt{charge}, and \ttt{configuration} must be set to their default values; and (2) the value of \ttt{element} must equal one of: \ttt{B}; \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; \ttt{S}; \ttt{F}; \ttt{Cl}; \ttt{Br}; \ttt{I}; or \ttt{None}.
-
-For unselected atoms, implicit hydrogen count equals subvalence, with one exception.
-An oxygen atom with one single bond has an implicit hydrogen count of one because subvalence is one ($2 - 1$).
-However, an oxygen atom with two singly-bonded neighbors has an implicit hydrogen count of zero ($2 - 2$).
-For atoms whose \ttt{element} attribute is \ttt{None}, implicit hydrogen count equals zero.
-
-For selected atoms, a modified procedure is used.
-If subvalence is greater than zero, then the implicit hydrogen count equals subvalence minus one.
-This subtraction models the unpaired electron that would would be required for delocalized bonding, or localized bonding if the atom were deselected.
+An atom whose \ttt{hydrogens} attribute equals \ttt{Implicit} signals that its hydrogens must be counted algorithmically.
+The exact algorithm depends on the value of the atoms's \ttt{selected} attribute.
+
+For an unselected atom, implicit hydrogen count equals subvalence.
+For example, an oxygen atom with one single bond has an implicit hydrogen count of one because subvalence is one ($2 - 1$).
+Similarly, an oxygen atom with two singly-bonded neighbors has an implicit hydrogen count of zero ($2 - 2$).
+
+For a selected atom, implicit hydrogen count equals subvalence minus one.
+This subtraction accounts for the extra valence implied by the atom's membership in the delocalization subgraph.
 Consider a selected carbon atom in benzene with a \ttt{hydrogens} attribute of \ttt{Implicit}.
-Subvalence equals two, so implicit hydrogen count equals one ($2 - 1$).
+Subvalence equals two ($4 - 2$), so implicit hydrogen count equals one ($2 - 1$).
+Likewise, the subvalence of a selected nitrogen atom in pyridine equals one ($3 - 2$) so implicit hydrogen count equals zero ($1 - 1$).
+Stated differently, the calculation ensures that an atom's implicit hydrogen count is identical before and after deselection.
 
 If the subvalence of a selected atom equals zero, then an implicit hydrogen count of zero is reported.
-The meaning of such an atomic state may seem suspect.
-If an atom has no unpaired electrons, how can it participate in delocalized bonding?
+The meaning of such an atomic state may seem suspect because an atom without a free valence can not perform the required promotion of a single bond during deselection.
 As will be explained in detail later (Pruning), this situation can arise for reasons of convenience or tradition.
@@
 
-An implicit hydrogen count may or may not correlate with chemical intuition or experimental data.
+An atom's implicit hydrogen count may or may not correlate with chemical intuition or experimental data.
 Consider the phosphorous-bearing atom of hypophosphorous acid (HOP(O)H\textsubscript{2}).
@@
 The implicit hydrogen count is therefore zero ($3 - 3$) rather than the expected two.
-The hydrogen count of this atom, and others like it, must be encoded as a virtual hydrogen count.
+To represent this atom and other like it, virtual hydrogens must be used.
 

### 2022-03-30 manuscript.tex
Improves the flow of Conformation.

@@
 
-To support the assignment and interpretation stereochemical features, each node is assigned a sequential, unique, non-negative integer \ttt{index} attribute (\enquote{index}).
-The value of an atom's index equals zero for a molecular's graph's first atom.
-The index for each subsequently-added atom equals the order (node count) of the graph prior to the addition.
-For example, the index for the second atom is one, the index for the third atom is two, and so on.
+To support the assignment and interpretation stereochemical features, each node carries a unique, non-negative integer \ttt{index} attribute (\enquote{index}).
+Assignment is sequential, meaning that the first atom gets index zero, the next one, and so on.
+For example, the index for the first atom is zero, for the second atom one, for the third atom two, and so on.
 The maximum value of an atomic index is 2\textsuperscript{32} - 1.
@@
 
-Organic chemistry uses the parity descriptors \textit{entgegen} (\textit{E}) and \textit{zusammen} (\textit{Z}) to label conformational parity.
+Organic chemistry uses the descriptors \textit{entgegen} (\textit{E}) and \textit{zusammen} (\textit{Z}) to label conformational parity.
 This system is based on prioritization of the atoms neighboring the double bond.
@@
 Constructing the full parity requires the double bond itself, and at least two PPBs.
-At least two PPBs must flank the double bond.
+With just a few exceptions (see below), a double bond with a neighbor at one terminal bound through a PPB must also have a neighbor at the opposite terminal bound through a PPB.
 
@@
 
-To support this system, each bond carries a nullable \ttt{direction} attribute (\enquote{direction}).
-When present, this attribute may assume one of the two values \ttt{Up} or \ttt{Down}.
+To support this system, each bond carries an optional \ttt{direction} attribute (\enquote{direction}).
+When present, direction may assume one of the two values \ttt{Up} or \ttt{Down}.
 These values refer to a geometrical model in which the two terminals of a double bond and their neighbors are assigned local relative coordinates in a plane.
-The double bond terminals are placed on the x-axis such that the terminal with the lowest \ttt{index} attribute appears to the left of the terminal with the higher atomic \ttt{index} attribute.
-These terminals are designated \enquote{left terminal} and \enquote{right terminal,} respectively.
-Each neighbor of a terminal is then assigned a relative coordinate based on the \ttt{index} attribute and the direction of its PPB bond.
+The terminals of the bond are placed on the x-axis such that the one on the right succeeds the one on the left.
+The neighbors at each terminal joined through a PPB are then assigned relative coordinates in the plane.
+
+Neighbors of the left terminal are positioned according to the relative ordering of atomic indexes and the bond's \ttt{direction} attribute.
+If the neighbor precedes the terminal then a \ttt{direction} attribute equal to \ttt{Up} locates the neighbor in the lower-left quadrant and a \ttt{direction} attribute of \ttt{Down} locates the neighbor in the upper-left quadrant.
+If the neighbor succeeds the terminal, these placements are reversed.
+
+An analogous procedure assigns relative coordinates to a neighbor of the right terminal.
+If a neighbor succeeds the right terminal, place it to the upper right given an \ttt{Up} direction, or to the lower right given a \ttt{Down} direction.
+Reverse these assignments if the neighbor precedes the right terminal.
 
@@
 \end{figure}
-
-The following procedure assigns a relative coordinate to a neighbor of the left terminal.
-First, determine the relative order of the indexes for the neighbor and the left terminal.
-If the neighbor succeeds the left terminal, place the neighbor to the upper left of the terminal if the bond direction is \ttt{Up}, or to the lower left if the bond direction is \ttt{Down}.
-If the terminal succeeds its neighbor, reverse these assignments.
-
-An analogous procedure assigns relative coordinates to a neighbor of the right terminal.
-If a neighbor succeeds the right terminal, place the neighbor to the upper right given an \ttt{Up} direction, or to the lower right given a \ttt{Down} direction.
-Reverse these assignments if the right terminal succeeds its neighbor.
 
@@
 The neighbor with the PPB can then be placed explicitly.
-Doing so allows the remaining neighbor without a PPB to be placed as well.
+Doing so allows the remaining neighbor without a PPB to be placed as well by elimination.
 For example, if the neighbor of a left terminal is placed in the upper left quadrant, then the sibling without a PPB can only occupy the lower left quadrant.
@@
 Consider the encoding of PPBs for (\textit{E})-2-butene.
-Assume that indexes are assigned sequentially from left to right and the goal is to arrive at a trans (or anti) substituent orientation.
-Begin by placing the two double bond terminals on the x-axis.
-Left and right terminals are placed on the basis of succession.
-The left terminal (Atom 1) succeeds its neighbor (Atom 0), which should be placed in the lower-left quadrant.
+Assume that indexes are assigned sequentially from the neighbor of the left terminal to the neighbor of the right terminal and that the goal is to arrive at a trans (or anti) substituent orientation.
+Begin by placing the two double bond terminals on the x-axis according to succession.
+The left terminal succeeds its neighbor, which should be placed in the lower-left quadrant.
 To achieve this, assign a PPB direction of \ttt{Up}.
-The right terminal (Atom 2) precedes its neighbor (Atom 3), which should be placed in the upper-right quadrant.
+The right terminal precedes its neighbor, which should be placed in the upper-right quadrant.
 To active this, assign a PPB direction of \ttt{Up}.
@@
 
-In a similar manner assigned PPBs can be decoded.
-The process starts by placing Atom 1 to the left of Atom 2 along the x-axis.
-The PPB between the left terminal and its neighbor (Atom 0) uses the \ttt{Up} direction.
-The neighbor precedes the left terminal, so Atom 0 should be placed in the lower-left quadrant.
-The PPB between the right terminal and its neighbor (Atom 3) uses the \ttt{Up} direction.
-The neighbor (Atom 3) succeeds the right terminal (Atom 2).
-Therefore, Atom 3 is placed in the upper right quadrant.
-This results in the expected conformation (\textit{E}).
+PPBs are decoded analogously.
+Start by placing the terminals on the x-axis according to succession.
+The PPB between the left terminal and its neighbor has a direction of \ttt{Up}.
+Because the neighbor precedes the terminal, the neighbor should be placed in the lower-left quadrant.
+The PPB between the right terminal and its neighbor has a direction of \ttt{Up}.
+Because the neighbor succeeds the terminal, the neighbor is placed in the upper-right quadrant.
+The result is the expected (\textit{E}) conformation.
 
@@
 
-Consider (\textit{E})-2-butene.
-The PPB spanning atoms 0 and 1 will typically be encoded as the tuple \ttt{(0, 1, Up)} because this orientation tracks the order of atom index.
-However, the PPB can be equivalently encoded as the tuple \ttt{(1, 0, Down)}.
-Before this representation can be used, it must first be inverted into the equivalent form represented by the tuple \ttt{(0, 1, Up)}.
-
 \begin{figure}
@@
 \begin{itemize}
-    \item{Overspecification. This state occurs when two or more PPBs share a common ancestor and simultaneously use the same parity, thus placing multiple neighbors into the same quadrant.}
+    \item{Overspecification. This state occurs when two or more PPBs share a common terminal and simultaneously use the same parity, thus placing multiple neighbors into the same quadrant.}
     \item{Underspecification.
@@
 An exception to the underspecification rule applies in the case of conjugated polyenes.
-In this case a double bond with a terminal bearing a PPB does not require the opposing terminal to also bear a neighbor with a PPB.
+Here a double bond with a terminal connected through a PPB does not require the opposing terminal to also bear a neighbor connected through a PPB.
 If the opposing terminal lacks neighbors, no error is generated and reading continues normally.
@@
 Cyclooctatetraene offers an example.
-Although the (\textit{E}, \textit{E}, \textit{E}, \textit{E}) conformation is accessible, the (\textit{Z}, \textit{E}, \textit{E}, \textit{E}) conformation is not.
+Although the (\textit{E}, \textit{E}, \textit{E}, \textit{E}) conformation is expressable, the (\textit{Z}, \textit{E}, \textit{E}, \textit{E}) conformation is not.
 This problem arises because the same PPB encodes a partial conformation of two different double bonds.

### 2022-03-30 manuscript.tex
Improves the flow of Configuration.

@@
 
-The third major element of molecular representation is \textit{configuration}.
-Configuration is the three-dimensional arrangement of neighbors about an atom.
-Dialect limits configuration to the special case of an atom with exactly four substituents placed at the vertexes of a tetrahedron.
-Here, \enquote{substituent} means an atomic neighbor; lone electron pairs are not considered substituents.
-One of the four substituents may be a virtual hydrogen.
+The third major element of molecular representation is \textit{configuration}, the three-dimensional arrangement of neighbors about a central atom.
+Dialect limits configuration to the special case of an atom with exactly four substituents placed at the vertices of a tetrahedron.
+Here, \enquote{substituent} means an atomic neighbor or virtual hydrogen.
+Lone electron pairs are not considered substituents.
 
@@
 Dialect supports two configurational descriptors: \ttt{Left} and \ttt{Right}.
-The context for the terms \enquote{left} and \enquote{right} will be explained shortly.
-
-The descriptors \ttt{Left} and \ttt{Right} restrict neighboring atoms to the vertices of a tetrahedron whose center is the central atom.
-First, the mate of the first bond to the central atom is placed at an arbitrary vertex.
-Then shift the frame of reference by sighting along the first bond from the neighbor toward the central atom.
-Finally, arrange the remaining neighbors using the ordering of bonds to the central atom.
-For the \ttt{Left} variant, use a counterclockwise distribution.
-For the \ttt{Right} variant, use a clockwise distribution.
+These descriptors constrain the neighbors of a central atom to the vertices of a tetrahedron.
+
+The \ttt{Left} variant works as follows.
+First, the bond having the index with the lowest value identified.
+The frame of reference is shifted by sighting down this bond from the neighbor to the central atom.
+An equilateral triangle with one vertex pointing upward is then inscribed about the central atom.
+The bond with the next lowest index is then identified and its neighbor is placed at the top vertex of the triangle.
+The neighbor attached to the succeeding bond is then placed at the left triangle vertex.
+Finally, the remaining neighbor is placed at the remaining vertex.
+The overall pattern is one of counterclockwise distribution.
+
+The \ttt{Right} variant follows the same pattern, except that placement of the neighbors attached to the last two bonds in succession are transposed.
+The overall pattern here is one of clockwise distribution.
 
@@
 
-If an atom bears one virtual hydrogen and three substituents, a modified procedure is used.
-The virtual hydrogen is first placed at an arbitrary vertex of the tetrahedron.
-The tetrahedron is then oriented along the axis pointing from the virtual hydrogen to the center.
-Neighbor atoms are then distributed in a counterclockwise (\ttt{Counterclockwise}) or clockwise (\ttt{Clockwise}) pattern to the remaining vertices.
+A modified procedure is used for atoms bearing one virtual hydrogen and three atomic substituents.
+The implied bond to the virtual hydrogen is considered to precede the others.
+Sighting down this bond, the atomic neighbors are then placed as previously at the vertices of a triangle inscribed about the central atom.
 
@@
 
-The association of a configurational descriptor with an atom implies a neighbor count of either four if no virtual hydrogens are present, or three if exactly one virtual hydrogen is present.
+An atom whose \ttt{configuration} attribute assumes a non-default value must have a neighbor count of either four if no virtual hydrogens are present or three if one virtual hydrogen is present.
 In any other context, the use of a configurational descriptor is considered an error.
@@
 This restriction places some constraints around the use of configuration.
-For example, the assignment of stereodescriptors to central carbon atom of allenes and other cumulenes might seem like a reasonable extension.
-However, to do so is an error because the central atom lacks four substituents.
-Similarly, lone pairs must not take the place of atomic substituents, as might be the case with sulfoxides for example.
+Some topologies may appear close enough to tetrahedral to stretch the meaning of the configurational descriptors.
+However, all such adaptations are invalid.
+Examples include the central atom of allenes and other odd cumulenes, and the sulfur atom of sulfoxides where the lone pair plays the role of a substituent.
 
@@
 In other words, single, double, triple, and partial parity bond types can all be present.
-The only requirement that must be satisfied is four substituents or, equivalently, three substituents and a virtual hydrogen.
+The only requirement that must be satisfied by a central atom is a neighbor count of four, or three when a virtual hydrogen is present.
 
@@
 This process is called \textit{transformation}.
-Five rules suffice to transform any configuration into any other:
+Five operations suffice to transform any configuration into any other:
 
@@
 In the context of Dialect, a stereocenter must be tetracoordinate.
+However, this is a necessary but insufficient condition.
+Molecular topology can result in permutation that does not produce stereoisomers.
 Readers should avoid the assignment of stereodescriptors to atoms that are not stereocenters.

### 2022-04-01 manuscript.tex
Improves the flow of Syntax.

@@
 
-A Dialect string (\enquote{string}) is a UTF-encoded sequence of zero or more characters chosen from the set: \ttt{A}; \ttt{B}; \ttt{C}; \ttt{D}; \ttt{E}; \ttt{F}; \ttt{G}; \ttt{H}; \ttt{I}; \ttt{K}; \ttt{L}; \ttt{M}; \ttt{N}; \ttt{O}; \ttt{P}; \ttt{R}; \ttt{S}; \ttt{T}; \ttt{U}; \ttt{V}; \ttt{W}; \ttt{X}; \ttt{Y}; \ttt{Z}; \ttt{a}; \ttt{b}; \ttt{c}; \ttt{d}; \ttt{e}; \ttt{f}; \ttt{g}; \ttt{h}; \ttt{i}; \ttt{k}; \ttt{l}; \ttt{m}; \ttt{n}; \ttt{o}; \ttt{p}; \ttt{r}; \ttt{s}; \ttt{t}; \ttt{u}; \ttt{v}; \ttt{0}; \ttt{1}; \ttt{2}; \ttt{3}; \ttt{4}; \ttt{5}; \ttt{6}; \ttt{7}; \ttt{8}; \ttt{9}; \ttt{+}; \ttt{-}; \ttt{.}; \ttt{\%}; \ttt{@}; \ttt{*}; \ttt{[}; \ttt{]}; \ttt{/}; \ttt{\char`\\}; \ttt{=}; and \ttt{\#}.
+A Dialect string is a UTF-8-encoded sequence of zero or more characters chosen from the set \{ \ttt{A}, \ttt{B}, \ttt{C}, \ttt{D}, \ttt{E}, \ttt{F}, \ttt{G}, \ttt{H}, \ttt{I}, \ttt{K}, \ttt{L}, \ttt{M}, \ttt{N}, \ttt{O}, \ttt{P}, \ttt{R}, \ttt{S}, \ttt{T}, \ttt{U}, \ttt{V}, \ttt{W}, \ttt{X}, \ttt{Y}, \ttt{Z}, \ttt{a}, \ttt{b}, \ttt{c}, \ttt{d}, \ttt{e}, \ttt{f}, \ttt{g}, \ttt{h}, \ttt{i}, \ttt{k}, \ttt{l}, \ttt{m}, \ttt{n}, \ttt{o}, \ttt{p}, \ttt{r}, \ttt{s}, \ttt{t}, \ttt{u}, \ttt{v}, \ttt{0}, \ttt{1}, \ttt{2}, \ttt{3}, \ttt{4}, \ttt{5}, \ttt{6}, \ttt{7}, \ttt{8}, \ttt{9}, \ttt{+}, \ttt{-}, \ttt{.}, \ttt{\%}, \ttt{@}, \ttt{*}, \ttt{[}, \ttt{]}, \ttt{/}, \ttt{\char`\\}, \ttt{=}, \ttt{\#} \}.
 The internal structure of a string reflects a depth-first traversal of the corresponding molecular graphs.
@@
 Additionally, LL(1) grammars expand the leftmost non-terminal first.
-This set of features makes LL(1) grammars such as the one used by Dialect a good fit for manually-written recursive descent parsers.
+These features make LL(1) grammars such as the one used by Dialect a good fit for manually-written recursive descent parsers.
 LL(1) grammars can also be used as a basis for auto-generated parsers through packages such as ANTLR~\cite{parr:2014}.
@@
 The quantifier (\ttt{*}) indicates that a text in this language consists of a sequence of zero or more instances of the production \ttt{<a>}.
-This production in turn is defined as a lowercase letter (`a`).
-Therefore, valid texts in this language include the empty string, \ttt{a}, \ttt{aa}, and \ttt{aaaaaa} to name a few.
+This production in turn is defined as a lowercase letter (\enquote{a}).
+Therefore, valid texts in this language include the empty string, \enquote{a}, \enquote{aa}, and \enquote{aaaaaa} to name a few.
 
@@
 
-Atoms carry most of the information in a Dialect string. The non-terminal \ttt{<atom>} is a selection among four alternatives.
+Atoms carry most of the information in a Dialect string. The non-terminal \ttt{<atom>} can be expressed using one of four variants.
 
@@
 
-The first atomic non-terminal, \ttt{<star>}, can assume one terminal value, the asterisk character (\ttt{*}).
-This \enquote{star atom} represents an atom in which every attribute assumes its default value.
-The implicit hydrogen count of a star atom is zero.
+The first variant, \ttt{<star>}, produces one terminal value, the asterisk character (\ttt{*}). This \enquote{star atom} represents an atom in which every attribute assumes its default value.
 
@@
 
-The next atomic production rule, \ttt{<shortcut>} is a non-terminal selected from the list: \enquote{B}; \enquote{C}; \enquote{N}; \enquote{O}; \enquote{P}; \enquote{S}; \enquote{F}; \enquote{Cl}; \enquote{Br}; and \enquote{I.} An atom encoded in this way (\enquote{shortcut atom}) assigns the corresponding symbol to the \ttt{element} attribute.
-All other attributes retain their default values.
-The implicit hydrogen count of a shortcut atom is determined by the algorithm previously given.
+The next atomic production rule, \ttt{<shortcut>} is a non-terminal selected from the list: \enquote{B}; \enquote{C}; \enquote{N}; \enquote{O}; \enquote{P}; \enquote{S}; \enquote{F}; \enquote{Cl}; \enquote{Br}; and \enquote{I.} An atom encoded in this way (\enquote{shortcut atom}) receives the corresponding symbol as the value for the \ttt{element} attribute.
+All other attributes retain their default values, except \ttt{hydrogens}, which is set to \ttt{Implicit}.
 
@@
 
-The third atomic production rule, \ttt{<selection>} is a non-terminal selected from the list: \enquote{c}; \enquote{n}; \enquote{o}; \enquote{p}; and \enquote{s.} An atom encoded in this way (\enquote{selected shortcut atom}) assigns the corresponding atom symbol to the \ttt{element} attribute and sets the \ttt{selected} attribute to `true`.
+The third atomic production rule, \ttt{<selection>} is a non-terminal selected from the list: \enquote{c}; \enquote{n}; \enquote{o}; \enquote{p}; and \enquote{s.} An atom encoded in this way (\enquote{selected shortcut atom}) assigns the corresponding atom symbol to the \ttt{element} attribute, sets the \ttt{selected} attribute to \ttt{true}, and sets the \ttt{hydrogens} attribute to \ttt{Implicit}.
 All other atomic attributes retain their default values.
-The implicit hydrogen count of a selected shortcut atom is determined as described previously.
 
@@
 The fourth and most complex atomic production rule is \ttt{<bracket>} (\enquote{bracket atom}).
-A bracket atom can be used to set any atomic attribute.
-A bracket atom must be used when the `element` attribute of an atom makes it ineligible for implicit hydrogen counting.
+Bracket can be used to set any atomic attribute.
+A bracket atom must be used for any atom whose \ttt{configuration}, or \ttt{charge} attributes deviate from default values, or whose \ttt{hydrogens} attributes equals anything other than \ttt{Implicit}.
 Attributes not set within the \ttt{<bracket>} production rule will leave the corresponding atomic values in their default states.
@@
 The values of a bracket atom's \ttt{element} and \ttt{selected} attributes are determined by the \ttt{<symbol>} production rule (\enquote{symbol}).
-Three non-terminal options are available: \ttt{<star>}; \ttt{<element>}; and \ttt{<selection>}.
+Three non-terminal variants are supported.
+These three variants influence the \ttt{element} and \ttt{selected} attributes in different ways.
+The \ttt{<star>} variant leaves both attributes as their default values.
+The \ttt{<element>} variant assigns the \ttt{element} attribute to the corresponding value while leaving the \ttt{selected} attribute as its default value.
+The \ttt{<selection>} non-terminal sets the \ttt{element} attribute to the corresponding element and the \ttt{selected} attribute to \ttt{true}.
+For example, the selection sequence \enquote{p} would assign the atomic \ttt{element} and \ttt{selected} attribute to \ttt{P} and \ttt{true}, respectively.
 
@@
 
-The \ttt{<element>} non-terminal option assigns the atomic \ttt{element} attribute to the corresponding value while leaving the \ttt{selected} attribute as its default value.
-Given the large number of choices within the \ttt{<element>} non-terminal, the following production rule only defines the first several.
-For a complete list of alternative, see the full grammar in the Supporting Information.
+Given the large number of variants within the \ttt{<element>} non-terminal, the following production rule only defines the first several.
+For a complete list of variants, see the full grammar in the Supporting Information.
 
@@
 
-A \ttt{<selection>} non-terminal found within a symbol sets the atomic \ttt{selected} attribute to \ttt{true}.
-The corresponding \ttt{element} attribute is assigned by capitalization.
-For example, the terminal \ttt{p} would assign the atomic \ttt{element} and \ttt{selected} attributes to the values \ttt{P} and \ttt{true}, respectively.
-
 The \ttt{stereodescriptor} attribute of a bracket atom is determined by the \ttt{<stereodescriptor>} non-terminal.
-Allowed values are \enquote{@} and \enquote{@@,} corresponding to the \ttt{stereodescriptor} values \ttt{Counterclockwise} and \ttt{Clockwise}, respectively.
+Allowed values are \enquote{@} and \enquote{@@,} corresponding to the values \ttt{Left} and \ttt{Right}, respectively.
 
@@
 The \ttt{hydrogens} attribute of a bracket atom is controlled by the \\ \ttt{<virtual{\_}hydrogen>} non-terminal.
-This non-terminal is comprised of the terminal \enquote{H} followed by an optional \ttt{<digit>} non-terminal.
+It is comprised of the terminal \enquote{H} followed by an optional \ttt{<digit>} non-terminal.
 A digit appearing after the \enquote{H} terminal sets \ttt{hydrogens} to the corresponding virtual hydrogen count.
@@
 The \ttt{<charge>} production rule sets the \ttt{charge} attribute of a bracket atom.
-This non-terminal begins with either the plus or minus terminals (\enquote{+} and \enquote{-,} respectively) and ends with an optional \ttt{<digit>} non-terminal.
+This non-terminal begins with either the plus or minus terminals (\enquote{+} and \enquote{-} respectively) and ends with an optional \ttt{<digit>} non-terminal.
 A missing digit causes the atomic \ttt{charge} attribute to be set to one.
@@
 
-Should an optional bracket non-terminal not be present, the atom must retain its corresponding default value.
-For example, the bracket \enquote{[C@H+]} lacks the \ttt{<isotope>} non-terminal so the value of the \ttt{isotope} attribute will remain as \ttt{None}.
-Similarly, the bracket \enquote{[13CH+]} lacks the \ttt{<stereodescriptor>} non-terminal, so the corresponding \ttt{configuration} attribute remains \ttt{None}.
+Values not defined within a bracket production leave the atomic attributes of the atom built from it in as their default values.
+For example, the bracket sequence \enquote{[C@H+]} lacks the \ttt{<isotope>} non-terminal so the value of the \ttt{isotope} attribute will remain as \ttt{None}.
+Similarly, the bracket sequence \enquote{[13CH+]} lacks the \ttt{<stereodescriptor>} non-terminal, so the corresponding \ttt{configuration} attribute remains \ttt{None}.
 
@@
 \begin{lstlisting}
-<bond> ::= "-" | "=" | "#" | | "/" | "\"
+<bond> ::= "-" | "=" | "#" | "/" | "\"
 \end{lstlisting}
@@
 The \ttt{<union>} non-terminal consists of an optional \ttt{<bond>} non-terminal followed by a mandatory non-terminal selected from the list: \ttt{<cut>} or \ttt{<sequence>}.
-If either of these latter non-terminals are detected but \ttt{<bond>} is not, bond elision is used.
+If either of these latter non-terminals are detected but \ttt{<bond>} is not, the resulting bond is elided.
 
@@
 Double- or single-digit cut indexes, supporting the values zero through 99 inclusive, are available by prepending the percent character (\ttt{\%}).
+For the latter variant, any leading zero is disregarded, meaning that the sequences "07" and "7" are both interpreted as the index seven.
 
@@
 Like union, branch joins a parent and child node through a bond.
-Wrapped by opening and closing parenthesis terminals (\enquote{(} and \enquote{),} respectively), branch encodes a sequence that may or may not be attached to its parent.
+Wrapped by opening and closing parenthesis terminals (\enquote{(} and \enquote{)} respectively), branch encodes a sequence that may or may not be attached to its parent.
 Attachment occurs if the \ttt{<bond>} non-terminal is included.
-Alternatively, the subgraph will be detached if the \ttt{<detachment>} non-terminal appears.
-If neither \ttt{<bond>} nor \ttt{<detachment>} non-terminals are detected between the parentheses but a sequence is, then bonding of parent and child occurs through an elided bond.
+Alternatively, the sequence will be detached if the \ttt{<detachment>} non-terminal (\enquote{.}) appears.
+If neither \ttt{<bond>} nor \ttt{<detachment>} are present then bond to the first atom of the sequence is elided.
 
@@
 
-The third option for associating a parent Atom with a child within a sequence is the \ttt{<split>} non-terminal (\enquote{split}).
-A split is a sequence of \ttt{<detachment>} and \ttt{<sequence>} non-terminals.
-The `<detachment>` non-terminal serves the same purpose as it does within a branch: the prevent the bonding of the parent atom with a child.
+The third option for adding atoms within a sequence is the \ttt{<split>} non-terminal (\enquote{split}).
+A split is a sequence of \ttt{<detachment>} and \ttt{<sequence>} productions.
+The \ttt{<detachment>} non-terminal serves the same purpose as it does within a branch: to prevent the bonding to the first atom of the sequence.
 
@@
 In other words, a Dialect string is either empty or contains a sequence.
-A string without a sequence encodes a molecular graph of zero nodes and zero edges.
+A string without a sequence encodes a molecular graph having zero nodes and zero edges.
 

### 2022-05-22 manuscript.tex
Reflows Reading Strings.

@@
 The second occurrence of a cut takes the entry at the cut index, attaching the atom in the entry to the current root atom.
-Either the leftmost or rightmost bond can be used to make the connection, provided they are \textit{compatible}.
+The two bonds spanning a cut must be \textit{compatible}.
+Compatibility is determined by comparing attributes using the following rules.
+If no rule applies to both bonds they are not compatible.
+It is an error to span a cut with incompatible bonds.
+
+\begin{enumerate}
+    \item An elided bond is compatible with any other bond, regardless of order or direction.
+    \item A bond with a defined direction is compatible with bond having the opposite direction if \ttt{source} precedes \ttt{target} (the usual case), or the same direction otherwise.
+    \item A bond without a direction is compatible with any bond having the same order.
+\end{enumerate}
 
@@
 
-Two cut bonds are compatible if they possess the same \ttt{order} attributes, and their \ttt{direction} attributes are complementary.
-Bonds with undefined \ttt{direction} are always complementary.
-A bond with a defined \ttt{direction} is complementary to any bond with an undefined state, or a bond with an opposing \ttt{direction}.
-For example, an elided bond is compatible with a single bond.
-Likewise, two elided bonds are compatible.
-A bond with a \ttt{direction} of \ttt{Up} is incompatible with another bond with a \ttt{direction} of \ttt{Up}, but compatible with one having a \ttt{direction} of \ttt{Down}.
-
 \begin{figure}
@@
 
-Decoding atomic configuration from a string requires the order of attachment from each atom to its neighbors to be recorded.
-Cuts are treated as if the neighboring atom itself were detected.
-In this regard, the first member of a cut pair requires special attention.
-The neighboring atom won't be available until the paired cut index is found.
-Nevertheless, it must ultimately appear before the neighbors in subsequent unions, branches, and splits.
-This can be accomplished though the use of a placeholder or some other mechanism that preserves the overall relative order of attachment.
-No such special treatment is needed for the second member of a cut pair.
+One further complication applies to decoding cuts.
+Atomic configuration is defined in terms of the relative order of attachment of neighbors about a central atom.
+In the first bond of a cut pair, the neighbor atom is undefined.
+It won't be known until the second bond of the cut pair is found.
+This might only occur after the addition of several intervening neighbors.
+Nevertheless, the neighbor atom across the cut must precede any subsequent neighbors.
+This requirement can be met by a placeholder or some other mechanism that preserves the relative order of neighbor attachment.
+No such special treatment is needed for the second bond of a cut pair.
 
@@
 This assumption is most likely to arise in the context of ad-hoc parsers using regular expressions, string matching, and the like.
-For example, the single molecule propane can be encoded using the string \enquote{C1C.C1}.
-
-The partial parity bond model is unusual in that it spreads conformational information over three or more bonds.
-Regardless, readers must report as an error any string containing an isolated PPB (e.g., \enquote{C/C}).
+For example, the connected molecular graph for propane can be encoded using the string \enquote{C1C.C1}.
+
+The use of PPBs conformation over three or more bonds.
+Readers must therefore report as an error any PPB not connected to an atom connected through a double bond.
+For example, an isolated PPB is invalid (e.g., \enquote{C/C} and \enquote{C/CC}).
 Beyond this check, generating (\textit{E}) and (\textit{Z}) stereodescriptors from PPBs is not a straightforward process.
-Performing such a conversion requires a method capable of translating PPB into a localized bond descriptor.
 One approach is to first consider the parity of each double bond terminal individually.

### 2022-05-25 manuscript.tex
Reflows Writing Strings.

@@
 \usepackage[USenglish]{babel}
+\usepackage{nameref}
 
@@
 \section*{Reading Strings}
+\label{reading-strings}
 
@@
 
-The presence of branches within an input data structure is encoded via the \ttt{<branch>} data structure.
+The presence of branches within an input data structure is encoded via the \ttt{<branch>} non-terminal.
 A useful tool for this purpose is a \ttt{stack}.
@@
 The presence of conformational restriction about a double bond presents a special challenge to writers.
-Few other molecular graph systems distribute a conformation descriptor over two or more bonds the way that PPBs does.
+Few other molecular graph systems distribute a conformation descriptor over multiple individual bonds the way that PPBs does.
 Instead, descriptors are more likely to be localized at the double bond.
 Producing the necessary PPBs from a descriptor localized at the double bond is likely to be non-trivial.
-An approach to the reverse problem was presented previously (Reading Strings section).
+An approach to the reverse problem was presented previously (\nameref{reading-strings} section).
 That approach can also be adapted to the reverse problem.
@@
 Writers should carefully weigh the non-negligible costs of atom selection.
-Algorithms for doing so sometimes involve the perceptions of cycles, and often exhibit superlinear time complexity.
-Very often a reader must perform a global deselection to arrive at a localized valence bond representation, which at the very least requires pruning and a maximal matching procedure.
+Algorithms for selection are likely to involve the perceptions of cycles, and could exhibit superlinear time complexity.
+Often, a reader must perform a global deselection to arrive at a localized valence bond representation, which at the very least requires pruning and a maximal matching procedure.
 In this sense atom selection imposes two sets of costs: one on the writer and the other on every subsequent reader forever into the future.
@@
 Consider pyrrole, erroneously encoded with a selected nitrogen atom (e.g., \enquote{n1cccc1}).
-The nitrogen atom can not be pruned because its subvalence is 1 (3 - 2).
+The nitrogen atom can not be pruned because of its subvalence is 1 (3 - 2).
 The DS therefore contains all five atoms and all five bonds.

### 2022-05-25 manuscript.tex
Refloaw Compatibility.

@@
 For maximum compatibility with existing software, Dialect was designed as a \textit{language subset} of SMILES.
-A language subset contains all of the syntax and semantics of its parent language, but adds none of its own.
+A language subset contains some of the syntax and semantics of its parent language, but adds none of its own.
 In principle this means that a feature present in Dialect should also be present in SMILES.
@@
 Language subsets arise for two main reasons.
-First, designing a language presents many opportunities for errors small and large.
+First, designing a language presents many opportunities to introduce errors small and large.
 Second, popular languages inevitably bring with them many users with diverse problems that demand new features.
@@
 The only remaining sources are The Paper and the Book Chapter.
-These are the two works that collectively define ProtoSMILES.
+Therefore, these two works collectively define ProtoSMILES.
 The remainder of this section compares and contrasts ProtoSMILES with Dialect.
@@
 
-The Book Chapter defines a recursive grammar for branching which, perhaps without the knowledge of the author, defines sequential parentheses~\cite[p.~86]{weininger:2003}.
+The Book Chapter defines a recursive grammar for branching which, perhaps without the knowledge of the author, allows sequentially nested parentheses~\cite[p.~86]{weininger:2003}.
 Dialect disallows such constructs (e.g., \enquote{*((*))*} and \enquote{*(((*)))*}).
@@
 ProtoSMILES places no upper bound on the charge notation (e.g., \enquote{++} and \enquote{+++}) and their corresponding negative forms: \enquote{The SMILES program also recognizes constructions of the form [Fe+++] as being synonymous with the form [Fe+3]}~\cite[p.~32]{weininger:1988}.
-Due to their redundancy, Dialect disallows duplicated plus and minus characters (\ttt{+} and \ttt{-}, respectively) from bracket atoms.
+Due to their redundancy, Dialect disallows duplicated plus and minus characters (\ttt{+} and \ttt{-}, respectively) within bracket atoms.
 
@@
 A detailed algorithm for determining implicit hydrogen count is provided, together with the required valence table.
-Unlike ProtoSMILES, Dialect explicitly considers the case of computing implicit hydrogens on selected atoms.
+Unlike ProtoSMILES, Dialect explicitly considers the case of computing implicit hydrogen counts on selected atoms.
 Dialect also provides detailed algorithms, absent in SMILES, for selection and deselection.
@@
         \hline
-        Feature & ProtoSMILES\footnote{Values of \enquote{maybe} indicate ambiguous source descriptions.} & Dialect \\
+        Feature & ProtoSMILES\footnote{Values of \enquote{maybe} indicate ambiguous and/or contradictory statements in the authoritative sources.} & Dialect \\
         \hline
@@
 If ProtoSMILES support tends toward \enquote{no,} then support by Dialect would break out of the ProtoSMILES subset.
-There are three such cases.
-The first is hydrogen having virtual hydrogens (e.g., \enquote{[HH]}).
-The second ambiguous case is whether or not the default valence of nitrogen includes the value five.
-The third case is whether or not ProtoSMILES recognizes acyclic \enquote{aromatic} atoms as valid.
+There are three such cases: hydrogen having virtual hydrogens (e.g., \enquote{[HH]}); whether or not the default valence of nitrogen includes the value five; and whether or not ProtoSMILES recognizes acyclic \enquote{aromatic} atoms as valid.
 

### 2022-05-25 manuscript.tex
Reflows Conclusion.

@@
 The line notation InChI might appear to compete with Dialect in this sense.
-Although InChI can be read and written, the InChI Technical manual notes that InChI's value lies in molecular identification.
-The authors of InChI have repeatedly noted that InChI is not a solution to the molecular serialization problem.
-The reason presumably lies with the fact that both serialization and deserialization require intimate knowledge of the InChI canonicalization algorithm.
+The authors of InChI have noted that InChI is not a solution to the molecular serialization problem but rather an identifier.\cite{heller:2015} The reason presumably lies with the fact that both serialization and deserialization require intimate knowledge of the InChI canonicalization algorithm.
 To date, no third-party reader or writer of InChI has ever been released.
 
-Despite its information density, Dialect can faithfully encode and decode most of what chemists would consider \enquote{organic molecules.} This is evidenced by the widespread presence of Dialect strings within large, public-facing databases such as PubChem, ChEMBL, ChEBI, and eMolecules.
+Despite its information density, Dialect can faithfully encode and decode most of what chemists would consider \enquote{organic molecules.} This is evidenced by the widespread presence of Dialect strings within large, public-facing databases such as PubChem,\cite{kim:2016} ChEMBL,\cite{gaulton:2012} ChEBI,\cite{degtyarenko:2007} and others.
 
@@
 Broader expansion of Dialect's capabilities could be possible through \textit{metaformats}.
-A metaformat embeds one or more Dialect strings within a surrounding serialization format.For example, two- or three-dimensional coordinates could be associated with each atom through a metaformat that includes a dictionary mapping atomic index to coordinate.
+A metaformat embeds one or more Dialect strings within a surrounding serialization format.
+For example, two- or three-dimensional coordinates could be associated with each atom through a metaformat that includes a dictionary mapping atomic index to coordinate.
 Collections of atoms or bonds could likewise be encoded to replicate the enhanced stereochemistry features of other formats.
@@
 The availability of a minimal yet highly functional, fully-specified core language offers many opportunities to improve data quality.
-One of the most important will be a reference implementation, the design and source code of which will be discussed later.
+One of the most important will be an open reference implementation, the design and source code of which will be discussed later.
 A freely-available reference implementation in turn makes automated validation suites possible.
@@
 
-Dialect's syntax is described in detail through both graphical diagrams and a formal grammar.
-The latter method concisely summarizes the complete set of strings that could be considered valid Dialect representations.
+Dialect's syntax is described in detail through a formal grammar.
+This method concisely summarizes the complete set of strings that could be considered syntactically valid Dialect representations.
 The formal grammar was deliberately structured to be used either directly with an automated parser generator or with a hand-written recursive-descent parser.

### 2022-05-25 manuscript.tex
Excludes star atoms from DS. closes #101.

@@
 For an atom to be selectable, its \ttt{element} attribute must be contained in the list: \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; and \ttt{S}.
-Atoms having other elements (or \ttt{None}) must not be selected.
+Atoms having other elements must not be selected.
 A bond will be added to the DS if both of its terminals are selected and the \ttt{elided} attribute equals \ttt{true}.
@@
 This rule allows arbitrary bonds between selected atoms to be excluded from the DS.
+
+Atoms whose \ttt{element} attribute equals \ttt{None} are not selectable, which prevents their entry into the DS.
+Bonds to such atoms are similarly excluded.
+The alternative, conditional inclusion, would complicate both data structures and algorithms for little gain.
+Exclusion avoids these complexities.
 

### 2022-05-25 manuscript.tex
Adds isotope to Table 2. closes #102
Adds reference for LL(1) grammar. closes #103

@@
     \ttt{charge} & Formal charge & \ttt{SignedTen} & 0 \\
+    \ttt{isotope} & Mass number & \ttt{Option<Thousand>} & \ttt{None} \\
     \ttt{selected} & Selection status & \ttt{boolean} & \ttt{false} \\
@@
 
-Strings conform to an \textit{LL(1) grammar}.
-An LL(1) grammar is a context-free grammar whose strings can be parsed one character at a time from left to right with at most one character of lookahead.
+Strings conform to an \textit{LL(1) grammar}.\cite{thain:2020} An LL(1) grammar is a context-free grammar whose strings can be parsed one character at a time from left to right with at most one character of lookahead.
 Additionally, LL(1) grammars expand the leftmost non-terminal first.

### 2022-05-25 manuscript.tex
Rephrases table entry for clarity. closes #106

@@
         detachments are bonds of \enquote{formal order zero} & maybe & no \\
-        upper and lower bounds on numerical atomic properties & no & yes \\
+        unbounded numerical atomic properties & yes & no \\
         nitrogen default valence includes 5 & maybe & yes \\

### 2022-05-25 manuscript.tex
Drops support for encoding zero for hydrogens, charge, and isotope attributes. closes #104

@@
     \hline
-    \ttt{Option<Type>} & Optional Value & \ttt{None}, \ttt{Type} \\
+    \ttt{Option<T>} & Optional Value & \ttt{None}, \ttt{T} \\
     \ttt{Index} & Unsigned integer & 0..2\textsuperscript{32} \\
     \ttt{Three} & Integer count & 1..4 \\
-    \ttt{Ten} & Unsigned integer & 0..10 \\
-    \ttt{SignedTen} & Integer count & -9..10 \\
-    \ttt{Thousand} & Integer count & 0..1000 \\
+    \ttt{Dec} & Integer count &0..10 \\
+    \ttt{SignedDec} & Integer count & -9..10 \\
+    \ttt{Thou} & Integer count & 1..1000 \\
     \ttt{Element} & IUPAC-\-approved element symbol & \ttt{Ac}, \ttt{Ag}, \ttt{Al}, {\ldots}\ttt{Zn}\\
-    \ttt{HydrogenCount} & Hydrogen count & \ttt{Implicit}, \ttt{Ten} \\
     \ttt{Configuration} & Configurational descriptor & \ttt{Right}, \ttt{Left} \\
     \ttt{Direction} & Conformational descriptor & \ttt{Up}, \ttt{Down} \\
+    \ttt{Implicit} & Implicit hydrogen count flag & - \\
     \ttt{boolean} & Boolean & \ttt{true}, \ttt{false} \\
@@
     \ttt{element} & Elemental symbol & \ttt{Option} \ttt{<Element>} & \ttt{None} \\
-    \ttt{hydrogens} & Hydrogen count & \ttt{HydrogenCount} & 0 \\
-    \ttt{configuration} & Configuration & \ttt{Configuration} & \ttt{None} \\
-    \ttt{charge} & Formal charge & \ttt{SignedTen} & 0 \\
-    \ttt{isotope} & Mass number & \ttt{Option<Thousand>} & \ttt{None} \\
+    \ttt{hydrogens} & Hydrogen count & \ttt{Implicit}, \ttt{Dec} & 0 \\
+    \ttt{configuration} & Configuration & \ttt{Option} \ttt{<Configuration>} & \ttt{None} \\
+    \ttt{charge} & Formal charge & \ttt{SignedDec} & 0 \\
+    \ttt{isotope} & Mass number & \ttt{Option<Thou>} & \ttt{None} \\
     \ttt{selected} & Selection status & \ttt{boolean} & \ttt{false} \\
@@
 Setting the \ttt{isotope} property to \ttt{None} means that the element's isotopic composition equals natural abundance.
-Otherwise, the lower bound on the value of \ttt{isotope} is zero.
-This lower limit allows for such physically nonsensical states such as atoms of zero implied mass (\ttt{isotope} equals zero) or negative implied neutron count (e.g., carbon-5).
+Otherwise, the lower bound on the value of \ttt{isotope} is one.
+This lower limit allows for such physically nonsensical states such as atoms of negative implied neutron count (e.g., carbon-5).
 Implementations may reject such atomic states as invalid.
@@
 The value of a bracket atom's \ttt{isotope} attribute is determined by the optional \ttt{<isotope>} non-terminal.
-It consists of between one and three digits encoding the integers 0-999.
-Leading zeros are disregarded when assigning the value of the \ttt{isotope} attribute.
-In other words, \enquote{007} and \enquote{7} are considered equivalent expressions of the value 7.
-
-\begin{lstlisting}
-<isotope> ::= <digit> <digit>? <digit>?
+It consists of between one and three digits encoding the integers 1-999.
+Leading zeros (e.g., \enquote{007} are disallowed.
+
+\begin{lstlisting}
+<isotope> ::= <nonzero> <digit>? <digit>?
 \end{lstlisting}
@@
 \begin{lstlisting}
-<virtual_hydrogen> ::= "H" <digit>?
+<virtual_hydrogen> ::= "H" <nonzero>?
 \end{lstlisting}
@@
 In other words, the plus terminal (\enquote{+}) sets the \ttt{charge} attribute to one and the minus terminal (\enquote{-}) sets it to minus one.
-A charge of zero (e.g., \enquote{+0} or \enquote{-0}) sets the \ttt{charge} attribute to zero.
-
-\begin{lstlisting}
-<charge> ::= ( "+" | "-" ) <digit>?
+Charge values of zero are obtained if \ttt{<charge>} production is present.
+
+\begin{lstlisting}
+<charge> ::= ( "+" | "-" ) <nonzero>?
 \end{lstlisting}
@@
 The \ttt{<cut>} non-terminal (\enquote{cut}) can takes two forms.
-A digit non-terminal can be used, enabling single-digit cut indexes.
-Double- or single-digit cut indexes, supporting the values zero through 99 inclusive, are available by prepending the percent character (\ttt{\%}).
-For the latter variant, any leading zero is disregarded, meaning that the sequences "07" and "7" are both interpreted as the index seven.
-
-\begin{lstlisting}
-<cut> ::= <digit> | "%" <digit> <digit>
+A digit non-terminal can be used, enabling single-digit cut indexes in the range one through nine, inclusive.
+Double- or single-digit cut indexes, supporting the values ten through 99 inclusive, are available by prepending the percent character (\ttt{\%}).
+For the latter variant, leading zeros are disallowed, meaning that the sequence \enquote{\%07} must generate an error.
+
+\begin{lstlisting}
+<cut> ::= <nonzero> | "%" <nonzero> <digit>
 \end{lstlisting}

### 2022-05-27 manuscript.tex
Re-works Goals to use molecular tree.

@@
 
-\section*{Goals, Design, and Tradeoffs}
+\section*{Goals and Overall Design}
 
@@
 
-Dialect molecules inherit behavior from an underlying \textit{molecular graph}~\cite{balaban:1985}.
-A graph is a data structure comprised of a set of nodes and a set of pairwise relationships between them called edges.
-A molecular graph is a specialized graph onto which chemically-relevant metadata have been overlaid.
+Dialect molecules inherit behavior from an underlying \textit{molecular tree}.
+A tree is a data structure comprised of nodes.
+Each node is associated with zero or more child nodes through a directed edge.
+A molecular tree is a specialized tree onto which chemically-relevant metadata have been overlaid.
 The presence of this overlay allows nodes and edges to represent atoms and bonds, respectively.
-Molecular graphs are ubiquitous in cheminformatics, appearing in both serialization formats in-memory data structures found in software toolkits.
-
-A Dialect molecule is a molecular graph having zero or more nodes and zero or more edges.
+
+A molecular tree is related to the more widely-used concept of a \textit{molecular graph}~\cite{balaban:1985}.
+There are two main difference: (1) bonds are undirected in molecular graphs but directed in molecular trees; and (2) cycles are possible in molecular graphs but not in molecular trees.
+The difference might appear to render trees unsuitable to represent molecules, in which cycles often play a prominent structural role.
+But as the next sections demonstrate, the differences between molecular trees and graphs are not so large as they might at first appear.
+
+A Dialect molecule is a molecular tree having zero or more nodes and zero or more edges.
 The empty molecule, devoid of atoms and bonds, is therefore allowed.
-Dialect imposes an upper bound on the number of nodes of $2^{32} - 1$ and the same limit for the number of edges.
-However, practical limitations related to memory, storage, and CPU time will likely further constrain these limits in practice.
-
-The edges of a Dialect molecular graph are defined as a set of unordered pairs of non-identical nodes.
-The practical consequences of this are threefold.
-First, \textit{loops} are not allowed.
-A loop is an edge joining a node to itself.
-Second, edges are \textit{undirected}.
-An undirected edge is equivalent to another edge whose terminals are transposed.
-With one exception that will be discussed in detail, an undirected edge between nodes A and B is therefore identical to an undirected edge between nodes B and A.
-Third, \textit{parallel edges} are disallowed.
-Edges are said to be parallel if they connect the same two nodes.
-
-A molecular graph contains zero or more \textit{connected components}.
-A connected component is a graph in which any two vertices are connected to each other by at least one path.
-The maximum number of connected components is equal to the maximum node count, although practical factors may reduce this limit.
+Dialect imposes no upper bound on the number of nodes or edges, although limitations in hardware and software may impose a practical upper bound on molecular size.
 
@@
 Dialect uses several abstractions collectively known as the \textit{data model}.
-The data model defines metadata attributes associated with the components of a molecular graph.
+The data model defines metadata attributes associated with the components of a molecular tree.
 These attributes are themselves composed of \textit{primitives} (Table~\ref{table:primitives}), or irreducible data types.
+
+% TODO: completely re-work this to fit with the exact model being used by the reference implementation
 
@@
     \ttt{Option<T>} & Optional Value & \ttt{None}, \ttt{T} \\
-    \ttt{Index} & Unsigned integer & 0..2\textsuperscript{32} \\
     \ttt{Three} & Integer count & 1..4 \\
@@
     \hline
-    \ttt{index} & Unique identifier &\ttt{Index} & - \\
     \ttt{element} & Elemental symbol & \ttt{Option} \ttt{<Element>} & \ttt{None} \\
@@
 \begin{enumerate}
-    \item{The value of \ttt{index} must be unique over the molecular graph.}
     \item{If the value of \ttt{hydrogens} equals \ttt{Implicit}, then the values of \ttt{isotope}, \ttt{configuration}, and \ttt{charge} must equal their respective defaults.
@@
     \hline
-    \ttt{index} & Unique identifier & \ttt{Index} & - \\
-    \ttt{source} & Source atom index & \ttt{Index} & - \\
-    \ttt{target} & Target atom index & \ttt{Index} & - \\
     \ttt{order} & Bond order & \ttt{Three} & 1 \\
@@
     \ttt{direction} & Partial conformational parity & \ttt{Option} \ttt{<Direction>} & \ttt{None} \\
+    \ttt{target} & Child atom & \ttt{Atom} & - \\
     \hline
@@
 \begin{enumerate}
-    \item{The value of \ttt{index} must be unique over the molecular graph.}
-    \item{The \ttt{source} and \ttt{target} attributes must reference atomic \ttt{index} attributes found elsewhere in the molecular graph.}
     \item{If the value of \ttt{direction} does not equal \ttt{None}, then the value of \ttt{order} must equal one.}
@@
 
-A molecular graph is described at the lowest level by its constitution.
-In Dialect, constitution comprises the molecular graph and those attributes locating nucleons and electrons at specific atoms and bonds.
+A molecular tree is described at the lowest level by its constitution.
+In Dialect, constitution comprises those attributes locating nucleons and electrons at specific atoms and bonds.
 All other attributes lie outside the scope of constitution.
-Atomic attributes related to constitution are: \ttt{index}; \ttt{element}; \ttt{hydrogens}; and \ttt{charge}.
-Bond attributes related to constitution are \ttt{source}, \ttt{target}, \ttt{order}, and \ttt{elided}.
+Atomic attributes related to constitution are: \ttt{element}; \ttt{hydrogens}; and \ttt{charge}.
+Bond attributes related to constitution are \ttt{order} and \ttt{elided}.
 

### 2022-05-28 manuscript.tex
Eliminates indexes on Atom and Bond. closes #107

@@
     \ttt{SignedDec} & Integer count & -9..10 \\
+    \ttt{Cent} & Integer count & 1..100 \\
     \ttt{Thou} & Integer count & 1..1000 \\
@@
     \ttt{direction} & Partial conformational parity & \ttt{Option} \ttt{<Direction>} & \ttt{None} \\
-    \ttt{target} & Child atom & \ttt{Atom} & - \\
     \hline
@@
 A molecular tree is described at the lowest level by its constitution.
-In Dialect, constitution comprises those attributes locating nucleons and electrons at specific atoms and bonds.
+In Dialect, constitution comprises those attributes that associate atoms in bonding relationships, and which locate nucleons and electrons at specific atoms and bonds.
 All other attributes lie outside the scope of constitution.
@@
 
-Each atom carries an \ttt{index} attribute, allowing bonds to reference atoms in a molecular graph.
-To do so, the \ttt{source} and \ttt{target} attributes are set with the respective indexes of the source and target atoms.
-The set of all bonds and the nodes they reference determines the connectivity of a molecular graph.
-In the context of determining connectivity, source and target indexes may be reversed.
+\section*{Node Order}
+
+Certain aspects of Dialect rely on \textit{node order}.
+An order is a property of a set that allows its members to be sorted through pairwise comparisons.
+Node order is the order in which the nodes of a molecular tree are added during depth-first construction.
+A node added before another node is said to \enquote{precede} it.
+A node added after another node is said to \enquote{succeed} it.
+Using node order, the nodes of a molecular tree can be sorted from first to last.
+
+\begin{figure}
+    \centering
+    \includegraphics{filler}
+    \caption{Node Order. Nodes in a molecular tree have the order in which they were added through depth-first construction.}
+    \label{fig:node-order}
+\end{figure}
+
+Some generalizations about node order follow from these definitions.
+The root atom, uniquely lacking a parent, precedes every other node.
+Every parent node precedes its children.
+The first child added to a given parent precedes all of its siblings.
+Moreover, this relationship is transitive.
+If a node precedes (or succeeds) a sibling, then all of its descendants will as well.
+
+\section*{Decoupling Bonding from Connectivity}
+
+A bond performs two important tasks within a molecular tree: it creates a bond between two atoms and it creates a connection between two nodes.
+Most of the time, performing both tasks together is exactly what's needed.
+In two cases, however, doing so limits the range of molecules that can be encoded.
+
+The first case for decoupling involves cycles.
+A molecular tree is by definition devoid of cycles, yet cycles are ubiquitous, crucial elements of molecular structure.
+To encode cycles, a molecular tree needs a way unambiguously record the presence of cycle-closure bonds without creating the corresponding connection.
+Dialect accomplishes this goal through \textit{cut bonding}.
+
+\begin{figure}
+    \centering
+    \includegraphics{filler}
+    \caption{Cut Bonding. A bonding relationship represented by two nodes and two edges.}
+    \label{fig:cut-bonding}
+\end{figure}
+
+Cut bonding represents a bond as two semi-independent node/bond pairs.
+These pairs are distinguished from each other with the designations \enquote{left} and \enquote{right.} The left cut node and its parent precede the right cut node and its parent.
+Left and right side cut nodes are attached to their respective atom parents through equivalent bonds, but not to each other.
+In this way a bonding arrangement is established without creating a connection.
+Unlike an atom, a cut node has zero children and must have an atomic parent.
+As such a cut node can not serve as the root of a molecular tree.
+
+Each cut node is associated with a single attribute, \ttt{index}.
+The value of this attribute ranges from one to 99, inclusive (\ttt{Cent} type).
+Left and right cut nodes identify themselves as a pair by using identical \ttt{index} attributes.
+A single molecular tree can support far more than 99 cuts through index reuse.
+Any left cut node succeeding the last right cut node having a particular index can reuse it.
+The next succeeding cut node with the index must by definition be the matching right hand side.
+
+\begin{figure}
+    \centering
+    \includegraphics{filler}
+    \caption{Index reuse. This allows for many more than 99 cuts to be used within a single molecular tree.}
+    \label{fig:index-reuse}
+\end{figure}
+
+The second case for decoupling the formation of bonds and connections is multiple molecules.
+Dialect supports these uses through \textit{split edges}.
+A split edge allows two atoms to be connected within a tree, but without creating a bonding relationship.
+Split edges can be used whenever multiple molecules need to be encoded within the same molecular tree.
+One application would be molecules with ionic bonding.
+The atoms are not bound covalently, and so fall outside the scope of the VB model.
+Another application would be non-ionic complexes.
+No bonding relationship whatsoever is implied between atoms connected through a split edge.
+
+\begin{figure}
+    \centering
+    \includegraphics{filler}
+    \caption{Split Edge. Connection without a bonding relationship.}
+    \label{fig:split-edge}
+\end{figure}
+
+Split edges and cut bonding accomplish the same goal, namely the decoupling of connectivity and bonding, but for opposite purposes.
+Whereas split edges connect two atoms without bonding them, cut bonds create a bonding relationship between two nodes without connecting them.
+This decoupling complicates the analysis of molecular trees.
+For example, the atom succeeding a split edge behaves in some situations as if it were a root node.
+When tallying the bonds surrounding such a node, only children are considered.
+Likewise, cut nodes behave as if they were atoms in the sense that they can be associated with a bond.
+Within these and several other situations, analysis of bonding and connectivity relationships within molecular trees need to explicitly account for the possibility of connection-free bonds and bond-free connections.
 
@@
 
-\section*{Atom Index}
-
-To support the assignment and interpretation stereochemical features, each node carries a unique, non-negative integer \ttt{index} attribute (\enquote{index}).
-Assignment is sequential, meaning that the first atom gets index zero, the next one, and so on.
-For example, the index for the first atom is zero, for the second atom one, for the third atom two, and so on.
-The maximum value of an atomic index is 2\textsuperscript{32} - 1.
-
-Indexes impose an ordering over the atoms in a molecular graph.
-An atom with an index less than another atom is said to \textit{precede} it.
-An atom with an index grater than another atom \textit{succeeds} it.
-No special significance is ascribed to the atom whose index is zero, except that every other atom succeeds it.
-
 \section*{Conformation}
@@
 To support this system, each bond carries an optional \ttt{direction} attribute (\enquote{direction}).
-When present, direction may assume one of the two values \ttt{Up} or \ttt{Down}.
-These values refer to a geometrical model in which the two terminals of a double bond and their neighbors are assigned local relative coordinates in a plane.
-The terminals of the bond are placed on the x-axis such that the one on the right succeeds the one on the left.
-The neighbors at each terminal joined through a PPB are then assigned relative coordinates in the plane.
-
-Neighbors of the left terminal are positioned according to the relative ordering of atomic indexes and the bond's \ttt{direction} attribute.
-If the neighbor precedes the terminal then a \ttt{direction} attribute equal to \ttt{Up} locates the neighbor in the lower-left quadrant and a \ttt{direction} attribute of \ttt{Down} locates the neighbor in the upper-left quadrant.
-If the neighbor succeeds the terminal, these placements are reversed.
-
-An analogous procedure assigns relative coordinates to a neighbor of the right terminal.
-If a neighbor succeeds the right terminal, place it to the upper right given an \ttt{Up} direction, or to the lower right given a \ttt{Down} direction.
-Reverse these assignments if the neighbor precedes the right terminal.
+When defined, direction may assume one of two values: \ttt{Up} or \ttt{Down}.
+These values refer to a geometrical model in which a parent node and its child are placed along a vertical axis.
+If the bond between parent and child has the direction \ttt{Up}, then the child is moved in the upward direction.
+If the bond between parent and child is \ttt{Down}, the child is moved downward.
+
+Using this protocol, the conformation for a double bond flanked by two or more PPBs can be determined.
+The process begins by identifying the parent node.
+This node has the lowest order of the conformational assembly, meaning that it was added first.
+The parent node is placed on a vertical axis.
+The children of the parent node are then examined in order.
+The child connected through the first directional bond is then positioned as described above, either above or below the parent as determined by the value of the bond's direction attribute.
+
+Next, grandchildren are examined in order.
+Should a PPB be found, its direction must match that of the bond between parent and child.
+Otherwise an error is generated.
+A double bond must be found, or an error is reported.
+On finding the double bond, the corresponding grandchild is placed to the right of the child.
+
+The process finishes by iterating the great-grandchildren of the double-bond grandchild, in order.
+The first PPB is placed as before, above the grandchild if the direction is \ttt{Up}, or below if the direction is \ttt{Down}.
+If an additional PPB is present, it must have a direction opposite to the one previously found.
+Otherwise, an error is reported.
+
+The result will be equivalent to one of two diagrams, which for convenience and to avoid confusion are given the labels \textit{syn} (both vertical lines above or below the double bond) and \textit{anti} (vertical lines on opposite sides of the double bond).
 
@@
 
-In some cases the placement of a terminal neighbor can be deduced from the placement of a sibling, without the presence of an explicit PPB.
-For example, a left terminal has two neighbors, but only one of them uses a PPB.
-The neighbor with the PPB can then be placed explicitly.
-Doing so allows the remaining neighbor without a PPB to be placed as well by elimination.
-For example, if the neighbor of a left terminal is placed in the upper left quadrant, then the sibling without a PPB can only occupy the lower left quadrant.
-
-\begin{figure}
-    \centering
-    \includegraphics{filler}
-    \caption{Implied neighbor placement. Bond directions can be omitted when inferrable.}
-    \label{fig:implied-neighbor-placement}
-\end{figure}
-
-The assignment of relative coordinates to both terminals and all neighbors yields a double bond conformation.
-
-Consider the encoding of PPBs for (\textit{E})-2-butene.
-Assume that indexes are assigned sequentially from the neighbor of the left terminal to the neighbor of the right terminal and that the goal is to arrive at a trans (or anti) substituent orientation.
-Begin by placing the two double bond terminals on the x-axis according to succession.
-The left terminal succeeds its neighbor, which should be placed in the lower-left quadrant.
-To achieve this, assign a PPB direction of \ttt{Up}.
-The right terminal precedes its neighbor, which should be placed in the upper-right quadrant.
-To active this, assign a PPB direction of \ttt{Up}.
+Consider the problem of encoding of PPBs for (\textit{E})-2-butene.
+The parent atom, a primary carbon, is first identified.
+A direction of \ttt{Up} for the PPB joining it to its child is arbitrarily chosen.
+A double bond connects child to grandchild.
+Just one bond connects grandchild to great-grandchild.
+Because the \textit{anti} arrangement is desired, the value for direction must be \ttt{Up}.
+In this example, the directions for both PPBs could also have been equivalently set to \ttt{Down}.
 
@@
 
-The above procedures assume that all PPBs are expressed in \textit{normal form}.
-In normal form, a PPB's source precedes it target.
-In other words, \ttt{source} is less than \ttt{target}.
-If a PPB is not in normal form, it must be \textit{inverted} before quadrants can be assigned.
-Inversion swaps the values of `source` and target attributes while simultaneously toggling the `direction` attribute.
-Inversion occurs in two contexts: conjugated polyenes and cycles.
-In the case of polyenes, the same PPB carries partial conformation for at least two different double bond systems.
-In the case of cycles, non-normal PPBs can be generated at bonds forming cycles.
-However, readers and writers must be capable of inverting PPBs wherever they appear.
-
-\begin{figure}
-    \centering
-    \includegraphics{filler}
-    \caption{Partial parity bond inversion. Direction must be inverted for bonds not expressed in normal form.}
-    \label{fig:ppb-inversion}
-\end{figure}
-
 The distributed nature of conformational encoding means that several additive error states are possible:
@@
 \begin{itemize}
-    \item{Overspecification. This state occurs when two or more PPBs share a common terminal and simultaneously use the same parity, thus placing multiple neighbors into the same quadrant.}
-    \item{Underspecification.
-A required and non-deducible PPB is absent.
-This state occurs when a double bond terminal connects to at least one neighbor through a PPB, but the terminal's mate does not.}
-    \item{ Misspecification. A PPB is used between atoms lacking at least one double bond.}
+    \item{Overspecification. This error occurs when children are forced into the same quadrant of a diagram.}
+    \item{Underspecification. A required PPB is missing.}
+    \item{ Misspecification. A PPB is used between atoms, neither of which connects to a double bond.}
 \end{itemize}
 
-A reader that detects one of these error conditions must abort the molecular graph being read and report the error state. It is possible for a conformation to be both overspecified and underspecified.
+A reader that detects any of these error conditions must abort the molecular tree being read and report the error state. It is possible for a conformation to be both overspecified and underspecified.
 
@@
 An exception to the underspecification rule applies in the case of conjugated polyenes.
-Here a double bond with a terminal connected through a PPB does not require the opposing terminal to also bear a neighbor connected through a PPB.
-If the opposing terminal lacks neighbors, no error is generated and reading continues normally.
-If the opposing terminal has at least one neighbor, the bond to it need not be a PPB.
-In either case the conformation of the double bond in question remains undefined.
+Here, a PPB may be present into the terminal double bond but not out of it.
+No error is reported in such a case and the conformation of the terminal double bond remains undefined.
 
@@
 
-Given that conformational specification is distributed over three or more bonds, some conformations will be difficult or possible to represent.
+Cyclic conjugated polyenes present special challenges to the use of PPBs.
 Cyclooctatetraene offers an example.
@@
 
-A configuration is comprised of two components: an ordering of bonds to a central atom; and a \textit{configurational descriptor}.
+A configuration is fully specified by its \textit{configurational descriptor}.
 A configurational descriptor is a template for the relative three-dimensional positioning of neighbors about a central atom.
@@
 
-The \ttt{Left} variant works as follows.
-First, the bond having the index with the lowest value identified.
-The frame of reference is shifted by sighting down this bond from the neighbor to the central atom.
-An equilateral triangle with one vertex pointing upward is then inscribed about the central atom.
-The bond with the next lowest index is then identified and its neighbor is placed at the top vertex of the triangle.
-The neighbor attached to the succeeding bond is then placed at the left triangle vertex.
-Finally, the remaining neighbor is placed at the remaining vertex.
-The overall pattern is one of counterclockwise distribution.
-
-The \ttt{Right} variant follows the same pattern, except that placement of the neighbors attached to the last two bonds in succession are transposed.
-The overall pattern here is one of clockwise distribution.
+A descriptor generates a tetrahedral arrangement as follows.
+First, the parent (predecessor) bound to an atom of interest is identified.
+Next, the frame of reference is shifted by sighting down the bond from parent to child.
+An equilateral triangle is then inscribed bout the central atom.
+Grandchildren are then iterated in order.
+If the configurational descriptor is \ttt{Left}, children are arranged counterclockwise.
+Otherwise, children are arranged clockwise.
 
@@
 The implied bond to the virtual hydrogen is considered to precede the others.
-Sighting down this bond, the atomic neighbors are then placed as previously at the vertices of a triangle inscribed about the central atom.
+Sighting down this bond, the parent node is placed at an arbitrary position.
+Children are then iterated counterclockwise (\ttt{Left}) or clockwise (\ttt{Right}).
 
@@
 \end{figure}
+
+A root atom bearing a configurational descriptor is handled with one more modification.
+The frame of reference is shifted to sight along the direction of the first child to the parent.
+The remaining children are then iterated in order either counterclockwise (\ttt{Left}) or clockwise (\ttt{Right}).
 
@@
 In other words, single, double, triple, and partial parity bond types can all be present.
-The only requirement that must be satisfied by a central atom is a neighbor count of four, or three when a virtual hydrogen is present.
+The only requirement that must be satisfied by a central atom is a connected neighbor count of four, or three when a virtual hydrogen is present.
 

### 2022-06-09 manuscript.tex
Replaces molecular graph with molecular tree.

--- 2022-05-27 manuscript.tex
@@
 \usepackage{nameref}
+\usepackage{makecell}
 
@@
 
-Dialect's purpose is to solve the \textit{molecular serialization} problem.
+Dialect's main goal is to solve the \textit{molecular serialization} problem.
 Serialization refers to the process of translating a data structure into a format that can be transmitted over a communication channel for later reconstruction.
-This is useful because it enables general-purpose devices and networks to convey complex pieces of domain-specific information.
+This allows general-purpose devices and networks to convey complex pieces of domain-specific information.
 Receiver and recipient may be separated by time, space, or both.
 Molecular serialization extends the concept of serialization to molecular structure.
-Examples of molecular serialization formats other than SMILES include Chemical Markup Language (CML)~\cite{murray-rust:2011}, CDXML~\cite{cdxml}, and CTfile ("molfile")~\cite{ctfileFormats}.
-
-Molecular serialization require careful attention to two complementary rule sets: \textit{syntax} and \textit{semantics}.
-Syntax defines the set of valid strings for a language, whereas semantics determine the meaning of valid strings.
-Incomplete specification of either syntax or semantics can lead to data loss and corruption.
-The semantics for Dialect are complex due to the brevity of its syntax and the inherently complex nature of molecular representation.
-
-As a molecular serialization scheme, Dialect must balance two opposing factors: generality and complexity.
-If the chemical space Dialect spans is too broad, complete implementation could become impractical due to language complexity.
-But if the language is made too simple, the chemical space it spans may be of little interest.
-An important goal for Dialect is therefore to address the most interesting chemical space with the simplest possible syntax and semantics.
-
-Toward this end, Dialect is built on four mutually-exclusive concepts:
+Examples of molecular serialization formats other than SMILES include Chemical Markup Language (CML)~\cite{murray-rust:2011}, CDXML~\cite{cdxml}, and CTfile (\enquote{molfile})~\cite{ctfileFormats}.
+
+A secondary goal is brevity.
+Molecules should be expressable using only a few characters for each atom and either zero or one character for each bond.
+Brevity benefits humans who manually encode chemical structures within interactive line-driven interfaces such as those found in terminal sessions and notebooks.
+Brevity also benefits computers, for which compact molecular representation can improve the performance when storage, bandwidth, or processor cycles are limiting factors.
+
+Molecular serialization schemes need to balance generality against complexity.
+As the diversity of the molecules that can be encoded increases, so do the rules needed to encode them and the software needed to interpret the rules.
+For this reason, Dialect's capabilities are directed at that region of chemical space populated by small organic molecules.
+Although Dialect can be used to represent a variety of other kinds of molecules, doing so in general risks data loss or corruption.
+
+Dialect's rules can be divided into two broad categories: \textit{syntax} and \textit{semantics}.
+Syntax defines the set of valid strings for a language, whereas semantics determine the meaning these strings.
+Dialect semantic are complex due to the brevity of its syntax and the inherently complex nature of molecular representation.
+
+Underpinning molecular serialization are conventions and abstractions for molecular representation. These are divided into three broad categories:
 
 \begin{itemize}
-\item \textit{Constitution}, the atoms present in a molecule and how they are connected with each other through bonds.
-Dialect supports elements having two-letter IUPAC designations and all bonding arrangements available through the valence bond model.
-\item \textit{Conformation}, the kind of stereoisomerism resulting from restricted rotation about a bond. Dialect support restricted rotation about double bonds leading to alkene isomerism.
-\item \textit{Configuration}, the kind of stereoisomerism resulting from the arrangement of neighboring atoms in three-dimensional space.
+\item \textit{Constitution}, the atoms present in a molecule and their bonding relationships. Dialect's extends the valence bond model with a limited form of delocalized bonding.
+\item \textit{Conformation}, a kind of stereoisomerism resulting from restricted rotation about a bond. Dialect supports restricted rotation about double bonds capable of producing alkene isomerism.
+\item \textit{Configuration}, a kind of stereoisomerism resulting from the arrangement of neighboring atoms in three-dimensional space.
 Dialect supports configurations for atoms having tetrahedral symmetry attached to four neighboring atoms.
-\item \textit{Delocalization}, a type of bonding in which electrons are shared among three or more atoms, or across a combination of atoms and bonds.
-Dialect supports a limited form of delocalization operating over paths of alternating single and double bonds.
 \end{itemize}
@@
 
-Dialect molecules inherit behavior from an underlying \textit{molecular tree}.
-A tree is a data structure comprised of nodes.
-Each node is associated with zero or more child nodes through a directed edge.
+\section*{Molecular Tree}
+\label{molecular-tree}
+
+Dialect represents molecular structure as a \textit{molecular tree}.
+A tree is a data structure comprised of a set of nodes and pairwise relationships between them called edges.
+Any two nodes in a tree are connected through exactly one path of alternating nodes and edges.
+As such, a tree contains no cycles and exactly one connected component.
 A molecular tree is a specialized tree onto which chemically-relevant metadata have been overlaid.
-The presence of this overlay allows nodes and edges to represent atoms and bonds, respectively.
-
-A molecular tree is related to the more widely-used concept of a \textit{molecular graph}~\cite{balaban:1985}.
-There are two main difference: (1) bonds are undirected in molecular graphs but directed in molecular trees; and (2) cycles are possible in molecular graphs but not in molecular trees.
-The difference might appear to render trees unsuitable to represent molecules, in which cycles often play a prominent structural role.
-But as the next sections demonstrate, the differences between molecular trees and graphs are not so large as they might at first appear.
-
 A Dialect molecule is a molecular tree having zero or more nodes and zero or more edges.
 The empty molecule, devoid of atoms and bonds, is therefore allowed.
-Dialect imposes no upper bound on the number of nodes or edges, although limitations in hardware and software may impose a practical upper bound on molecular size.
-
-\section*{Data Model}
-
-Dialect uses several abstractions collectively known as the \textit{data model}.
-The data model defines metadata attributes associated with the components of a molecular tree.
-These attributes are themselves composed of \textit{primitives} (Table~\ref{table:primitives}), or irreducible data types.
-
-% TODO: completely re-work this to fit with the exact model being used by the reference implementation
+There is no upper bound on the number of nodes or edges, although limitations in hardware and software may restrict maximum molecular size.
+
+A related but more common concept is the \textit{molecular graph}~\cite{balaban:1985}.
+Unlike molecule graphs, molecular trees do not support cycles or disconnected components.
+This difference might appear to make molecular trees unsuitable for molecular representation, where both rings and aggregation play important structural roles.
+However, these apparent shortcomings can be mitigated and even harnessed.
+
+The main advantage of trees over graphs is information density.
+Graphs must accommodate cycles, which is accomplished by assigning a unique identifier to each node.
+An edge references two of these identifiers.
+Graph representations therefore incur the storage overhead of identifiers associated with both nodes and edges.
+In contrast, a tree can be directly encoded as a set of nested, alternating paths of nodes and edges.
+This simplifies the representation and eliminates the overhead of assigning and using node identifiers.
+
+\begin{figure}
+    \centering
+    \includegraphics{filler}
+    \caption{Tree vs. Graph.}
+    \label{fig:tree-vs-graph}
+\end{figure}
+
+When discussing molecular trees, it's often useful to refer to \textit{node order}.
+An order is a property of a set that allows its members to be sorted through pairwise comparisons.
+Node order is the order in which the nodes of a molecular tree are added through depth-first construction.
+A node added before another node is said to \enquote{precede} it.
+A node added after another node is said to \enquote{succeed} it.
+In this way nodes can be prioritized.
+The root atom, uniquely lacking a parent, has highest priority because it precedes every other node.
+Similarly, a parent always precedes its children.
+Finally, the first child added to a given parent precedes all of its siblings.
+This relationship is transitive.
+If a node precedes (or succeeds) a sibling, then so do all of its descendants.
+
+\begin{figure}
+    \centering
+    \includegraphics{filler}
+    \caption{Node Order. Nodes in a molecular tree assume the order in which they were added through depth-first construction.}
+    \label{fig:node-order}
+\end{figure}
+
+To support cycles and disconnected components, a molecular tree decouples connectivity from bonding.
+This means that nodes and edges don't generally map to atoms and bonds as they often do in molecular graphs.
+This indirection enables rings and disconnections to be encoded despite the fact that trees can contain no cycles or disconnected components.
+
+Cycles are supported through \textit{bridges}.
+A bridge consists of two bond-node pairs that together represent one bond.
+One ach side of a bridge and atom connects to a bond, but the two atoms are not connected to each other.
+Instead, each atom is succeeded by a \textit{bridge node}.
+Unlike an atom, a bridge node has no children and so may not serve as the root of the tree.
+A bridge preserves the immediate atomic environment around each atom, allowing many of the same atomic computations possible in molecular graphs.
+Any bond, whether it closes a ring or not, can be represented as a bridge.
+
+Each bridge node serves as a proxy to a \textit{referenced atom}.
+Should a bridge node be present during an operation that requires an atom, the referenced is used instead.
+Although a bridge node always succeeds its parent, the referenced atom may precede or succeed that parent.
+
+Multiple molecules are supported through \textit{gaps}.
+A gap is an edge devoid of bonding.
+Gaps can be used whenever two or more molecules need to be encoded within the same molecular tree.
+Specific applications include ionic bonding and non-ionic complexes.
+However, no bonding relationship whatsoever is implied by the presence of a gap.
+In particular, the zero-order bond interpretation is excluded \cite{clark:2011}.
+
+\begin{figure}
+    \centering
+    \includegraphics{filler}
+    \caption{Nodes and Edge Associations.
+Three node/edge associations are possible within a molecular tree.
+A parent atom can be connected to a child atom through a bond (left).
+A parent atom can also be connected to a child bridge node through a bond (center).
+Finally, an atom can be connected, without bonding, to a child atom through a gap (right).}
+    \label{fig:nodes-and-edges}
+\end{figure}
+
+Gaps and bridges decouple connectivity from bonding, but for opposite purposes.
+Whereas a gap connects two atoms without bonding them, a bridge creates a bond between two atoms without connecting them.
+These relationships can complicate the analysis of connectivity and bonding.
+For example, the atom succeeding a gap behaves in most situations as if it were a root node.
+When tallying the bonds surrounding such a node, only children are considered.
+Similar considerations apply when a gap succeeds an atom.
+Likewise, a bridge node behaves as if it were an atom in the sense that it connects to an atom through a bond.
+Nevertheless, a bridge node merely stands in for another atom found elsewhere in the tree.
+Bonding and connectivity analyses within molecular trees must therefore account for the possibility of connection-free bonds and bond-free connections.
+
+\section*{Atoms and Bonds}
+\label{atoms-and-bonds}
+
+Within a molecular tree, atoms and bonds play complementary roles.
+Atoms control nuclear particle counts, valence electron counts, and configuration.
+Bonds determine bonding electron counts and conformation.
+Constitution, configuration, and conformation are all influenced by the individual and collective behavior of atoms and bonds.
+
+At the lowest level, atom and bond behavior are governed by their respective \textit{attributes}.
+An attribute is a key/value pair in which the key is a name composed of a character sequence and the value is constrained by a single, immutable \textit{type}.
+Types can be composed and decomposed, a feature supported by many programming languages.
 
@@
 \renewcommand*\footnoterule{}
-\caption{Data Model Primitives.}
+\caption{Primitives.}
 \label{table:primitives}
 \centering
-\begin{tabular}{l l l}
-    \hline
-    Name & Description & Values\footnote{Ranges defined as \ttt{a..b} include \ttt{a} and exclude \ttt{b}.} \\
-    \hline
-    \ttt{Option<T>} & Optional Value & \ttt{None}, \ttt{T} \\
-    \ttt{Three} & Integer count & 1..4 \\
-    \ttt{Dec} & Integer count &0..10 \\
-    \ttt{SignedDec} & Integer count & -9..10 \\
-    \ttt{Thou} & Integer count & 1..1000 \\
-    \ttt{Element} & IUPAC-\-approved element symbol & \ttt{Ac}, \ttt{Ag}, \ttt{Al}, {\ldots}\ttt{Zn}\\
-    \ttt{Configuration} & Configurational descriptor & \ttt{Right}, \ttt{Left} \\
-    \ttt{Direction} & Conformational descriptor & \ttt{Up}, \ttt{Down} \\
-    \ttt{Implicit} & Implicit hydrogen count flag & - \\
-    \ttt{boolean} & Boolean & \ttt{true}, \ttt{false} \\
+\begin{tabular}{l l l l}
+    \hline
+    Name & Description & Notation & Values \\
+    \hline
+    \ttt{Option} & Optional value & \ttt{\{T\}} & \ttt{None}, $t$ \\
+    \ttt{List} & Ordered list of values & \ttt{[T]} & $t_0, t_1, ... t_n$ \\
+    \ttt{Range} & Bounded integer range & m..n & from \ttt{m} to \ttt{n}, excluding \ttt{n} \\
+    \ttt{Boolean} & Boolean value & \ttt{?} & \ttt{true}, \ttt{false} \\
     \hline
@@
 
-An atom possesses several attributes (Table~\ref{table:atomic-state}), which collectively define an \textit{atomic state}.
-Not all atomic states are valid.
-Implementations must ensure either the impossibility of an invalid state, or an error condition in the event that one is created.
-Atomic state is restricted in the following ways:
+Dialect types are composed of four \textit{primitives} (Table~\ref{table:primitives}).
+The first, \ttt{Option}, represents a value of a particular type that may either be present or absent.
+If not present, the special value \ttt{None} is used.
+A \ttt{List} is a possibly empty ordered collection of values of the same type.
+The \ttt{Range} type represents an integer value constrained by a lower inclusive bound (\ttt{m}) and an upper, exclusive bound (\ttt{n}).
+The \ttt{Boolean} type assumes one of two values: \ttt{true} or \ttt{false}.
+
+An atom is composed of seven attributes (Table~\ref{table:atomic-attributes}).
+Default values are applied at the time of atom creation if no other values are used.
+The specific values assigned to an atom collectively define an \textit{atomic state}.
 
 \begin{table}
-\caption{Atomic State.}
-\centering
-\begin{tabular}{l L L l}
+\caption{Atomic Attributes.}
+\centering
+\begin{tabular}{l L l l}
     \hline
@@
     \hline
-    \ttt{element} & Elemental symbol & \ttt{Option} \ttt{<Element>} & \ttt{None} \\
-    \ttt{hydrogens} & Hydrogen count & \ttt{Implicit}, \ttt{Dec} & 0 \\
-    \ttt{configuration} & Configuration & \ttt{Option} \ttt{<Configuration>} & \ttt{None} \\
-    \ttt{charge} & Formal charge & \ttt{SignedDec} & 0 \\
-    \ttt{isotope} & Mass number & \ttt{Option<Thou>} & \ttt{None} \\
-    \ttt{selected} & Selection status & \ttt{boolean} & \ttt{false} \\
+    \ttt{element} & Elemental symbol & \ttt{\{Element\}} & \ttt{None} \\
+    \ttt{isotope} & Mass number & \ttt{\{1..1000\}} & \ttt{None} \\
+    \ttt{hydrogens} & Hydrogen count & \ttt{Implicit}, \ttt{0..10} & 0 \\
+    \ttt{charge} & Formal charge & \ttt{-9..10} & 0 \\
+    \ttt{selected} & Selection status & \ttt{?} & \ttt{false} \\
+    \ttt{parity} & Configuration & \ttt{\{AtomParity\}} & \ttt{None} \\
+    \ttt{edges} & Edges directed toward child nodes & \ttt{[Edge]} & \ttt{[]} \\
     \hline
 \end{tabular}
-\label{table:atomic-state}
+\label{table:atomic-attributes}
 \end{table}
 
-\begin{enumerate}
-    \item{If the value of \ttt{hydrogens} equals \ttt{Implicit}, then the values of \ttt{isotope}, \ttt{configuration}, and \ttt{charge} must equal their respective defaults.
-Furthermore, the value of \ttt{element} must equal one of \ttt{B}; \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; \ttt{S}; \ttt{F}; \ttt{Cl}; \ttt{Br}; \ttt{I}; or \ttt{None}.}
-    \item{If the value of \ttt{selected} equals \ttt{true}, then the value of \ttt{element} must equal one of: \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; or \ttt{S}.}
-\end{enumerate}
-
-Like atoms, bonds carry metadata in the form of attributes (Table \ref{table:bond-state}), which collectively define a \textit{bond state}.
-As with atomic attributes, not all bond states are valid.
-Bond state is restricted in the following ways:
-
-\begin{table}
-\caption{Bond State.}
-\centering
-\begin{tabular}{l L L l}
-    \hline
-    Attribute & Description & Type & Default \\
-    \hline
-    \ttt{order} & Bond order & \ttt{Three} & 1 \\
-    \ttt{elided} & Elision status & \ttt{boolean} & \ttt{false} \\
-    \ttt{direction} & Partial conformational parity & \ttt{Option} \ttt{<Direction>} & \ttt{None} \\
-    \ttt{target} & Child atom & \ttt{Atom} & - \\
-    \hline
-\end{tabular}
-\label{table:bond-state}
-\end{table}
-
-\begin{enumerate}
-    \item{If the value of \ttt{direction} does not equal \ttt{None}, then the value of \ttt{order} must equal one.}
-    \item{If the value of \ttt{elided} equals \ttt{true}, the values of \ttt{order} and \ttt{direction} must equal one and \ttt{None}, respectively.}
-\end{enumerate}
-
-The data model will be referenced throughout this paper as a way to connect syntax and semantics.
-Dialect-compatible software may, of course, use any suitable internal data model.
-Implementations must, however, ensure that any internal data model is consistent with the one provided here.
-
-\section*{Constitution}
-
-A molecular tree is described at the lowest level by its constitution.
-In Dialect, constitution comprises those attributes locating nucleons and electrons at specific atoms and bonds.
-All other attributes lie outside the scope of constitution.
-Atomic attributes related to constitution are: \ttt{element}; \ttt{hydrogens}; and \ttt{charge}.
-Bond attributes related to constitution are \ttt{order} and \ttt{elided}.
-
-The attributes \ttt{element} and \ttt{isotope} characterize an atom's nucleus.
-The \ttt{element} attribute is an optional one- or two-letter character sequence selected from the set designated by the IUPAC~\cite{periodicTable}.
-When IUPAC authorizes additional symbols, they will become valid values for the \ttt{element} attribute.
-If the elemental identity of an atom is unknown, its \ttt{element} attribute must equal \ttt{None}.
-The \ttt{isotope} attribute is an optional integer value representing an atom's nuclear mass number, where mass number is the sum of proton and neutron count.
+The attributes \ttt{element} (\enquote{element}) and \ttt{isotope} (\enquote{isotope}) define atomic number and neutron count, respectively.
+Elements are selected from a subset of those approved by IUPAC \cite{periodicTable} (Figure~\ref{fig:elements}).
+An atom whose element is unknown must have an \ttt{element} attribute equal to \ttt{None}.
+The \ttt{isotope} attribute is an optional integer value representing an atom's nuclear mass number, defined as the sum of proton and neutron count.
 Setting the \ttt{isotope} property to \ttt{None} means that the element's isotopic composition equals natural abundance.
-Otherwise, the lower bound on the value of \ttt{isotope} is one.
-This lower limit allows for such physically nonsensical states such as atoms of negative implied neutron count (e.g., carbon-5).
+Otherwise, the lower bound of the value of \ttt{isotope} is one.
+This lower limit allows for physically nonsensical states such as atoms of negative implied neutron count (e.g., carbon-5).
 Implementations may reject such atomic states as invalid.
+
+\begin{figure}
+    \centering
+    \includegraphics{filler}
+    \caption{Elements. The element that may be associated with an atom.}
+    \label{fig:elements}
+\end{figure}
 
@@
 Even so, a suppressed hydrogen and its bond are understood to be present.
-The relative abundance and regular bonding of hydrogen atoms in molecular graphs makes hydrogen suppression an effective technique for increasing information density.
+The high relative abundance and regular bonding of hydrogen atoms makes hydrogen suppression an effective technique for increasing information density.
 Two mutually-exclusive types of suppressed hydrogen are supported: \textit{virtual hydrogen} and \textit{implicit hydrogen}.
@@
 A virtual hydrogen is one that appears as an integer unit contribution associated with the parent atom.
-For example, methane can be represented by a molecular graph having five nodes and four edges.
-But methane can also be represented as a graph of one node whose \ttt{hydrogens} attribute equals four.
-When using this mode, the \ttt{hydrogens} attribute may assume integer values from zero to nine, inclusive.
-The process of replacing atomic hydrogens with virtual hydrogen is called \enquote{virtualization.} The opposite process, replacing virtual hydrogen with atomic hydrogen, is called \enquote{reification.}
+For example, methane can be represented using five atoms and four bonds.
+But methane can also be represented as a single atom whose \ttt{hydrogens} attribute equals four.
+Virtual hydrogen count defaults to zero, and may be set to a maximum value of nine.
+By default, atoms use virtual hydrogen counting.
 
 An implicit hydrogen is similar to a virtual hydrogen in that both the hydrogen node and bond have been replaced with a tally.
-Unlike a virtual hydrogen, however, this tally is not present as attribute on the parent atom but rather as a computation.
-Subject to certain limitations, an atom identifies itself as having an implicit hydrogen count by setting its \ttt{hydrogens} attribute to \ttt{Implicit}.
-The computation and its limitations will be described later in this paper.
-
-The atomic attribute, \ttt{charge}, enables electron counting.
-The purpose of electron counting is to associate each electron in a molecular graph with either an atom or a bond.
-This is a fundamental requirement for the principle of mass conversion.
-In a more practical sense, electron counting lays the foundation for calculations of formal charge and bond order~\cite{dietz:1995}.
+Unlike a virtual hydrogen, however, this tally is not present as an attribute on the parent atom but rather as a computation.
+Setting the \ttt{hydrogens} attribute to \ttt{Implicit} activates implicit hydrogen counting.
+However, not all atoms are eligible to do so (see: \nameref{computing-implicit-hydrogen-count}).
+
+The \ttt{charge} (\enquote{charge}) and \ttt{selected} (\enquote{selected}) attributes determine valence electron count (see: \nameref{electron-counting}).
+Charge refers formal charge, or the difference between the valence electron count of the element and the sum of bonding and nonbonding electrons of the bound atom.
+By default, an atom is assigned a charge of zero.
+Minimum and maximum values for charge are -9 and +9, respectively.
+The \ttt{selected} attribute adjusts an atom's selection status.
+Selecting an atom allows it to participate in extended bonding (see: \nameref{delocalization-subgraph}).
+
+Configuration is determined by the \ttt{parity} attribute (\enquote{atom parity}).
+The type of this value, \ttt{AtomParity} is an enumeration supporting two values: \ttt{Clockwise} and \ttt{Counterclockwise} (Table~\ref{table:atom-parity-variants}).
+The interpretation and encoding of atom parity will be described in detail later (see: \nameref{configuration}).
+
+\begin{table}
+\caption{Atom Parity Variants.}
+\centering
+\begin{tabular}{l l}
+    \hline
+    Variant & Description  \\
+    \hline
+    \ttt{Clockwise} & Configuration runs with node order. \\
+    \ttt{Counterclockwise} & Configuration runs against node order. \\
+    \hline
+\end{tabular}
+\label{table:atom-parity-variants}
+\end{table}
+
+The \ttt{edges} ("edges") attribute is a possibly empty ordered list of edges associated with an atom.
+As noted in the previous section, an edge may represent either a bond or a gap (Table~\ref{table:edge-variants}).
+The mere presence of an edge does not guarantee a bonding relationship.
+
+\begin{table}
+\caption{Edge Variants.}
+\centering
+\begin{tabular}{l l}
+    \hline
+    Variant & Description  \\
+    \hline
+    \ttt{Bond} & A bonding and connectivity relationship. \\
+    \ttt{Gap} & A connectivity relationship only. \\
+    \hline
+\end{tabular}
+\label{table:edge-variants}
+\end{table}
+
+Not all atomic states are valid.
+Implementations must ensure either the impossibility of creating an invalid state, or an error condition in the event that one is created.
+Atomic state is restricted in the following two ways:
+
+\begin{enumerate}
+    \item{If implicit hydrogen counting is used, default values of isotope, charge, and parity must also be used.
+Implicit hydrogen counting is only supported for elements in the list: B; C, N; O; P; S; F; Cl; Br; or I.}
+    \item{A selected atom must use one of the following five elements: C; N; O; P; or S.}
+\end{enumerate}
+
+A bond is composed of four attributes (Table~\ref{table:bond-attributes}).
+The \ttt{order} attribute (\enquote{order}) represents the concept of formal bond order, or the number of bonding electrons divided by two.
+A bond's order may only assume the values one, two, or three.
+The \ttt{elided} attribute (\enquote{elided}) allows a bond to be omitted during serialization and participate in extended bonding.
+The \ttt{direction} attribute (\enquote{direction}) is used with other information to conformationally restrict a bond.
+The two allowed variants for \ttt{Direction}, \ttt{Up} and \ttt{Down}, will be described in detail later (see: \nameref{conformation}).
+The \ttt{target} attribute identifies the child node at which the bond is directed (Table~\ref{table:target-variants}).
+
+\begin{table}
+\caption{Bond Attributes.}
+\centering
+\begin{tabular}{l p{3.6cm} l l}
+    \hline
+    Attribute & Description & Type & Default \\
+    \hline
+    \ttt{order} & Formal bond order & \ttt{1..4} & 1 \\
+    \ttt{elided} & Elision status & \ttt{boolean} & \ttt{false} \\
+    \ttt{direction} & Partial conformational parity & \ttt{\{Direction\}} & \ttt{None} \\
+    \ttt{target} & An atom or bridge node & \ttt{Target} & - \\
+    \hline
+\end{tabular}
+\label{table:bond-attributes}
+\end{table}
+
+\begin{table}
+\caption{Target Variants.}
+\centering
+\begin{tabular}{l p{6.6cm}}
+    \hline
+    Variant & Description \\
+    \hline
+    \ttt{Atom} & A parent and child atom are bound. \\
+    \tt{Bridge} & A parent atom and bridge node are bound. \\
+    \hline
+\end{tabular}
+\label{table:target-variants}
+\end{table}
+
+As with atomic state, not all bond states are valid. The following restrictions apply:
+
+\begin{enumerate}
+    \item{An elided bond must have an order of one and default direction.}
+    \item{A bond with a non-default direction must have an order of one.}
+\end{enumerate}
+
+The atom and bond attributes defined here will be referenced throughout this paper as a way to connect syntax and semantics.
+Implementations may, of course, use any suitable internal data model.
+They must, however, ensure that any internal data model is consistent with the one provided here.
+
+\section*{Electron Counting}
+\label{electron-counting}
+
+A molecular tree explicitly encodes the presence of protons and neutrons through the atomic attributes \ttt{element} and \ttt{isotope}, respectively.
+The presence of electrons, in contrast, is merely implied.
+The principles of charge and mass conservation nevertheless require a method to determine explicit electron counts.
+This is possible through \textit{electron counting}.
 
 Electron counting in Dialect is based on the well-known \textit{valence bond model} (VB Model)~\cite{lewis:1916}.
-Although not commonly considered as such, the VB model can be thought of as an algorithm for molecular assembly.
-This perspective is especially useful given that Dialect itself can be viewed as a programming language for molecular assembly.
-
-A simplified algorithm for electron counting can be summarized as follows.
-A molecular graph starts as a set of disconnected atoms and an integer electron count equal to the atomic number of the atom's element.
-Two nodes are selected as the terminals of a bond.
-From each atom, the same positive electron count is deducted.
-The total electron count deducted from each atom is simultaneously credited to the bond's electron count.
-For example, if each atom contributes one electron to a bond, then the bond's electron count will increase by two.
-
-\begin{figure}
-    \centering
-    \includegraphics{filler}
-    \caption{Generalized Valence Bond Electron Counting.
-Two unconnected nodes are chosen for bonding (left).
-One electron is subtracted from each atom, and both electrons are added to the bond's count (right).}
+This model assumes that a bond spans exactly two atoms and is associated with a nonzero, positive, and even number of electrons.
+These electrons are drawn in equal quantities from each atom's set of valence electrons.
+A single bond draws one electron from each atom (two total), a double bond two (four total), and a triple bond three (six total).
+The electron count of an atom whose \ttt{element} attribute equals \ttt{None} is always zero.
+Bonding to such an atom adds the required number of electrons directly to the bond.
+
+Electrons are counted as follows.
+A molecular tree starts as a single root atom.
+The electron count of this atom equals the difference between proton count (atomic number) and charge.
+Adding a bond is a three-step process.
+First, a child atom is constructed with an electron count determined in the same way as the root atom.
+Next, the child atom and the bond are added to the molecular tree.
+Finally, the electron counts of each atom and the bond are updated.
+If parent and child are connected through a gap edge, no electrons are deducted from either atom.
+Should a parent atom be bound to a child bridge node, then the electron count of the referenced atom is reduced.
+
+\begin{figure}
+    \centering
+    \includegraphics{filler}
+    \caption{Electron Counting.
+A root atom and its child are added to a tree (left).
+Next, a bond is added (center).
+Finally, the appropriate electron count is deducted from each atom and added to the bond (right).}
     \label{fig:electron-counting}
 \end{figure}
-
-For convenience, the actual model used by Dialect differs slightly from the above description.
-Neither nodes nor edges carry an explicit electron count attribute.
-Instead, edges carry an \ttt{order} attribute, and nodes carry a \ttt{charge} attribute.
-Bond order equals electron count divided by two.
-Formal charge equals the number of electrons gained or lost due to ionization events outside of bond formation.
-Gaining an electron decrements the \ttt{charge} attribute, whereas losing an electron increments it.
-If no electrons are gained or lost to ionization outside of bond formation, the formal charge equals zero.
-
-\begin{figure}
-    \centering
-    \includegraphics{filler}
-    \caption{Electron counting shorthand.
-Rather than explicit electron counts, atoms and bonds possess charge and order attributes.
-The atomic charge attribute captures deviations from generalized counting.}
-    \label{fig:electron-counting-shorthand}
-\end{figure}
-
-From these rules follow some practical consequences.
-For one, bond order must be an integer greater than zero.
-This follows from the definition of bond formation.
-The electron count deducted from each atom is $n$, an integer greater than zero.
-Given that bond formation deducts an equal number of electrons from each atom, the total number of electrons associated with the newly-formed bond is $2n$.
-By definition, bond order is the bond's electron count divided by two, therefore $n$ is the bond order.
-Fractional, negative, and zero bond orders are all disallowed by Dialect for this reason.
-
-Dialect imposes an upper limit of three on formal bond order.
-The relative scarcity of bond orders greater than this limit in molecules conforming the the VB Model makes this restriction unlikely to meaningfully restrict applicability.
 
@@
 Such an arrangement may be physically meaningless, but Dialect explicitly supports it.
-Software using Dialect may or may not reject such species for semantic reasons.
-
-\begin{figure}
-    \centering
-    \includegraphics{filler}
-    \caption{Hypervalence. Depletion of atomic electron count through excessive bonding. Hypervalent atoms are valid by default, although chemically nonsensical.}
-    \label{fig:hypervalence}
-\end{figure}
-
-Each atom carries an \ttt{index} attribute, allowing bonds to reference atoms in a molecular graph.
-To do so, the \ttt{source} and \ttt{target} attributes are set with the respective indexes of the source and target atoms.
-The set of all bonds and the nodes they reference determines the connectivity of a molecular graph.
-In the context of determining connectivity, source and target indexes may be reversed.
-
-\section*{Bond Elision}
-
-A bond whose attributes are assigned default values can participate in \textit{bond elision}.
-In bond elision, a bond's \ttt{elided} attribute is set to \ttt{true}.
-When written in string form, elided bonds require no characters.
-This makes bond elision useful for data compression.
-The second application for elided bonds will be described in the next section.
+Software using Dialect may or may not reject such species on semantic grounds.
 
 \section*{Delocalization Subgraph}
+\label{delocalization-subgraph}
 
 A molecular representation based solely on the valence bond model can yield artifacts resulting from \textit{delocalization induced molecular equality} (DIME).
-DIME arises when a molecular graph exists in two or more unequal forms, each differing from the others only through electron delocalization.
-In this context, two graphs are equal if they are isomorphic and each matched pair of atoms or bonds possess equivalent non-index attributes.
-If shifting electrons through delocalization causes two graphs to become equal, then DIME has occurred.
+DIME arises when a given molecule can be represented by two forms, each differing from the other only in double bond placement.
 DIME is sometimes referred to as \enquote{resonance} or \enquote{aromaticity,} but those terms are avoided here due to their ambiguity~\cite{randic:2018,kerber:2006}.
@@
 
-DIME can interfere with \textit{canonicalization}, or the selection of a single representation for a molecular graph.
-The presence of multiple equivalent molecular graphs differing only in electron delocalization complicates the formulation of selection rules and invariants, which must be adapted to account for the artificial asymmetry.
-
-To eliminate DIME and thereby streamline canonicalization, the molecular graph is augmented with a \textit{delocalization subgraph} (DS).
-A DS is a possibly empty node-induced subgraph of a molecular graph.
-The membership of a DS may be drawn from the set of atoms and bonds that participate in DIME within a given molecular graph.
+The main problem with DIME is its interference with \textit{canonicalization}, or the selection of a single molecular representation among many alternatives.
+The presence of multiple equivalent molecular forms differing only in electron delocalization complicates the formulation of selection rules and invariants, which must be adapted to account for the artificial asymmetry.
+
+To eliminate DIME and thereby streamline canonicalization, each molecular tree is augmented with a \textit{delocalization subgraph} (DS).
+A DS is a possibly empty node-induced subgraph over a molecular tree.
+The membership of a DS will typically be drawn from the set of atoms and bonds that participate in DIME within a given molecular tree.
 
@@
 An atom is added to the DS by setting its \ttt{selected} attribute to \ttt{true} through a process called \textit{selection}.
-For an atom to be selectable, its \ttt{element} attribute must be contained in the list: \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; and \ttt{S}.
-Atoms having other elements must not be selected.
-A bond will be added to the DS if both of its terminals are selected and the \ttt{elided} attribute equals \ttt{true}.
-No other bond will be added to the DS.
-This rule allows arbitrary bonds between selected atoms to be excluded from the DS.
-
-Atoms whose \ttt{element} attribute equals \ttt{None} are not selectable, which prevents their entry into the DS.
-Bonds to such atoms are similarly excluded.
-The alternative, conditional inclusion, would complicate both data structures and algorithms for little gain.
-Exclusion avoids these complexities.
+As noted previously (see: \nameref{atoms-and-bonds}), only some atomic states are compatible with selection.
+An edge will be added to the DS if both of the atoms it spans are selected and the bond itself is elided.
+This allows arbitrary bonds between selected atoms to be excluded from the DS.
+If a bond's target is a bridge node, the corresponding edge will be added to the DS if both the parent atom and the equivalent target atom are selected.
 
@@
 Equivalently, a matching is a subgraph in which all nodes have degree one.
-A perfect matching includes all the nodes of its parent graph.
+A perfect matching includes every node present in the parent graph.
 Every atom added to a DS must therefore become part of its perfect matching.
@@
 First, a perfect matching over the DS is found.
-Next, each matched edge is replaced by a double bond.
+Next, the elided bond corresponding to each matched edge is replaced by a double bond.
 Because the presence of a filled DS implies a perfect matching over it, kekulization always succeeds.
@@
 The opposite operation can be accomplished with a \textit{selection algorithm}.
-A selection algorithm selects two or more atoms, thereby adding them to the DS.
+A selection algorithm selects two or more nodes, thereby adding them to the DS.
 The only requirement for a selection algorithm is that the resulting DS must have a perfect matching.
@@
 \section*{Computing Implicit Hydrogen Count}
+\label{computing-implicit-hydrogen-count}
 
@@
 
-\section*{Atom Index}
-
-To support the assignment and interpretation stereochemical features, each node carries a unique, non-negative integer \ttt{index} attribute (\enquote{index}).
-Assignment is sequential, meaning that the first atom gets index zero, the next one, and so on.
-For example, the index for the first atom is zero, for the second atom one, for the third atom two, and so on.
-The maximum value of an atomic index is 2\textsuperscript{32} - 1.
-
-Indexes impose an ordering over the atoms in a molecular graph.
-An atom with an index less than another atom is said to \textit{precede} it.
-An atom with an index grater than another atom \textit{succeeds} it.
-No special significance is ascribed to the atom whose index is zero, except that every other atom succeeds it.
-
 \section*{Conformation}
-
-The second major component of molecular representation in Dialect is \textit{conformation}.
-Conformation is a rotational restriction about one or more bonds.
-Dialect limits conformation to just one of many possible types: rotational restrictions occurring at individual double bonds.
-
-Organic chemistry uses the descriptors \textit{entgegen} (\textit{E}) and \textit{zusammen} (\textit{Z}) to label conformational parity.
-This system is based on prioritization of the atoms neighboring the double bond.
-One pattern of priority yields the \textit{E} parity, and the other yields \textit{Z}.
-Conveniently, the parity descriptor can be localized at the double bond.
-
-\begin{figure}
-    \centering
-    \includegraphics{filler}
-    \caption{Traditional conformational stereodescriptors.}
-    \label{fig:traditional-conformational-stereodescriptors}
-\end{figure}
-
-Dialect takes a different approach by introducing \textit{partial parity bonds} (PPB).
-As the name implies, a PPB expresses a portion of the parity characterizing a conformationally-restricted double bond.
-Constructing the full parity requires the double bond itself, and at least two PPBs.
-With just a few exceptions (see below), a double bond with a neighbor at one terminal bound through a PPB must also have a neighbor at the opposite terminal bound through a PPB.
+\label{conformation}
+
+Restricted rotation about a bond leads to conformational isomers (\enquote{conformers}).
+When the bond in question is a double bond, two conformers result.
+They can be distinguished through \textit{partial parity bonds} (PPBs).
+As the name implies, a PPB encodes some of the parity information characterizing a conformationally restricted double bond.
+Reconstruction of the full parity requires the double bond itself and at least one flanking PPB at each terminal.
 
@@
 
-To support this system, each bond carries an optional \ttt{direction} attribute (\enquote{direction}).
-When present, direction may assume one of the two values \ttt{Up} or \ttt{Down}.
-These values refer to a geometrical model in which the two terminals of a double bond and their neighbors are assigned local relative coordinates in a plane.
-The terminals of the bond are placed on the x-axis such that the one on the right succeeds the one on the left.
-The neighbors at each terminal joined through a PPB are then assigned relative coordinates in the plane.
-
-Neighbors of the left terminal are positioned according to the relative ordering of atomic indexes and the bond's \ttt{direction} attribute.
-If the neighbor precedes the terminal then a \ttt{direction} attribute equal to \ttt{Up} locates the neighbor in the lower-left quadrant and a \ttt{direction} attribute of \ttt{Down} locates the neighbor in the upper-left quadrant.
-If the neighbor succeeds the terminal, these placements are reversed.
-
-An analogous procedure assigns relative coordinates to a neighbor of the right terminal.
-If a neighbor succeeds the right terminal, place it to the upper right given an \ttt{Up} direction, or to the lower right given a \ttt{Down} direction.
-Reverse these assignments if the neighbor precedes the right terminal.
-
-\begin{figure}
-    \centering
-    \includegraphics{filler}
-    \caption{Interpreting partial parity bonds.}
-    \label{fig:interpret-ppb}
-\end{figure}
-
-In some cases the placement of a terminal neighbor can be deduced from the placement of a sibling, without the presence of an explicit PPB.
-For example, a left terminal has two neighbors, but only one of them uses a PPB.
-The neighbor with the PPB can then be placed explicitly.
-Doing so allows the remaining neighbor without a PPB to be placed as well by elimination.
-For example, if the neighbor of a left terminal is placed in the upper left quadrant, then the sibling without a PPB can only occupy the lower left quadrant.
-
-\begin{figure}
-    \centering
-    \includegraphics{filler}
-    \caption{Implied neighbor placement. Bond directions can be omitted when inferrable.}
-    \label{fig:implied-neighbor-placement}
-\end{figure}
-
-The assignment of relative coordinates to both terminals and all neighbors yields a double bond conformation.
-
-Consider the encoding of PPBs for (\textit{E})-2-butene.
-Assume that indexes are assigned sequentially from the neighbor of the left terminal to the neighbor of the right terminal and that the goal is to arrive at a trans (or anti) substituent orientation.
-Begin by placing the two double bond terminals on the x-axis according to succession.
-The left terminal succeeds its neighbor, which should be placed in the lower-left quadrant.
-To achieve this, assign a PPB direction of \ttt{Up}.
-The right terminal precedes its neighbor, which should be placed in the upper-right quadrant.
-To active this, assign a PPB direction of \ttt{Up}.
-
-\begin{figure}
-    \centering
-    \includegraphics{filler}
-    \caption{Assigning partial parity bonds to *trans*-2-butene.}
+A PPB is designated by setting its \ttt{direction} attribute to a non-default variant (\ttt{Up} or \ttt{Down}).
+The names of these variants refer to a two-dimensional geometrical model in which a parent node and its child are placed along a vertical axis.
+If the child lies above the parent, the \ttt{Up} variant is used.
+If the child lies below the parent, the \ttt{Down} variant is used.
+
+The parity of a double bond is determined by the relative direction of the PPBs neighboring it.
+First, the parent atom is identified.
+This atom has the lowest order in the assembly, meaning that it was added first to the molecular tree.
+The parent is placed on a vertical axis.
+Bonds from the parent atom are then iterated.
+If a PPB is found, its child atom is placed above the parent for a direction of \ttt{Up} or below the parent for a direction of \ttt{Down}.
+The child's bonds are then iterated in order.
+Should an additional PPB be found, its direction must match that of the bond between parent and child.
+Otherwise an error must be reported.
+An error must also be reported if more than two PPBs are present.
+If a double bond is found, its target, a grandchild, is placed to the right of the parent.
+The bonds of the grandchild are then iterated in order.
+The target of the first PPB, a great-grandchild, is placed as before, above the grandchild if the direction is \ttt{Up}, or below if the direction is \ttt{Down}.
+If a second PPB is present, it must have a direction opposite the first.
+Otherwise an error must be reported.
+
+The placement of parent, child, grandchild, and great-grandchild will yield one of two parities (Figure~\ref{fig:ppb-placement}).
+For convenience and to avoid confusion, these parities are given the labels  \textit{syn} (vertical lines both lie above or below the double bond) and \textit{anti} (vertical lines lie on opposite sides of the double bond).
+
+\begin{figure}
+    \centering
+    \includegraphics{filler}
+    \caption{Partial Parity Bond Placement.}
+    \label{fig:ppb-placement}
+\end{figure}
+
+To illustrate, consider the placement for (\textit{E})-2-butene.
+The parent atom, a methyl group, is placed on a vertical axis.
+Bonds are iterated and a PPB with a direction of \ttt{Down} is found.
+The child attached to the bond is therefore placed below the parent.
+Iterating the child's bonds reveals one bond of order two.
+The atom attached to it is placed to the right of the previous child.
+Bonds are again iterated, revealing a PPB whose direction is \ttt{Down}.
+The atom attached to this bond is then placed below its parent.
+The completed diagram reveals the \textit{anti} conformation.
+
+The same procedure works in reverse when encoding the \ttt{direction} attribute.
+2-butene having the \textit{anti} conformation is modeled using a diagram whose left- and right-hand sides point downward and upward, respectively.
+The parent atom, a primary carbon, is identified.
+It is bonded to one neighbor through a PPB.
+To match the diagram, a direction of \ttt{Up} is assigned.
+The child atom is attached to its own child through a double bond.
+Finally, the grandchild is attached to its own child through a PPB whose direction is consistent with the diagram.
+A direction of \ttt{Up} is therefore assigned.
+
+\begin{figure}
+    \centering
+    \includegraphics{filler}
+    \caption{Assigning partial parity bonds to \textit{anti}-2-butene.}
     \label{fig:assigning-ppb}
@@
 
-PPBs are decoded analogously.
-Start by placing the terminals on the x-axis according to succession.
-The PPB between the left terminal and its neighbor has a direction of \ttt{Up}.
-Because the neighbor precedes the terminal, the neighbor should be placed in the lower-left quadrant.
-The PPB between the right terminal and its neighbor has a direction of \ttt{Up}.
-Because the neighbor succeeds the terminal, the neighbor is placed in the upper-right quadrant.
-The result is the expected (\textit{E}) conformation.
-
-The above procedures assume that all PPBs are expressed in \textit{normal form}.
-In normal form, a PPB's source precedes it target.
-In other words, \ttt{source} is less than \ttt{target}.
-If a PPB is not in normal form, it must be \textit{inverted} before quadrants can be assigned.
-Inversion swaps the values of `source` and target attributes while simultaneously toggling the `direction` attribute.
-Inversion occurs in two contexts: conjugated polyenes and cycles.
-In the case of polyenes, the same PPB carries partial conformation for at least two different double bond systems.
-In the case of cycles, non-normal PPBs can be generated at bonds forming cycles.
-However, readers and writers must be capable of inverting PPBs wherever they appear.
-
-\begin{figure}
-    \centering
-    \includegraphics{filler}
-    \caption{Partial parity bond inversion. Direction must be inverted for bonds not expressed in normal form.}
-    \label{fig:ppb-inversion}
-\end{figure}
-
-The distributed nature of conformational encoding means that several additive error states are possible:
+Spreading conformational parity over more than one bond in this way means that several error states are possible (Figure~\ref{fig:bond-error-states}):
 
 \begin{itemize}
-    \item{Overspecification. This state occurs when two or more PPBs share a common terminal and simultaneously use the same parity, thus placing multiple neighbors into the same quadrant.}
-    \item{Underspecification.
-A required and non-deducible PPB is absent.
-This state occurs when a double bond terminal connects to at least one neighbor through a PPB, but the terminal's mate does not.}
-    \item{ Misspecification. A PPB is used between atoms lacking at least one double bond.}
+    \item{Overspecification. Two children are forced into the same direction along an axis.}
+    \item{Underspecification. A required PPB is missing.}
+    \item{ Misspecification. A PPB is used between atoms, neither of which connects to a double bond.}
 \end{itemize}
 
-A reader that detects one of these error conditions must abort the molecular graph being read and report the error state. It is possible for a conformation to be both overspecified and underspecified.
+An error must be reported should any of these states be encountered. Note that it is possible for a conformation to be both overspecified on one side and underspecified on the other.
 
@@
 
-An exception to the underspecification rule applies in the case of conjugated polyenes.
-Here a double bond with a terminal connected through a PPB does not require the opposing terminal to also bear a neighbor connected through a PPB.
-If the opposing terminal lacks neighbors, no error is generated and reading continues normally.
-If the opposing terminal has at least one neighbor, the bond to it need not be a PPB.
-In either case the conformation of the double bond in question remains undefined.
+An exception to the underspecification rule applies in the case of conjugated polyenes (Figure~\ref{fig:bond-error-state-exceptions}).
+Here, a PPB may run from parent to child but not from grandchild to great-grandchild.
+No error is reported in such a case and the conformation of the terminal double bond remains undefined.
 
@@
 
-Given that conformational specification is distributed over three or more bonds, some conformations will be difficult or possible to represent.
-Cyclooctatetraene offers an example.
-Although the (\textit{E}, \textit{E}, \textit{E}, \textit{E}) conformation is expressable, the (\textit{Z}, \textit{E}, \textit{E}, \textit{E}) conformation is not.
-This problem arises because the same PPB encodes a partial conformation of two different double bonds.
+The presence of a bridge node within a PPB complicates the assignment of direction (Figure~\ref{fig:bridged-ppb-bond}).
+A bridge node has no children and therefore can never be a parent.
+At the same time, a bond always points from parent to child.
+This means that a bridged PPB yields a pair of bonds with opposite directions.
+The first (or "left") bond, has the direction that would result from a direct bond to an atomic child.
+The second (or "right") bond has the opposite direction.
+
+\begin{figure}
+    \centering
+    \includegraphics{filler}
+    \caption{Bridged PPB Bond. A bridge across a PPB yields two bonds of opposite direction.}
+    \label{fig:bridged-ppb-bond}
+\end{figure}
+
+Cyclic conjugated polyenes present special challenges to the use of PPBs, as exemplified by cyclooctatetraene.
+Although the (\textit{anti}, \textit{anti}, \textit{anti}, \textit{anti}) conformation is expressable, the (\textit{syn}, \textit{anti}, \textit{anti}, \textit{anti}) conformation is not.
+This problem arises because the same PPB encodes partial parity for two different double bonds.
 
@@
 \section*{Configuration}
-
-The third major element of molecular representation is \textit{configuration}, the three-dimensional arrangement of neighbors about a central atom.
+\label{configuration}
+
 Dialect limits configuration to the special case of an atom with exactly four substituents placed at the vertices of a tetrahedron.
 Here, \enquote{substituent} means an atomic neighbor or virtual hydrogen.
-Lone electron pairs are not considered substituents.
-
-A configuration is comprised of two components: an ordering of bonds to a central atom; and a \textit{configurational descriptor}.
-A configurational descriptor is a template for the relative three-dimensional positioning of neighbors about a central atom.
-Dialect supports two configurational descriptors: \ttt{Left} and \ttt{Right}.
-These descriptors constrain the neighbors of a central atom to the vertices of a tetrahedron.
-
-The \ttt{Left} variant works as follows.
-First, the bond having the index with the lowest value identified.
-The frame of reference is shifted by sighting down this bond from the neighbor to the central atom.
-An equilateral triangle with one vertex pointing upward is then inscribed about the central atom.
-The bond with the next lowest index is then identified and its neighbor is placed at the top vertex of the triangle.
-The neighbor attached to the succeeding bond is then placed at the left triangle vertex.
-Finally, the remaining neighbor is placed at the remaining vertex.
-The overall pattern is one of counterclockwise distribution.
-
-The \ttt{Right} variant follows the same pattern, except that placement of the neighbors attached to the last two bonds in succession are transposed.
-The overall pattern here is one of clockwise distribution.
+Lone electron pairs are not considered to be substituents.
+
+The configuration about an atom can specified through its \ttt{parity} attribute.
+This attribute, together with a general convention, determine the relative three-dimensional positioning of substituents about a central atom.
+First, the parent of the central atom is identified.
+Then the frame of reference is shifted by sighting down the bond from the central atom's parent to the central atom.
+An equilateral triangle is then inscribed about the central atom.
+The bonds from the central atom are then iterated in order.
+If the parity of the central atom equals \ttt{Clockwise}, the child atoms attached through each bond are placed around the triangle in clockwise order.
+Otherwise the children are arranged in counterclockwise order.
 
@@
 
-A modified procedure is used for atoms bearing one virtual hydrogen and three atomic substituents.
+The procedure is modified for atoms bearing one virtual hydrogen and three atomic substituents.
 The implied bond to the virtual hydrogen is considered to precede the others.
-Sighting down this bond, the atomic neighbors are then placed as previously at the vertices of a triangle inscribed about the central atom.
+Sighting down this bond, the central atom's parent is placed on the triangle at an arbitrary position.
+Children are then iterated counterclockwise or clockwise relative to it.
 
@@
 
-An atom whose \ttt{configuration} attribute assumes a non-default value must have a neighbor count of either four if no virtual hydrogens are present or three if one virtual hydrogen is present.
-In any other context, the use of a configurational descriptor is considered an error.
-
-This restriction places some constraints around the use of configuration.
-Some topologies may appear close enough to tetrahedral to stretch the meaning of the configurational descriptors.
-However, all such adaptations are invalid.
-Examples include the central atom of allenes and other odd cumulenes, and the sulfur atom of sulfoxides where the lone pair plays the role of a substituent.
+One additional modification applies to a central atom without a parent.
+The frame of reference is shifted to sight along the direction of the first child.
+The remaining children are then placed around the triangle in either counterclockwise or clockwise order according to the value of the central atom's \ttt{parity} attribute.
+If a central root atom bears a virtual hydrogen, it is considered to be the first child.
+
+A central atom whose parity attribute assumes a non-default value must have either four bonds if no virtual hydrogens are present, or three bonds if one virtual hydrogen is present.
+Some topologies might appear to be similar enough that the \ttt{parity} attribute can be used, but they all must be reported as errors.
+Examples include the central atom of allenes and other odd cumulenes, and the sulfur atom of sulfoxides where the lone pair is misconstrued as a substituent.
 
@@
 
-A configurational descriptor may be applied without regard to bond type.
-In other words, single, double, triple, and partial parity bond types can all be present.
-The only requirement that must be satisfied by a central atom is a neighbor count of four, or three when a virtual hydrogen is present.
+Configurational descriptors may be assigned without regard to bond order.
+In other words, single, double, triple, and PPBs can all be present.
+The only requirement that must be satisfied by a central atom is an atomic neighbor count of four, or three when a virtual hydrogen is present.
 
@@
 A stereocenter is an atom whose \enquote{ligand permutation produces stereoisomers,} as defined by Mislow and Siegel~\cite{mislow:1984}.
-In the context of Dialect, a stereocenter must be tetracoordinate.
-However, this is a necessary but insufficient condition.
-Molecular topology can result in permutation that does not produce stereoisomers.
-Readers should avoid the assignment of stereodescriptors to atoms that are not stereocenters.
+The presence of four bonds around a central atom is therefore a necessary, but insufficient condition.
+Molecular topology can result in permutation that does not yield stereoisomers.
+The assignment of non-default parities to such atoms should therefore be avoided.
 
 Special handling is required for \textit{undefined stereocenters}.
-A stereocenter is undefined if it lacks a stereodescriptor.
-Omitting a stereodescriptor conveys to readers that no information about the stereocenter's configuration is known.
-This could imply that either a single configuration or a configurational mixture are present.
+A stereocenter is undefined if it lacks a parity.
+Undefined stereocenters indicate that no information about a central atom's configuration is known.
+This could indicate the presence of either a single configuration or a configurational mixture.
 This interpretation is consistent with the one used by molfile format~\cite{ctfileFormats}.
@@
 
-Strings conform to an \textit{LL(1) grammar}.\cite{thain:2020} An LL(1) grammar is a context-free grammar whose strings can be parsed one character at a time from left to right with at most one character of lookahead.
+Strings conform to an \textit{LL(1) grammar} \cite{thain:2020}.
+An LL(1) grammar is a context-free grammar whose strings can be parsed one character at a time from left to right with at most one character of lookahead.
 Additionally, LL(1) grammars expand the leftmost non-terminal first.
@@
 
-The third atomic production rule, \ttt{<selection>} is a non-terminal selected from the list: \enquote{c}; \enquote{n}; \enquote{o}; \enquote{p}; and \enquote{s.} An atom encoded in this way (\enquote{selected shortcut atom}) assigns the corresponding atom symbol to the \ttt{element} attribute, sets the \ttt{selected} attribute to \ttt{true}, and sets the \ttt{hydrogens} attribute to \ttt{Implicit}.
+The third atomic production rule, \ttt{<selection>} is a non-terminal selected from the list: \enquote{c}; \enquote{n}; \enquote{o}; \enquote{p}; and \enquote{s.} An atom encoded in this way (\enquote{selected shortcut atom}) assigns the corresponding capitalized atom symbol to the \ttt{element} attribute, sets the \ttt{selected} attribute to \ttt{true}, and sets the \ttt{hydrogens} attribute to \ttt{Implicit}.
 All other atomic attributes retain their default values.
@@
 Bracket can be used to set any atomic attribute.
-A bracket atom must be used for any atom whose \ttt{configuration}, or \ttt{charge} attributes deviate from default values, or whose \ttt{hydrogens} attributes equals anything other than \ttt{Implicit}.
+A bracket atom must be used for any atom whose \ttt{configuration}, or \ttt{charge} attributes assume non-default values, or whose \ttt{hydrogens} attributes equals anything other than \ttt{Implicit}.
 Attributes not set within the \ttt{<bracket>} production rule will leave the corresponding atomic values in their default states.
@@
 \begin{lstlisting}
-<bracket> ::= "[" <isotope>? <symbol> <stereodescriptor>?
+<bracket> ::= "[" <isotope>? <symbol> <parity>?
               <virtual_hydrogen>? <charge>? "]"
@@
 It consists of between one and three digits encoding the integers 1-999.
-Leading zeros (e.g., \enquote{007} are disallowed.
+Leading zeros (e.g., \enquote{007}) are disallowed.
 
@@
 
-The \ttt{stereodescriptor} attribute of a bracket atom is determined by the \ttt{<stereodescriptor>} non-terminal.
+The \ttt{parity} attribute of a bracket atom is determined by the \ttt{<parity>} non-terminal.
 Allowed values are \enquote{@} and \enquote{@@,} corresponding to the values \ttt{Left} and \ttt{Right}, respectively.
@@
 \begin{lstlisting}
-<stereodescriptor> ::= "@" "@"?
+<parity> ::= "@" "@"?
 \end{lstlisting}
@@
 The \ttt{hydrogens} attribute of a bracket atom is controlled by the \\ \ttt{<virtual{\_}hydrogen>} non-terminal.
-It is comprised of the terminal \enquote{H} followed by an optional \ttt{<digit>} non-terminal.
+It is comprised of the terminal \enquote{H} followed by an optional \ttt{<nonzero>} non-terminal.
 A digit appearing after the \enquote{H} terminal sets \ttt{hydrogens} to the corresponding virtual hydrogen count.
-A digit of 0 (i.e., \enquote{H0}) sets the \ttt{hydrogens} attribute to zero.
 If no digit is present, \ttt{hydrogens} is set to one.
+The virtual hydrogen count is set to zero by omitting the \ttt{<virtual{\_}hydrogen} non-terminal.
 
@@
 The \ttt{<charge>} production rule sets the \ttt{charge} attribute of a bracket atom.
-This non-terminal begins with either the plus or minus terminals (\enquote{+} and \enquote{-} respectively) and ends with an optional \ttt{<digit>} non-terminal.
+This non-terminal begins with either the plus or minus terminals (\enquote{+} and \enquote{-} respectively) and ends with an optional \ttt{<nonzero>} non-terminal.
 A missing digit causes the atomic \ttt{charge} attribute to be set to one.
 In other words, the plus terminal (\enquote{+}) sets the \ttt{charge} attribute to one and the minus terminal (\enquote{-}) sets it to minus one.
-Charge values of zero are obtained if \ttt{<charge>} production is present.
+Charge values of zero are obtained if \ttt{<charge>} production is omitted.
 
@@
 For example, the bracket sequence \enquote{[C@H+]} lacks the \ttt{<isotope>} non-terminal so the value of the \ttt{isotope} attribute will remain as \ttt{None}.
-Similarly, the bracket sequence \enquote{[13CH+]} lacks the \ttt{<stereodescriptor>} non-terminal, so the corresponding \ttt{configuration} attribute remains \ttt{None}.
-
-\subsection*{Bond}
-
-An atom may be connected to zero or more neighbors through a bond, encoded with the non-terminal \ttt{<bond>}.
-
-\begin{lstlisting}
-<bond> ::= "-" | "=" | "#" | "/" | "\"
-\end{lstlisting}
-
-Five terminals are available (\enquote{-,} \enquote{=,} \enquote{\#,} \enquote{/,} and \enquote{\textbackslash}).
+Similarly, the bracket sequence \enquote{[13CH+]} lacks the \ttt{<parity>} non-terminal, so the corresponding \ttt{parity} attribute remains \ttt{None}.
+
+\subsection*{Sequence}
+
+Children are added to an atom through the \ttt{<sequence>} non-terminal (\enquote{sequence}).
+A sequence starts with a required \ttt{<atom>} non-terminal.
+If no allowed non-terminals follow, the corresponding atom will have no children.
+Allowed non-terminals are chosen from the list: \ttt{<union>}; \ttt{<branch>}; and \ttt{<split>}.
+
+\begin{lstlisting}
+<sequence> ::= <atom> ( <union> | <branch> | <split> )*
+\end{lstlisting}
+
+The \ttt{<union>} non-terminal consists of an optional \ttt{<bond>} non-terminal followed by a mandatory non-terminal selected from the list: \ttt{<gap>} or \ttt{<sequence>}.
+If either of these latter non-terminals are detected but \ttt{<bond>} is not, the resulting bond is elided.
+
+\begin{lstlisting}
+<union> ::= <bond>? ( <bridge> | <sequence> )
+\end{lstlisting}
+
+The \ttt{<bond>} non-terminal supports five variants (\enquote{-,} \enquote{=,} \enquote{\#,} \enquote{/,} and \enquote{\textbackslash}).
 The first three (\enquote{-,} \enquote{=,} and \enquote{\#}) set the \ttt{order} attribute of a bond to one, two, or three, respectively.
@@
 
-\subsection*{Sequence}
-
-An atom may be assigned zero or more children through the \ttt{<sequence>} non-terminal (\enquote{sequence}).
-A sequence starts with a required \ttt{<atom>} non-terminal.
-If no allowed non-terminals follow, the corresponding atom will have no children.
-Allowed non-terminals are chosen from the list: \ttt{<union>}; \ttt{<branch>}; and \ttt{<sequence>}.
-
-\begin{lstlisting}
-<sequence> ::= <atom> ( <union> | <branch> | <split> )*
-\end{lstlisting}
-
-The \ttt{<union>} non-terminal consists of an optional \ttt{<bond>} non-terminal followed by a mandatory non-terminal selected from the list: \ttt{<cut>} or \ttt{<sequence>}.
-If either of these latter non-terminals are detected but \ttt{<bond>} is not, the resulting bond is elided.
-
-\begin{lstlisting}
-<union> ::= <bond>? ( <cut> | <sequence> )
+\begin{lstlisting}
+<bond> ::= "-" | "=" | "#" | "/" | "\"
 \end{lstlisting}
@@
 
-The \ttt{<cut>} non-terminal (\enquote{cut}) can takes two forms.
-A digit non-terminal can be used, enabling single-digit cut indexes in the range one through nine, inclusive.
-Double- or single-digit cut indexes, supporting the values ten through 99 inclusive, are available by prepending the percent character (\ttt{\%}).
+The \ttt{<bridge>} non-terminal (\enquote{bridge}) can take two forms.
+A digit non-terminal can be used, enabling single-digit bridge indexes in the range one through nine, inclusive.
+Double-digit bridge indexes, supporting the values ten through 99 inclusive, are available by prepending the percent character (\ttt{\%}).
 For the latter variant, leading zeros are disallowed, meaning that the sequence \enquote{\%07} must generate an error.
@@
 \begin{lstlisting}
-<cut> ::= <nonzero> | "%" <nonzero> <digit>
-\end{lstlisting}
-
-A cut signifies the disconnection of an atom from its neighbor.
-The most common use for cut is to encode a cycle.
-The cut index serves as a placeholder for the neighboring atom, which is located by scanning either forward or backward along a string.
-If the number of previous appearances of the cut index is even, then the string is scanned right.
-Otherwise, the string is scanned left.
-Although often used for cycles, cuts can be used for any bond, regardless of cycle membership.
-The only requirement is that a cut must be paired with another cut having the same index.
-
-\begin{figure}
-    \centering
-    \includegraphics{filler}
-    \caption{Cut. Each cycle yields a bond that must be cut.}
-    \label{fig:cut}
-\end{figure}
-
-To prevent overflow, a cut index may be reused provided that it appears to the right of its last paired appearance.
+<bridge> ::= <nonzero> | "%" <nonzero> <digit>
+\end{lstlisting}
+
+As noted previously, the purpose of a bridge is to create a bond without creating a connection within the molecular tree (see: \nameref{molecular-tree}).
+A bridge will most commonly be used to encode a ring-closure bond, but can be used anywhere within a string.
+The only requirement is that a bridge must be paired with another bridge having the same index.
+To prevent overflow, a bridge index may be reused.
 
@@
 Like union, branch joins a parent and child node through a bond.
-Wrapped by opening and closing parenthesis terminals (\enquote{(} and \enquote{)} respectively), branch encodes a sequence that may or may not be attached to its parent.
-Attachment occurs if the \ttt{<bond>} non-terminal is included.
-Alternatively, the sequence will be detached if the \ttt{<detachment>} non-terminal (\enquote{.}) appears.
-If neither \ttt{<bond>} nor \ttt{<detachment>} are present then bond to the first atom of the sequence is elided.
-
-\begin{lstlisting}
-<branch> ::= "(" ( <detachment> | <bond> )?
-             <sequence> ")"
-\end{lstlisting}
-
-The third option for adding atoms within a sequence is the \ttt{<split>} non-terminal (\enquote{split}).
-A split is a sequence of \ttt{<detachment>} and \ttt{<sequence>} productions.
-The \ttt{<detachment>} non-terminal serves the same purpose as it does within a branch: to prevent the bonding to the first atom of the sequence.
-
-\begin{lstlisting}
-<split> ::= <detachment> <sequence>
+Wrapped by opening and closing parenthesis terminals (\enquote{(} and \enquote{)} respectively), branch encodes a sequence that may or may not be bonded to its parent.
+Bonding occurs if the \ttt{<bond>} non-terminal is included.
+Alternatively, the sequence will not be bonded if the \ttt{<dot>} non-terminal (\enquote{.}) appears.
+If neither \ttt{<bond>} nor \ttt{<dot>} are present then bond to the first atom of the sequence is elided.
+
+\begin{lstlisting}
+<branch> ::= "(" ( <dot> | <bond> )? <sequence> ")"
+\end{lstlisting}
+
+The third option for adding atoms within a sequence is the \ttt{<gap>} non-terminal (\enquote{gap}).
+A gap is a sequence of \ttt{<dot>} and \ttt{<sequence>} productions.
+The \ttt{<gap>} non-terminal serves the same purpose as it does within a branch: to enable connection within a molecular tree without creating a bonding relationship.
+
+\begin{lstlisting}
+<gap> ::= <dot> <sequence>
 \end{lstlisting}
@@
 In other words, a Dialect string is either empty or contains a sequence.
-A string without a sequence encodes a molecular graph having zero nodes and zero edges.
+A string without a sequence encodes a molecular tree having zero nodes and zero edges.
 
@@
 On reading the first complete \ttt{<atom>} non-terminal, the corresponding atom is constructed and set as the root atom.
-Subsequently processing a complete \ttt{<union>}, \ttt{<branch>}, or \ttt{<split>} non-terminal triggers three changes: (1) a child atom is constructed; (2) the child is connected to the current root atom, unless a \ttt{<detachment>} non-terminal intervenes; and (3) the root atom is replaced with the child atom.
+Subsequently processing a complete \ttt{<union>}, \ttt{<branch>}, or \ttt{<gap>} non-terminal triggers three changes: (1) a child atom is constructed; (2) the child is connected to the current root atom, unless a \ttt{<dot>} non-terminal intervenes; and (3) the root atom is replaced with the child atom.
 
@@
 
-Further workflow adjustments can handle cuts.
-The necessary state can be maintained by a 100-element array.
-The first occurrence of a cut creates an entry in the array at the cut index consisting of the bond and the root atom.
-The second occurrence of a cut takes the entry at the cut index, attaching the atom in the entry to the current root atom.
-The two bonds spanning a cut must be \textit{compatible}.
-Compatibility is determined by comparing attributes using the following rules.
-If no rule applies to both bonds they are not compatible.
-It is an error to span a cut with incompatible bonds.
+Additional modifications allow bridge bonds to be tested for \textit{compatibility}. Compatibility is tested by comparing the attributes of each bridge bond. This test has three parts:
 
 \begin{enumerate}
-    \item An elided bond is compatible with any other bond, regardless of order or direction.
-    \item A bond with a defined direction is compatible with bond having the opposite direction if \ttt{source} precedes \ttt{target} (the usual case), or the same direction otherwise.
-    \item A bond without a direction is compatible with any bond having the same order.
+    \item An elided bond is compatible with any other bond.
+    \item A bond with a non-default direction is only compatible with an elided bond or a bond of opposite direction.
+    \item A bond with default direction is only compatible with an elided bond or a bond of identical order and default direction.
 \end{enumerate}
 
-\begin{figure}
-    \centering
-    \includegraphics{filler}
-    \caption{Array for cuts. The array's fixed length reflects the syntactic constraint on cut index.}
-    \label{fig:cuts-array}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics{filler}
-    \caption{Bond compatibility. Two non-elided bonds across a cut must have compatible `order` and \ttt{direction} attributes.}
-    \label{fig:bond-compatibility}
-\end{figure}
-
-One further complication applies to decoding cuts.
-Atomic configuration is defined in terms of the relative order of attachment of neighbors about a central atom.
-In the first bond of a cut pair, the neighbor atom is undefined.
-It won't be known until the second bond of the cut pair is found.
-This might only occur after the addition of several intervening neighbors.
-Nevertheless, the neighbor atom across the cut must precede any subsequent neighbors.
-This requirement can be met by a placeholder or some other mechanism that preserves the relative order of neighbor attachment.
-No such special treatment is needed for the second bond of a cut pair.
-
-\begin{figure}
-    \centering
-    \includegraphics{filler}
-    \caption{Order of attachment. A placeholder target index preserves the order of attachment.}
-    \label{fig:order-of-attachment}
-\end{figure}
-
-Readers must not assume that detachment (the period terminal, \enquote{.}) implies the presence of disconnected components.
+A workflow modification allows bridge bonds to be compared for compatibility (Figure~\ref{fig:testing-bridge-bond-compatibility}).
+A 100-element array is used to maintain state.
+The first occurrence of a bridge causes the bridge bond's attributes to be copied to the array at the bridge index.
+The second occurrence of a bridge causes the entry to be removed.
+The attributes stored in the entry are then compared with those of the current bridge bond.
+If the two sets of attributes are incompatible, an error must be reported.
+
+\begin{figure}
+    \centering
+    \includegraphics{filler}
+    \caption{Testing bridge bond compatibility.}
+    \label{fig:testing-bridge-bond-compatibility}
+\end{figure}
+
+An elided bond is compatible with any other bond (Rule 1).
+Unless both bridge bonds are elided, however, the presence of compatible but unmatched bridge bonds could interfere with atom-centered computations.
+One solution is to set the attributes of the elided bridge bond explicitly.
+The array used for compatibility testing can be re-used to set the attributes of an unmatched bridge bond pair.
+
+Readers must not assume that a dot (the period terminal, \enquote{.}) implies the presence of disconnected components.
 This assumption is most likely to arise in the context of ad-hoc parsers using regular expressions, string matching, and the like.
 For example, the connected molecular graph for propane can be encoded using the string \enquote{C1C.C1}.
-
-The use of PPBs conformation over three or more bonds.
-Readers must therefore report as an error any PPB not connected to an atom connected through a double bond.
-For example, an isolated PPB is invalid (e.g., \enquote{C/C} and \enquote{C/CC}).
-Beyond this check, generating (\textit{E}) and (\textit{Z}) stereodescriptors from PPBs is not a straightforward process.
-One approach is to first consider the parity of each double bond terminal individually.
-A separate step would then relate the individual terminal parities to a double bond conformation parity.
-This intermediate parity can then be used to perform the final conversion.
-
-\begin{figure}
-    \centering
-    \includegraphics{filler}
-    \caption{Localized configurational descriptor. Such a representation can be useful when bridging Dialect to other representations.}
-    \label{fig:localized-configurational-descriptor}
-\end{figure}
 
@@
     \item{Unexpected end-of-line. Input ended unexpectedly.}
-    \item{Unbalanced cut (position). A cut with a given index appears an odd number of times.}
-    \item{Incompatible cut bonds (position, position). The bonds to a pair of cuts are incompatible.}
+    \item{Unbalanced bridge (position). A bridge with a given index appears an odd number of times.}
+    \item{Incompatible bridge bonds (position, position). The bonds to a pair of cuts are incompatible.}
     \item{Delocalization subgraph lacks perfect matching. Before reporting this error, steps to remove unnecessarily selected atoms should be taken as described in the next section.}
@@
 As noted previously, a delocalization subgraph is invalid if it lacks a perfect matching.
-The one exception is when a selected atom can be deleted from the delocalization subgraph through \textit{pruning}.
+The one exception is when a selected atom and its associated bonds can be deleted from the delocalization subgraph through \textit{pruning}.
 Pruning unsets the \ttt{selected} attribute of a selected atom without corresponding promotion of any attached bonds.
@@
 And so on.
-If no default valences are found in this way (e.g., \enquote{[c+2]}), a reader must generate an error.
+If no default valences are found in this way (e.g., \enquote{[c+2]}), a reader must report an error.
 Writers must not encode such atoms.
@@
 Selecting any atom is unnecessary because furan does not exhibit DIME.
-But selecting the oxygen atom is particularly unnecessary because it lacks an unpaired electron and so will never lead to DIME.
+This applies doubly to the the oxygen atom because it lacks an unpaired electron and so will never lead to DIME.
 It is nevertheless convenient to select the carbon atoms because all bonds can then be elided.
@@
 Whereas a reader transforms a string into a data structure, the goal of a writer is the opposite: to transform a data structure into a valid string.
-This data structure will most commonly take the form of a molecular graph, but other forms are possible.
-Although a graph itself could be used as the basis for an input representation, a more primitive representation such as an adjacency list could also be used.
+This data structure will most commonly take the form of a molecular graph or tree, but other forms are possible.
 For practical reasons, the input data structure is likely to resemble the output from a reader.
@@
 Traversal proceeds by successively replacing each node as the center of focus, or root.
-Each new root is selected from the untraversed neighbors of the current root.
-Iteration eventually selects all nodes, at which point the traversal ends.
+Each new root is chosen from the untraversed neighbors of the current root.
+Eventually all nodes are processed, ending the traversal.
 
@@
 
-A writer intercepts the depth-first traversal of an input data structure to write node and edge representations.
-The \ttt{<atom>} and \ttt{<bond>} non-terminals are used for this purpose.
+A writer intercepts the depth-first traversal of an input data structure to write an output string compatible with the formal grammar.
 There are no requirements around style.
 For example, it's equally valid to represent the carbon atom of methane using either implicit or virtual hydrogens (e.g., \enquote{C} or \enquote{[CH4]}).
-Single bonds may be elided or not.
+Single bonds may be elided or not, all other things being equal.
 Similarly, the DS may or may not be empty.
-Although an organization may seek to standardize certain styles of string output, any syntactically-valid string must be considered valid by a reader.
+Although an organization may seek to standardize certain styles of string output, any syntactically and semantically valid string must be considered valid by a reader.
 
@@
 
-Cycles are encoded using the \ttt{<cut>} non-terminal.
-This one- or two-digit integer replaces one terminal of a cycle closure bond.
-The challenge is to unambiguously supply these integers while allowing re-use to avoid overflow.
-This can be accomplished with a \textit{pool}.
-A pool is a data structure that yields a numerical index given an ordered pairing of atomic indexes.
-The numerical index will not be re-issued until the pool receives the corresponding reversed pairing.
-
-\begin{figure}
-    \centering
-    \includegraphics{filler}
-    \caption{Pool for assigning cut indexes.}
-    \label{fig:pool}
-\end{figure}
-
-A pool can be used by a writer in the following way.
-The presence of a cycle during depth-first traversal is indicated by an atom that has already been traversed.
-On encountering a cycle, a writer requests an index from the pool, submitting the corresponding atomic indexes as an ordered pair.
-Later, the same bond must be reversed in the reverse direction.
-When it is, the writer once again requests an index, but this time using a reversed pairing.
-Doing so yields the same index, while simultaneously freeing it for later use.
-
-The encoding of cuts is likely to pose special challenges for writers.
-For a cut across a PPB, care must be taken to report the correct parity.
-Consider \textit{trans}-cyclooctene.
-The correct encoding accounts for the reverse in parity at the left-hand side (e.g., \enquote{C{\textbackslash}1C=C/CCCCC/1}).
-Moreover, the double-encoding of the PPB bond type can lead to inconsistencies that must be reported by a reader (e.g., the invalid string \enquote{C/1C=C/CCCCC/1}).
-Such errors can be avoided by encoding the bond type of one side or the other side of a cut, but not both.
-(e.g., \enquote{C{\textbackslash}1C=C/CCCCC/1} or \enquote{C1C=C/CCCCC/1}).
-A cut across a bond attached to an atom with a stereodescriptor must take bond ordering into consideration.
-For example, the two strings \enquote{O[C@H]1NC1} and \enquote{O[C@H](C1)N1} encode the same configuration, but the string \enquote{O[C@H](N1)C1} encodes the \textit{opposite} configuration.
-
-\begin{figure}
-    \centering
-    \includegraphics{filler}
-    \caption{Difficult cuts.}
-    \label{fig:difficult-cuts}
-\end{figure}
-
-The presence of conformational restriction about a double bond presents a special challenge to writers.
-Few other molecular graph systems distribute a conformation descriptor over multiple individual bonds the way that PPBs does.
-Instead, descriptors are more likely to be localized at the double bond.
-Producing the necessary PPBs from a descriptor localized at the double bond is likely to be non-trivial.
-An approach to the reverse problem was presented previously (\nameref{reading-strings} section).
-That approach can also be adapted to the reverse problem.
-
-\begin{figure}
-    \centering
-    \includegraphics{filler}
-    \caption{Writing partial parity bonds through a localized conformation descriptor.}
-    \label{fig:writing-ppb}
-\end{figure}
-
-Writers should carefully weigh the non-negligible costs of atom selection.
+Writers should consider the non-negligible costs of atom selection.
 Algorithms for selection are likely to involve the perceptions of cycles, and could exhibit superlinear time complexity.
 Often, a reader must perform a global deselection to arrive at a localized valence bond representation, which at the very least requires pruning and a maximal matching procedure.
-In this sense atom selection imposes two sets of costs: one on the writer and the other on every subsequent reader forever into the future.
+In other words, atom selection imposes two sets of costs: one on the writer and the other on every subsequent reader forever into the future.
 
@@
 Consider pyrrole, erroneously encoded with a selected nitrogen atom (e.g., \enquote{n1cccc1}).
-The nitrogen atom can not be pruned because of its subvalence is 1 (3 - 2).
+The nitrogen atom can not be pruned because of its non-zero subvalence (3 - 2).
 The DS therefore contains all five atoms and all five bonds.
@@
 Doing so leaves a DS with four nodes, four edges, and a perfect matching.
-Readers will consider the string valid.
-Nevertheless, the larger question of gratuitous selection remains.
+Readers will therefore consider the string valid.
+The larger question of gratuitous selection nevertheless remains.
+
+\section*{Working with Molecular Graphs}
+\label{working-with-molecular-graphs}
+
+Molecular trees offer some important advantages, but they are used less often than molecular graphs.
+Sooner or later, it will be necessary to interconvert the two representations when using Dialect.
+Readers will need to transform a tree into a graph, and writers will need to transform a graph into a tree.
+Although there are only three main differences between molecular trees and molecular graphs, these differences can lead to subtle effects during transformation.
+Understanding these effects is the key to lossless conversion.
+
+The first difference is the simplest to account for: atoms in a molecular tree have two kinds of edges: the one entering from an optional parent; and the zero or more exiting to children.
+A molecular graph, in contrast, maintains a single set of edges for an atom.
+Transformation of an atom in a tree to one in a graph must account for a possible inbound bond.
+The converse transformation must place the inbound bond.
+
+The second difference is that molecular trees have gaps, but molecular graphs do not.
+An edge may or may not represent a bonding relationship.
+Transformation from tree to graph must recognize the presence of gaps and place no corresponding bond.
+Gaps can occur at any outbound edge, even those within branches.
+Transformation from graph to tree must locate an appropriate pair of atoms to host the gap.
+A convenient pair would be the last atom of the first connected component and the first atom of the next connected component, but other choices are possible.
+
+The third difference leads to the most complex reconciliation: molecular trees have bridges but molecular graphs do not.
+When transforming tree to graph, the two sides of a bridge bond must be identified, spliced, and added to the graph.
+When transforming graph to tree, a ring closure bond must be detected and split into a pair compatible bridge bonds.
+Compounding the problem in either sense of conversion is index reuse.
+Transformation from graph to tree will reuse gap indexes to avoid overflow.
+Transformation from tree to graph must account for the possibility that gap index reuse has occurred.
+
+Transformation of tree to graph can be aided by a \textit{fastener} (Figure~\ref{fig:fastener}).
+A fastener is a key/value store whose keys are bridge indexes and whose values are an aggregation of atom rank with the bond attributes \ttt{order}, \ttt{elided}, and \ttt{direction}.
+Odd insertions (first, third, etc.) associate the bridge index with the aggregation and return \ttt{None}.
+Even insertions (second, fourth, etc.) remove the associated aggregation and return it.
+A fastener is paired with a depth-first traversal of the tree to be transformed.
+On encountering a gap an insertion is made.
+If the return value is an aggregation, it is used together with the current gap bond attributes and atom rank to create a bond and place it within the molecular graph.
+
+\begin{figure}
+    \centering
+    \includegraphics{filler}
+    \caption{Fastener. Use of a fastener with the depth-first traversal of a molecular tree aids transformation to a molecular graph.}
+    \label{fig:fastener}
+\end{figure}
+
+Transformation of graph to tree can be aided by a \textit{pool} (Figure~\ref{fig:pool}).
+A pool issues a bridge index given an unordered pairing of atomic indexes.
+The gap index will not be re-issued until the pool receives the same pairing.
+A pool is used during a depth-first traversal of a molecular graph, where the presence of a cycle is indicated by the discovery of an index that has already been traversed.
+On encountering a cycle, a bridge index is requested from the pool by submitting the source and target atomic indexes as an unordered pair.
+The bridge index is then used to construct the first half of the gap bond.
+Later, the same bond will be traversed in the opposite direction.
+When it is, a gap index is requested from the pool by re-submitting a source and target index pairing.
+Doing so yields the bridge index, while simultaneously freeing it for later use.
+
+\begin{figure}
+    \centering
+    \includegraphics{filler}
+    \caption{Pool. When used with the depth-first traversal of a molecular graph, pool enables bridge index reuse.}
+    \label{fig:pool}
+\end{figure}
 
@@
 
-These problems can be avoided by considering as a base language not SMILES itself, but a \textit{protolanguage}.
-A protolanguage is a hypothetical language from which a family of languages can be derived.
-A suitable SMILES protolanguage would ideally contain all of the features consistent with the many ways in which the term \enquote{SMILES} is used today.
-
-Accordingly, Dialect's base language is \enquote{ProtoSMILES,} a SMILES protolanguage.
+These problems can be avoided by considering as a base language not SMILES itself, but a \textit{proto-language}.
+A proto-language is a hypothetical language from which a family of languages can be derived.
+A suitable SMILES proto-language would ideally contain all of the features consistent with the many ways in which the term \enquote{SMILES} is used today.
+
+Accordingly, Dialect's base language is \enquote{ProtoSMILES,} a SMILES proto-language.
 ProtoSMILES was developed through a process of elimination that started with every SMILES document ever published and every SMILES software implementation, Open Source or not.

### 2022-06-09 manuscript.tex
Drops The Article as a source for ProtoSMILES. closes #109

@@
 Implicit hydrogen counting is only supported for elements in the list: B; C, N; O; P; S; F; Cl; Br; or I.}
-    \item{A selected atom must use one of the following five elements: C; N; O; P; or S.}
+    \item{A selected atom must use one of the following five elements: B; C; N; O; P; or S.}
 \end{enumerate}
@@
 \section*{Valence and Subvalence}
+\label{valence-and-subvalence}
 
@@
 Accordingly, Dialect's base language is \enquote{ProtoSMILES,} a SMILES proto-language.
-ProtoSMILES was developed through a process of elimination that started with every SMILES document ever published and every SMILES software implementation, Open Source or not.
+ProtoSMILES was identified through a process of elimination that started with every SMILES document ever published and every SMILES software implementation.
 The field was narrowed through two limitations: (1) the source must be authoritative; and (2) the source must be publicly accessible.
-In this context, authoritative means that the source was created by either Weininger himself or Daylight.
-They alone can speak most authoritatively about what SMILES is.
-The requirement for public accessibility ensures that, even if inconvenient, claims of compatibility can be tested.
+"Authoritative" means that the source was created by either Weininger himself or Daylight, who alone can speak most authoritatively about what SMILES is.
+The requirement for public accessibility ensures that claims of compatibility can be tested.
 The following documents were identified as candidate ProtoSMILES sources:
@@
 \begin{itemize}
-    \item Weininger's report to the US Environmental Protection Agency (\enquote{The Article})~\cite{anderson:1987}.
+    \item Weininger's report to the US Environmental Protection Agency (\enquote{The Report})~\cite{anderson:1987}.
     \item Weininger's original publication (herein \enquote{The Article})~\cite{weininger:1988}.
-    \item Weininger's 2003 book chapter, revised in 2008 (\enquote{The Book Chapter})~\cite{weininger:2003}.
+    \item Weininger's 2003 book chapter, revised in 2008 (\enquote{The Chapter})~\cite{weininger:2003}.
     \item An online manual maintained by Daylight (\enquote{The Manual}).
@@
 
-The Report was eliminated mainly because the contents of this older document contradict the three remaining sources.
-For example, The Report contains no mention of bracket syntax, using markedly different notation instead.
-The report also supports unusual characters including the exclamation mark (\ttt{!}).
-Semantically, The Report supports \enquote{special states} for nitrogen-containing atoms.
-
-The Manual was disregarded as a ProtoSMILES source because it merely recapitulates content already present in The Paper and The Book Chapter.
+The Report was eliminated mainly because the contents of this older document contradict at least two of the remaining sources (The Article and the Chapter).
+The Report contains no mention of bracket syntax, using markedly different notation instead.
+The Report also supports unusual characters including the exclamation mark (\ttt{!}).
+Unspecified \enquote{special states} for nitrogen-containing atoms are also supported by the Report.
+
+The Manual was disregarded as a ProtoSMILES source because it merely restates content already present in The Paper and The Chapter.
 Removing this source simplifies the construction of ProtoSMILES without changing its definition.
@@
 
-The only remaining sources are The Paper and the Book Chapter.
-Therefore, these two works collectively define ProtoSMILES.
+Careful reading of The Article and The Chapter reveals the former to contain all of the material in the latter.
+Moreover, The Chapter contains information not present in The Article.
+Therefore, the Paper was eliminated as a basis for ProtoSMILES.
+
+This process of elimination leaves just The Chapter as the basis for ProtoSMILES.
 The remainder of this section compares and contrasts ProtoSMILES with Dialect.
-The main difficulty with this approach is that the works defining ProtoSMILES not only contradict each other in some ways, but they also contradict themselves internally.
-These contradictions will be noted when meaningful.
-
-The Book Chapter defines a fixed set of element symbols which even at the time of publication was obsolete~\cite[p.~83]{weininger:2003}.
-The symbol for element 105 is given as Ha (Hahnium), despite IUPAC's resolution of this naming controversy in favor of the symbol Db (Dubnium) years prior~\cite{sageson:1997}.
-Accordingly, Dialect disallows the symbol Ha.
-To support the full set of currently-named transuranium elements, Dialect adds the symbols: Db; Sg; Bh; Hs; Mt; Rg; Cn; Nh; Fl; Mc; Lv; Ts; and Og.
-Moreover, Dialect explicitly accepts all future element symbols approved by IUPAC.
-The addition of these element symbols means that Dialect can't be considered a strict subset of SMILES.
-The alternative would, however, leave dialect without the ability to express every kind of IUPAC-recognized element.
-
-The Article implies that comma (\ttt{,}) is either an element symbol or a valid SMILES character: \enquote{This subset uses only the symbols H, C, N, O, P, S, F, Cl, Br, I, and (,) and digits, with the following four rules...}~\cite[p.~33]{weininger:1988}.
-The comma character is explicitly disallowed in Dialect.
-
-The Book Chapter defines a recursive grammar for branching which, perhaps without the knowledge of the author, allows sequentially nested parentheses~\cite[p.~86]{weininger:2003}.
+The main difficulty with this approach is that the work defining ProtoSMILES occasionally contains contradictions.
+These will be noted when relevant.
+
+ProtoSMILES defines a fixed set of element symbols that even at the time of publication was partially obsolete~\cite[p.~83]{weininger:2003}.
+The symbol for element 105 is given as Ha (Hahnium), despite IUPAC's resolution of the naming controversy of the element in favor of the symbol Db (Dubnium) years prior~\cite{sageson:1997}.
+Dialect therefore disallows the symbol Ha.
+To maintain compatibility with ProtoSMILES and avoid future changes to Dialect's grammar, elements after element 105 are also excluded.
+
+ProtoSMILES defines a recursive grammar for branching which, perhaps without the knowledge of the author, allows sequentially nested parentheses~\cite[p.~86]{weininger:2003}.
 Dialect disallows such constructs (e.g., \enquote{*((*))*} and \enquote{*(((*)))*}).
 
-\enquote{Reaction SMILES} is an extension described in the Book Chapter~\cite[p.~89]{weininger:2003}, but disallowed in Dialect.
+ProtoSMILES supports reactions but Dialect does not.
 Specifically, the greater than symbol (\ttt{$>$}) is not a valid Dialect character.
-This precludes strings such as \enquote{*$>$$>$*} in Dialect.
+This restriction precludes strings such as \enquote{*$>$$>$*}.
 Furthermore, Dialect lacks the \enquote{map} attribute, used for atom-atom mapping in reactions, and its accompanying syntax.
 
-Dialect only supports two stereodescriptors, encoded with the terminals \enquote{@} and \enquote{@@.} The Book Chapter provides a recursive grammar for stereodescriptors that allows multi-symbol descriptors such as \enquote{@@@} and \enquote{@@@@}~\cite[p.~94]{weininger:2003}.
-The book chapter also discusses non-tetrahedral descriptors including \enquote{@AL1,} \enquote{@AL2,} \enquote{@1,} and \enquote{@SP1.} None of these are supported by Dialect.
-Nor does Dialect support the application of tetracoordinate stereodescriptors to odd cumulenes as described in the book chapter.
-Although some of these forms of configuration are supported today, it's unusual to find them all supported.
-The reason is simple: in the context in which Dialect will most often used, these forms of stereochemistry are rare.
-To support them burdens implementors with little payoff.
-
-ProtoSMILES supports the concept of \enquote{aromatic bond} (\enquote{:}), but Dialect does not.
-Despite clear inclusion within the syntax of ProtoSMILES, no precise guidance on using this feature is provided.
-The Book Chapter notes \enquote{Adjacent atoms without an intervening bond symbol are connected by a valence-dictated bond (typically a single or aromatic bond) \enquote{-} (single) and \enquote{:} (aromatic) bond symbols may always be omitted on input}~\cite[p.~85]{weininger:2003}.
+Dialect only supports two stereodescriptors, encoded with the terminals \enquote{@} and \enquote{@@.} ProtoSMILES provides a recursive grammar for stereodescriptors that allows multi-symbol descriptors such as \enquote{@@@} and \enquote{@@@@}~\cite[p.~94]{weininger:2003}.
+ProtoSMILES also supports non-tetrahedral descriptors including \enquote{@AL1,} \enquote{@AL2,} \enquote{@1,} and \enquote{@SP1.} None of these are supported by Dialect.
+Nor does Dialect support the application of tetracoordinate stereodescriptors to odd cumulenes.
+These other forms of atomic configuration are relatively rare.
+Their inclusion in Dialect would add substantial cost for minor payoff.
+
+ProtoSMILES supports \enquote{aromatic bonds} (\enquote{:}), but Dialect does not.
+Despite clear inclusion within the syntax of ProtoSMILES, no precise guidance on using aromatic bonds is provided.
+Weininger notes that \enquote{Adjacent atoms without an intervening bond symbol are connected by a valence-dictated bond (typically a single or aromatic bond) \enquote{-} (single) and \enquote{:} (aromatic) bond symbols may always be omitted on input}~\cite[p.~85]{weininger:2003}.
 This passage implies that aromatic bonds serve no purpose.
@@
 
-Along these lines, Dialect does not recognize atoms bearing either arsenic (As) or selenium (Se) as selectable.
-Doing so would increase the complexity of the language unnecessarily.
-Although The Paper implies selectability (\enquote{Elements other than C, N, O, P, S, As, and Se are not yet dealt with in an aromatic context.})~\cite[p.~35]{weininger:1988}, default valences for the latter two atoms are not provided.
-Even if they were, the elements As and Se must be bracketed.
-As such, special rules for handling them in the context of pruning would be required.
-Moreover, molecules bearing the elements As or Se are often made \enquote{aromatic} unnecessarily in SMILES.
-Finally, allowing the selection of atoms bearing As or Se does nothing to address molecules with atoms bearing silicon, germanium, tin, or other heavy main group elements that may be discovered in the future to participate in DIME.
+% ProtoSMILES appears to allow ANY element to be made lower case!
+
+ProtoSMILES sets no limits on the atoms that can be marked as selectable (i.e., \enquote{aromatic}), but Dialect does.
+As described previously, pruning the DS requires a method to determine whether or nor an atom is subvalent (see: \nameref{valence-and-subvalence}).
+Making this determination requires at least one default valence.
+The only reasonable way to allow selection of all atoms regardless of associated element would be to provide default valences.
+As a result, only those atoms associated with a subset of elements may be selected (\enquote{B}, \enquote{C}, \enquote{N}, \enquote{O}, \enquote{P}, \enquote{S}).
+This list is a subset of the elements whose default valences are defined.
+Halogens were excluded because they rarely appear in rings.
+This restriction only marginally limits the capabilities of Dialect because atoms other than those supported are rarely involved in DIME.
 
 It's unclear whether ProtoSMILES allows a virtual hydrogen count on hydrogen itself.
-The Article implies that the element symbol H (hydrogen) must not have an implicit hydrogen count when used within a bracket sequence (\enquote{There are few exceptions to the hydrogen-suppression convention, the most obvious being specification of a proton, [H+], and molecular hydrogen, [H][H].})~\cite[p.~34]{weininger:1988}.
-If so, the author may have erroneously believed that parsers would be incapable of processing strings encoding bracket hydrogen with a virtual hydrogen count (e.g., \enquote{[HH]}).
-The Book Chapter appears to reiterate this view, noting that \enquote{explicit hydrogen specification is required} in the case of \enquote{hydrogens connected to other hydrogens, e.g.
+Weininger notes that \enquote{explicit hydrogen specification is required} in the case of \enquote{hydrogens connected to other hydrogens, e.g.
 [H][H], molecular hydrogen}~\cite[p.~97]{weininger:2003}.
-To be clear, Dialect supports any valid virtual hydrogen count on hydrogen itself (e.g., \enquote{[HH]} and \enquote{[HH2]}).
-Subject to interpretation of the original literature, this feature may or may not render Dialect a superset of SMILES.
-
-ProtoSMILES ascribes ambiguous meaning to detachments.
-As noted in the Book Chapter, \enquote{...
+This may reflect an erroneous belief that parsers would be incapable of processing strings such as \enquote{[HH]}.
+To be clear, Dialect supports virtual hydrogen counts on all atoms.
+
+ProtoSMILES ascribes ambiguous meaning to gaps.
+As noted by Weininger, \enquote{...
 In terms of the valence model being represented, the dot literally represents a bond of formal order zero: the atoms on either side of the dot are explicitly not bonded to each other.}~\cite[p.~88]{weininger:2003}.
 Other statements appear to contradict this statement.
-To be clear, the state of not being bonded and the state of having a zero-order bond are not necessarily equivalent~\cite{clark:2011}.
-Dialect explicitly disallows the zero bond order interpretation of detachments and the dot character.
+Dialect explicitly disallows the zero bond order interpretation.
 
 ProtoSMILES assigns no explicit upper or lower bounds to numerical atomic attributes.
-These boundaries are nevertheless crucial for implementors, who seek the data types necessary to prevent underflow and overflow.
-For this reason, Dialect sets both upper and lower bounds on the following atomic attributes: isotope (0 $\leq$ value \textless 1000); charge (-10, \textless value \textless 10); and hydrogens, 0 $\leq$ value \textless 10).
+These boundaries are nevertheless crucial for implementors, who must determine the data types necessary to efficiently prevent underflow and overflow.
+For this reason, Dialect sets both upper and lower bounds, respectively, on the following atomic attributes: isotope (0 $<$ value \textless 1000); charge (-10, \textless value \textless 10); and hydrogens, 0 $\leq$ value \textless 10).
 Although it might be argued that lower bounds on physical quantities such as isotope and hydrogen count should be implicit, the lack of precision forces implementors to improvise, which can lead to divergent behavior.
-ProtoSMILES places no upper bound on the charge notation (e.g., \enquote{++} and \enquote{+++}) and their corresponding negative forms: \enquote{The SMILES program also recognizes constructions of the form [Fe+++] as being synonymous with the form [Fe+3]}~\cite[p.~32]{weininger:1988}.
+ProtoSMILES places no upper bound on the charge notation (e.g., \enquote{++} and \enquote{-{}-}): \enquote{\dots \enquote{+} meaning +2 \dots}~\cite[p.~94]{weininger:2003}.
 Due to their redundancy, Dialect disallows duplicated plus and minus characters (\ttt{+} and \ttt{-}, respectively) within bracket atoms.
@@
 ProtoSMILES does not restrict partial parity bonds to those atoms also possessing at least one double bond, whereas Dialect does.
-Finally, Dialect syntax is based on a formal grammar and tooling rooted in decades of computer science research, whereas ProtoSMILES syntax is based for the most part on imprecise natural language descriptions.
-
-Dialect also resolves some internal contradictions within ProtoSMILES.
-The Article reports the only default valence for nitrogen as three (p.
-34)~\cite[p.~34]{weininger:1988}, whereas the Book Chapter reports two default valences: 3 and 5~\cite[p.~84]{weininger:2003}.
-Dialect uses the latter default valences.
-Atoms whose \ttt{hydrogens} attribute may be set to \ttt{Implicit} (\enquote{the organic subset}) are reported in the Article as: B; C; N; O; P; S; F; Cl; Br; and I~\cite[p.~32]{weininger:2003}.
-However, an example later in the Article indicates that the string \enquote{Hn1cccc1} with its unbracketed hydrogen valid~\cite[p.~35]{weininger:1988}.
-Dialect explicitly rejects strings containing unbracketed hydrogen atoms.
-On the topic of atom selection (\enquote{aromaticity}), The Article implies that selected atoms must be part of a cycle: \enquote{To qualify as aromatic, all atoms in the ring must be sp2 hybridized and the number of available 'excess' $\pi$ electrons must satisfy Hückel's 4{\textit{N}} + 2 criterion}~\cite[p.~34]{weininger:1988}.
-The Book Chapter, however, presents two examples of valid strings that do not encode cycles (\enquote{cc} and \enquote{cccc})~\cite[p.~85]{weininger:2003}.
-Consistent with the latter view, Dialect explicitly allows the selection of atoms regardless of cycle membership.
+Finally, Dialect syntax is based on a formal grammar and tooling rooted in decades of computer science research, whereas ProtoSMILES syntax is based for the most part on imprecise natural language descriptions and only a partial formal grammar.
 
@@
         \hline
-        Feature & ProtoSMILES\footnote{Values of \enquote{maybe} indicate ambiguous and/or contradictory statements in the authoritative sources.} & Dialect \\
+        Feature & ProtoSMILES\footnote{Values of \enquote{maybe} indicate ambiguous and/or contradictory statements by Weininger.} & Dialect \\
         \hline
         element symbol Ha & yes & no \\
-        element symbols Db; Sg; Bh; Hs; Mt; Rg; Cn; Nh; Fl; Mc; Lv; Ts; and Og & no & yes \\
-        future element symbols approved by IUPAC & no & yes \\
-        comma symbol (\enquote{,}) & maybe & no \\
         multiple branching (e.g., \enquote{*((*))*}) & yes & no \\
@@
         atomic \enquote{map} attribute & yes & no \\
-        extended stereodescriptors (e.g., \enquote{@@@} and \enquote{@OH1}) & yes & no \\
+        extended stereodescriptors (e.g., \enquote{@OH1}) & yes & no \\
         use of stereodescriptors on odd cumulene centers & yes & no \\
@@
         unbounded numerical atomic properties & yes & no \\
-        nitrogen default valence includes 5 & maybe & yes \\
-        unbracketed hydrogen atom & maybe & no \\
-        acyclic atom selection & maybe & yes \\
-        atoms with elements As and Se are selectable & yes & no \\
+        atoms associated with any element are selectable & yes & no \\
         non-elided bonds may be promoted during selection & maybe & no \\
         isolated partial parity bond & yes & no \\
-        repeated charge notation (e.g., \enquote{++} and \enquote{+++}) & yes & no \\
+        repeated charge notation (e.g., \enquote{++} and \enquote{-{}-}) & yes & no \\
         aromatic bond (\enquote{:}) & yes & no \\
@@
 
-With only one clear-cut exception, Table~\ref{table:feature-comparison} shows Dialect to be a subset of ProtoSMILES.
-The exceptional case deals with the validity certain transfermium element symbols already recognized by IUPAC and to be recognized in the future by Dialect.
-Ambiguous exceptions include those for which the ProtoSMILES column feature support is uncertain (\enquote{maybe}) but the Dialect column is affirmative (\enquote{yes}).
-If ProtoSMILES support tends toward \enquote{no,} then support by Dialect would break out of the ProtoSMILES subset.
-There are three such cases: hydrogen having virtual hydrogens (e.g., \enquote{[HH]}); whether or not the default valence of nitrogen includes the value five; and whether or not ProtoSMILES recognizes acyclic \enquote{aromatic} atoms as valid.
+Compatibility between Dialect and ProtoSMILES is summarized in Table~\ref{table:feature-comparison}.
+Barring elements of ProtoSMILES that are either underspecified or internally inconsistent (entries marked \enquote{maybe}), Dialect can be seen to be a subset of ProtoSMILES.
+Output from Dialect writers should therefore be readable by SMILES readers.
+However, output from SMILES writers may or may not be readable by Dialect readers.
 

### 2022-06-10 manuscript.tex
Drops atom rank. closes #110

@@
 Transformation of tree to graph can be aided by a \textit{fastener} (Figure~\ref{fig:fastener}).
-A fastener is a key/value store whose keys are bridge indexes and whose values are an aggregation of atom rank with the bond attributes \ttt{order}, \ttt{elided}, and \ttt{direction}.
-Odd insertions (first, third, etc.) associate the bridge index with the aggregation and return \ttt{None}.
-Even insertions (second, fourth, etc.) remove the associated aggregation and return it.
+A fastener is a key/value store whose keys are bridge indexes and whose values are a tuple composed of: (1) the identifier of the last atom added to the graph; and (2) the bridge bond attributes \ttt{order}, \ttt{elided}, and \ttt{direction}.
+Odd insertions (first, third, etc.) associate the bridge index with the tuple and return \ttt{None}.
+Even insertions (second, fourth, etc.) remove the previously-associated tuple and return it.
 A fastener is paired with a depth-first traversal of the tree to be transformed.
-On encountering a gap an insertion is made.
-If the return value is an aggregation, it is used together with the current gap bond attributes and atom rank to create a bond and place it within the molecular graph.
+On encountering a bridge, an insertion is made.
+If the return value is a tuple, it is used together with the current gap bond attributes and atom identifier to create a bond and place it within the molecular graph.
 

### 2022-06-10 manuscript.tex
Replaces "root atom" with "hub". closes #111

@@
 
-Readers that capture atom-atom connectivity will typically maintain a reference to a \textit{root atom}.
-The root atom is initially undefined.
-On reading the first complete \ttt{<atom>} non-terminal, the corresponding atom is constructed and set as the root atom.
-Subsequently processing a complete \ttt{<union>}, \ttt{<branch>}, or \ttt{<gap>} non-terminal triggers three changes: (1) a child atom is constructed; (2) the child is connected to the current root atom, unless a \ttt{<dot>} non-terminal intervenes; and (3) the root atom is replaced with the child atom.
+Readers that capture atom-atom connectivity will typically maintain a reference to a \textit{hub}.
+A hub is an atom to which the next node will be connected.
+On reading the first complete \ttt{<atom>} non-terminal, the corresponding atom is constructed and set as the hub.
+Subsequently processing a complete \ttt{<union>}, \ttt{<branch>}, or \ttt{<gap>} non-terminal triggers three changes: (1) a child node is constructed; (2) the child is connected to the current hub, unless a \ttt{<dot>} non-terminal intervenes; and (3) the hub is replaced with the child node if it is an atom.
 
 The presence of a branch adds some nuances over a union.
-The leading open parenthesis terminal (\enquote{(}) signifies that the current root atom will later be re-exposed.
+The leading open parenthesis terminal (\enquote{(}) signifies that the current hub will later be re-exposed.
 This operation can be supported by a stack.
-At the start of a branch, the current root is pushed to the stack.
-At the end of the branch, the stack is popped and its top value is assigned as the new root atom.
+At the start of a branch, the current hub is pushed to the stack.
+At the end of the branch, the stack is popped and its top value is assigned as the new hub.
 

### 2022-06-19 manuscript.tex
Adds figure for reprepsentative dialect examples. closes #114
Adds figure tree vs. graph. closes #115

@@
 
+\begin{figure}
+\centering
+\includegraphics[width=\columnwidth]{representative-examples.pdf}
+\caption{Representative Dialect examples.}
+\label{fig:myfig}
+\end{figure}
+
+\section*{Molecular Tree}
+\label{molecular-tree}
+
 Underpinning molecular serialization are conventions and abstractions for molecular representation. These are divided into three broad categories:
@@
 \end{itemize}
-
-\begin{figure}
-\centering
-\includegraphics{filler}
-\caption{Representative Dialect examples.}
-\label{fig:myfig}
-\end{figure}
-
-\section*{Molecular Tree}
-\label{molecular-tree}
 
@@
     \centering
-    \includegraphics{filler}
-    \caption{Tree vs. Graph.}
-    \label{fig:tree-vs-graph}
+    \includegraphics[width=6cm]{molecular-graph-vs-tree.pdf}
+    \caption{Molecular Graph vs.
+Molecular Tree.
+The bonds in a molecular graph reference atoms indirectly through unique indexes (lower left).
+In contrast, a bond in a molecular tree defines its target atom inline (lower right).
+This increases the potential efficiency of trees over graphs for molecular representation.
+To simulate ring closure, a molecular tree introduces bridge nodes, as described later (solid grey circles).}
+    \label{fig:molecular-graph-vs-tree}
 \end{figure}

### 2022-06-19 manuscript.tex
Adds figure for node order. closes #116
Adds figure for node and edge associations. closes #117

@@
     \centering
-    \includegraphics{filler}
-    \caption{Node Order. Nodes in a molecular tree assume the order in which they were added through depth-first construction.}
+    \includegraphics[width=2.5cm]{node-order.pdf}
+    \caption{Node Order.
+Nodes in a molecular tree assume the order in which they were added through depth-first construction.
+Parents precede children, and children added first preceded those added later.}
     \label{fig:node-order}
@@
     \centering
-    \includegraphics{filler}
+    \includegraphics[width=4in]{node-and-edge-associations.pdf}
     \caption{Nodes and Edge Associations.
@@
 Finally, an atom can be connected, without bonding, to a child atom through a gap (right).}
-    \label{fig:nodes-and-edges}
+    \label{fig:nodes-and-edge-associations}
 \end{figure}

### 2022-06-20 manuscript.tex
Adds Elements figure. Closes #118

@@
 The attributes \ttt{element} (\enquote{element}) and \ttt{isotope} (\enquote{isotope}) define atomic number and neutron count, respectively.
-Elements are selected from a subset of those approved by IUPAC \cite{periodicTable} (Figure~\ref{fig:elements}).
+Elements are selected from a subset of those approved by IUPAC \cite{periodicTable} (Figure~\ref{fig:periodic-table}).
 An atom whose element is unknown must have an \ttt{element} attribute equal to \ttt{None}.
@@
     \centering
-    \includegraphics{filler}
-    \caption{Elements. The element that may be associated with an atom.}
-    \label{fig:elements}
+    \includegraphics[width=4in]{periodic-table.pdf}
+    \caption{Elements.
+Available chemical elements and their allowed roles.
+Not all IUPAC-approved elements are available.
+Hydrogen counts of atoms associated with elements on blue background can be computed algorithmically.
+Atoms associated with elements on green background can be selected.}
+    \label{fig:periodic-table}
 \end{figure}
